---
title: "Tidy Data"
subtitle: Data Analysis for Psychology in R 1
author: Tom Booth
---

```{r premable, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(kableExtra)
```

## Today
- Principles of tidy data
- Data sets (or frames in R language)
- Discuss assessments
    - Homework quizzes
    - Lab test dates

## Learning objectives
- Be aware of what tidy data is and why it is preferable.
- Be familiar with some key R terminology `tidyverse` and `tibble`

## Tidy data
1. Each variable must have its own column.
2. Each observation must have its own row.
3. Each value must have its own cell.

- This means that each individual value belongs to both a variable and an observation.

## How do we know if data is tidy?
- To know if we meet points 1, 2 and 3, we need to;
    - Identify variables
    - Identify observations

## Is this data tidy?

```{r, echo=FALSE}
data1 <- tibble(
    ID = paste("ID", 101:110, sep=""),
    TestScore = round(sample(20:50, 10, replace=T),0),
    Degree = as_factor(sample(c("Psychology", "Linguistics", "Philosophy"), 10, replace = T))
)

kable(data1) %>%
    kable_styling(bootstrap_options = "striped", full_width = F)
```

## Is this data tidy?

```{r, echo=FALSE}
data2 <- tibble(
    ID = paste("ID", 101:105, sep=""),
    TestScore_time1 = round(sample(20:50, 5, replace=T),0),
    TestScore_time2 = round(sample(20:50, 5, replace=T),0),
    TestScore_time3 = round(sample(20:50, 5, replace=T),0)
)

kable(data2) %>%
    kable_styling(bootstrap_options = "striped", full_width = F)
```


## Why do we need tidy data?
- If we have a common structure to data, we can have a common set of tools (and language) to do everything we need to do with that data. 
- For this course, in R, these tools are the R functions within the `tidyverse`
- These tools allow us to...
    - visualize
    - score and transform
    - summarise
    - analyse

## Don't study tools do this for us?
- No!

## Common issues with data

```{r, echo=FALSE}
data3 <- tibble(
    Degree = as_factor(c("Psychology", "Linguistics", "Philosophy")),
    "£20-30k" = c(28, 33, 45),
    "£30-40k" = c(40, 32, 21)
)

kable(data3) %>%
    kable_styling(bootstrap_options = "striped", full_width = F)
```

- Variables are rows and columns

## Common issues with data

```{r, echo=FALSE}
data4 <- tibble(
    ID = c("ID101", "ID102", "ID103", "ID101", "ID102", "ID103"),
    Score = c("Psych50", "Psych20", "Psych30", "stats10", "stats30", "stats25")
)

kable(data4) %>%
    kable_styling(bootstrap_options = "striped", full_width = F)
```

- Multiple variables in a column

## Common issues with data
- Other things we may need to do to data
    - Rename variables
    - Give new labels to levels of nominal variables
    - Code missing data
    - Select certain columns or rows
- Over the rest of this year, we will learn how to do all of these things.

## Let's make a data set
- Data sets in `tidyverse` are called `tibbles`
- Variables:
    - ID = nominal category - unique label for each person in data.
    - Height(cm) = continuous ratio
    - Age(years) = continuous ratio
    - Chocolate preference = nominal category
    - Uni so far = order category (1=rubbish, 2 = OK, 3 = Great)
    - Younger than Tom = logical (Yes/No)


## Class .Rmd: let's build a `tibble`

## Tasks for this week...
1. Finish any existing tasks from lab 2.
2. Reading: Linked at the top of lab 3.
    - Additional reading on final slide. 
3. Quiz 3 - measurement.
    - **This quiz counts**
    - Live now (as of Monday at 09:00).
    - Closes Sunday at 17:00

## Further reading (not compulsory)
- Wickham, H. (2014). Tidy data. *Journal of Statistical Software*, 59(10), 1-23.

## Recommendations of the week
- Podcast: [*In Our Time* on Gauss](https://www.bbc.co.uk/programmes/b09gbnfj) + [*More of Less*](https://www.bbc.co.uk/programmes/b006qshd)  
- Recipe: [Carrot, cougette and sweetcorn fritters](https://www.foodtolovemagazine.co.uk/recipe-articles/courgette-carrot-sweetcorn-fritters)  
- Book: [Flatland: A Romance of Many Dimensions, Edwin Abbott](https://www-cambridge-org.ezproxy.is.ed.ac.uk/core/books/flatland/71E876947B2318B1B023975C7EAA5E69)  
- Thing to do: Arthur's Seat  
- Coffee/food Place: [Gaia, Leith Walk](http://gaiadeliedinburgh.co.uk/) 
    - Best coffee in Edinburgh