---
title: "<b>Semester 2 Welcome</b>"
subtitle: "<small>Data Analysis for Psychology in R 1<br>Semester 2, Week 1</small>"
author: "<br>"
institute: "Department of Psychology<br/>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: jk_libs/libs
    css: 
      - un-xaringan-themer.css
      - jk_libs/tweaks.css
    nature:
      beforeInit: "jk_libs/macros.js"
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(digits=4,scipen=2)
options(knitr.table.format="html")
xaringanExtra::use_xaringan_extra(c("tile_view","animate_css","tachyons"))
xaringanExtra::use_extra_styles(
  mute_unhighlighted_code = FALSE
)
library(knitr)
library(tidyverse)
library(ggplot2)
knitr::opts_chunk$set(
  dev = "svg",
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
knitr::opts_chunk$set(fig.asp=.9)
#source('R/myfuncs.R')
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro"),
  outfile = "un-xaringan-themer.css"
)
```

class: inverse, center, middle
# Semester 1 recap  

---

.pull-left[

## S1 weeks 1 - 5  

- Collecting Data & Types of Data  

- Describing and Visualising Categorical Variables   

- Describing and Visualising Numeric Variables  

- Describing and Visualising Relationships  

- Types of Relations (functions)  

<br>

__Formative report A__
]

--

.pull-right[
## S1 weeks 7 - 11  

- Probability Theory  

- Probability Rules  

- Discrete Random Variables   

- Continuous Random Variables  

- Sampling Distributions  

<br>

__Formative report B__
]

---

# Things we hope you did last semester

.pull-left[

### Engage with course materials

- Complete the weekly quizzes

- Attend lectures and complete any lecture activities or readings

- Work through the lab tasks and complete the formative reports

- Attend office hours for one-to-one support

- Write your questions on the discussion forum

] 
.pull-right[

### RStudio skills

- Basic arithmetic operations   

- RStudio functionality 

- Data exploration in R 

- RMarkdown reporting

- ggplot!

]

---

class: inverse, center, middle, animated, rotateInDownLeft

# A map of semester 2

---

.pull-left[

## S2 weeks 1 - 5

- Confidence Intervals

- Hypothesis Testing: P-values  

- Hypothesis Testing: Critical Values  

- Hypothesis Testing & Confidence Intervals

- Making Decisions (Effect Sizes, Power, Errors) 

<br>

Lecturer: Dr Umberto Noe (weeks 1-5)
]

.pull-right[

## S2 weeks 6 - 10

- One Sample t-Test  

- Independent Samples t-test  

- Paired Samples t-test

- Chi-square test

- Covariance and correlation

<br>

Lecturer: Dr Emma Waterston (weeks 6-8)

Lecturer: Dr Monica Truelove-Hill (weeks 9-10)

]


---

# S2 Weeks 1-5

- Formative report C

- Group-based, as in the previous 2 practice reports

- This is the last practice report

- Due: 12 noon, Friday 17th February 2023

---

# S2 Weeks 6-10

- Assessed report, worth 30% of final DAPR1 grade

- To be done in groups (same groups as Formative Report C)

- Only one person per group submits on behalf of the entire group

- No extensions allowed as this is group work based

- Due: 12 noon, Friday 31st March 2023

- Based on your feedback from semester 1:

    + Peer adjusted marking
    + Report grade: 90% from content, 10% from contribution rated by your peers

- Don't plagiarise!

    + You cannot share information with a different group
    + You cannot share code with a different group
    + You cannot use code from another group and add your own write-up
    + You cannot plagiarise example solutions


---

# Exam information

- When and where: May exam diet. The exact date will be announced later by the University timetabling unit. 
The location will also be announced later by the University timetabling unit

- Format:

    + In-person written exam in an exam hall
    
    + 2 hours long
    
    + 60 multiple choice questions

- An equation sheet will be provided in the exam hall, you cannot bring your own one

    + You can preview the equation sheet as a PDF file on LEARN
    
    + LEARN > Course Information > Final Exam Information

- Bring multiple pens (blue or black ink) and a pocket calculator

    + Your phone is not allowed as a calculator

---

# Equation sheet

- See Course Information on LEARN

- Familiarise yourself with the equation sheet, you will be provided with a copy in the exam hall

- Note: the sheet may change slightly over the course of the semester. Changes will be announced!

    + Make sure you check the Piazza discussion forum and LEARN announcements regularly.
    
    + The up-to-date version will be in folder shown.


---

# Mock exam

- A file with a few example questions will be added to the Final Exam Information folder. 

    + LEARN > Course information > Final Exam Information

- An announcement will be sent when this has been added.

- Remember, the exam will have 60 multiple choice questions!


---

class: inverse, center, middle, animated, rotateInDownLeft


