---
title: "Assumptions of t-tests"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
```

We can only trust the outcome of our t-test if the assumptions of the test are met.

## The three assumptions

1. The data are continuous;
2. The data are independent; 
3. The data are normally distributed __OR__ the sample size is large enough (rule-of-thumb $n \geq 30$) and the data are not strongly skewed;


## Checking the assumptions

1. Can the data take on any numeric values?
2. Have the observations been randomly sampled from the population?
3. If the sample size is below 30, check visually using plots and/or use a Shapiro-Wilk test.

### Plots for normality

- Plot the data using a histogram or density plot and check that it looks like a bell curve.
- Make a Q-Q plot (Quantile-Quantile plot), which orders the data and plots it against the equivalent quantile of the normal distribution. We want the observed quantiles to follow the diagonal line of the ideal quantiles.


### Shapiro-Wilk test for normality

We can also conduct a formal hypothesis test for normality, such as the **Shapiro-Wilk** test. 

The null hypothesis of the Shapiro-Wilk test is that the sample came from a population that is normally distributed.  
The alternative hypothesis is that the sample came from a population that is *not* normally distributed.  

The test returns a test statistic $W$, and a p-value $p$. The $W$ test statistic is slightly complex to compute by hand, so we will use R to compute it for us. The test statistic $W$ measures how much the sample data depart from normality (the null hypothesis).

The p-value corresponds to the probability of observing data of this shape of distribution, assuming the data are drawn from a normally distributed population (i.e., assuming the null hypothesis to be true).  

In R:  

```{r eval=F}
shapiro.test(mydata$myvariable)
```

If the p-value is greater than .05, then we fail to reject the H0 that the sample came from a population that is normally distributed.

In other words, we want a non-significant result here—if the result is significant, then the data is significantly non-normally-distributed.