---
title: "Describe relationships using mathematical functions"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```


## Functions  

You have seen by now how to visualise the distribution of a variable, and how to visualise a relationship between two variables. Relationships between two variables can look very different, and can follow different patterns. These patterns can be expressed mathematically in the form of __functions.__  

:::{.callout-note}
__Functions__

A function is a mapping between two sets of numbers ($x$ and $y$) - associating every element of $x$ with an element in $y$.  
We often denote functions using the letter $f$, in that we state that $y = f(x)$ ("y equals f of x").     

For example, there is a mapping between these two sets of numbers: 

- $x$ = 1, 2, 3, 4, 5, 6, 7, 8
- $y$ = 3, 4, 5, 6, 7, 8, 9, 10

To get $y$, you take $x$ and add two.

Formally we'll represent two lists of numbers like this:

$$x=\begin{bmatrix} 1 \\ 2 \\ 3 \\ 4 \\ 5 \\ 6 \\ 7 \\ 8 \\ \end{bmatrix}, \qquad
y=\begin{bmatrix} 3 \\ 4 \\ 5 \\ 6 \\ 7 \\ 8 \\ 9 \\ 10 \\ \end{bmatrix}$$

And mathematically, we can write the mapping between them as:  

$$f(x) = x + 2$$

And we could visualise this relationship between $x$ and $y$:

```{r echo=FALSE}
tibble(
    X = 1:8,
    Y = X+2
) |> 
    ggplot(aes(X,Y)) +
    geom_point() +
    theme_classic()
```

:::

### Linear functions {-}  

In statistics, we often attempt to summarise the pattern that is present in the data using _linear functions_.  

Imagine that we plant 10 trees, and measure their heights each year for 10 years. We could visualise this data (the relationship between _time_ and _tree height_) on a scatterplot (we have added some lines to the plot to show which tree is which):  
<br>
```{r echo=FALSE}
set.seed(677)
tibble(
    tree = rep(1:10,each=11),
    time = rep(0:10,10),
    growth = abs(round(rnorm(110,10,2)))
) |> 
    mutate(growth = ifelse(time==0, round(growth/2), growth)) |>
    group_by(tree) |> 
    mutate(height=cumsum(growth)) |> 
    ungroup() -> treedat 


ggplot(treedat,aes(x=time, y=height))+
    geom_point() + 
    geom_line(aes(group=tree,col=factor(tree)), alpha=.4)+
    xlab("Time (years)") +
    ylab("Tree Height (cm)") +
    ggtitle("Tree height") + 
    theme_classic()+
    theme(legend.position="none")+
    scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0))
```

We might sensibly choose to describe this pattern as a line: 
<br>
```{r echo=FALSE}
ggplot(treedat,aes(x=time, y=height))+
    geom_point(alpha=.2) + 
    geom_line(aes(group=tree,col=factor(tree)), alpha=.2)+
    xlab("Time (years)") +
    ylab("Tree Height (cm)") +
    ggtitle("Tree height") + 
    theme_classic()+
    theme(legend.position="none")+
    scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0))+
    geom_abline(intercept=5, slope = 10, lwd=1)
```

And in order to describe a line like this, we require two things:  

1. The starting point (i.e., where it crosses the y-axis)
2. The amount it goes up every year. 

When we planted the trees (at year 0), they were on average about 5cm tall. So this is where our line starts.  
For every year, the trees grew by about 10cm on average. So we can now describe tree height _as a function of_ time:  
$$\textrm{Tree height} = 5 + (10 \times \textrm{Years})$$

We can write this in terms of $x$ and $y$:  

1. $y = f(x)$ &nbsp; &nbsp; &nbsp; &nbsp; "$y$ is some function $f$ of $x$"
1. $f(x) = 5 + 10x$ &nbsp; &nbsp; &nbsp; &nbsp; "the function $f$ maps each value $x_i$ to $5 + (10 \times x_i)$"


### Non-linear functions {-}  

Functions don't have to be linear. Often, we might want to describe relationships which appear to be more complex than a straight line.

For example, it is often suggested that for difficult tasks, _some_ amount of stress may [improve performance](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1906714/) (but not too little or too much). We might think of the relationship between performance and stress as a curve (@fig-ydodson).

<br>

```{r ydodson, echo=FALSE}
#| label: fig-ydodson
#| fig-cap: "Yerkes Dodson Law"
tibble(
    stress = -100:100,
    performance = -(stress^2)
) |> 
    ggplot(aes(x=stress,y=performance))+
    geom_path()+
    scale_x_continuous(breaks=c(-100,100),labels=c("Low","High"))+
    theme_classic()+
    theme(axis.ticks = element_blank(), axis.text.y = element_blank(), axis.line = element_blank())+
    labs(x="Stress",y="Performance")
```

<br>

One way to describe curves is to use polynomials ($x^2$, $x^3$, etc.).   
For example, in the following two sets, $y$ can be described as $f(x)$ where $f(x)=x^2$:  

$$x=\begin{bmatrix} -5 \\ -4 \\ -3 \\ -2 \\ -1 \\ 0 \\ 1 \\ 2 \\ 3 \\ 4 \\ 5 \end{bmatrix}, \qquad
y=\begin{bmatrix} 25 \\ 16 \\ 9 \\ 4 \\ 1 \\ 0 \\ 1 \\ 4 \\ 9 \\ 16 \\ 25 \end{bmatrix}$$

and when we plot each value of $x$ against the corresponding value of $y$: 
```{r echo=FALSE}
tibble(x = -5:5,
       y = x^2
) |> ggplot(aes(x = x, y = y)) + 
    geom_point()
```

:::{.callout-note collapse="true"}
### The code to create the above plot

```{r eval=FALSE}
# the tibble() function can be used to create a dataframe
# here, we create one with two variables, x and y.
# x is the sequence from -5 to 5,
# y is equal to x squared. 
# we save this data as "poly"
poly <- 
    tibble(
        x = c(-5,-4,-3,-2,-1,0,1,2,3,4,5),
        y = x^2
    )
# create a plot with "poly", with variable "x" on the x-axis,
# and variable "y" on the y-axis. Add geom_points for each entry
ggplot(data = poly, aes(x = x, y = y)) + 
    geom_point()
```
:::

 
## Glossary

+ __Function:__ A mapping between two sets of numbers, associating every element of the first set with an element in the second.  