---
title: "Store data with vectors"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
```


To store a sequence of numbers into R, we combine the values using the combine function `c()` and give the sequence a name. A sequence of elements all of the same type is called a **vector**.
To view the stored content, simply type the name of the vector.

```{r}
myfirstvector <- c(1, 5, 3, 7)
myfirstvector
```

We can perform arithmetic operations on each value of the vector. For example, to add five to each entry:
```{r}
myfirstvector + 5
```

Recall that vectors are sequences of elements **all of the same type**. They do not have to be always numbers; they could be words such as real or fictional animals.
Words need to be written inside quotations, e.g. "anything", and instead of being of numeric type, we say they are characters.

```{r}
wordsvector <- c("cat", "dog", "parrot", "peppapig")
wordsvector
```

:::frame
**NOTE**

You can use either double-quote or single-quote:
```{r}
c("cat", "dog", "parrot", "peppapig")
c('cat', 'dog', 'parrot', 'peppapig')
```
:::

The function `class()` will tell you the type of the object. In this case, it is a character vector.
```{r}
class(wordsvector)
```

It does not make sense to add a number to words, hence some operations like addition and multiplication are only defined on vectors of numeric type. If you make a mistake, R will warn you with a red error message.

```{r eval=FALSE}
wordsvector + 5
```
<p style="color:red">
Error in wordsvector + 5 : non-numeric argument to binary operator
</p>

Finally, it is important to notice that if you combine together in a vector a number and a word, R will transform all elements to be of the same type.
Why? Recall: vectors are sequences of elements **all of the same type**.
Typically, R chooses the most general type between the two. 
In this particular case, it would make everything a character, check the "", as it would be harder to transform a word into a number!

```{r}
mysecondvector <- c(4, "cat")
mysecondvector
```
