---
title: "Block 4 Week 5: Fitness & Health Report"
output: bookdown::pdf_document2
toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{r include = FALSE}
######## LOAD LIBRARIES & DATA ######## 
library(tidyverse)
library(kableExtra)
library(psych)
library(patchwork)
library(kableExtra)


dat <- read_csv("https://uoepsy.github.io/data/fitnessProgram.csv")

###################### RQ1 ########################


###################### RQ2 ########################


```

# Introduction

The data available at https://uoepsy.github.io/data/fitnessProgram.csv were provided by a Personal Trainer who specializes in fitness programs, specifically running programs. Their dataset contained information on 8 variables - the number of days their clients have been enrolled on the program (`daysInProgram`), the maximum distance they can run (in miles; `maxDistance`), their base fitness level (`baseFitness`), their cardiovascular health rating (`cvHealth`), their average resting pulse rate (`pulse`), their age (in years; `age`), their difficulty rating of the program (on a scale of 1-5; `Difficulty`), and their overall rating of satisfaction with the program (on a scale of 1-5; `Satisfaction`). 

## Research Questions


# Analysis 


## Research Question 1


## Research Question 2


# Appendix

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```

