---
title: "Block 4: Weeks 1-3 Reaction Time Report Template"
output: bookdown::pdf_document2
toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{r include = FALSE}
######## LOAD LIBRARIES & DATA ######## 

#load packages
library(tidyverse)
library(patchwork)
library(kableExtra)
library(psych)
library(effectsize)

#read in data
rt_data <- read_csv("https://uoepsy.github.io/data/dapr1_reaction_time_2526.csv")

########################### INSPECT, CLEAN, TIDY, & MANAGE DATA ###########################


###### STEP 1: INSPECT AND CLEAN ###### 

# examine dataset
head(rt_data)
str(rt_data) # data that should be factors/numeric not currently coded as so
summary(rt_data) # we do not have values less than 0cm or over 30cm (as expected)

#make numeric / factors
rt_data <- rt_data |> 
    mutate(age = as.numeric(age),
           sex = as.factor(sex),
           handedness = as.factor(handedness),
           participation_order = as.factor(participation_order),
           lab_session = as.factor(lab_session),
           first_condition = as.factor(first_condition),
           visualcue_attempt1 = as.numeric(visualcue_attempt1),
           visualcue_attempt2 = as.numeric(visualcue_attempt2),
           visualcue_attempt3 = as.numeric(visualcue_attempt3),
           visualcue_attempt4 = as.numeric(visualcue_attempt4),
           visualcue_attempt5 = as.numeric(visualcue_attempt5),
           auditorycue_attempt1 = as.numeric(auditorycue_attempt1),
           auditorycue_attempt2 = as.numeric(auditorycue_attempt2),
           auditorycue_attempt3 = as.numeric(auditorycue_attempt3),
           auditorycue_attempt4 = as.numeric(auditorycue_attempt4),
           auditorycue_attempt5 = as.numeric(auditorycue_attempt5)
           )

###### STEP 2: DISTANCE TO TIME CONVERSION (ON PSYCH 1A LAB SHEET) ###### 

#function to convert distance to time
rt_data <- rt_data |>
  mutate(across(visualcue_attempt1:auditorycue_attempt5, .fns = function(y) sqrt(2 * y / 980)) )

###### STEP 3: CREATE AVERAGE SCORES (Q ON PSYCH 1A LAB SHEET) AND SUBSET / FILTER ###### 

#create average scores
rt_data$VisualCue_AvgRT <- rowMeans(rt_data[ , c(8:12)], na.rm = TRUE)
rt_data$AuditoryCue_AvgRT <- rowMeans(rt_data[ , c(13:17)], na.rm = TRUE)

#subset
rt_data2 <- rt_data[ , c(1:7, 18:19)]

#retain only complete cases
rt_data2 <- rt_data2 |>
  filter(complete.cases(VisualCue_AvgRT, AuditoryCue_AvgRT))

########################### WEEK 1 ###############################


###### STEP 1: DESCRIBE & VIZ ######


###### STEP 2: CONDUCT ANALYSIS ######


###### STEP 3: CHECK ASSUMPTIONS ######

```


# Introduction

The data available at https://uoepsy.github.io/data/dapr1_reaction_time_2526.csv comprises of measurements of reaction times (RT) across two types of conditions - visual and auditory - from a sample of 175 Psychology 1A students at the University of Edinburgh. Data was collected across morning (AM) and afternoon (PM) lab sessions (`lab_session`) that took place on Wednesdays, Thursdays, and Fridays. Participants worked in pairs, and switched roles after one person had completed both conditions (`participant_order`), and noted which condition was completed first (`first_condition`). The participant was tasked with catching a falling ruler as quickly as possible across 5 trials in the visual condition (`visualcue_attempt1` - `visualcue_attempt5`) and 5 trials in the auditory condition (`auditorycue_attempt1` - `auditorycue_attempt5`). Measurements were originally recorded in distance ($cm$). 

## Research Questions

+ RQ1: Did the average reaction time to catch a falling ruler in the visual condition significantly differ from 0.16 seconds?  

# Analysis 

First we transformed our distance into reaction time ($t$; measured in seconds) for each of the 5 trials in both the visual and auditory conditions. To do so, we applied the following conversion (where $y$ is the distance the ruler fell (in $cm$), and $g$ is the acceleration caused by gravity (which is known to be 980 $cm/s^2$)):

$$
y = \frac{1}{2}gt^2
$$

and so 

$$
t = \sqrt\frac{2y}{g}
$$


We only retained cases where participants had average RTs for both the visual and auditory conditions. This resulted in a final sample size of 176. 

## Research Question 1


# Discussion


# Appendix

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```

