---
title: "Block 4 Week 4: Personal Training & Marathon Report"
output: bookdown::pdf_document2
toc: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r include = FALSE}
######## LOAD LIBRARIES & DATA ######## 
library(tidyverse)
library(patchwork)
library(kableExtra)
library(psych)
library(effectsize)

dat1 <- read_csv("https://uoepsy.github.io/data/marathonDat1.csv")
dat2 <- read_csv("https://uoepsy.github.io/data/marathonDat2.csv")

###################### RQ1 ########################



###################### RQ2 ########################


```


# Introduction

The data available at https://uoepsy.github.io/data/marathonDat1.csv and https://uoepsy.github.io/data/marathonDat2.csv were provided by a Personal Trainer who specializes in workouts that involve running. Their first data set contained information on the base fitness levels (`BaseFitness`; levels = Low, Moderate, High) of 90 clients; whilst the second contained information regarding the training methods (`Group`; levels = New Method, Typical, Solo) they use with their clients and whether they could complete a marathon (`Marathon`; levels = None, Half, Full).

## Research Questions


# Analysis


## Research Question 1


## Research Question 2


# Appendix

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```


