---
title: "Research Design & Data"
subtitle: "Semester 1 - Week 1"
bibliography: references.bib
biblio-style: apalike
link-citations: yes
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('assets/setup.R')
```

---

# Data

**Lexical Decision Task**. In the next few weeks of the course you will be working with a simulated dataset arising from a 2x2 mixed design lexical decision task. One hundred participants were asked to decide whether a presented word was a real word or a non-word. The data include 100 rows (one for each participant) and 7 variables:

```{r, echo=F}
require(tidyverse)
tribble(
    ~'Variable Name', ~'Description',
    '`id`', 'Participant ID',
    '`age`', 'Age',
    '`language`', 'Language group (1 = monolingual, 2 = bilingual)',
    '`rt_word`', 'Reaction time (ms) for word trials',
    '`rt_nonword`', 'Reaction time (ms) for non-word trials',
    '`acc_word`', 'Accuracy for word trials',
    '`acc_nonword`', 'Accuracy for non-word trials'
) %>% 
    knitr::kable()
```

You can download the data from this link: https://uoepsy.github.io/data/ldt_data.csv.


# Formative Report A

By the end of week 5 you should have produced a PDF report using Rmarkdown for which you will receive formative feedback in week 6. For formative report A, you will be asked to perform the following tasks, each related to a week of teaching in this course:

A1. read the data into R and inspect it

A2. display and describe the categorical data

A3. display and describe the numerical data

A4. display and describe a relationship of interest between two or three variables of your choice

This week you will learn how to do task A1.


# Exercises

1. Read the lexical decision task data into R.

1. How many observations are there? 

1. How many variables are there? 

1. What is the type of each variable?

1. What's the minimum and maximum age in the sample? What about the average age?

1. Do you notice any issues with the ages? Look at your data. Think about a possible reason why the results are strange. What could be the reason the researchers 

1. The `language` variable appears to contain numbers. Do those numbers correspond to an amount or a type? Convert that variable to the appropriate type and provide more descriptive labels.


## Hints

__Useful functions__. You may find some of these functions useful: `read_csv()` or `read.csv()`, `head()`, `View()`, `dim()`, `nrow()`, `ncol()`, `min()`, `max()`, `range()` `mean()`, `summary()`, `str()`, `%>%`, `summarise()`, `factor()`, `fct_recode()`.

The above list contains functions that can lead you to the same conclusion, so you don't have to do the same thing in 2 ways, pick the way you find easiest.

Make sure to look at the help for the functions, either by running `?functionname` in R or by searching on the web, for example `R read_csv`. The R will ensure you will get results for the R language.

If you are stuck, please have a look at the worked example below and try to apply the same rationale to the data above.


# Worked Example

```{r, echo=F, eval=F}
set.seed(1)
df = tibble(
    StudentID = paste("s", 829040 + 1:30, sep = ""),
    Height = runif(30, 152, 192), 
    DegreeType = sample(c("UG", "PG"), 30, replace = T),
    IQ = rnorm(30, 100, 15)
)
df[11, "Height"] = 0
df[27, "Height"] = 0
df[4, "IQ"] = NA

write_csv(df, "../../data/acme_student_data.csv")
```

Consider the dataset available at the address https://uoepsy.github.io/data/acme_student_data.csv, in which the following variables were recorded for 30 students of ACME university, a fictitious university in the UK:

```{r, echo=F}
tribble(
    ~'Variable Name', ~'Description',
    '`StudentID`', 'Student ID provided by the university',
    '`Height`', 'Height in cm',
    '`DegreeType`', 'Type of degree (UG = undegraduate; PG = postgraduate)',
    '`IQ`', 'IQ score',
) %>% 
    knitr::kable()
```

Let's read the data in R. If you download the data, you will see that the file name has extension ".csv", meaning that in the file the values are separated by commas. If you don't believe me, try opening the file in Notepad on your laptop!
As such, the file is called a "comma separated value" file and we read such data in R using the function `read_csv()` from the tidyverse package or `read.csv()` that comes out of the box with your R installation. Those functions will convert the commas into column separators and this way we will get a proper data table: 

```{r}
library(tidyverse)
acme_students <- read_csv("https://uoepsy.github.io/data/acme_student_data.csv")
head(acme_students)
```

We can find the number of cases and variables as follows:

```{r}
dim(acme_students)
```

There are many equivalent ways to inspect the variable types in the dataset. One it to look at the output of this function:

```{r}
sapply(acme_students, data.class)
```

From this we can see that `StudentID` and `DegreeType` are categorical variables (they store characters / words), while `Height` and `IQ` are numerical variables.

Another way is to look at the data structure:

```{r}
str(acme_students)
```

The data comprise 5 measurements on 30 students. As such, the number of variables is 5 and the number of cases or units is 30. The variables `StudentID`, `FirstName`, and `DegreeType` are categorical, while `Height` and `IQ` are numerical. This can be seen from the output of the structure function (`str`), showing the descriptors `chr` = character and `num` = numeric next to the variable names; you can ignore all the other lines in the output.

```{r}
summary(acme_students)
```


We can see that the heights in the sample range from 0 to 191.1 cm. A height of 0 is implausible and as such we doubt that the researchers used 0 to encode nonresponses or missing values. There do not seem to be any other values that may indicate inputting errors or missing values.
IQ scores in the sample seem to range from 71 to 127, with one missign entry which was already correctly recognised and displays as NA. The following fixes the NAs for Height:

```{r}
acme_students$Height[acme_students$Height == 0] <- NA
```

```{r}
summary(acme_students)
```

Finally, as student ID and degree type are a categorical variable, we will change their type in R:

```{r}
acme_students$StudentID <- factor(acme_students$StudentID)
acme_students$DegreeType <- factor(acme_students$DegreeType)
```

```{r}
summary(acme_students)
```


# Student Glossary

To conclude the lab, create a glossary of R functions. You can do so by opening Microsoft Word, Excel, or OneNote and creating a table with two columns: one where you should write the name of an R function, and the other column where you should provide a brief description of what the function does.

This "do it yourself" glossary is an opportunity for you to revise what you have learned in today's lab and write down a few take-home messages. You will find this glossary handy as a reference to keep next to you when you will be doing the assessed weekly quizzes.



---

# References

_Data adapted from:_ Nordmann, E., McAleer, P., Toivo, W., Paterson, H. & DeBruine, L. (2022). Data visualisation using R, for researchers who don't use R. Advances in Methods and Practices in Psychological Science. https://doi.org/10.1177/25152459221074654



<!-- Page style -->

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
