---
title: "Categorical data"
subtitle: "Semester 1 - Week 2"
callout-appearance: simple
---


```{r setup}
#| include: false

source('assets/setup.R')
library(tidyverse)
```



## Formative report A


::: {.callout-important}
Instructions and data were released in [week 1](1_01_design_and_data.html).
:::


### This week's task

**Task A2**

> A2) Display and describe the categorical variables


**Sub-steps**

Below there are sub-steps you need to consider to complete this week's task.

::: {.column-margin}
::: {.callout-tip}
#### Tip

To see the hints, hover your cursor on the superscript numbers.
:::
:::

- Reopen last week's Rmd file, as you will continue last week's work and build on it.^[
Hint: access the Rmd file from the Group Discussion Space. <br>If last week's driver hasn't uploaded it yet, please ask them to share it with the group via the Group Discussion Space, email, or Teams.
]

:::{.callout-note collapse="true"}
#### Selecting a subset of columns

Consider a table of toy data comprising a participant identifier (id: 1 to 5), the participant age, the course (A or B) they are enrolled into, and their height:

```{r}
# This code creates some toy data. 
#   tibble() creates a dataset
#   each column is specied as column_name = values
#   the function c() is used to concatenate the values going into a column
toy_data <- tibble(
    id = 1:5,
    age = c(18, 20, 25, 22, 19),
    course = c("A", "B", "A", "B", "A"),
    height = c(171, 180, 168, 193, 174)
)
toy_data
```

To select columns to keep you can either (1) specify the range `from`:`to`, if the columns are sequential, or (2) list the columns one by one.

::: {.panel-tabset}
##### Range `from`:`to`

If the columns you want to keep are sequential, you can just specify the first and last by using numbers:

```{r}
toy_data |>
    select(1:3)
```

or using their names:

```{r}
toy_data |>
    select(id:course)
```

Either option keeps columns `id` up to `course`.

##### Listing all columns

If the columns you want to keep are not in sequential order, you have to list all of the columns you want to keep. This can be tedious if you have many. 

You can do so using numbers:

```{r}
toy_data |>
    select(1, 2, 3)
```

Or column names:

```{r}
toy_data |>
    select(id, age, course)
```

:::

However, if you check the data in `toy_data`, those didn't change. The result of the above computation was only printed to the screen but not stored. 

```{r}
toy_data
```

To store it, we need to assign the result to an object. By using the same name `toy_data` we overwrite the data:

```{r}
toy_data <- toy_data |>
    select(id:course)
```

```{r}
toy_data
```

By doing the above, we have overwritten the data stored in `toy_data` with the selected columns.
:::

- In Formative Report A you will only work with the variables (i.e., columns) `Movie` up to, and including, `Year`. Overwrite the data to only include the first 15 variables.^[Hint: the `select()` function from `tidyverse`.<br>For an explanation of the function, did you read the drop down "Selecting a subset of columns"?]

- Create a plot displaying the frequency distribution of movie genres.[^geom-bar]

[^geom-bar]: Hint: we display categorical variables with barplots. Consider the `geom_bar()` function.

    Example: For the `toy_data` from above, the frequency distribution of course enrollment is:

    ```{r}
    #| code-line-numbers: true
    #| eval: false
    ggplot(toy_data, aes(x = course)) +
        geom_bar() +
        labs(x = "Enrollment per course", y = "Frequency")
    ```
    
    Line 1 sets the plotting canvas: it tells R to make a plot of the data `toy_data` and to put the column `course` on the x-axis of the plot.  
    Line 2 tells R to plot the data as a frequency barplot.  
    Line 3 provides user-friendly labels for the x-axis and y-axis.
    
    ```{r}
    #| echo: false
    ggplot(toy_data, aes(x = course)) +
        geom_bar() +
        labs(x = "Enrollment per course", y = "Frequency")
    ```
    
- Create a plot displaying the frequency distribution of the lead studios.^[
Hint: similar to above, change the column to `LeadStudio`.
]

- _Thinking question:_ Would it make sense to plot the frequency distribution of movie titles (`Movie`)?^[
Hint: what would be the height of each bar? Would adding such plot to a report give any useful insights to decision makers?<br><br>
In the data, `Movie` stores the movie titles. This variable is what is known in statistics as an "identifier" or "ID" variable as it uniquely identifies each unit in the study. 
If your study involved several participants, your ID would be the unique participant identifier. 
Plotting the frequency distribution of an identifier variable doesn't convey insights or summarise the data as all vertical bars in the frequency plot will have height equal 1.
]

::: {.callout-tip}
#### Tip

Before applying a function to your data, you should always ask yourself if what you are about to do is going to convey insights about the data, as opposed to directly looking at the data.  
The goal of data analysis is to to go from a multitude of values to insights that provide actionable information.
:::

- Describe the distribution of movie genres. You may want to include both the frequency and the percentage frequency.[^movies-freq-distr]

[^movies-freq-distr]: Hint: We describe categorical variables with frequency distributions.

    Consider using the `count()` function from tidyverse and `mutate()` for adding percentages.
    
    Example:
    
    ```{r}
    toy_data |>
        count(course) |>
        mutate(
            Percent = round(n / sum(n) * 100, digits = 2)
        )
    ```
    
    __Advanced__: `count(course)` is equivalent to `group_by(course) |> summarise(n = n())`. See the box below for more details.

:::{.callout-note collapse="true"}
#### An alternative to `count()`

Consider the code below, which creates a table of absolute frequencies (or counts):

```{r}
toy_data |>
    count(course)
```

An alternative to the above involves using the `group_by()`, `summarise()`, and `n()` functions from `tidyverse`:

```{r}
#| code-line-numbers: true
toy_data |>
    group_by(course) |>
    summarise(
        n = n(),
    )
```

Line 1 takes `toy_data` and then does something to it (`|>`).  
Line 2 specifies to do any computations which follow separately for each `course` (the groups).  
Lines 3-5 summarise the data by creating a column named `n` (the name goes before the = sign) that stores the sizes of each group. The group size is returned by the tidyverse function `n()`.

:::

- Describe the distribution of lead studios. You may want to include both the frequency and the percentage frequency.^[
Hint: similar to above, but replacing `Genre` with `LeadStudio`
]

- What is the most common genre and the most common lead studio?^[
Hint: What is the mode of `Genre` and `LeadStudio`? In other words, which category in each of those frequency distributions has the highest frequency?<br><br>
Tip: You may want to order the barplots and/or frequency tables in descending order to help you identify the mode.  
For barplots, use `aes(x = fct_infreq(VARIABLE))` instead of `aes(x = VARIABLE)`. The function `fct_infreq()` orders a categorical variable according to the frequencies.  
For tables, add `|> arrange(desc(n))` at the end of the table code.
]

- Format your frequency tables properly using the `kbl()` function from the `kableExtra` package.^[Hint: See the worked example below.]

- Summarise your findings in the Analysis section. For each categorical variable, show either the frequency table or frequency plot in the Analysis section, not both. This avoids duplication of information.


<!-- Solution -->

```{r, echo=FALSE, eval=FALSE}
# week 1
library(tidyverse)
movies <- read_csv("https://uoepsy.github.io/data/hollywood_movies_subset.csv")
head(movies)
nrow(movies)
ncol(movies)
dim(movies)
glimpse(movies)
summary(movies)

# week 2
library(patchwork)

movies <- movies |>
    select(1:15)

freqGenre <- ggplot(movies, aes(x = Genre)) +
    geom_bar() +
    labs(x = "Movie genre", y = "Frequency") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
freqLeadStudio <- ggplot(movies, aes(x = LeadStudio)) +
    geom_bar() +
    labs(x = "Lead studio", y = "Frequency") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
freqGenre | freqLeadStudio

movies |> 
    count(Genre) |> 
    mutate(Perc = round(n / sum(n) * 100, digits = 2)) |> 
    arrange(desc(n))

movies |> 
    count(LeadStudio, name = 'Freq') |> 
    mutate(Perc = round(n / sum(n) * 100, digits = 2)) |> 
    arrange(desc(Freq))

```

<!-- End -->


## Worked example {#worked-example}

Consider the dataset available at <https://uoepsy.github.io/data/RestaurantTips.csv>, containing 157 observations on the following 7 variables:

```{r, echo=FALSE}
library(tidyverse)
library(kableExtra)

tribble(
    ~'Variable Name', ~'Description',
    'Bill', 'Size of the bill (in dollars)',
    'Tip', 'Size of the tip (in dollars)',
    'Credit', 'Paid with a credit card? n or y',
    'Guests', 'Number of people in the group',
    'Day', 'Day of the week: m=Monday, t=Tuesday, w=Wednesday, th=Thursday, or f=Friday',
    'Server', 'Code for specific waiter/waitress: A, B, or C',
    'PctTip', 'Tip as a percentage of the bill'
) |>
    kbl() |>
    kable_styling(full_width = FALSE)
```

These data were collected by the owner of a bistro in the US, who was interested in understanding the tipping patterns of their customers. The data are adapted from @lock2020.

```{r}
library(tidyverse)
```

:::{.column-margin}
We load the `tidyverse` package as we will use the functions `read_csv` and `glimpse` from this package.
:::

```{r}
tips <- read_csv("https://uoepsy.github.io/data/RestaurantTips.csv")
```

:::{.column-margin}
`read_csv` is the function to read CSV (comma separated values) files. Once we have read the file, it is stored into an object called tips using the arrow (`<-`).
:::

```{r}
head(tips)
```

:::{.column-margin}
`head()` shows the top 6 rows of data. Use the `n = ...` option to change the default behaviour, e.g. `head(<data>, n = 10)`.
:::

```{r}
glimpse(tips)
```

:::{.column-margin}
`glimpse` is part of the tidyverse package and is used to check the type of each variable.
:::

We can use better and more descriptive labels for the categorical variables:

```{r}
tips$Day <- factor(tips$Day, 
                   levels = c("m", "t", "w", "th", "f"),
                   labels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
```

:::{.column-margin}
`tips$Day`, i.e. the column `Day` within the data `tips`, is converted to a factor in R (the appropriate storage mode for categorical variables). Furthermore, it replaces the level "m" with the new label "Monday", "t" with the new label "Tuesday", and so on.
:::

```{r}
tips$Credit <- factor(tips$Credit, 
                      levels = c("n", "y"),
                      labels = c("No", "Yes"))
```

We don't have better labels for Server (current values A , B, or C), so we will just convert it to a factor by keeping the current levels:

```{r}
tips$Server <- factor(tips$Server)
```

Check the relabelled columns:

```{r}
glimpse(tips)
```

<!-- :::{.column-margin} -->
<!-- The categorical variable `course` in `toy_data` with levels "A" and "B" can have the values relabelled to "Year1" and "Year2" via: -->

<!-- ``` -->
<!-- toy_data$course <- factor( -->
<!--   toy_data$course,  -->
<!--   levels = c("A", "B"),  -->
<!--   labels = c("Year1", "Year2") -->
<!-- ) -->
<!-- ``` -->
<!-- ::: -->

Last week, we also saw that if someone tipped more than 100% of the bill size, it was likely a data input error and we decided to replace that value with NA (not available):

::: {.column-margin}
The `mutate` function takes as arguments:

- column name  
- =  
- how to compute that column
:::

::: {.column-margin}
The syntax for `ifelse` is:
```
ifelse(test_condition, 
       value_if_true, 
       avalue_if_false)
```
:::

```{r}
tips <- tips |>
    mutate(
        PctTip = ifelse(PctTip > 100, NA, PctTip)
    )
```

This displays the frequency distribution of credit card payers:

```{r}
plt_credit <- ggplot(tips, aes(x = Credit)) +
    geom_bar() +
    labs(x = "Paid by credit card?", y = "Count")
plt_credit
```

You can even flip the coordinates, if you wish to, using the `coord_flip()` function:

```{r}
ggplot(tips, aes(x = Credit)) +
    geom_bar() +
    labs(x = "Paid by credit card?") +
    coord_flip()
```

You can use the `patchwork` package to place graphs side by side. Simply create an object for each graph, and concatenate the objects with `|` for horizontal concatenation and `/` for vertical concatenation of graphs. You can even combine this by using parentheses, e.g. (plot1 | plot2) / (plot3 | plot4) for 2 rows and 2 columns.

::: {.column-margin}
Run `install.packages("patchwork")` first in your R console
:::

We can display the frequency distribution of all the categorical variables: Credit, Day, and Server:

::: {.callout-note collapse="true"}
## Rotate x-axis labels

To rotate x-axis labels by 90 degrees, you can use this code:  
`theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))`  
To rotate the labels by 45 degrees, you can use:
`theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))`  
Don't worry, no one remembers it. People always google "rotate x-axis labels ggplot" to find it.
:::

```{r}
#| column: page
#| fig-height: 5
#| fig-width: 12

library(patchwork)

plt1 <- ggplot(tips, aes(x = Credit)) +
    geom_bar() +
    labs(x = "Paird by credit card?", y = "Count") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

plt2 <- ggplot(tips, aes(x = Day)) +
    geom_bar() +
    labs(x = "Day of week", y = "Count") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

plt3 <- ggplot(tips, aes(x = Server)) +
    geom_bar() +
    labs(y = "Count") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

plt1 | plt2 | plt3
```


:::{.callout-note collapse="true"}
### Sorting the barplot

If wanted, you can sort the bars in order of frequency by using the `fct_infreq()` function. 

In the last plot, plt3, this involves changing the first row from `ggplot(tips, aes(x = Server))` to `ggplot(tips, aes(x = fct_infreq(Server)))`.

In these plot I have preferred not to do so, as changing the order of levels may confuse the reader when the factors have easily understood ordering: credit (No/Yes), day (Mon,Tue,Wed,Thu,Fri), server (A,B,C)

```{r}
#| fig-height: 5
#| fig-width: 12

library(patchwork)

plt1 <- ggplot(tips, aes(x = fct_infreq(Credit))) +
    geom_bar() +
    labs(x = "Paird by credit card?", y = "Count") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

plt2 <- ggplot(tips, aes(x = fct_infreq(Day))) +
    geom_bar() +
    labs(x = "Day of week", y = "Count") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

plt3 <- ggplot(tips, aes(x = fct_infreq(Server))) +
    geom_bar() +
    labs(x = "Server", y = "Count") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

plt1 | plt2 | plt3
```
:::


A frequency table can be obtained using:

```{r}
tbl_credit <- tips |>
    count(Credit) |>
    mutate(
        Percent = round((n / sum(n)) * 100, digits = 2)
    )
tbl_credit
```

```{r}
tbl_day <- tips |>
    count(Day) |>
    mutate(
        Percent = round((n / sum(n)) * 100, digits = 2)
    )
tbl_day
```

```{r}
tbl_server <- tips |>
    count(Server) |>
    mutate(
        Percent = round((n / sum(n)) * 100, digits = 2)
    )
tbl_server
```

You can create nice tables with the `kbl` command from the `kableExtra` package.

:::{.column-margin}
Run `install.packages("kableExtra")` first in your R console
:::

```{r, eval=FALSE}
library(kableExtra)

kbl(list(tbl_credit, tbl_day, tbl_server), booktabs = TRUE)
```

<!-- Test -->

```{r}
#| tbl-cap: Frequency tables of categorical variables
#| tbl-subcap: 
#|   - "Paid with a credit card"
#|   - "Day of the week"
#|   - "Server"
#| layout-ncol: 3
#| echo: false

kbl(tbl_credit) |> kable_styling(full_width = FALSE)
kbl(tbl_day) |> kable_styling(full_width = FALSE)
kbl(tbl_server) |> kable_styling(full_width = FALSE)
```

<!-- End test -->

:::{.callout-note collapse="true"}
#### How do I arrange by descending frequency order?

Add `arrange(desc(<column_of_freq>))`. For example:

```{r}
tbl_day <- tips |>
    count(Day) |>
    mutate(
        Percent = round((n / sum(n)) * 100, digits = 2)
    ) |>
    arrange(desc(n))
tbl_day
```

If you just did `arrange(n)`, it would be in ascending order.

:::

:::{.callout-note collapse="true"}
#### How do I rename the frequency and percent columns?

You can specify a different name for the column of counts by using `name = "new name"`. If you don't specify it, the default is `n`.

You can specify any valid name for the percentages inside of mutate.

For example:

```{r}
tbl_day <- tips |>
    count(Day, name = "Freq") |>
    mutate(
        Perc = round((Freq / sum(Freq)) * 100, digits = 2)
    ) |>
    arrange(desc(Freq))
tbl_day
```

:::

From the univariate distribution (or marginal distribution) of each categorical variable we see that the most common payment method was not a credit card, and the most common day of the week to dine at that restaurant was Wednesday, followed by Thursday and Friday. Finally, most parties were waited on by server B. 

::: {.column-margin}
The __mode__ of a variable is the value that appears most often.

The term comes from the French expression "à la mode", i.e. in fashion. If you think about it, something is considered to be in fashion if it's worn very often.
:::


::: {.callout-tip collapse="true"}
### Referencing tables

To reference a table in text you first give the code chunk a unique label, e.g. `tableLabel`, and a caption to the table, e.g. "My table caption is this"

````
```{{r tableLabel}}
tbl_credit |>
    kbl(digits = 2, booktabs = TRUE, caption = "My table caption is this")
```
````

This creates

```{r tableLabel, echo=FALSE}
#| label: tbl-label
#| tbl-cap: "My table caption is this"
tbl_credit |>
    kbl(digits = 2, caption = "My table caption is this") |>
    kable_styling(full_width = FALSE)
```

Then you reference it in text using `\@ref(tab:tableLabel)`. For example:

> `Table \@ref(tab:tableLabel) displays etc.`

Which renders as:

@tbl-label displays etc.


:::

## Student Glossary

To conclude the lab, add the new functions to the glossary of R functions that you started last week. 

| Function    | Use and package                                                                                                                                                                                                 |
|:-------------|:-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| `factor`      | Base R function for creating and managing categorical variables. Converts text or numbers into a categorical variable with specific levels.                                                                     |
| `|>`          | The pipe operator (base R) that passes the output from one function to the input of another. Makes code more readable by avoiding nested functions. You might encounter an older version of this, `%>%`, which works the same way and comes from the tidyverse package.                                               |
| `geom_bar`    | Function from tidyverse that creates bar plots. Used to display counts of categorical variables.                                                              |
| `labs`        | Function from tidyverse that adds or modifies labels on plots (titles, axis labels, etc.).                                                                      |
| `count`       | Function from tidyverse that counts observations by group. Simplifies the process of creating frequency tables.                                                  |
| `mutate`      | Function from tidyverse that creates new variables or modifies existing ones.                                                    |
| `sum`         | Base R function that adds numbers together. Can handle vectors and ignore missing values with na.rm = TRUE.                                                              |
| `round`       | Base R function that rounds numbers to a specified number of decimal places.                                                                                             |
| `coord_flip`  | Function from tidyverse that switches the x and y axes. Useful for making horizontal bar charts.                                                              |
| `kbl`         | Function from kableExtra package that creates nicely formatted tables for reports and presentations.                                                                     |
| `arrange`     | Function from tidyverse that sorts data rows based on values in specified columns.                                                                               |
| `desc`        | Function from tidyverse used inside arrange() to sort in descending order rather than the default ascending order.                                               |

