---
title: "Formatting resources"
---


```{r setup}
#| include: false

source('assets/setup.R')
library(tidyverse)
```


This page contains formatting resources that will help you finalise your report formatting prior to submission.



## Successful knitting checklist

If you encounter errors when knitting the Rmd file, go through the following checklist to try finding the source of the errors.

- [Successful knitting checklist](https://uoepsy.github.io/dapr1/2526/useful_files/checklist_knitting.pdf)



## APA style reporting

Check the following guide for reporting numbers and statistics in APA style (7th edition).

- [Numbers and statistics guide, APA style 7th edition](https://apastyle.apa.org/instructional-aids/numbers-statistics-guide.pdf)



## Hiding code and/or output

:::{.panel-tabset}
### Hide R code

To not show the code of an R code chunk, and only show the output, write:

````
```{{r, echo=FALSE}}
# code goes here
```
````

### Hide R output

To show the code of an R code chunk, but hide the output, write:

````
```{{r, results='hide'}}
# code goes here
```
````

To hide both text output and figures, use:

````
```{{r, results='hide', fig.show='hide'}}
# code goes here
```
````

### Hide both R code and output

To hide both code and output of an R code chunk, write:

````
```{{r, include=FALSE}}
# code goes here
```
````
:::



## Figures


### Combining plots

You can combine multiple plots into a single figure using the functions `|` and `/` from `library(patchwork)`.

Suppose you stored four plots into objects named `plt1, plt2, plt3, plt4`. To combine the four plots into a single figure with 2 rows and 2 columns, you can use:

```
(plt1 | plt2) / (plt3 | pl4)
```

The vertical bar `|` places figures side by side. The forward slash `/` starts a new row.


### Reducing figure size

You can adjust the figure height and width with the following code chunk options: `fig.height = ?, fig.width = ?`.  
Trying different numbers by trial-and-error, substitute numbers where the `?` are, and adjust as needed: for example, 5 and 4, 12 and 8, etc. Always keep in mind, that the figure labels should still be legible in your resulting plots.

````
```{{r, fig.height = 5, fig.width = 4}}
# your code to display the figure here
```
````


### Referencing figures

Step 1. Create a **unique** label for the code chunk that displays the figure, in this case `UniqueFigureLabel` but you should use a more descriptive name.

````
```{{r UniqueFigureLabel, fig.cap = "Write a figure caption here"}}
library(tidyverse)
library(palmerpenguins)   # for the penguins example data we are using here
pltSpecies <- ggplot(penguins, aes(x = species)) + 
    geom_bar()
pltSpecies
```
````

```{r UniqueFigureLabel, fig.cap = "Write a figure caption here", echo=FALSE, fig.height=4, fig.width=4}
library(tidyverse)
library(palmerpenguins)   # for the penguins example data we are using here
pltSpecies <- ggplot(penguins, aes(x = species)) + 
    geom_bar()
pltSpecies
```


To reference a figure in the Rmd file, for example the one above, you would write:

> ```
> Figure \@ref(fig:UniqueFigureLabel) displays...
> ```

which, when you Knit to PDF, becomes:

> Figure 1 displays...



## Tables

We will use example data from `library(palmerpenguins)` about the `body_mass_g` of different `species` of penguins.

### Pretty PDF tables

Suppose you created a descriptives table and stored it into `tbl`:

```{r}
library(tidyverse)
library(palmerpenguins)   # for the penguins example data we are using here

tbl <- penguins |>
    group_by(species) |>
    summarise(M = mean(body_mass_g, na.rm = TRUE),
              SD = sd(body_mass_g, na.rm = TRUE))
tbl
```

It wouldn't be appropriate to show this R printout in a paper/report/dissertation. Instead, you can create a properly formatted table for a PDF report using the `kbl()` or `kable()` functions from `library(kableExtra)`.

````
```{{r, echo=FALSE}}
library(kableExtra)
tbl |>
    kbl(digits = 2, booktabs = TRUE, 
        caption = "Write a table caption here")
```
````

```{r echo=FALSE}
library(kableExtra)
tbl |>
    kbl(digits = 2, 
        caption = "Write a table caption here") |>
    kable_styling(full_width = FALSE)
```

The provided options are:

- `digits = 2`, to format numbers to only have 2 decimal places
- `booktabs = TRUE`, to create "book-style" tables, i.e. tables with horizontal rows only
- `caption = "Write a table caption here"`, to provide a table caption


### Referencing tables

This is a continuation of the previous example, where we created a descriptives table and stored it into `tbl`. Please note that `tbl` is not a `kbl()`/`kable()` yet.

To reference a table you need to pick a **unique** label for the code chunk that displays the table, in this case `UniqueTableLabel` but you should use a more descriptive name. 

You must also ensure that the table has a caption for the referencing to work.

````
```{{r UniqueTableLabel, echo=FALSE}}
library(kableExtra)
tbl |>
    kbl(digits = 2, booktabs = TRUE, 
        caption = "Write here a table caption")
```
````

```{r echo=FALSE}
library(kableExtra)
tbl |>
    kbl(digits = 2, 
        caption = "Write here a table caption") |>
    kable_styling(full_width = FALSE)
```


In the Rmd file, the table is referenced as:

> ```
> Table \@ref(tab:UniqueTableLabel) displays...
> ```

which, when you knit to PDF, is displayed as:

> Table 1 displays...

For details on styling PDF tables, see [this link](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_pdf.pdf).


