---
title: "example_lec"
---

# Course Overview {.smaller}

:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block1_name = "block  1 name"
block1_lecs = c("week 1 topic",
                "week 2 topic",
                "week 3 topic",
                "week 4 topic",
                "week 5 topic")
block2_name = "block 2 name"
block2_lecs = c("week 7 topic",
                "week 8 topic",
                "week 9 topic",
                "week 10 topic",
                "week 11 topic")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")

course_table(block1_name,block2_name,block1_lecs,block2_lecs,week=2)
```

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block3_name = "block 3 name"
block3_lecs = c("week 1 topic",
                "week 2 topic",
                "week 3 topic",
                "week 4 topic",
                "week 5 topic")
block4_name = "block 4 name"
block4_lecs = c("week 7 topic",
                "week 8 topic",
                "week 9 topic",
                "week 10 topic",
                "week 11 topic")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")

course_table(block3_name,block4_name,block3_lecs,block4_lecs,week=0)
```

:::

::::

## This week

1. quick recap
2. straight lines
3. outcome = model + error
4. using models to answer questions

# the story so far... 

## block 1


## block 2


## block 3
testing relationships two variables - associations

# straight lines

## correlation

correlation - two continuous variables

## correlation 

different correlations - patterns

## correlation ++ 

prev slide but with lines

## what is a line?

a line is an intercept and a slope - where start, how much does it increase?

# shifting focus

## a line is a model

modelling an outcome

## the only equation you'll ever need

outcome = model + error

## the only equation you'll ever need (2)

1. deterministic vs statistical
2. terminology: outcome/response/DV , predictor/explanatory/IV

## the linear regression equation

outcome = model + error
y = b0 + b1 x + error

## lines

slide correlation++ with values of b0, b1 on them

## aside

how does the model get fitted? 
	1. minimise resid SS (just an animation to show the idea of RSS differing depending on line)

# Activity break! 
quiz - b0 b1 neg/pos, values etc. 

# using models to answer questions

## some models are useful

	1. correlation describes a pattern but lm is a model you can *use*
	2. use how? to estimate a quantity of interest! 
	
## example: continuous RCT. 

	1. wordle_score ~ sleep deprivation. randomised 1 to 7 days sleep disturbance
	2. correlation = when X is higher, Y is higher

## example: continuous RCT:

	1. model predictions (use faces as units of interest - see old dapr3 refresh)
		1. wordle score with 1 day of disrupted sleep: hat y = b0 + b1 (x=1)
		2. wordle score with 1 week (7 days) of disrupted sleep: hat y = b0 + b1 (x=7)
		3. difference that an extra day makes: hat y(1) - hat y(0) 
			1. oh look, it's the coefficient b1! 

## the how to (fitting models in R)
	1. formula notation  outcome ~ model
	2. lm(wordle ~ sleep_dep)

## the how to (model predictions in R)

	1. predict.lm




# End


