---
title: "Lines as Models"
editor_options: 
  chunk_output_type: console
---


```{r}
#| label: setup
#| echo: false
library(tidyverse)
library(gganimate)
library(patchwork)
```


# Course Overview {.smaller}
:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block1_name = "block  1 name"
block1_lecs = c("week 1 topic",
                "week 2 topic",
                "week 3 topic",
                "week 4 topic",
                "week 5 topic")
block2_name = "block 2 name"
block2_lecs = c("week 7 topic",
                "week 8 topic",
                "week 9 topic",
                "week 10 topic",
                "week 11 topic")

source("../../../../uoepsy_other/junk/R/course_table.R")

course_table(block1_name,block2_name,block1_lecs,block2_lecs,week=11)
```

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block3_name = "block 3 name"
block3_lecs = c("week 1 topic",
                "week 2 topic",
                "week 3 topic",
                "week 4 topic",
                "week 5 topic")
block4_name = "block 4 name"
block4_lecs = c("week 7 topic",
                "week 8 topic",
                "week 9 topic",
                "week 10 topic",
                "week 11 topic")

source("../../../../uoepsy_other/junk/R/course_table.R")

course_table(block3_name,block4_name,block3_lecs,block4_lecs,week=7)
```

:::

::::

## This week

1. where my p-values at?
2. tests and confidence intervals for the intercept and slope
3. the other bits.. 

# Where did we get to last week?

## the linear model {.smaller}

::::{.columns}
:::{.column width="50%"}
$$
\begin{align}
\color{red}{\text{outcome}} &= \color{blue}{\text{model}} &+& \quad \quad \color{black}{\text{error}} \\
\quad \\
\quad \\
\quad \\
\color{red}{y} &= \color{blue}{\underset{\underset{\text{intercept}}{\uparrow}}{b_0} + \underset{\underset{\text{slope}}{\uparrow}}{b_1} \cdot x} & + & \quad \quad \color{black}{\epsilon}
\end{align}
$$
:::
:::{.column width="50%" .fragment fragment-index=1}
```{r}
#| include: false
somedata <- tibble(
  x = round(runif(10,4,20)),
  y = round(rnorm(10,40+1*x,4))
)
# egfit <- lm(y~x,somedata)
# egsegs <- tibble(x1 = 0:6, x2 = x1+1) |> 
#   mutate(
#     y1 = map_dbl(x1,~(c(1,.)%*%coef(egfit))[1]),
#     y2 = map_dbl(x2,~(c(1,.)%*%coef(egfit))[1]),
#     xlp = (x1+x2)/2,
#     ylp = (y1+y2)/2
#   )
# 
# ggplot(somedata,aes(x=x,y=y))+
#   geom_point(alpha=.1,size=3)+
#   stat_smooth(method=lm,se=FALSE,color="#0000FF",lwd=1) +
#   geom_point(x=0,y=coef(egfit)[1],col="#0000FF",size=5)+
#   geom_segment(data = egsegs, aes(x=x1,xend=x2,y=y1,yend=y1),
#                lty="dashed",col="#0000FF")+
#   geom_segment(data = egsegs, aes(x=x2,xend=x2,y=y1,yend=y2),
#                lty="dashed",col="#0000FF") 
```
```{r}
#| echo: true
#| eval: false
my_model <- lm(y ~ x, data = somedata)

my_model
```
```
Call:
lm(formula = y ~ x, data = somedata)

Coefficients:
(Intercept)            x  
     ....           ....  
```
:::
::::


## using models {.smaller}

```{r}
#| echo: false
set.seed(65683728)
sleepwordle <- tibble(
  n_disrupted = rep(0:7,e=10),
  time = rgamma(80,(n_disrupted+1)/2,1)*60,
  guesses = rbinom(80, 5, plogis(.6+.5*scale(n_disrupted)))+1,
  wordle = time + (60*guesses)
) |> slice_sample(prop=1)
```



> **RQ:** How does an additional night of interrupted sleep affect cognitive skills?

What quantity will answer my question?  
  - expected change in Score as the night of disrupted sleep increases by 1.  


::::{.columns}
:::{.column width="50%"}
```{r}
head(sleepwordle)

sleep_mdl <- lm(wordle ~ n_disrupted, data = sleepwordle)

sleep_mdl
```
:::
:::{.column width="50%"}

```{r}
#| echo: false
fitw = lm(wordle~n_disrupted,sleepwordle)

segsw <- tibble(x1 = 0:6, x2 = x1+1) |> 
  mutate(
    y1 = map_dbl(x1,~(c(1,.)%*%coef(fitw))[1]),
    y2 = map_dbl(x2,~(c(1,.)%*%coef(fitw))[1]),
    xlp = (x1+x2)/2,
    ylp = (y1+y2)/2
  )
sleepwordle |> 
  ggplot(aes(x=n_disrupted,y=wordle))+
  geom_point(size=3,col="tomato1",alpha=.5)+
  stat_smooth(method=lm,se=FALSE,color="#0000FF",lwd=1) +
  geom_point(x=0,y=coef(fitw)[1],col="#0000FF",size=5)+
  geom_segment(data = segsw, aes(x=x1,xend=x2,y=y1,yend=y1),
               lty="dashed",col="#0000FF")+
  geom_segment(data = segsw, aes(x=x2,xend=x2,y=y1,yend=y2),
               lty="dashed",col="#0000FF") 
```

:::
::::


:::{.aside}
We're basing these predictions on groups of _different_ people. i.e., to answer our question we're comparing: 
  person_A with 4 nights sleep dep 
  vs 
  personB with 5 nights sleep dep

this only targets our quantity _if_ we can think of persons A and B as exchangeable!  
the random allocation of people to number of nights disrupted gives us this!  
:::


# where'd the tests go?  

## summary()

```{r}
summary(sleep_mdl)
```


## coefficients

the intercept and slope (b0, b1)
or, as we've seen
- intercept is predict(x=0)
- slope of x is predict(x=x+1)-predict(x=x)

## what are the hypotheses being tested?  

**The intercept:**  
$$H_0: b_0 = 0 \quad\quad\quad\quad H_1: b_0 \neq 0$$   

Often not that interesting.  
Why would we ever believe $H_0$ in first place?  

**The slope:**  

$$H_0: b_1 = 0 \quad\quad\quad\quad H_1: b_1 \neq 0$$   
$b_1$ corresponds to the quantity we're interested in!  

## under the null hypothesis

- pretend $b_1 = 0$.  
- what would lines from 80 people look like in that universe?  

:::{.fragment fragment-index=1}
- what did we actually see?  
:::

:::{.r-stack}
```{r}
#| echo: false
sims = MASS::mvrnorm(n=1000,mu=c(coef(sleep_mdl)[1],0),
                     Sigma = vcov(sleep_mdl)) |>
  as.data.frame() |>
  rename(int = `(Intercept)`,slp = V2)

ggplot(sleepwordle, aes(x = n_disrupted,y=wordle))+
  geom_point(alpha=0)+
  geom_abline(data = sims, aes(intercept = int, slope = slp), alpha=.05)+
  ylim(150,700)
```
:::{.fragment fragment-index=1}
```{r}
#| echo: false
sims = MASS::mvrnorm(n=1000,mu=c(coef(sleep_mdl)[1],0),
                     Sigma = vcov(sleep_mdl)) |>
  as.data.frame() |>
  rename(int = `(Intercept)`,slp = V2)

ggplot(sleepwordle, aes(x = n_disrupted,y=wordle))+
  geom_point(alpha=0)+
  geom_abline(data = sims, aes(intercept = int, slope = slp), alpha=.05)+
  geom_abline(intercept=coef(sleep_mdl)[1],slope=coef(sleep_mdl)[2],
              col="#0F4C81",lwd=1)+
  ylim(150,700)
```
:::
:::

## a confidence interval

Using that same estimated uncertainty:  

```{r}
#| code-fold: true
#| eval: false
ggplot(sleepwordle, aes(x = n_disrupted,y=wordle))+
  geom_smooth(method = lm)+
  ylim(150,700)
```

:::{.r-stack}

```{r}
#| echo: false
sims = MASS::mvrnorm(n=1000,mu=coef(sleep_mdl),
                     Sigma = vcov(sleep_mdl)) |>
  as.data.frame() |>
  rename(int = `(Intercept)`,slp = n_disrupted)

ggplot(sleepwordle, aes(x = n_disrupted,y=wordle))+
  geom_point(alpha=0)+
  geom_abline(data = sims, aes(intercept = int, slope = slp), alpha=.05)+
  geom_abline(intercept=coef(sleep_mdl)[1],slope=coef(sleep_mdl)[2],
              col="#0F4C81",lwd=1)+
  ylim(150,700)
```

:::{.fragment fragment-index=1}
```{r}
#| echo: false
ggplot(sleepwordle, aes(x = n_disrupted,y=wordle))+
  geom_smooth(method = lm)+
  geom_abline(intercept=coef(sleep_mdl)[1],
              slope=coef(sleep_mdl)[2],
              col="blue",lwd=1)+
  ylim(150,700)
```
:::

# activity break


# the other bits

## ...

```{r}
summary(sleep_mdl)
```

## all the variance in the outcome

```{r}
#| echo: false

sleepwordle |>
  mutate(
         my = mean(wordle),
         f1 = mean(wordle),
         f2 = predict(sleep_mdl),
         xjit = rnorm(n(),n_disrupted,.1)) |>
  pivot_longer(f1:f2,names_to="frame",values_to="pred") |>
  filter(frame=="f1") |>
  ggplot(aes(x=xjit,y=wordle))+
  geom_point(size=3)+
  geom_hline(yintercept = mean(sleepwordle$wordle),
             col="blue",lwd=1,alpha=1,lty="dashed")+
  # geom_abline(intercept=coef(sleep_mdl)[1],
  #             slope=coef(sleep_mdl)[2],
  #             col="blue",lwd=1)+
  geom_segment(aes(x=xjit,xend=xjit,
                   y=pred,yend=my),
               col="black",lty="dotted",lwd=.75)+
  geom_segment(aes(x=xjit,xend=xjit,
                   y=pred,yend=wordle),
               col="tomato1",lty="dotted",lwd=.75)
```


plot flat line, resids
line is just the mean. so it's sort of a model, but the simplest model possible. 

all the variability in the outcome is distance back to mean

## explaining variance in the outcome

```{r}
p_anim <-
  sleepwordle |>
  mutate(
         my = mean(wordle),
         f1 = mean(wordle),
         f2 = predict(sleep_mdl),
         xjit = rnorm(n(),n_disrupted,.1)) |>
  pivot_longer(f1:f2,names_to="frame",values_to="pred") |>
  mutate(int = ifelse(frame=="f1",my,coef(sleep_mdl)[1]),
         slp = ifelse(frame=="f1",0,coef(sleep_mdl)[2])) |>
  #filter(frame=="f2") |>
  ggplot(aes(x=xjit,y=wordle))+
  geom_point(size=3)+
  geom_hline(yintercept = mean(sleepwordle$wordle),
             col="blue",lwd=1,alpha=1,lty="dashed")+
  geom_abline(aes(intercept=int,slope=slp),
              col="blue",lwd=1)+
  geom_segment(aes(x=xjit,xend=xjit,
                   y=pred,yend=my),
               col="black",lty="dotted",lwd=.75)+
  geom_segment(aes(x=xjit,xend=xjit,
                   y=pred,yend=wordle),
               col="tomato1",lty="dotted",lwd=.75)+
  theme_minimal() +
  transition_states(frame, transition_length = 2, state_length = 1)

animp <- animate(p_anim, nframes = 180,
        fps = 15, end_pause = 45, res=150,
        width = 800, height = 500, 
        renderer = gifski_renderer())
```

regression line. 
resids are smaller. 

explained/model 
unexplained/residual 

## R2

total = model + residual

model explained variance / total variance

"proportion of variance in Y accounted for by X"  

## the test bit... 

explained variance / 2 things (int + slope)
vs
residual variance / all the other bits (n-2)

if model isn't accounting for anything more than randomness would, then 1
if model is, then >1

## note it's all the same!  

slope changing animation, resids shrinking animation 

even p-value is identical

## so what's the point

in this case 

outcome ~ predictor

there isn't one.

in future cases (dapr2)

outcome ~ predictor1 + predictor2

it might sometimes become more useful

## quiz q

what is R2 for lm(y~1)? 

