---
title: "Lines as Models"
editor_options: 
  chunk_output_type: console
---


```{r}
#| label: setup
#| echo: false
library(tidyverse)
library(gganimate)
library(patchwork)
```


# Course Overview {.smaller}
:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block1_name = "block  1 name"
block1_lecs = c("week 1 topic",
                "week 2 topic",
                "week 3 topic",
                "week 4 topic",
                "week 5 topic")
block2_name = "block 2 name"
block2_lecs = c("week 7 topic",
                "week 8 topic",
                "week 9 topic",
                "week 10 topic",
                "week 11 topic")

source("../../../../uoepsy_other/junk/R/course_table.R")

course_table(block1_name,block2_name,block1_lecs,block2_lecs,week=11)
```

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block3_name = "block 3 name"
block3_lecs = c("week 1 topic",
                "week 2 topic",
                "week 3 topic",
                "week 4 topic",
                "week 5 topic")
block4_name = "block 4 name"
block4_lecs = c("week 7 topic",
                "week 8 topic",
                "week 9 topic",
                "week 10 topic",
                "week 11 topic")

source("../../../../uoepsy_other/junk/R/course_table.R")

course_table(block3_name,block4_name,block3_lecs,block4_lecs,week=6)
```

:::

::::

## This week

1. where my p-values at?
2. NHS tests and confidence intervals for the intercept and slope
3. the other bits.. 

# Where did we get to last week?

## outcome = model + error

model equation

in R

## using models

what quantity will answer my question?  
e.g. 

> **RQ:** How does an additional day of interrupted sleep affect cognitive skills?

predicted change in Y when X increases by 1



# where'd the tests go?  

## summary()

there's lots of stuff

## coefficients

the intercept and slope (b0, b1)
or, as we've seen
- intercept is predict(x=0)
- slope of x is predict(x=x+1)-predict(x=x)

## the hypotheses tested

$H_0: b_0 = 0 \,\,\,\, H_1: b_0 \neq 0$   
probably not that interesting?  
why would we ever assume $H_0$ in first place?  

$H_0: b_1 = 0 \,\,\,\, H_1: b_1 \neq 0$
testing the 

## under the null hypothesis:  

```{r}
sims = MASS::mvrnorm(n=1000,mu=c(coef(sleep_mdl)[1],0),
                     Sigma = vcov(sleep_mdl)) |>
  as.data.frame() |>
  rename(int = `(Intercept)`,slp = V2)

ggplot(sleepwordle, aes(x = n_disrupted,y=wordle))+
  geom_point(alpha=0)+
  geom_abline(data = sims, aes(intercept = int, slope = slp), alpha=.05)+
  geom_abline(intercept=coef(sleep_mdl)[1],slope=coef(sleep_mdl)[2],
              col="#0F4C81",lwd=1)+
  ylim(150,700)
```

## a confidence interval

uses that same estimated variability:  

```{r}
sims = MASS::mvrnorm(n=1000,mu=coef(sleep_mdl),
                     Sigma = vcov(sleep_mdl)) |>
  as.data.frame() |>
  rename(int = `(Intercept)`,slp = n_disrupted)

ggplot(sleepwordle, aes(x = n_disrupted,y=wordle))+
  geom_point(alpha=0)+
  geom_abline(data = sims, aes(intercept = int, slope = slp), alpha=.05)+
  geom_abline(intercept=coef(sleep_mdl)[1],slope=coef(sleep_mdl)[2],
              col="#0F4C81",lwd=1)+
  ylim(150,700)
```


```{r}
ggplot(sleepwordle, aes(x = n_disrupted,y=wordle))+
  geom_smooth(method = lm)+
  ylim(150,700)
```

# activity break


# the other bits

## all the variance in the outcome

plot flat line, resids
line is just the mean. so it's sort of a model, but the simplest model possible. 

all the variability in the outcome is distance back to mean

## explaining variance in the outcome

regression line. 
resids are smaller. 

explained/model 
unexplained/residual 

## R2

total = model + residual

model explained variance / total variance

"proportion of variance in Y accounted for by X"  

## the test bit... 

explained variance / 2 things (int + slope)
vs
residual variance / all the other bits (n-2)

if model isn't accounting for anything more than randomness would, then 1
if model is, then >1

## note it's all the same!  

slope changing animation, resids shrinking animation 

even p-value is identical

## so what's the point

in this case 

outcome ~ predictor

there isn't one.

in future cases (dapr2)

outcome ~ predictor1 + predictor2

it might sometimes become more useful

## quiz q

what is R2 for lm(y~1)? 

