---
title: "Binary Predictors"
editor_options: 
  chunk_output_type: console
---


```{r}
#| label: setup
#| echo: false
library(tidyverse)
library(gganimate)
library(patchwork)
source("_theme/helpers.R")
```


# Course Overview {.smaller}
:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block1_name = "block  1 name"
block1_lecs = c("week 1 topic",
                "week 2 topic",
                "week 3 topic",
                "week 4 topic",
                "week 5 topic")
block2_name = "block 2 name"
block2_lecs = c("week 7 topic",
                "week 8 topic",
                "week 9 topic",
                "week 10 topic",
                "week 11 topic")

source("../../../../uoepsy_other/junk/R/course_table.R")

course_table(block1_name,block2_name,block1_lecs,block2_lecs,week=11)
```

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block3_name = "block 3 name"
block3_lecs = c("week 1 topic",
                "week 2 topic",
                "week 3 topic",
                "week 4 topic",
                "week 5 topic")
block4_name = "block 4 name"
block4_lecs = c("week 7 topic",
                "week 8 topic",
                "week 9 topic",
                "week 10 topic",
                "week 11 topic")

source("../../../../uoepsy_other/junk/R/course_table.R")

course_table(block3_name,block4_name,block3_lecs,block4_lecs,week=7)
```

:::

::::

## This week

1. What if my predictor is binary?
2. Moving between means

## lines again

model equation
plot
r summary

## doing maths with words

head data


y ~ b0 + b1 * "disrupted"

computer says no... 

## expressing groups as numbers

cols: 
original, isA 1/0

## ??

example with 3 levels to show that this isn't just 
A=0,B=1,C=2,D=3... 

showing this to help understand the logic, but for dapr1 we're just dealing with binary

## R does it for us

remember, types of data in R:
numeric, factor, character

when x is characters or factors, `lm()` will do this all for you:  

## what are b0 and b1?  

show plot, x axis has labels, and underneath 0 and 1

## the underlying maths is still a line!  

show last plot with line

**but** we can't have values other than 0 and 1. 
some people don't like to think of this as a line because a line aross X implies a continuum

## moving between categories









## Block 4 skills involved in this week's lecture and/or workshop

- **A10** I can represent binary predictors using treatment coding.
- **A7** I can use a fitted simple linear model to predict an outcome value for a given predictor value
- **A8** I can appropriately interpret the parameters of a simple linear model.
- **C3** I can describe in writing the results of a given simple linear model.
- **I8** I can identify the null hypothesis being tested for the parameters of a linear model.



