---
title: "Basic Tests as Linear Models"
author: "Emma Waterston (Emma.Waterston@ed.ac.uk)"
editor_options: 
  chunk_output_type: console
format:
  revealjs:
    smaller: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)
library(kableExtra)
library(simglm)

source('_theme/theme_quarto.R')

```


# Course Overview 

<br>

:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block1_name = "Exploratory Data Analysis"
block1_lecs = c("W1",
                "W2",
                "W3",
                "W4",
                "W5")
block2_name = "Inference"
block2_lecs = c("W7",
                "W8",
                "W9",
                "W10",
                "W11")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")
course_table(block1_name,block2_name,block1_lecs,block2_lecs, week = 0)
```

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block3_name = "Basic Tests"
block3_lecs = c("W1",
                "W2",
                "W3",
                "W4",
                "W5")
block4_name = "Intro Linear Models"
block4_lecs = c("W6",
                "W7",
                "W8",
                "W9",
                "W10")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")
course_table(block3_name,block4_name,block3_lecs,block4_lecs,week=9)
```

:::

::::


## This Week's Learning Objectives

<br>


::: {style="font-size: 125%;"}

::: {.fragment}
::: {.dapr3callout}
XXXXXXXXXXXX 
:::
:::


::: {.fragment}
::: {.dapr3callout}
XXXXXXXXXXXX 
:::
:::

::: {.fragment}
::: {.dapr3callout}
XXXXXXXXXXXX 
:::
:::

::: {.fragment}
::: {.dapr3callout}
XXXXXXXXXXXX 
:::
:::

:::


# Linear Models Recap

## Intercept Only & Simple Linear Regression Models

For the last few weeks, we've fitted variants of these models:

- Intercept Only Model:

$$
y_i = \beta_0 + \epsilon_i
$$

- Simple Linear Model: 

$$
y_i = \beta_0 + \beta_1 \cdot x_{i} + \epsilon_i
$$

<br>

But how does this follow from what we learned about basic statistical tests?

# Comparing to a Known Mean

## One-Sample T-Test: Recap 

- Used when we want to test the difference in the mean scores between a sample and a known mean 

$$
t = \frac{\bar x - \mu_0}{SE_{\bar x}} = \quad  \frac{\text{sample mean} - \text{hypothesised mean value}}{\text{variation in sample mean due to random sampling}}
$$


<br>

- Sampling distribution is a $t$-distribution with $n-1$ degrees of freedom
  - $n-1$ degrees of freedom as we have 1 group of participants

*This is just an intercept only model*

## Example: Balancing App

A cognitive psychologist is developing a new app-based game to test 'balance' in older adults. They measured how well participants could maintain a virtual object on a moving surface. 

The game gives participants a score after each playing session. A score of 0 represented perfect balance, whilst scores above or below zero indicate deviations. Researchers wanted to test whether the average score of participants deviated from zero.


## Hypotheses

- I elect to use a two-tailed test with alpha $(\alpha)$ of .05

<br>

- I specify the following hypotheses:

<br>

:::: {.columns}

::: {.column width="50%"}

$$
\begin{matrix}
H_0: \mu = \mu_0\\
H_1: \mu \neq \mu_0\\
\end{matrix}
$$

:::

::: {.column width="50%"}


$$
\begin{matrix}
H_0: \beta_0 = 0\\
H_1: \beta_0 \neq 0\\
\end{matrix}
$$

:::

::::

## Data

```{r}
#| echo: false

set.seed(7284)
app_dat <- tibble(
  ID = c(paste("ID", 1:30, sep ="")),
  balance = round(rnorm(30, mean = 0, sd = 4),0)
)
```

```{r}
#| echo: true

app_dat
```

## Data Visualisation {.smaller}

```{r}
#| echo: true

ggplot(data = app_dat, aes(x = balance)) +
  geom_histogram() + 
  geom_vline(aes(xintercept = mean(balance)), color = "red", linetype = "dashed", size = 2) +
  geom_vline(aes(xintercept = 0), color = "green", linetype = "dotted", size = 2) +
  labs(title = "Histogram of Balance Scores \nwith Mean (Red Line) and Hypothesised Mean (Green Line)",
       x = "Balance Score",
       y = "Frequency")
```


## Conduct Tests **in R**

:::: {.columns}

::: {.column width="50%"}

**One Sample T-Test** 

```{r}
#| echo: true

tres1 <- t.test(app_dat$balance, 
              mu=0, 
              alternative="two.sided",
              conf.level = 0.95)
tres1
```


:::

::: {.column width="50%"}

**Intercept Only Linear Model**


```{r}
#| echo: true

mres1 <- lm(balance ~ 1, data = app_dat)
summary(mres1)

```


:::

::::


## Results

- Mean of balance is equal to the intercept -0.53
- t-statistic, df, and p-value equivalent across approach 


# Comparing Two Means

## Independent Samples T-Test: Recap 

- Used when we want to test whether the mean of one group is statistically different from/greater than/less than the mean of another group

$$
t \quad = \quad \frac{(\bar{x}_1 - \bar{x}_2) - \delta_0}{SE_{(\bar{x}_1 - \bar{x}_2)}} \quad = \quad \frac{\text{difference between sample means} - \text{hypothesised difference in means}}{\text{variation in sample mean differences due to random sampling}}
$$

<br>

- Sampling distribution is a $t$-distribution with $n-2$ degrees of freedom
  - $n-2$ degrees of freedom as we have 2 group of participants, where $n$ = $n_1 + n_2$

*This is just a simple linear model using dummy coding*

## Example: Recall

**Is recall success different depending on the memory technique used?**

- Comparing the effectiveness of two different memory techniques

- Want to determine whether the method of loci or the chunking technique is more effective in helping individuals recall a list of 25 words 

- Data is collected from 30 participants, with 15 individuals randomly assigned to each memory technique 


## Hypotheses

- I elect to use a two-tailed test with alpha $(\alpha)$ of .05

<br>

- I specify the following hypotheses:

<br>

:::: {.columns}

::: {.column width="50%"}

$$
\begin{matrix}
H_0: \mu_1 = \mu_2\\
H_1: \mu_1 \neq \mu_2\\
\end{matrix}
$$

:::

::: {.column width="50%"}

$$
\begin{matrix}
H_0: \beta_1 = 0\\
H_1: \beta_1 \neq 0\\
\end{matrix}
$$
:::

::::


## Data

```{r}
#| echo: false

Memory <- tibble(
  Method = as_factor(c(rep("MoL", 15), rep("Chunking", 15))),
  Words_Recalled = c(18, 20, 19, 21, 20, 19, 22, 20, 21, 19, 18, 20, 21, 22, 20, 15, 14, 16, 14, 15, 14, 16, 15, 14, 16, 15, 13, 16, 15, 14)
)
```

```{r}
#| echo: true

Memory
```


## Data Visualisation

```{r}
#| echo: true

ggplot(data = Memory, aes(x = Method, y = Words_Recalled, fill = Method)) +
    geom_violin() + 
    geom_jitter(alpha = 0.5) + 
    labs(x = "Method", y = "No. of Words Recalled")
```


## Conduct Tests **in R**

:::: {.columns}

::: {.column width="50%"}


**Independent Samples T-Test** 

```{r}
#| echo: true

tres2 <- t.test(Memory$Words_Recalled ~ Memory$Method, 
       mu = 0,
       var.equal = TRUE,
       conf.level = 0.95)
tres2
```



:::

::: {.column width="50%"}


**Simple Linear Model**


```{r}
#| echo: true

mres2 <- lm(Words_Recalled  ~ Method, data = Memory)
summary(mres2)

```


:::

::::

## Results

- Mean in group MoL = 20, which is equivalent to our intercept as it is the baseline/reference level
- Mean in group chunking = 14.8, which we can calculate as the difference between the intercept $\beta_0$ and $\beta_1$ (i.e., 20-5.2 = 14.8)
- t-statistic, df, and p-value equivalent across approach 

# Is There a Difference Between Pairs?

## Paired Samples T-Test: Recap 

- Used when we want to test the difference in mean scores for a sample comprising matched (or naturally related) pairs 

$$
t \quad = \quad  \frac{\overline{d} - \mu_{d_0}}{SE_\bar{d}} \quad = \quad  \frac{\text{mean difference between pairs} - \text{hypothesised mean difference}}{\text{variation in sample mean differences due to random sampling}}
$$

<br>

- Sampling distribution is a $t$-distribution with $n-1$ degrees of freedom
  - $n-1$ degrees of freedom as we have 1 group of participants


*This is just an intercept only model where the difference is our dependent variable*

## Example: Community & Attitudes

- Researchers want to assess the impact of a three-month community garden engagement programme on attitudes towards volunteerism. 

- They recruited 40 participants, who:
    - Completed a self-report attitudes toward volunteerism questionnaire, measured in March (pre-programme)
    - Participated in the community garden programme on a monthly basis for three months (April-June)
    - Completed a self-report attitudes toward volunteerism questionnaire, measured in July (post-programme)

- Attitude scores were measured on a scale of 1 to 100, with higher scores reflecting more positive attitudes towards volunteerism

- Researchers calculated the difference between post- and pre-scores


## Hypotheses

- I elect to use a two-tailed test with alpha $(\alpha)$ of .01

<br>

- I specify the following hypotheses:

<br>

:::: {.columns}

::: {.column width="50%"}

$$
\begin{matrix}
H_0: \mu_{d} = \mu_{d_0}\\
H_1: \mu_{d} \neq \mu_{d_0}\\
\end{matrix}
$$

:::

::: {.column width="50%"}

$$
\begin{matrix}
H_0: \beta_0 = 0\\
H_1: \beta_0 \neq 0\\
\end{matrix}
$$
:::

::::

## Data

```{r}
#| echo: false
set.seed(007)
volunteer <- tibble(
  PID = rep(paste("ID", 1:100, sep="")),
  pre_score = round(rnorm(100, 50, 4.2)),
  post_score = round(rnorm(100, 75, 3.7))
)

volunteer <- volunteer |>  
  mutate(
    dif = post_score - pre_score)
```

```{r}
#| echo: true

volunteer 
```


## Conduct Tests **in R**

:::: {.columns}

::: {.column width="50%"}


**Paired Samples T-Test** 

```{r}
#| echo: true

tres3 <- t.test(volunteer$dif, 
       mu = 0,
       alternative = "two.sided",
       conf.level = 0.95)

tres3
```

:::

::: {.column width="50%"}


**Intercept Only Model**


```{r}
#| echo: true

mres3 <- lm(dif ~ 1, data = volunteer)
summary(mres3)

```


:::

::::


## Results

- Mean of dif is equal to the intercept 24.99
- t-statistic, df, and p-value equivalent across approach 


# Is there a Linear Association between 2 Variables?

## Correlation: Recap 

- Used when we want to assess the direction and degree of association between two variables 

$$
r=\frac{\frac{\sum_{i=1}^n(x_i-\bar{x})(y_i-\bar{y})}{n-1}}{\sqrt{\frac{\sum_{i=1}^n(x_i-\bar{x})^2}{n-1}}\sqrt{\frac{\sum_{i=1}^n(y_i-\bar{y})^2}{n-1}}} = \frac{Cov_{xy}}{s_xs_y}
$$

+ The $t$-statistic for a correlation is calculated as:

$$
t=r\sqrt{\frac{n-2}{1-r^2}}
$$


- Sampling distribution is a $t$-distribution with $n-2$ degrees of freedom
  - $n-2$ degrees of freedom as we calculate the means of *two* variables 
  
  *This is just a simple linear model where the variables are standardised*

## Example: Social Media & Loneliness

- Researchers want to assess the influence of social media usage on feelings of loneliness in young people. 

- They recruited 50 participants, who:
    - Self-reported the number of hours spent per week on social media (`social_hrs`)
    - Completed a self-report loneliness questionnaire, measured on a scale 0-100 (`loneliness`; higher scores indicated greater feelings of loneliness)



## Hypotheses

- I elect to use a two-tailed test with alpha $(\alpha)$ of .05

<br>

- I specify the following hypotheses:

<br>

:::: {.columns}

::: {.column width="50%"}

$$
\begin{matrix}
H_0: r = 0 \\
H_1: r \neq 0 \\
\end{matrix}
$$

:::

::: {.column width="50%"}

$$
\begin{matrix}
H_0: \beta_1 = 0\\
H_1: \beta_1 \neq 0\\
\end{matrix}
$$
:::

::::

## Data

```{r}
#| echo: false

set.seed(007)

lonely <- tibble(
  PID = rep(paste("ID", 1:50, sep="")),
  loneliness = round(rnorm(50, 40, 9)),
  social_hrs = round(rnorm(50, 44, 3))) |>
  mutate(
    social_hrs_z = (social_hrs - mean(social_hrs)) / sd(social_hrs),
    loneliness_z = (loneliness - mean(loneliness)) / sd(loneliness)
  )
```

```{r}
#| echo: false

lonely
```


## Data Visualisation

```{r}
#| echo: true

ggplot(data = lonely, aes(x=social_hrs, y=loneliness)) +
  geom_point() +
  geom_smooth(method = "lm")+
  xlab("Social Media Hours (per week)") +
  ylab("Loneliness Score (0-100)")
```


## Conduct Tests **in R**

:::: {.columns}

::: {.column width="50%"}

**Correlation**

```{r}
cor.test(lonely$loneliness, lonely$social_hrs)
```

:::

::: {.column width="50%"}

**Standardised Simple Linear Model**

```{r}
m1z <- lm(loneliness_z ~ social_hrs_z, data = lonely)
summary(m1z)

```

:::

::::


## Results

Since our linear model had a single continuous predictor, the standardised slope $(\hat \beta_1^*)$ is exactly the same as the correlation coefficient $(r)$ between the predictor and the outcome!

Why?

+ They are equivalent:
  + $r$ is a standardised measure of linear association
  + $\hat \beta_1^*$ is a standardised measure of the linear slope


## Why is the Linear Model a Better Choice?

- We have a much wider class of problems than what can be assessed with basic tests, and linear models can be extended to other forms 
- We often have many predictors we want to include
- We have variables we want to control for (e.g., age and sex)
- More robust to assumptions
- Predictive power



## Revisiting This Week's Learning Objectives

<br>

::: {}
::: {.dapr3callout}
**XXXXXXXXXX**


:::
:::

::: {}
::: {.dapr3callout}
**XXXXXXXXXX**


:::
:::

## Revisiting This Week's Learning Objectives

<br>

::: {}
::: {.dapr3callout}
**XXXXXXXXXX**


  
:::
:::

::: {}
::: {.dapr3callout}
**XXXXXXXXXX**


:::
:::


## This Week 

