---
title: "Successful knitting checklist"
author: "DAPR1 staff"
output: bookdown::pdf_document2
toc: false
geometry: "left=2.5cm,right=2.5cm,top=1.5cm,bottom=2cm"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Fixing errors is an iterative process

If you are getting errors when knitting an Rmd file, please try the following steps in order. 

Once you have identified an error and fixed it, knit again to spot the next error in the file. 

R will always stop at the first error, rather than listing all errors in your file. As such, fixing errors is an iterative process of spotting the first error to appear in your file, then fixing it, and continuing this way.


# Checklist

1. Did you install `tinytex`? This is required for knitting to a PDF file.

    If not, or if you are unsure, try installing it by typing the following in the Console, and then pressing Enter:
        
    ```
    install.packages("tinytex")
    ```
    
    Next, type the following in the console, press Enter, then Y, the Enter:
    
    ```
    tinytex::install_tinytex()
    ```

1. Ensure you have an empty line before and after every code chunk.

1. Ensure you have an empty line before and after every section title.

1. Ensure the top of the Rmd file is structured as follows:
    
    ```
    ---
    title: "Title goes here"
    author: "Author goes here"
    date: "Date goes here"
    output: bookdown::pdf_document2
    toc: false
    ---
    ```

    The file won't knit if you forget to open and close the two delimiters, `---`.
    
    Similarly, the file won't knit if you use uppercase incorrectly. For example, `toc: FALSE` or `toc: False` will lead to errors.


1. Your code chunk labels should not have any underscores or spaces.
    
    Replace the code chunk labels having space or underscores, for example:
    
    ````{verbatim}
    ```{r plt freq distr}
    
    ```
    ````
    
    ````{verbatim}
    ```{r plt_freq_distr}
    
    ```
    ````
    
    to either use dashes or camel case:
    
    ````{verbatim}
    ```{r plt-freq-distr}
    
    ```
    ````
    
    ````{verbatim}
    ```{r pltFreqDistr}
    
    ```
    ````

1. Ensure that you don't have two code chunks with the same label. The two code chunks below have the same label, and R doesn't like this and can't tell them apart:

    ````{verbatim}
    ```{r pltHistogram}
    
    ```
    ````
    
    ````{verbatim}
    ```{r pltHistogram}
    
    ```
    ````

1. There should be no comma before the code chunk label. 
    
    Replace:

    ````{verbatim}
    ```{r, plt-freq-distr}
    
    ```
    ````
    
    to be:
    
    ````{verbatim}
    ```{r plt-freq-distr}
    
    ```
    ````


1. Ensure that all mathematical symbols only appear inside equations blocks (delimited by dollars), i.e. `$ equation $`. Equations should only appear in text, not inside of code chunks.

    - Example 1: Consider $\cup$ in $P(A \cup B)$. This should be written as `$P(A \cup B)$` and you should not directly copy and paste the $\cup$ symbol into a comment or text.
    
    - Example 2: Approximately equal to ($\approx$) should also only appear inside of an equation. You should not directly copy and paste this symbol into a comment or text.
    
    - Example 3: $\alpha$ should be written within an equation as `$\alpha$`. You should not directly copy and paste this symbol into a comment or text.
        
    - Example 4: $H_{0}$ should be written `$H_{0}$`. You should not directly copy and paste this symbol into a comment or text.

1. Ensure that there are no unicode characters in your file. Unicode characters are symbols beyond those included in a standard PC keyboard, and can cause problems with tinytex.

1. Don't worry about the order of tables or figures.

    + A reader will be able to find your figure or table as long as they are correctly referenced and numbered!

