---
title: "Checklist: successful knitting"
author: "DAPR1 staff"
output: bookdown::pdf_document2
toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Checklist for ensuring an Rmd file successfully knits to PDF

- Did you install `tinytex`? This is required for knitting to a PDF file.

    If not, or if you are unsure, try installing it with:
        
    ```
    install.packages("tinytex")
    tinytex::install_tinytex()
    ```

- Ensure the metadata block at the top of the Rmd file is written correctly:
    
    ```
    ---
    title: "Write your title here"
    author: "Your exam number"
    output: bookdown::pdf_document2
    toc: false
    ---
    ```

    The file won't knit if you forget one of the two delimiters, `---`.
    
    Similarly, the file won't knit if you use uppercase incorrectly. For example, `toc: FALSE` or `toc: False` will lead to errors.


- Your code chunk labels should not have any underscores or spaces.
    
    Replace the code chunk labels having space or underscores, for example:
    
    ````{verbatim}
    ```{r plt freq distr}
    
    ```
    ````
    
    ````{verbatim}
    ```{r plt_freq_distr}
    
    ```
    ````
    
    to either use dashes or camel case:
    
    ````{verbatim}
    ```{r plt-freq-distr}
    
    ```
    ````
    
    ````{verbatim}
    ```{r pltFreqDistr}
    
    ```
    ````


- There should be no comma before the code chunk label. 
    
    Replace:

    ````{verbatim}
    ```{r, plt-freq-distr}
    
    ```
    ````
    
    to be:
    
    ````{verbatim}
    ```{r plt-freq-distr}
    
    ```
    ````


- Ensure there are no unicode symbols copied and pasted in your file.

    - Example 1: $\alpha$ should be written within an equation as `$\alpha$`. All mathematical symbols should be written within equations blocks (delimited by dollars) among the text part of the report (__not inside of code chunks__).
    
    - Example 2: Consider $\cup$ in $P(A \cup B)$. This should be written as `$P(A \cup B)$`, you should not directly copy and paste the "u" symbol.
    
    - Example 3: $H_{0}$ should be written `$H_{0}$`



- Don't worry about the order of tables or figures.

    + A reader will be able to find your figure or table as long as they are correctly referenced and numbered!

