---
title: "Formatting descriptives tables"
format:
    html:
        theme: cosmo
        toc: true
        toc-depth: 3
        number-sections: false
        embed-resources: true
html-table-processing: none
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE
)
```

## The data

The data we will use come from a randomised controlled trial comparing
treatments for Major Depressive Disorder (MDD) (DeRubeis et al., 2005)^[DeRubeis RJ, Hollon SD, Amsterdam JD, Shelton RC, Young PR, Salomon RM, O'Reardon JP, Lovett ML, Gladis MM, Brown LL, Gallop R. Cognitive therapy vs medications in the treatment of moderate to severe depression. Arch Gen Psychiatry. 2005 Apr;62(4):409-16. doi: 10.1001/archpsyc.62.4.409. PMID: 15809408.].
Patients with moderate to severe depression were randomly assigned to
one of three groups: a placebo, anti-depressant medication (`ADM`), or
cognitive therapy (`CT`). The outcome of interest, `HDRS8`, is each
patient's total score on the Hamilton Depression Rating Scale after 8
weeks of treatment, where higher scores indicate more severe
depressive symptoms.

The dataset is available online, so you could read it directly from
the web:

```{r}
library(tidyverse)
mdd_data <- read_csv("https://uoepsy.github.io/data/dapr1-mdd-data.csv")
```

Preview the data structure:

```{r}
glimpse(mdd_data)
```

Tell R that the variable Treatment is a factor (i.e. a categorical variable):

```{r}
mdd_data <- mdd_data |>
    mutate(Treatment = factor(Treatment))
glimpse(mdd_data)
```

## Formatting tables with the `kableExtra` package

So far we've printed our summary tables directly to the screen as
plain tibbles. This is fine for checking your work, but it is **not**
how tables should look in a dissertation or journal paper. Raw
tibble output shows column types (e.g. `<dbl>`, `<fct>`), too many
decimal places, and non-user-friendly column names, none of which you want
a reader to see.

The `kable()` function from the package `kableExtra`converts a tibble into a clean, readable table.

First, we must load the `kableExtra` package:

```{r}
library(kableExtra)
```

A few things you will want to control every time you want to make a publication-ready table:

1. **Rounding**: use the `digits` argument of `kable()` to round
  numeric columns (e.g. `digits = 2` for 2 decimal places), rather
  than rounding the values in your data. This keeps your underlying
  numbers precise while only rounding what is displayed.
2. **Column names**: use the `col.names` argument to replace raw
  variable names (e.g. `HDRS8`) with readable labels (e.g. `"HDRS score at 8 weeks"`).
3. **Caption**: use the `caption` argument to give every table a
  short description of what it shows.

::: {.callout-tip}
## Tip: round the displayed table, not your underlying variables

Round only at the point of display using `kable(digits = ...)`.
Avoid rounding the variables themselves with `round()` earlier in your
workflow, since this can introduce small errors if those rounded
values are then used in further calculations.
:::

## Frequency tables

Frequency tables summarise how many (and what percentage of)
observations fall into each category of a categorical variable. Let's
build one for `Treatment`:

```{r}
freq_table <- mdd_data |>
    count(Treatment) |>
    mutate(f = n / sum(n),
           Percent = f * 100)

freq_table
```

Printed as a tibble, the `f` column has far too many decimal
places, the `Percent` column is not rounded to 2 decimal places, and the column names are not very user friendly. Passing
this to `kable()` with `digits = 2` rounds all numeric columns to 2
decimal places, and `col.names` lets us relabel the columns:

```{r}
freq_table |>
    kable(
        digits = 2,
        col.names = c("Treatment Group", "n", "Rel. frequency","Percent (%)"),
        caption = "Frequency, relative frequency, and percentage of participants in each treatment group."
    )
```

We can add `kableExtra` styling by piping the result of `kable()` into
`kable_styling()`:

```{r}
freq_table |>
    kable(
        digits = 2,
        col.names = c("Treatment Group", "n", "Rel. frequency", "Percent (%)"),
        caption = "Frequency, relative frequency, and percentage of participants in each treatment group."
    ) |>
    kable_styling(
        full_width = FALSE,
        position = "center"
    )
```

Some useful `kable_styling()` options:

- `full_width = FALSE`: the table only takes up as much width as it
  needs, rather than stretching across the whole page.
- `position = "center"`: centers the table on the page.

## Descriptives tables: n, M, and SD

For a numerical variable such as `HDRS8`, the standard descriptive
statistics to report are the sample size ($n$), mean ($M$), and
standard deviation ($SD$).

So far, when using `summarise()` we have calculated these statistics
across **all** participants at once:

```{r}
mdd_data |>
    summarise(
        n = n(),
        M = mean(HDRS8),
        SD = sd(HDRS8)
    )
```

But what we are interested in, is comparing the effectiveness of the differen treatments for depression. To do so, we want these statistics calculated **separately for each group** (e.g. for each `Treatment`), so that they can be compared. 
To do this, we add a `group_by()` step before `summarise()`. This tells R to compute the number of values, mean HDRS8, and standard deviation of HDRS8 separately for each treatment group.

```{r}
descriptives_table <- mdd_data |>
    group_by(Treatment) |>
    summarise(
        n = n(),
        M = mean(HDRS8),
        SD = sd(HDRS8)
    )

descriptives_table
```

`group_by(Treatment)` tells R to split the data into groups according
to the values of `Treatment`, and then `summarise()` calculates `n`,
`M`, and `SD` separately within each group, rather than across the
whole sample. This combination of `group_by()` and `summarise()` is
how we will create descriptives broken down by group across your DAPR courses.

::: {.callout-tip}
## `group_by()` + `summarise()`

`group_by()` on its own doesn't change how the data looks, it just
tells R to do grouped computations. It's only when you follow it with
`summarise()` (or another function like `mutate()`) that you see the
effect as the calculation is done separately for each group.
:::

Again, before putting this in a report we round to 2 decimal places,
relabel the columns, and add a caption:

```{r}
descriptives_table |>
    kable(
        digits = 2,
        col.names = c("Treatment", "n", "M", "SD"),
        caption = "Sample size, mean, and standard deviation of HDRS scores at 8 weeks by treatment group."
    ) |>
    kable_styling(
        full_width = FALSE,
        position = "center"
    )
```


::: {.callout-important}
## Explore further

We've only covered a handful of `kableExtra` options here. The
package can do much more including grouped headers, cell highlighting,
footnotes, and more.

Have a look at the 
[`kableExtra` vignette](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html),
and try playing around with some of the extra settings shown there on
your own tables (e.g. `descriptives_table` or `freq_table`). Being
comfortable experimenting with formatting options like this will make
your life a lot easier when polishing tables for your dissertation.
:::


::: {.callout-note}
## Why this matters for your dissertation

Most dissertations and papers in
Psychology and related disciplines are expected to follow
[**APA style**](https://apastyle.apa.org/instructional-aids/numbers-statistics-guide.pdf) (currently 7th edition), the formatting conventions of the American Psychological Association.
APA style gives precise rules for how numbers and tables should be
presented, and markers will check your work against these
conventions. Some of the most relevant rules for descriptives tables
are:

- Report means and standard deviations to **2 decimal places**
  (e.g. `M = 15.34`, `SD = 4.02`), and use no more decimal places
  than the precision of your measurement justifies.
- Report sample sizes ($n$) as **whole numbers**, with no decimal
  places.
- Use italicised statistical symbols (e.g. *M*, *SD*, *n*) as column
  headings, rather than plain English or raw variable names.
- Tables should be **numbered** (e.g. "Table 1") with a clear,
  descriptive caption/title, and should be understandable without
  reading the main text.

This is precisely why we control **rounding**, **column names**, and
**captions** through `kable()`: it lets you match APA formatting
requirements directly from your R code, rather than manually editing
numbers in Word afterwards (which is error-prone and hard to
reproduce if your data changes).

Get into the habit of piping your summary tables through `kable()`
(and `kableExtra` where useful) every time, so that producing
publication-ready tables becomes automatic by the time you write up
your dissertation.
:::
