---
title: "W2: Data Types"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
format: html
---

```{r}
#| include: false
library(tidyverse)
source('assets/dropdowns.R')
source('assets/setup.R')
```

<!-- MAX 5 QUESTIONS -->
<!-- make each question ~10 minute work -->
<!-- MAX 5 QUESTIONS -->
<!-- MAX 5 QUESTIONS -->

:::frame
This week's workshop we're going to start working together in groups. 

Pick one of you to be the 'driver'. They will type the code. The rest of you are navigators, you need to tell the driver where to go (i.e., what to type!).  

Driver instructions:

1. Log in to the bigger screen at your table
2. Open [RStudio Server](https://rstudio.ppls.ed.ac.uk){target="_blank"}
3. Listen to your navigators

Navigator instructions:

1. Tell your driver where to go (i.e., what to type)!

:::



`r qbegin(qcounter())`
Open a new R script for this week.  

The first thing we're going to do is to get some practice with packages.  

1. load the "tidyverse" (it should be already installed on the server)
2. install and load the "skimr" package

::: {.callout-tip collapse="true"}
#### Hints

For help, check out [LINK P2 flashcard]. 

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

To load the tidyverse, we just need to do: 
```{r}
library(tidyverse)
```

To install the "skimr" package, we need to use `install.packages("skimr")`. BUT, we probably don't want to keep this in our script, because then it would re-install the package every time we ran the whole script.  

Installing packages is one of the few things we might type specifically in the console
```{r}
#| eval: false
install.packages("skimr")
```

And then back in our script we will want: 
```{r}
library(skimr)
```


`r solend()`

`r qbegin(qcounter())`
Now that we've loaded packages, we want to read in some data!  
Last week, we put a little survey up for you to fill out, so that we could have data for you to play with.  

The data are available at: TODO LINK  

Read it in to your R session now. Be sure to assign it a name! 

::: {.callout-tip collapse="true"}
#### Hints

Here's the basic usage of `read_csv()` and how you can assign its output to a variable named `mydata`.

Replace the `...` with the actual location of the data (the location could be somewhere stored on your computer, or the URL address of where it is located online!)  

```{r}
#| eval: false
mydata <- read_csv("...")
```

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
sdata <- read_csv("../../data/usmr2025.csv")
```

`r solend()`

`r qbegin(qcounter())`
How many people do we have data for? And on how many variables?  

::: {.callout-tip collapse="true"}
#### Hints

Functions like `dim(data)`, `nrow(data)`, `ncol(data)` will come in handy here.  

You can also try using the `skim(data)` function (from the "skimr" package we installed a minute ago) --- it will give you loads of information!  

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`


`r solend()`


`r qbegin(qcounter())`
Go through each variable and write down what **type** of variable it is (e.g., is it continuous, categorical, or ordinal?)  

::: {.callout-tip collapse="true"}
#### Hints

See flashcard S4 TODO link

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`



`r solend()`


`r qbegin(qcounter())`

R will be default treat a variable with only numbers in it as a numeric variable, and anything else as a "character" variable ("character" = any letters/numbers/symbols).  

Change the eye colour variable to a factor in R, that only has the options "brown", "blue", "green" or "hazel".    

What happens when you now do `skim(data)`? How does it treat this variable differently?  

::: {.callout-tip collapse="true"}
#### Hints

Here is the basic code structure for taking a dataframe named `mydata` and adding a new variable (or changing an existing one).  

- Line 2 takes the data, and passes it to line 3.  
- Lines 3-5 "mutates" the data, adding a variable called `new_variable`. The `new_variable` is going to contain the values of an existing variable (`old_variable`) but with some function applied to them. For example, if we had `new_variable = old_variable/2` then it would contain the same values but halved. 
- It feels odd, but it helps to think about the first line of code once we've thought of the others. This line **re-assigns** the name `mydata` to the mutated `mydata`, thereby making the change save to the object in R's memory.   

```{r}
#| eval: false
#| code-line-numbers: true
mydata <-
  mydata |> 
  mutate(
    new_variable = function(old_variable)
  )
```

To make a variable a factor, the function we want is `factor()`. 
This will consider every different value that we have to be a possible category: 
```{r}
#| eval: false
factor(variable)
```

If we want to have a *specific* set of possible values, we can do so with the `levels` part of the function, replacing the `"..."` with the options we consider valid:  
```{r}
#| eval: false
factor(variable, levels = c("...","...","...", ....... ))
```

You can see more of this stuff in Flashcard P3 TODO link.  

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
sdata <-
  sdata |>
  mutate(
    eyecolour = factor(eyecolour, levels = c("brown","blue","green","hazel"))
  )


skim(sdata)
```

`r solend()`












