---
title: "W2: Data Types"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
format: html
---

```{r}
#| include: false
library(tidyverse)
source('assets/dropdowns.R')
source('assets/setup.R')
```

<!-- MAX 5 QUESTIONS -->
<!-- make each question ~10 minute work -->
<!-- MAX 5 QUESTIONS -->
<!-- MAX 5 QUESTIONS -->

:::frame
This week's workshop we're going to start working together in groups. 

Pick one of you to be the 'driver'. They will type the code. The rest of you are navigators, you need to tell the driver where to go (i.e., what to type!).  

Driver instructions:

1. Log in to the bigger screen at your table
2. Open [RStudio Server](https://rstudio.ppls.ed.ac.uk){target="_blank"}
3. Listen to your navigators

Navigator instructions:

1. Tell your driver where to go (i.e., what to type)!

:::



`r qbegin(qcounter())`
Open a new R script for this week.  

The first thing we're going to do is to get some practice with packages.  

1. load the "tidyverse" (it should be already installed on the server)
2. install and load the "skimr" package

::: {.callout-tip collapse="true"}
#### Hints

For help, check out [LINK P2 flashcard]. 

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

To load the tidyverse, we just need to do: 
```{r}
library(tidyverse)
```

To install the "skimr" package, we need to use `install.packages("skimr")`. BUT, we probably don't want to keep this in our script, because then it would re-install the package every time we ran the whole script.  

Installing packages is one of the few things we might type specifically in the console
```{r}
#| eval: false
install.packages("skimr")
```

And then back in our script we will want: 
```{r}
library(skimr)
```


`r solend()`

`r qbegin(qcounter())`
Now that we've loaded packages, we want to read in some data!  
Last week, we put a little survey up for you to fill out, so that we could have data for you to play with.  

The data are available at: TODO LINK  

Read it in to your R session now.  

::: {.callout-tip collapse="true"}
#### Hints

The function you'll need is `read_csv()`, and be sure to assign it a name!  

```{r}
#| eval: false
mydataframe <- read_csv("...")
```

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
sdata <- read_csv("../../data/usmr2025.csv")
```

`r solend()`

`r qbegin(qcounter())`
How many people do we have data for? And on how many variables?  

::: {.callout-tip collapse="true"}
#### Hints

functions like `dim(data)`, `nrow(data)`, `ncol(data)` will come in handy here. You could also try using the `skim(data)` function (from the "skimr" package we installed a minute ago) --- it will give you loads of information!  

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`


`r solend()`


`r qbegin(qcounter())`
Go through each variable and write down what **type** of variable it is (e.g., is it continuous, categorical, or ordinal?)  

::: {.callout-tip collapse="true"}
#### Hints

See flashcard S4 TODO link

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`



`r solend()`


`r qbegin(qcounter())`

R will be default treat a variable with only numbers in it as a numeric variable, and anything else as a "character" variable ("character" = any letters/numbers/symbols).  

Change the eye colour variable to a factor in R, that only has the options "brown", "blue", "green" or "hazel".    

What happens when you now do `skim(data)`? How does it treat this variable differently?  

::: {.callout-tip collapse="true"}
#### Hints

You can see an example of this in Flashcard P3 TODO link

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
sdata <-
  sdata |>
  mutate(
    eyecolour = factor(eyecolour, levels = c("brown","blue","green","hazel"))
  )


skim(sdata)
```

`r solend()`












