---
title: "W3: Visualising Distributions"
params: 
    # SHOW_SOLS: TRUE
    SHOW_SOLS: FALSE
    TOGGLE: TRUE
format: html
---

<!--
WHEN RENDERING: 

- tutor-facing: 
  - SHOW_SOLS: TRUE
  - render in RStudio, or in Terminal using `quarto render FILE.qmd`
  - resulting file will land in `docs/YEAR/workshops/tutors/`
  
- student-facing:
  - SHOW_SOLS: FALSE
  - render in Terminal using `quarto render FILE.qmd --profile production`
  - resulting file will land in `docs/YEAR/workshops/`
-->

```{r}
#| include: false
library(tidyverse)
source('assets/dropdowns.R')
source('assets/setup.R')
```

<!-- MAX 5 QUESTIONS -->
<!-- make each question ~10 minute work -->
<!-- MAX 5 QUESTIONS -->
<!-- MAX 5 QUESTIONS -->

:::frame
Pick someone new to be the driver this week. They will type the code. 
The rest of you are navigators, you need to tell the driver where to go (i.e., what to type!).  

Driver instructions:

1. Log in to the bigger screen at your table
2. Open [RStudio Server](https://rstudio.ppls.ed.ac.uk){target="_blank"}
3. Open a new .R script, and load the **tidyverse** package.
4. Listen to your navigators

Navigator instructions:

1. Tell your driver where to go (i.e., what to type)!

:::

`r qbegin(qcounter())`
For this week, we are going to use the same data as last week (the survey that many of you completed in the first week). The data are stored at [https://uoepsy.github.io/data/dapr1_2627_survey.csv](https://uoepsy.github.io/data/dapr1_2627_survey.csv){target="_blank"}  
(Check back to your work from last week to remind yourself how to get the data into R). 

One of the questions in the survey asked people to choose between Dog and Cat (with options for "both" and "neither" as well).  
Create a frequency table for this variable (`catdog`).  

:::flash
🗂️
See [Read in and inspect data](https://uoepsy.github.io/dapr1/2627/flashcards/block1/read-data.html){target="_blank"} flash card.

🗂️
See [Make a frequency table](https://uoepsy.github.io/dapr1/2627/flashcards/block1/freq-table.html){target="_blank"} flash card.
:::


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
library(tidyverse)
sdata <- read_csv("https://uoepsy.github.io/data/dapr1_2627_survey.csv")

sdata |>
  count(catdog) |>
  mutate(
    Percent = round((n/sum(n))*100, 2)
  )

```

`r solend()`


`r qbegin(qcounter())`
Based on the previous question, what do you think a visualisation of the distribution of the `catdog` variable will look like?

To check if your guess is right, plot the `catdog` variable.

:::flash
🗂️
See [Plot a variable](https://uoepsy.github.io/dapr1/2627/flashcards/block1/intro-ggplot.html){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
ggplot(sdata, aes(x = catdog)) +
  geom_bar()
```


`r solend()`


`r qbegin(qcounter())`
Another question in the survey asked people how far away (in kilometers) they were born (the `distance_born` variable in the dataset).  

Visualise this variable using a histogram.  

:::flash
🗂️
See [Plot a variable](https://uoepsy.github.io/dapr1/2627/flashcards/block1/intro-ggplot.html){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
ggplot(sdata, aes(x = distance_born)) + 
  geom_histogram()
```


`r solend()`

`r qbegin(qcounter())`
Now visualise this same variable using a density plot.

:::flash
🗂️
See [Plot a variable](https://uoepsy.github.io/dapr1/2627/flashcards/block1/intro-ggplot.html){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
ggplot(sdata, aes(x = distance_born)) + 
  geom_density()
```


`r solend()`


`r qbegin(qcounter())`
Write a sentence or two commenting on the shape of the distribution of birthplace distances from Edinburgh, noting the presence of any outliers.  

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Distances from Edinburgh were strongly positively skewed, with a distinct peak near zero, and a long right tail extending to about ~15,000, with a possible smaller peaks around ~2,000 and 9,000.  
`r solend()`







:::frame
__Finishing up__  

Once you've finished, don't forget to: 

1. Save your R script
2. Post the R script to the group discussion space on Learn, to ensure that everyone has access to it! (For step-by-step instructions on how to do this, click [here](00serverfiles.html){target="_blank"})  


:::








