---
title: "W3: Visualising Distributions"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
format: html
---

```{r}
#| include: false
library(tidyverse)
source('assets/dropdowns.R')
source('assets/setup.R')
```

<!-- MAX 5 QUESTIONS -->
<!-- make each question ~10 minute work -->
<!-- MAX 5 QUESTIONS -->
<!-- MAX 5 QUESTIONS -->

:::frame
Pick someone new to be the driver this week. They will type the code. 
The rest of you are navigators, you need to tell the driver where to go (i.e., what to type!).  

Driver instructions:

1. Log in to the bigger screen at your table
2. Open [RStudio Server](https://rstudio.ppls.ed.ac.uk){target="_blank"}
3. Listen to your navigators

Navigator instructions:

1. Tell your driver where to go (i.e., what to type)!

:::

`r qbegin(qcounter())`
For this week, we are going to use the same data as last week (the survey that many of you completed in the first week). The data are stored at [https://uoepsy.github.io/data/dapr1_2526_survey.csv](https://uoepsy.github.io/data/dapr1_2526_survey.csv){target="_blank"}  
(Check back to your work from last week to remind yourself how to get the data into R). 

One of the questions in the survey asked people to choose between Dog and Cat (with options for "both" and "neither" as well).  
Create a frequency table for this variable (`catdog`).  

::: {.callout-tip collapse="true"}
#### Hints

TODO flashcard frequency table?

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
library(tidyverse)
sdata <- read_csv("https://uoepsy.github.io/data/dapr1_2526_survey.csv")

sdata |>
  count(catdog) |>
  mutate(
    Percent = round((n/sum(n))*100,2)
  )

```

`r solend()`


`r qbegin(qcounter())`
Create a visualisation of the distribution of the `catdog` variable.  

Before you do so, can you work out what it will look like, based on the previous question?  

::: {.callout-tip collapse="true"}
#### Hints

TODO flashcard ggplot

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
ggplot(sdata, aes(x = catdog)) +
  geom_bar()
```


`r solend()`


`r qbegin(qcounter())`
Another question in the survey asked people how far away (in kilometers) they were born (the `distance_born` variable in the dataset).  

Visualise this variable using a boxplot.  

::: {.callout-tip collapse="true"}
#### Hints

TODO flashcard ggplot

:::
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
ggplot(sdata, aes(x = distance_born)) + 
  geom_boxplot()
```


`r solend()`

`r qbegin(qcounter())`
Now visualise this same variable using a histogram.  

::: {.callout-tip collapse="true"}
#### Hints

TODO flashcard ggplot

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
ggplot(sdata, aes(x = distance_born)) + 
  geom_histogram()
```


`r solend()`


`r qbegin(qcounter())`
Write a sentence or two commenting on the shape of the distribution of birthplace distances from Edinburgh, noting the presence of any outliers.  

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`


`r solend()`











:::frame
__Finishing up__  

Once you've finished, don't forget to: 

1. Save your R script
2. Post the R script to the group discussion space on Learn, to ensure that everyone has access to it! (For step-by-step instructions on how to do this, click [here](00serverfiles.html){target="_blank"})  


:::








