---
title: "W4: Summarising Distributions I"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
format: html
---

```{r}
#| include: false
library(tidyverse)
source('assets/dropdowns.R')
source('assets/setup.R')
```

<!-- MAX 5 QUESTIONS -->
<!-- make each question ~10 minute work -->
<!-- MAX 5 QUESTIONS -->
<!-- MAX 5 QUESTIONS -->

:::frame
Pick someone new to be the driver this week. They will type the code. 
The rest of you are navigators, you need to tell the driver where to go (i.e., what to type!).  

Driver instructions:

1. Log in to the bigger screen at your table
2. Open [RStudio Server](https://rstudio.ppls.ed.ac.uk){target="_blank"}
3. Listen to your navigators

Navigator instructions:

1. Tell your driver where to go (i.e., what to type)!

:::

`r qbegin(qcounter())`
We're continuing to use the same dataset for the next couple of weeks (the survey that many of you completed in the first week). The data are stored at [https://uoepsy.github.io/data/dapr1_2526_survey.csv](https://uoepsy.github.io/data/dapr1_2526_survey.csv){target="_blank"} (Check back to your work from the previous two weeks to remind yourself how to get the data into R).  

What was the most common month in which survey respondents were born? 

:::flash
🗂️
See [Measure central tendency > Mode](https://uoepsy.github.io/dapr1/2627/flashcards/block1/central-tendency.html#mode){target="_blank"} flash card.
:::


::: {.callout-tip collapse="true"}
#### Hint

The variable that holds this information is the one called `birthmonth`. You can check what different variables are in your data by quickly using `names(data)`, replacing "data" with whatever name you used to store your dataset in R.  

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
library(tidyverse)
sdata <- read_csv("https://uoepsy.github.io/data/dapr1_2526_survey.csv")

sdata |> 
  count(birthmonth)
```

`r solend()`


`r qbegin(qcounter())`
One of the questions in the survey asked about how spicy people liked their food.  
The possible responses were:  

<small>

- Anti-spice: Mayonnaise is too spicy for me; I avoid any black pepper.
- Mild: I like a tiny hint of warmth, but flavour comes first.
- Medium: I enjoy a solid kick that makes the meal interesting.
- Hot: I want to feel the burn.
- Very Hot: I want to sweat and question my life choices.
- Extreme: I want to transcend mortal existence and taste the fury of a thousand suns.  

</small> 

This has been recoded for you already so that the dataset contains both a variable with the descriptive word ("medium", "hot" etc), and a variable with a numeric value (1 being assigned to "anti-spice" and 6 being assigned to "extreme").  

For what level of spicy-ness does half of the sample fall above and half fall below?  

:::flash
🗂️
See [Measure central tendency > Median](https://uoepsy.github.io/dapr1/2627/flashcards/block1/central-tendency.html#median){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
sdata |> 
  summarise(
    spicy_Med = median(spicy_num, na.rm = TRUE)
  )
```


`r solend()`

`r qbegin(qcounter())`
Another question asked respondents to rate the quality of their sleep (from 0 to 100). 
On average, how well did respondents sleep?  
What was the average sleep quality rating 

:::flash
🗂️
See [Measure central tendency > Mean](https://uoepsy.github.io/dapr1/2627/flashcards/block1/central-tendency.html#mean){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
sdata |>
  summarise(
    sleep_M = mean(sleepqual, na.rm = TRUE)
  )
```

`r solend()`


`r qbegin(qcounter())`
The code below creates a visualisation of the "distance-from-Edinburgh" that respondents were born (you may need to replace the `sdata` with whatever you named your dataset in R):    

```{r}
#| out-width: "450px"
#| out-height: "250px"
ggplot(sdata, aes(x = distance_born)) +
  geom_histogram()
```

Do you think the median or the mean of this variable will be higher?

Calculate both the mean and median for this variable to see if you were right.  

(Challenge: calculate both summary statistics within the same instance of `summarise()`)

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
sdata |>
  summarise(
    dist_M = mean(distance_born),
    dist_Med = median(distance_born)
  )
```


`r solend()`


`r qbegin(qcounter())`
Choose any of the variables in the data that we haven't used so far (excluding the `pseudonym` and `threewords` variables!).  

For your chosen variable, calculate an appropriate summary statistic for the central tendency.  

Can you do this separately for the people who responded "cat", "dog", "both" or "neither" to the question "Dog or Cat?" (this is the `catdog` variable in the dataset).  

:::flash
🗂️
See [Measure central tendency](https://uoepsy.github.io/dapr1/2627/flashcards/block1/central-tendency.html){target="_blank"} flash card - it also covers how to calculate measures by-groups.  
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
sdata |>
  group_by(catdog) |>
  summarise(
    sleepqual = mean(sleepqual)
  )
```


`r solend()`












:::frame
__Finishing up__  

Once you've finished, don't forget to: 

1. Save your R script
2. Post the R script to the group discussion space on Learn, to ensure that everyone has access to it! (For step-by-step instructions on how to do this, click [here](00serverfiles.html){target="_blank"})  


:::








