---
title: "W5: Summarising Distributions II"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
format: html
---

```{r}
#| include: false
library(tidyverse)
source('assets/dropdowns.R')
source('assets/setup.R')
```

<!-- MAX 5 QUESTIONS -->
<!-- make each question ~10 minute work -->
<!-- MAX 5 QUESTIONS -->
<!-- MAX 5 QUESTIONS -->

:::frame
Pick someone new to be the driver this week. They will type the code. 
The rest of you are navigators, you need to tell the driver where to go (i.e., what to type!).  

Driver instructions:

1. Log in to the bigger screen at your table
2. Open [RStudio Server](https://rstudio.ppls.ed.ac.uk){target="_blank"}
3. Listen to your navigators

Navigator instructions:

1. Tell your driver where to go (i.e., what to type)!

:::

`r qbegin(qcounter())`
We're continuing to use the same dataset for the next couple of weeks (the survey that many of you completed in the first week). The data are stored at [https://uoepsy.github.io/data/dapr1_2526_survey.csv](https://uoepsy.github.io/data/dapr1_2526_survey.csv){target="_blank"} (Check back to your work from the previous weeks to remind yourself how to get the data into R).  

Calculate the Inter Quartile Range (IQR) for: 

a. `distance_born` - the distances from Edinburgh that the respondents were born. The IQR is appropriate here because the variable is highly skewed.  
b. `spicy_num` - the levels of spice that respondents like. The IQR is appropriate here because the variable is ordinal.  

:::flash
🗂️
See [Measure variability > IQR](https://uoepsy.github.io/dapr1/2627/flashcards/block1/variability.html#inter-quartile-range-iqr){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
library(tidyverse)
sdata <- read_csv("https://uoepsy.github.io/data/dapr1_2526_survey.csv")

sdata |> 
  summarise(
    dist_IQR = IQR(distance_born, na.rm = TRUE),
    spice_IQR = IQR(spicy_num, na.rm = TRUE)
  )
```

`r solend()`


`r qbegin(qcounter())`
The survey also asked people to rate, on scales of 0 to 100, three things:

| variable | question | 
| -------- |---------- |
| `sleepqual` | "how well do you tend to sleep on average?" | 
| `procrast` | "how often do you tend to procrastinate on tasks or responsibilities on average?" |
| `multitask` | "how well do you think you can handle multiple tasks or responsibilities at the same time on average?" | 

Calculate the mean and standard deviation for one of these variables.  

:::flash
🗂️
See [Measure variability > SD](https://uoepsy.github.io/dapr1/2627/flashcards/block1/variability.html#standard-deviation-sd){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
sdata |> 
  summarise(
    sleep_m = mean(sleepqual, na.rm = TRUE),
    sleep_s = sd(sleepqual, na.rm = TRUE),
    procr_m = mean(procrast, na.rm = TRUE),
    procr_s = sd(procrast, na.rm = TRUE),
    mutli_m = mean(multitask, na.rm = TRUE),
    multi_s = sd(multitask, na.rm = TRUE)
  )
```

`r solend()`

`r qbegin(qcounter())`
For the sleep-quality, procrastination, and multitasking variables, respondents were given the scale from 0 to 100. That doesn't necessarily mean that people used the whole scale.  

For the same variable you chose in the previous question, calculate its range.  

:::flash
🗂️
See [Measure variability > Range](https://uoepsy.github.io/dapr1/2627/flashcards/block1/variability.html#range){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
sdata |> 
  summarise(
    sleep_min = min(sleepqual, na.rm = TRUE),
    sleep_max = max(sleepqual, na.rm = TRUE)
  )
```

`r solend()`



`r qbegin(qcounter())`
In one call to `summarise()` (that is, using the function `summarise()` only once), calculate descriptive statistics of central tendency and variability for all three variables `sleepqual`, `procrast` and `multitask`.  

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
sdata |> 
  summarise(
    sleep_m = mean(sleepqual, na.rm = TRUE),
    sleep_s = sd(sleepqual, na.rm = TRUE),
    procr_m = mean(procrast, na.rm = TRUE),
    procr_s = sd(procrast, na.rm = TRUE),
    multi_m = mean(multitask, na.rm = TRUE),
    multi_s = sd(multitask, na.rm = TRUE)
  )
```


`r solend()`


`r qbegin(qcounter())`
Using the summary statistics that you computed in the previous questions, write up a description of the respondents to the survey. 

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

`r solend()`






:::frame
__Finishing up__  

Once you've finished, don't forget to: 

1. Save your R script
2. Post the R script to the group discussion space on Learn, to ensure that everyone has access to it! (For step-by-step instructions on how to do this, click [here](00serverfiles.html){target="_blank"})  


:::








