---
title: "W4: Summarising Distributions"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
format: html
---

```{r}
#| include: false
library(tidyverse)
source('assets/dropdowns.R')
source('assets/setup.R')
```

<!-- MAX 5 QUESTIONS -->
<!-- make each question ~10 minute work -->
<!-- MAX 5 QUESTIONS -->
<!-- MAX 5 QUESTIONS -->

:::frame
Pick someone new to be the driver this week. They will type the code. 
The rest of you are navigators, you need to tell the driver where to go (i.e., what to type!).  

Driver instructions:

1. Log in to the bigger screen at your table
2. Open [RStudio Server](https://rstudio.ppls.ed.ac.uk){target="_blank"}
3. Listen to your navigators

Navigator instructions:

1. Tell your driver where to go (i.e., what to type)!

:::

`r qbegin(qcounter())`
We're continuing to use the same dataset for the next couple of weeks (the survey that many of you completed in the first week). The data are stored at [https://uoepsy.github.io/data/dapr1_2526_survey.csv](https://uoepsy.github.io/data/dapr1_2526_survey.csv){target="_blank"} (Check back to your work from the previous two weeks to remind yourself how to get the data into R).  

What is the Inter Quartile Range (IQR) of distances from Edinburgh that the respondents were born?  

::: {.callout-tip collapse="true"}
#### Hints

TODO! 

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
library(tidyverse)
sdata <- read_csv("https://uoepsy.github.io/data/dapr1_2526_survey.csv")


```

`r solend()`


`r qbegin(qcounter())`

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

`r solend()`



`r qbegin(qcounter())`

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

`r solend()`



`r qbegin(qcounter())`

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

`r solend()`


`r qbegin(qcounter())`

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

`r solend()`








:::frame
__Finishing up__  

Once you've finished, don't forget to: 

1. Save your R script
2. Post the R script to the group discussion space on Learn, to ensure that everyone has access to it! (For step-by-step instructions on how to do this, click [here](00serverfiles.html){target="_blank"})  


:::








