---
title: "W8: Sampling Distributions & Standard Error"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
format: html
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false
library(tidyverse)
source('assets/dropdowns.R')
source('assets/setup.R')
```

<!-- MAX 5 QUESTIONS -->
<!-- make each question ~10 minute work -->
<!-- MAX 5 QUESTIONS -->
<!-- MAX 5 QUESTIONS -->

:::frame
Pick someone new to be the driver this week. If everyone in the group has already had a week as the driver, start going round again!  

The driver will type the code. The rest of you are navigators, you need to tell the driver where to go (i.e., what to type!). 

Driver instructions:

1. Log in to the bigger screen at your table
2. Open [RStudio Server](https://rstudio.ppls.ed.ac.uk){target="_blank"}
3. Open a new .R script, and load the **tidyverse** package.
4. Listen to your navigators

Navigator instructions:

1. Tell your driver where to go (i.e., what to type)!

:::



`r qbegin(qcounter())`
For the exercises this week, we are going to pretend that we know that the distribution of `outlook` scores (pessimism/optimism measured on the -100 to 100 scale) for the wider population has a mean of 35 and a standard deviation of 20.  

So **if** we could ask every person in the world to fill out our questionnaire, we would see something like this:^[(the "8e+08" on the y-axis is a way of writing an eight followed by eight zeros, or 800,000,000 - there are 8 Billion people in the world, so the numbers in each of these histogram bins are going to be huge!)]
```{r}
#| echo: false
#| out-width: "600px"
#| out-height: "400px"
tibble(
  outlook = round(rnorm(1e6,35,20))
) |>
  ggplot(aes(x=outlook))+geom_histogram() +
  scale_y_continuous(labels = scales::label_scientific(scale = 8000))
  
```

We can generate a pretend sample of outlook scores by using the `rnorm()` function.  
Try playing around with this code. Run each line a couple of times to see how the output changes.  

```{r}
#| echo: true
fake_sample <- rnorm(n = 20, mean = 35, sd = 20)
fake_sample
mean(fake_sample)
```


**Important Note:** None of this is real data. When we do real research, we won't know what the distribution looks like in the wider population, and we won't be able to make pretend samples. We're doing this here because it is a useful exercise to help demonstrate how sampling variability works.  

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`


`r solend()`



`r qbegin(qcounter())`
From the previous question, you have hopefully noticed 3 things: 

1. We can make a pretend sample using `rnorm()`.  
2. Each time we do 1, we get a new, different, sample.  
3. Each time we do 1, the new sample we get has a slightly different mean.  

What we now want to start thinking about is "all the samples we *could* have taken", and how their means will vary.  

The function `replicate()` can let us do something in R lots of times. 

Can you figure out what these two lines of code do?  
Raise your hand if you want a tutor to come and help explain.  
```{r}
means20 <- replicate(1000, mean(rnorm(n = 20, mean = 35, sd = 20)))

means100 <- replicate(1000, mean(rnorm(n = 100, mean = 35, sd = 20)))
```

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`


`r solend()`



`r qbegin(qcounter())`
The objects we made in the previous question are "vectors". Unlike the dataframes we normally work with, these don't have rows and columns, they just have one dimension.  

In order to plot them, we need to put them into the "rows and column" format. We can make a dataframe in R using the `tibble()` function.  

Inside `tibble()`, we give names of variables and the values we want those variables to take:  
```{r}
#| eval: false
mydata <- 
  tibble(
    variable_name = values,
    variable2_name = values,
    ...
  )
```

**Task** Make two dataframes, call them something like `data20` and `data100`. Each one should have just one column, which contains the values in `means20` and `means100`. 

:::flash
🗂️
See [TODO making a tibble/df flashcard](https://uoepsy.github.io/dapr1/2627/flashcards/){target="_blank"} flash card.
:::

::: {.callout-tip collapse="true"}
#### Hint

Because the `means20` and `means100` are vectors, that means they are just a bunch of values. 
Which means we can follow this logic: 

```{r}
myvector <- c(1,2,5,7,3,5)

mydataframe <- 
  tibble(
    thing = myvector
  )
```
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
data20 <- 
  tibble(
    estimate = means20
  )

data100 <- 
  tibble(
    estimate = means100
  )
```

`r solend()`



`r qbegin(qcounter())`
We should now have two dataframes. 
One contains the means from 1000 random samples of size 20 (taken from our supposed population of optimism scores that are assumed to be normally distributed with mean of 35 and standard deviation of 20).  
The second contains the means from 1000 random samples of size 100.  

Make a plot of each of these sets of 1000 estimated means.  
(What sort of plot? That's up to you!).  

When you have made them, let's display them side by side. To do this, we can use the **patchwork** package (you might need to install it first!). 
Then we can arrange multiple plots by 1. assigning each plot a name, and 2. using things like `+` and `/` to but them next to/above one another.  

Each of the plots shows "means from 1000 samples we might see". Because each sample is a random selection people's optimism scores, the means of the samples differed (some samples were full of more optimistic people, some with less optimistic people).  

Where can you see differences between the two plots?   Bonus challenge: why?  



:::flash
🗂️
See [Use R packages](https://uoepsy.github.io/dapr1/2627/flashcards/block1/packages.html){target="_blank"} flash card.

🗂️
See [TODO patchwork](https://uoepsy.github.io/dapr1/2627/flashcards/){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
library(patchwork)

p20 <- ggplot(data20, aes(x = estimate)) +
  geom_histogram()

p100 <- ggplot(data100, aes(x = estimate)) +
  geom_histogram()

p20 + p100
```
`r solend()`



`r qbegin(qcounter())`
We saw in the lecture that we can characterise "how much a sample estimate will vary due to random sampling" by using the "standard error". And we saw a formula for this for the sample mean: 

$$
SE_{\text{mean}} = \frac{\sigma}{\sqrt{n}}
$$

In the two scenarios we have been creating, we have seen how the means from samples of $n=20$ vary, as well as means from samples of $n=100$.  

Because we started this by assuming that optimism scores in the population have mean of 35 and standard deviation of 20, we can calculate:  

::::{.columns}
:::{.column width="50%"}
$$
\begin{align}
SE_{\text{mean, n=20}} &= \frac{20}{\sqrt{20}} \\
\quad \\
&= 4.47 \\
\end{align}
$$
:::
:::{.column width="50%"}

$$
\begin{align}
SE_{\text{mean, n=100}} &= \frac{20}{\sqrt{100}} \\
\quad \\
&= 2 \\
\end{align}
$$

:::
::::

What **we** have done today is actually _generate_ lots of samples and their means, and stored them in R.  

It's really cool that statisticians figured out how to estimate the sampling variability, because it save us from having to do all this rigmarole. 
The standard error calculations above give us estimates of how much the sample estimates will vary because of random sampling. 
So we can simply use the calculations to get at a *theoretical* probability distribution (the red curves below), rather than having to actually simulate stuff.  

```{r}
#| echo: false
#| fig-height: 4
#| fig-width: 12
mean_20 <- mean(data20$estimate, na.rm = TRUE)
sd_20   <- sd(data20$estimate, na.rm = TRUE)
bw=1

p1 <- ggplot(data20, aes(x = estimate)) +
  geom_histogram(binwidth = bw,alpha=.5) +
  stat_function(
    fun = function(x) dnorm(x, mean = mean_20, sd = sd_20) * 1000 * bw,
    color = "tomato", linewidth = 1
  ) + 
  geom_vline(xintercept = mean_20, color = "blue", linetype = "solid", linewidth = 0.8) +
  geom_vline(xintercept = c(mean_20 - sd_20, mean_20 + sd_20),
             color = "blue", linetype = "dashed", linewidth = 0.8)+
  geom_vline(xintercept = c(mean_20 - 2*sd_20, mean_20 + 2*sd_20), 
             color = "blue", linetype = "dotted", linewidth = 0.8) +
  annotate("text", x = mean_20 - 2*sd_20, y = 100, label = "Mean - 4.47*2",
           angle=90,
           vjust = -0.5, color = "blue", size = 3.5, fontface = "bold") +
  annotate("text", x = mean_20 - sd_20, y = 100, label = "Mean - 4.47",
           angle=90,
           vjust = -0.5, color = "blue", size = 3.5, fontface = "bold") +
  annotate("text", x = mean_20, y = 100, label = "Mean",
           angle=90,
           vjust = -0.5, color = "blue", size = 3.5, fontface = "bold") +
  annotate("text", x = mean_20 + sd_20, y = 100, label = "Mean + 4.47",
           angle=90,
           vjust = -0.5, color = "blue", size = 3.5, fontface = "bold") +
  annotate("text", x = mean_20 + 2*sd_20, y = 100, label = "Mean + 4.47*2",
           angle=90,
           vjust = -0.5, color = "blue", size = 3.5, fontface = "bold")+
  scale_y_continuous("count",limits=c(0,115))+
  labs(x="estimates\n(sample n = 20)")



mean_100 <- mean(data100$estimate, na.rm = TRUE)
sd_100   <- sd(data100$estimate, na.rm = TRUE)
bw2=.45

p2 <- ggplot(data100, aes(x = estimate)) +
  geom_histogram(binwidth = bw2,alpha=.5) +
  stat_function(
    fun = function(x) dnorm(x, mean = mean_100, sd = sd_100) * 1000 * bw2,
    color = "tomato", linewidth = 1
  ) + 
  geom_vline(xintercept = mean_100, color = "blue", linetype = "solid", linewidth = 0.8) +
  geom_vline(xintercept = c(mean_100 - sd_100, mean_100 + sd_100),
             color = "blue", linetype = "dashed", linewidth = 0.8)+
  geom_vline(xintercept = c(mean_100 - 2*sd_100, mean_100 + 2*sd_100), 
             color = "blue", linetype = "dotted", linewidth = 0.8) +
  annotate("text", x = mean_100 - 2*sd_100, y = 100, label = "Mean - 2*2",
           angle=90,
           vjust = -0.5, color = "blue", size = 3.5, fontface = "bold") +
  annotate("text", x = mean_100 - sd_100, y = 100, label = "Mean - 2",
           angle=90,
           vjust = -0.5, color = "blue", size = 3.5, fontface = "bold") +
  annotate("text", x = mean_100, y = 100, label = "Mean",
           angle=90,
           vjust = -0.5, color = "blue", size = 3.5, fontface = "bold") +
  annotate("text", x = mean_100 + sd_100, y = 100, label = "Mean + 2",
           angle=90,
           vjust = -0.5, color = "blue", size = 3.5, fontface = "bold") +
  annotate("text", x = mean_100 + 2*sd_100, y = 100, label = "Mean + 2*2",
           angle=90,
           vjust = -0.5, color = "blue", size = 3.5, fontface = "bold")+
  scale_y_continuous("count",limits=c(0,115))+
  labs(x="estimates\n(sample n = 100)")


p1 + p2
```


Just to give ourselves confirmation, we can compare the calculations above with our simulation, by getting the standard deviation of all the means from samples of size 20, and of all the means from samples of size 100.  
They won't *exactly* match the formula calculations above, but they should be close.  
Check them now.  

::: {.callout-tip collapse="true"}
#### Hint

This is just like calculating the standard deviation for any other variable! 

:::flash
🗂️
[Measuring Variability (standard deviation)](https://uoepsy.github.io/dapr1/2627/flashcards/block1/variability.html#standard-deviation-sd){target="_blank"} flash card.
:::

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
getwd()

data20 |>
  summarise(
    sd = sd(estimate)
  )

data100 |>
  summarise(
    sd = sd(estimate)
  )
```

`r solend()`



:::frame
__Finishing up__  

Once you've finished, don't forget to: 

1. Save your R script
2. Post the R script to the group discussion space on Learn, to ensure that everyone has access to it! (For step-by-step instructions on how to do this, click [here](00serverfiles.html){target="_blank"})  


:::








