---
title: "W9: Confidence Intervals"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
format: html
---

<!--
WHEN RENDERING: 

- tutor-facing: 
  - SHOW_SOLS: TRUE
  - render in RStudio, or in Terminal using `quarto render FILE.qmd`
  - resulting file will land in `docs/YEAR/workshops/tutors/`
  
- student-facing:
  - SHOW_SOLS: FALSE
  - render in Terminal using `quarto render FILE.qmd --profile production`
  - resulting file will land in `docs/YEAR/workshops/`
-->

```{r}
#| include: false
library(tidyverse)
source('assets/dropdowns.R')
source('assets/setup.R')
```

<!-- MAX 5 QUESTIONS -->
<!-- make each question ~10 minute work -->
<!-- MAX 5 QUESTIONS -->
<!-- MAX 5 QUESTIONS -->

:::frame
Pick someone new to be the driver this week. If everyone in the group has already had a week as the driver, start going round again!  

The driver will type the code. The rest of you are navigators, you need to tell the driver where to go (i.e., what to type!). 

Driver instructions:

1. Log in to the bigger screen at your table
2. Open [RStudio Server](https://rstudio.ppls.ed.ac.uk){target="_blank"}
3. Open a new .R script, and load the **tidyverse** package.
4. Listen to your navigators

Navigator instructions:

1. Tell your driver where to go (i.e., what to type)!

:::

`r qbegin(qcounter())`

The code below gives us the means from lots and lots of samples. Last week we saw this used to get estimated means from 1000 samples, but this week we'll up it to 10,000!  

```{r}
means100 <- replicate(10000, mean(rnorm(n = 100, mean = 35, sd = 20)))
```

::: {.callout-tip collapse="true"}
#### reminder/breakdown of how the code works

This bit of code takes a sample of size 100, from a distribution that has a mean of 35 and a standard deviation of 20.  
It gives 100 numbers --- in our example, each one of them is a persons' "outlook" score.  
```{r}
#| eval: false
rnorm(n = 100, mean = 35, sd = 20)
```

Wrapping the whole bit inside `mean()` makes it so that instead of returning all 100 scores in the sample, it returns just the mean of the 100 scores
```{r}
#| eval: false
mean( rnorm(n = 100, mean = 35, sd = 20) )
```

Finally, the function `replicate()` is used to "run code n times". In this case, we're running the `mean(rnorm(...))` bit of code 10000 times. Each time, it takes a sample of 100 people's scores and returns the estimated mean. So we get back 10000 means.  

```{r}
#| eval: false
replicate(10000, mean( rnorm(n = 100, mean = 35, sd = 20) ) )
```

Finally, it stores those 10000 means in R using the name `means100`  

:::

Because `means100` is a vector (it just has one dimension), if we want to find the 10th entry, we can use `<vector name>[<entry-number>]` to print it.  
*(Note, if you do this, you will get a different number, because you have a different set of 10000 samples!)*  
```{r}
means100[10]
```

We also saw how we can create a dataframe object (with rows and columns), by using code such as the below. We've added a little extra bit here. Line 5 takes the dataframe created in lines 2-4, and arranges it (i.e. sorts it) according to the value in the `estimate` column. Which means the first row is now the smallest estimate, and the last row (row 10000) is the largest. 
```{r}
#| code-line-numbers: true
data100 <- 
  tibble(
    estimate = means100
  ) |> 
  arrange(estimate)
```


**Task:** What is the estimate for the 5000th row of the dataframe?  
Bonus challenge: can we find this out another way?  

:::flash
🗂️
See [TODO df\[r,c\] flashcard](https://uoepsy.github.io/dapr1/2627/flashcards/){target="_blank"} flash card.
:::


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
Here is the 5000th row for me. 
Yours will be different, because each time we run the `means100 <- replicate(10000, mean(rnorm......` bit of code, it takes a new 10000 random samples.  
```{r}
data100[5000, ]
```

Because we have estimates from 10000 samples, ordered from smallest to largest, then the 5000th row is the one where 50% are above, and 50% are below - it's the median! 
```{r}
data100 |>
  summarise(
    med_est = median(estimate)
  )
```

`r solend()`



`r qbegin(qcounter())`

Create a new column of standardised (z-scored) estimates. 


:::flash
Look back to how we did this in week 7! See also...

🗂️
[TODO z score calc flashcard](https://uoepsy.github.io/dapr1/2627/flashcards/){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
data100 <- 
  data100 |>
  mutate(
    est_Z = (estimate - mean(estimate)) / sd(estimate)
  )
```


`r solend()`



`r qbegin(qcounter())`

last week we saw two ways to quantify how much variability in an estimated mean we would expect due to sampling:  

1. We could generate lots of samples of size $n$, estimate their means, and calculate the standard deviation of all those estimated means
2. We could use the formula $\frac{\sigma}{\sqrt{n}}$, where $\sigma$ is the standard deviation of scores in the population, and $n$ is the size of the sample.  

Both of these give us the "standard error", or 'how estimates from samples of size n vary around the population mean'. Another very useful way to express the precision in our estimate is to give the range between which 95% of the sample estimates will fall. This is known as a "confidence interval".  

We can, again, do this by either:  

1. Generate lots of samples of size $n$, estimate their means, and see where the middle 95% of the means fall
2. Use the formula $\frac{\sigma}{\sqrt{n}}$, along with the fact that 95% of a normal distribution falls within 1.96 SD of the mean.  

We already have our "lots and lots of samples", so let's do 1 now.  

**Task:** Because the data are already ordered by size of the estimate, and we have 10000 of them, we simply need to get the middle 9500 estimates. Extract _two_ rows of the data, one that is the first row of the middle 9500, and one that is the last row of the middle 9500.. 
The value of `estimate` for these rows will give us the interval!  

Bonus-- what do you notice about the values of the z-scored/standardised column for these rows? They should be close to +/-1.96!

::: {.callout-tip collapse="true"}
#### Hint

if we have 10000 things, then the middle 9500 are between 250 and 9750!  

:::flash
🗂️
See [TODO df\[r,c\] flashcard](https://uoepsy.github.io/dapr1/2627/flashcards/){target="_blank"} flash card.
:::


:::


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
data100[c(250,9750), ]
```


`r solend()`



`r qbegin(qcounter())`
Let's compare the interval we got in the previous question, to the one we get when we: 

2. Use the formula $\frac{\sigma}{\sqrt{n}}$, along with the fact that 95% of a normal distribution falls within 1.96 SD of the mean.  

We know that the mean is 35, and the standard deviation ($\sigma$) is 20. And we are taking samples of size $n = 100$.
All this means that we can calculate a confidence interval using:  

$$
\begin{align}
\mu \pm 1.96 \cdot \frac{\sigma}{\sqrt{n}} \\
\quad \\
35 \pm 1.96 \cdot \frac{20}{\sqrt{2}} \\
\end{align}
$$

Check that it matches the interval from the previous question (it might not be *exact*, but it should be very close!)


::: {.callout-tip collapse="true"}
#### Hint

This is just plugging in numbers and using R like a basic calculator.  

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
35 - (1.96 * 20/sqrt(100))
35 + (1.96 * 20/sqrt(100))
```

`r solend()`



`r qbegin(qcounter())`
In all of what we have done this week and last week, we've been assuming that **we already know the population mean and population standard deviation**!! 

When we have been generating samples, we knew to put 35 and 20 into the code `rnorm(100, mean = .., sd = ..)`.  

And when we have been using the formula, we have used those values in place of $\mu$ and $\sigma$.  

But in reality, **we don't know anything about the population**. That's what we're actually trying to learn about!  

In real research, all we have is a single sample, and we have to go from there... 

Our *real* sample is all of you! 

```{r}
sdata <- read_csv("https://uoepsy.github.io/data/dapr1_2526_survey.csv") |>
  select(pseudonym, outlook)
```

what do we do? well, we use our best guesses for $\mu$ and $\sigma$ - we use the mean ($\bar x$) and standard deviation ($s$) from our sample!  

$$
\begin{align}
\bar x \pm 1.96 \cdot \frac{s}{\sqrt{n}} \\
\end{align}
$$

**Task:** calculate a confidence interval for the mean outlook score using the sample of `r length(sdata$outlook)` DAPR1 students.  

:::flash
🗂️
See [flashcard](https://uoepsy.github.io/dapr1/2627/flashcards/){target="_blank"} flash card.
:::

::: {.callout-tip collapse="true"}
#### Hint

We'll need to replace parts of the calculation we used in the previous question. Instead of $n=100$ we now have $n = `r length(sdata[['outlook']])`$, and we will also need to replace the numbers 35 and 20 with code to calculate sample estimate: `mean(sdata$outlook)` and `sd(sdata$outlook)`

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
mean(sdata$outlook) - (1.96 * sd(sdata$outlook)/sqrt(41))
mean(sdata$outlook) + (1.96 * sd(sdata$outlook)/sqrt(41))
```

`r solend()`




:::frame
__Finishing up__  

Once you've finished, don't forget to: 

1. Save your R script
2. Post the R script to the group discussion space on Learn, to ensure that everyone has access to it! (For step-by-step instructions on how to do this, click [here](00serverfiles.html){target="_blank"})  


:::






