---
title: "W10: Null Hypothesis Significance Testing (NHST)"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
format: html
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false
library(tidyverse)
source('assets/dropdowns.R')
source('assets/setup.R')
```

<!-- MAX 5 QUESTIONS -->
<!-- make each question ~10 minute work -->
<!-- MAX 5 QUESTIONS -->
<!-- MAX 5 QUESTIONS -->

:::frame
Pick someone new to be the driver this week. If everyone in the group has already had a week as the driver, start going round again!  

The driver will type the code. The rest of you are navigators, you need to tell the driver where to go (i.e., what to type!). 

Driver instructions:

1. Log in to the bigger screen at your table
2. Open [RStudio Server](https://rstudio.ppls.ed.ac.uk){target="_blank"}
3. Open a new .R script, and load the **tidyverse** package.
4. Listen to your navigators

Navigator instructions:

1. Tell your driver where to go (i.e., what to type)!

:::

TODO CHECK UN - are you okay with all of this being called Z even though it should technically be t? the idea being that the explanation (at least a very brief one) comes next week
TODO CHECK UN - would you prefer this is two-sided?

`r qbegin(qcounter())`

This week, we're going to draw on the ideas we covered over the last few weeks and go step-by-step through the process of performing a 'null hypothesis significance test'.  

We start with a research question: 

> **Research Question:** Are first year undergraduate students more optimistic than adults in general?  
>   *Operationalisation*: We surveyed students on a first year undergraduate statistics course (that's you!), and asked them to indicate their level of pessimism/optimism on a sliding scale from -100 (pessimist) to +100 (optimist). Previous research using this measure suggests that the overall adult population has an average score of 35.  
> **Data:** [https://uoepsy.github.io/data/dapr1_2526_survey.csv](https://uoepsy.github.io/data/dapr1_2526_survey.csv){target="_blank"}  

**Task:** Define the null and alternative hypotheses.  


:::flash
🗂️
See [TODO defining h0 h1? flashcard](https://uoepsy.github.io/dapr1/2627/flashcards/){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`


$H_0: \mu - 35 = 0$
$H_1: \mu - 35 > 0$

`r solend()`



`r qbegin(qcounter())`
In future, we're going to learn and use functions in R that do all the steps of NHST for us, but for this week we're going to break down the process, so we're going to need all the different parts.   

We're going to need:  

a. the size of our sample
b. the mean that we observe in our sample
c. the hypothesised value that we are testing if our sample mean is equal to
d. (an estimate of) how much we expect samples of our size to vary

Write down or calculate each of these now.  

::: {.callout-tip collapse="true"}
#### Hints

for d., this is the "standard error". We have seen the formula for this a few times, it's $\frac{\sigma}{\sqrt{n}}$. As this is real research, we don't know $\sigma$ (the standard deviation of scores in the population), so we'll estimate it using $s$, the standard deviation in our sample.   

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
library(tidyverse)
sdata <- read_csv("https://uoepsy.github.io/data/dapr1_2526_survey.csv")

nrow(sdata)
mean(sdata$outlook)
35
sd(sdata$outlook) / sqrt(nrow(sdata))
```

`r solend()`


`r qbegin(qcounter())`

Together, the information from the previous question allow us to build up a picture of "what would sample means be like **if** the null hypothesis ($H_0$) were true".  
We can then calculate the probability of seeing a sample mean at least as extreme as the one we _have_ seen, if the null hypothesis were true:    

```{r}
#| echo: false
#| label: fig-nhst
#| fig-cap: "Null hypothesis testing in a nutshell. SE stands for 'standard error'."
set.seed(2394)
samplemeans <- replicate(2000, mean(rnorm(n=100, mean=0, sd=5)))
ss = sd(samplemeans)
g <- ggplot(data=tibble(samplemeans),aes(x=samplemeans))+
  #geom_histogram(alpha=.3)+
  stat_function(geom="line",fun=~dnorm(.x, mean=0,sd=sd(samplemeans))*270,lwd=1)

ld <- layer_data(g) |> filter(x <= sd(samplemeans)+.02 & x >= (-sd(samplemeans)-.02))
ld2 <- layer_data(g) |> filter(x <= 2*sd(samplemeans)+.02 & x >= (-2*sd(samplemeans))-.02)
ld3 <- layer_data(g) |> filter(x >= 1.2)
ld4 <- layer_data(g) |> filter(x <= -1.2)

g + geom_area(data=ld,aes(x=x,y=y),fill="grey30",alpha=.3) + 
  geom_area(data=ld2,aes(x=x,y=y),fill="grey30",alpha=.1) +
  geom_area(data=ld3,aes(x=x,y=y),fill="tomato1",alpha=.3) +
  geom_area(data=ld4,aes(x=x,y=y),fill="tomato1",alpha=.3) +
  geom_segment(aes(x=0,xend=0,y=0,yend=dnorm(0,0,sd=sd(samplemeans))*280), lty="dashed")+
  geom_segment(aes(x=1.2,xend=1.2,y=0,yend=180), lty="dashed", col="#2b8ce0")+ 
  # geom_vline(aes(xintercept=1.2),lty="dashed",col="tomato1")+
  labs(x = "statistic")+
  scale_y_continuous(NULL, breaks=NULL, limits=c(0,275))+
  annotate("text",x=0, y=260, label="population mean\naccording to\nnull hypothesis", col="grey30")+
  annotate("text",x=-1.5, y=210, label="means we would expect from\nsamples of size n if the\nnull hypothesis is true", col="grey30")+
  annotate("text",x=1.2, y=220, label="mean we observe\nin our sample", col="#2b8ce0")+
  annotate("text",x=1.5, y=40, label="probability of seeing\na mean at least\nas extreme\nif null is true", col="tomato1", hjust=0,size=4)+
  
  annotate("text",x=-1.5, y=40, label="probability of seeing\na mean at least\nas extreme (in other\ndirection) if null is true", col="tomato1", hjust=1,size=4)+
  
  
  geom_curve(aes(x=0, xend=0, y=240, yend=225), col="grey30", size=0.5, curvature = 0, arrow = arrow(length = unit(0.03, "npc")))+
  geom_curve(aes(x=-1, xend=-0.5, y=190, yend=150), col="grey30", size=0.5, curvature = 0.2, arrow = arrow(length = unit(0.03, "npc")))+
  geom_curve(aes(x=1.2, xend=1.2, y=205, yend=185), col="#2b8ce0", size=0.5, curvature = 0, arrow = arrow(length = unit(0.03, "npc"))) +
  geom_curve(aes(x=1.45, xend=1.3, y=35, yend=2.5), col="tomato1", size=0.5, curvature = 0, arrow = arrow(length = unit(0.02, "npc")))+
  geom_curve(aes(x=-1.45, xend=-1.3, y=35, yend=2.5), col="tomato1", size=0.5, curvature = 0, arrow = arrow(length = unit(0.02, "npc")))+
  scale_x_continuous(NULL, breaks=c(-2:2)*ss, 
                     labels=c("-2*SE","-1*SE","","1*SE","2*SE"),
                     limits=c(-2.5,2.5))+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  
```


To calculate the probability areas, we're going to need to get convert our mean into a "test statistic". 

This is called a "Z-statistic", and it bears a similar resemblance to what we did a few weeks back in calculating "Z-scores". In this case, however, we are doing the process on the weird abstract theoretical distribution of "the means from all the possible samples we _could have taken_".  

$$
Z = \frac{M - \mu_0}{SE}
$$

**Task:** Calculate Z 

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
zstat <- 
  ( mean(sdata$outlook) - 35 ) / 
    ( sd(sdata$outlook) / sqrt(nrow(sdata)) )

zstat
```

`r solend()`


`r qbegin(qcounter())`
Now that we have our test statistic, we can calculate the "p-value".  
This is "probability of seeing a sample mean at least as extreme as the one we _have_ seen, if the null hypothesis were true", and corresponds to one, or both, of the shaded red areas in @fig-nhst, depending on how we phrased $H_1$.  

Z statistics are normally distributed with a mean of 0 and standard deviation of 1, so we just need to use `pnorm()` to find the area of the normal distribution curve that is beyond *our* Z statistic.  

::: {.callout-tip collapse="true"}
#### Hint

think i need a hint here about direction

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Area greater:  
```{r}
1-pnorm(zstat)
```


`r solend()`



`r qbegin(qcounter())`

Fill in the blanks 

TODO take example write up from Lecture when available, and leave blanks


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`


`r solend()`




:::frame
__Finishing up__  

Once you've finished, don't forget to: 

1. Save your R script
2. Post the R script to the group discussion space on Learn, to ensure that everyone has access to it! (For step-by-step instructions on how to do this, click [here](00serverfiles.html){target="_blank"})  


:::








