---
title: "W13: Chi-squared test of independence"
format: html
---

<!--
WHEN RENDERING: 

- tutor-facing: 
  - either use the build button, or render specific files. or in the terminal using `quarto render`
  - resulting file will land in `docs/YEAR/workshops/tutors/`
  - solutions will always be visible
  
- student-facing:
  - set current_block: to the block we're in
  - render in Terminal using either: 
      - `quarto render --profile production`
      - `quarto render FILE.qmd --profile production`
  - resulting file will land in `docs/YEAR/workshops/`
  - solutions will be available for previous blocks.  
  
-->

```{r}
#| include: false
library(tidyverse)
source('assets/dropdowns.R')
source('assets/setup.R')

vars <- readRDS("block_sols.rds")
# set params manually rather than yaml params
params <- list(
  SHOW_SOLS = tryCatch({ vars$sols[[tools::file_path_sans_ext(knitr::current_input(dir=FALSE))]]},
                       error = function(e) { TRUE } ),
  TOGGLE = TRUE
)
```



```{r}
attach_data <- read_csv('https://uoepsy.github.io/data/attachmentstyle_sexuality.csv')

# **RQ: Do homosexual and heterosexual people show different patterns of attachment styles?**

attach_data |>
  ggplot(aes(x = sexuality, fill = sexuality)) +
  geom_bar() +
  facet_wrap(~ style, nrow = 1) +
  theme(
    legend.position = 'none'
  )
```

This example uses real data from [Ridge & Feeney (1998)](https://onlinelibrary.wiley.com/doi/pdf/10.3109/00048679809073875){target="_blank"}.

---


:::frame
Pick someone new to be the driver this week. They will type the code. 
The rest of you are navigators, you need to tell the driver where to go (i.e., what to type!).  

Driver instructions:

1. Log in to the bigger screen at your table
2. Open [RStudio Server](https://rstudio.ppls.ed.ac.uk){target="_blank"}
3. Listen to your navigators

Navigator instructions:

1. Tell your driver where to go (i.e., what to type)!

:::

<!-- # q1 -->

`r qbegin(qcounter())`

TODO

:::flash
🗂️
See [TODO](TODO){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
TODO
`r solend()`



:::frame
__Finishing up__  

Once you've finished, don't forget to: 

1. Save your R script
2. Post the R script to the group discussion space on Learn, to ensure that everyone has access to it! (For step-by-step instructions on how to do this, click [here](00serverfiles.html){target="_blank"})  


:::








