---
title: "W14: Correlations"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
format: html
---

<!--
WHEN RENDERING: 

- tutor-facing: 
  - SHOW_SOLS: TRUE
  - render in RStudio, or in Terminal using `quarto render FILE.qmd`
  - resulting file will land in `docs/YEAR/workshops/tutors/`
  
- student-facing:
  - SHOW_SOLS: FALSE
  - render in Terminal using `quarto render FILE.qmd --profile production`
  - resulting file will land in `docs/YEAR/workshops/`
-->

```{r}
#| include: false
library(tidyverse)
source('assets/dropdowns.R')
source('assets/setup.R')
```



```{r}
rmp_data <- read_csv('https://uoepsy.github.io/data/ratemyprof_stars.csv')

# **RQ: When students rate instructors as higher difficulty, do they also tend to give those instructors lower star ratings?**

# a task promised in the lecture: recreate this plot yourself.
# the main trick: will need to convert student_star into a factor for the fill to work right
rmp_data |>
  mutate(student_star = factor(student_star)) |>
  ggplot(aes(x = student_difficult, fill = student_star)) +
  geom_bar()
```

This example uses real data from [He (2020)](https://data.mendeley.com/datasets/fvtfjyvw7d/2){target="_blank"}.


---


:::frame
Pick someone new to be the driver this week. They will type the code. 
The rest of you are navigators, you need to tell the driver where to go (i.e., what to type!).  

Driver instructions:

1. Log in to the bigger screen at your table
2. Open [RStudio Server](https://rstudio.ppls.ed.ac.uk){target="_blank"}
3. Listen to your navigators

Navigator instructions:

1. Tell your driver where to go (i.e., what to type)!

:::

<!-- # q1 -->

`r qbegin(qcounter())`

TODO

:::flash
🗂️
See [TODO](TODO){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
TODO
`r solend()`



:::frame
__Finishing up__  

Once you've finished, don't forget to: 

1. Save your R script
2. Post the R script to the group discussion space on Learn, to ensure that everyone has access to it! (For step-by-step instructions on how to do this, click [here](00serverfiles.html){target="_blank"})  


:::








