---
title: "W17: Chi-squared test of independence"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
format: html
---

```{r}
#| include: false
library(tidyverse)
source('assets/dropdowns.R')
source('assets/setup.R')
```


```{r include=F}
# Frequencies from RidgeFeeney1998, Table 2, p. 853

ridgefeeney1998_tab2 <- tribble(
  ~style,   ~sex, ~sexuality, ~count,
  'Secure', 'M',  'Homo', 26,
  'Secure', 'F',  'Homo', 41,
  'Fearful', 'M', 'Homo', 19,
  'Fearful', 'F', 'Homo', 22,
  'Preoccupied', 'M', 'Homo', 16,
  'Preoccupied', 'F', 'Homo', 12,
  'Dismissing', 'M', 'Homo', 16,
  'Dismissing', 'F', 'Homo', 25,
  'Secure', 'M',  'Hetero', 15,
  'Secure', 'F',  'Hetero', 53,
  'Fearful', 'M', 'Hetero', 8,
  'Fearful', 'F', 'Hetero', 26,
  'Preoccupied', 'M', 'Hetero', 10,
  'Preoccupied', 'F', 'Hetero', 11,
  'Dismissing', 'M', 'Hetero', 6,
  'Dismissing', 'F', 'Hetero', 21
)

# aggregating over sex for simplicity's sake
attach_data <- ridgefeeney1998_tab2 |>
  group_by(style, sexuality) |>
  summarise(
    count = sum(count)
  ) |>
  ungroup()

# TODO write_csv and save to data repo
```


:::frame
Pick someone new to be the driver this week. They will type the code. 
The rest of you are navigators, you need to tell the driver where to go (i.e., what to type!).  

Driver instructions:

1. Log in to the bigger screen at your table
2. Open [RStudio Server](https://rstudio.ppls.ed.ac.uk){target="_blank"}
3. Listen to your navigators

Navigator instructions:

1. Tell your driver where to go (i.e., what to type)!

:::

<!-- # q1 -->

`r qbegin(qcounter())`

TODO

:::flash
🗂️
See [TODO](TODO){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
TODO
`r solend()`


```{r}
# **RQ: Do homosexual and heterosexual people show different patterns of attachment styles?**

attach_data |>
  ggplot(aes(x = sexuality, y = count, fill = sexuality)) +
  geom_col() +
  facet_wrap(~ style, nrow = 1) +
  theme(
    legend.position = 'none'
  )
```

This example uses real data from [Ridge & Feeney (1998)](https://onlinelibrary.wiley.com/doi/pdf/10.3109/00048679809073875){target="_blank"}.



:::frame
__Finishing up__  

Once you've finished, don't forget to: 

1. Save your R script
2. Post the R script to the group discussion space on Learn, to ensure that everyone has access to it! (For step-by-step instructions on how to do this, click [here](00serverfiles.html){target="_blank"})  


:::








