---
title: "<b>Correlation</b>"
subtitle: "Data Analysis for Psychology in R 1<br><br> "
author: "dapR1 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: TRUE
      countIncrementalSlides: FALSE
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
baseColour <- '#0F4C81'
knitr::opts_chunk$set(
  dev = "svg",
  warning = FALSE,
  message = FALSE
)

xaringanExtra::use_xaringan_extra(c("tile_view","animate_css","tachyons"))
xaringanExtra::use_extra_styles(
  mute_unhighlighted_code = FALSE
)

library(tidyverse)
library(kableExtra)
library(faux)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
    base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

# Week's Learning Objectives

After this week, you should be able to:

1. Describe the difference between variance, covariance, and correlation

2. Calculate both covariance and correlation

3. Interpret the correlation coefficient

4. Perform and interpret the results of a significance test of your correlation

5. Understand which form of correlation is most appropriate to use with your data

---
# Data for today's class

.center[

```{r, echo=F, message = F, fig.width=7, fig.height=5}
set.seed(526)
dat <- rnorm_multi(100, mu = c(90, 13), sd = c(30,  7), varnames = c('daysInProgram', 'maxDistance'), r = 0.2)

dat$daysInProgram[dat$daysInProgram<0] <- dat$daysInProgram[dat$daysInProgram<0]-min(dat$daysInProgram)
dat$maxDistance[dat$maxDistance<0] <- dat$maxDistance[dat$maxDistance<0]-min(dat$maxDistance)

ggplot(dat, aes(daysInProgram, maxDistance)) + geom_point(alpha = 0.5) + geom_smooth(method = 'lm', se=F, color = baseColour) +
  labs(x='Days In Program', y='Max Distance (miles)') + 
  theme(axis.text = element_text(size=12), axis.title = element_text(size=14, face = 'bold'))
```
]

---

class: inverse, center, middle
# Part 1: The Correlation Coefficient

---

# Variance Recap

.pull-left[
$$s^2_x = \frac{\sum_{i=1}^{n}{(x_i - \bar{x})}^2}{n-1}$$

+ **Variance:** Deviance around the mean of a single variable
]

.pull-right[
```{r, echo = F, fig.height=3, fig.width=5}
sampleDat <- tibble(Names=c('Alfred', 'Bernard', 'Clarence', 'Dorothy', 'Edna', 'Flora', 'Geraldine'),
                    daysInProgram=c(121.57, 99.73, 114.63, 78.37, 136.41, 126.42, 108.94),
                    maxDistance=c(8.53, 10.57, 8.01, 14.63, 27.03, 22.22, 6.77))

p1 <- ggplot(sampleDat, aes(Names, daysInProgram)) + geom_point(size=2, colour = baseColour) +
  labs(x='Participant Name', y = 'Days in Program') + 
  theme(axis.text = element_text(size=10), axis.title = element_text(size=12, face = 'bold')) +
  scale_y_continuous(breaks=seq(80, 140, by = 20), limits = c(70, 145))

p1
```

]

---
count: false

# Variance Recap

.pull-left[
$$s^2_x = \frac{\sum_{i=1}^{n}{(x_i - \bar{x})}^2}{n-1}$$

+ **Variance:** Deviance around the mean of a single variable

+ Raw deviation is the distance between each person's days in the program and the mean number of days in the program. 
]

.pull-right[
```{r, echo = F, fig.height=3, fig.width=5}
p1 + geom_hline(yintercept = mean(sampleDat$daysInProgram)) +
  labs(x='Participant Name', y = 'Days In Program') + 
  geom_segment(x = sampleDat$Names, y = sampleDat$daysInProgram, 
               xend = sampleDat$Names, yend = mean(sampleDat$daysInProgram), linetype = "dashed",
               colour= baseColour)
```
]

---
count: false

# Variance Recap

.pull-left[
$$s^2_x = \frac{\sum_{i=1}^{n}{(x_i - \bar{x})}^2}{n-1}$$

+ **Variance:** Deviance around the mean of a single variable

+ Raw deviation is the distance between each person's days in the program and the mean number of days in the program. 

+ To get the variance, we:
  1. Square the values to get rid of the negative
  2. Sum them up and divide by $n-1$ to get the average deviation of the group from its mean.
]

.pull-right[
```{r, echo = F, fig.height=3, fig.width=5}
p1 + geom_hline(yintercept = mean(sampleDat$daysInProgram)) +
  geom_segment(x = sampleDat$Names, y = sampleDat$daysInProgram, 
               xend = sampleDat$Names, yend = mean(sampleDat$daysInProgram), linetype = "dashed",
               colour= baseColour)
```
]


---
# Covariance Recap

.pull-left[
+ **Covariance:** A value that represents how two variables change together

+ Does $y$ differ from its mean in a similar way to $x$?
]

.pull-right[

```{r, echo = F, fig.height=2.5, fig.width=5}
p1 + geom_hline(yintercept = mean(sampleDat$daysInProgram)) +
  geom_segment(x = sampleDat$Names, y = sampleDat$daysInProgram, 
               xend = sampleDat$Names, yend = mean(sampleDat$daysInProgram), linetype = "dashed",
               colour= baseColour)
```

```{r, echo = F, fig.height=2.5, fig.width=5}
p2 <- ggplot(sampleDat, aes(Names, maxDistance)) + geom_point(size=2, colour = '#BF1932') +
  labs(x='Participant Name', y ='Max Distance (miles)') +
  theme(axis.text = element_text(size=10), axis.title = element_text(size=12, face = 'bold')) + 
  geom_hline(yintercept = mean(sampleDat$maxDistance)) + 
  geom_segment(x = sampleDat$Names, y = sampleDat$maxDistance, 
               xend = sampleDat$Names, yend = mean(sampleDat$maxDistance), linetype = "dashed",
               colour= '#BF1932') +
  scale_y_continuous(breaks=seq(0, 30, 10), limits = c(0, 33))

p2
```

]


---
count: false

# Covariance Recap

.pull-left[
+ **Covariance:** A value that represents how two variables change together

+ Does $y$ differ from its mean in a similar way to $x$?

+ Mathematically similar to variance:


**Variance**
$$s^2_x = \frac{\sum_{i=1}^{n}{(x_i-\bar{x})^2}}{n-1} = \frac{\sum_{i=1}^{n}{(x_i-\bar{x})(x_i-\bar{x})}}{n-1}$$



**Covariance**

$$Cov_{xy} = \frac{\sum_{i=1}^{n}{\color{#0F4C81}{(x_i-\bar{x})}\color{#BF1932}{(y_i-\bar{y})}}}{n-1}$$
]

.pull-right[
```{r, echo = F, fig.height=2.5, fig.width=5}
p1 + geom_hline(yintercept = mean(sampleDat$daysInProgram)) +
  geom_segment(x = sampleDat$Names, y = sampleDat$daysInProgram, 
               xend = sampleDat$Names, yend = mean(sampleDat$daysInProgram), linetype = "dashed",
               colour= baseColour)
```

```{r, echo = F, fig.height=2.5, fig.width=5}
p2
```
]

---

# Covariance Recap

.pull-left[
+ It's possible two variables are related if their observations differ proportionally from their means in a consistent way

+ Covariance gives us a sense of this...

  + High covariance suggests a stronger relationship than a lower covariance

  + Why can't we stop here? 
  
  + Why is correlation necessary? 

]


.pull-right[
```{r, echo = F, fig.height=2.5, fig.width=5}
p1 + geom_hline(yintercept = mean(sampleDat$daysInProgram)) +
  geom_segment(x = sampleDat$Names, y = sampleDat$daysInProgram, 
               xend = sampleDat$Names, yend = mean(sampleDat$daysInProgram), linetype = "dashed",
               colour= baseColour)
```

```{r, echo = F, fig.height=2.5, fig.width=5}
p2
```
]

---

# The Trouble with Covariance

.pull-left[

```{r, echo = F}
sampleDat$varX <- sampleDat$daysInProgram-mean(sampleDat$daysInProgram)
sampleDat$varY <- sampleDat$maxDistance-mean(sampleDat$maxDistance)
sampleDat$varXY <- sampleDat$varX*sampleDat$varY
row.names(sampleDat) <- NULL
```

$$Cov_{xy}=\frac{\sum_{i=1}^n\color{#BF1932}{(x_i-\bar{x})}(y_i-\bar{y})}{n-1}$$
<br>

|        $x_i - \bar{x}$          |
|--------------------------------:|
| `r round(sampleDat$varX[1], 2)` |
| `r round(sampleDat$varX[2], 2)` |
| `r round(sampleDat$varX[3], 2)` |
| `r round(sampleDat$varX[4], 2)` |
| `r round(sampleDat$varX[5], 2)` |
| `r round(sampleDat$varX[6], 2)` |
| `r round(sampleDat$varX[7], 2)` |

]



.pull-right[

```{r, echo = F, fig.width=5, fig.height=2.5}
devX <- round((sampleDat$daysInProgram - mean(sampleDat$daysInProgram)), 2)
p3 <- p1 + geom_hline(yintercept = mean(sampleDat$daysInProgram)) +
  geom_segment(x = sampleDat$Names, y = sampleDat$daysInProgram, 
               xend = sampleDat$Names, yend = mean(sampleDat$daysInProgram), linetype = "dashed",
               colour= baseColour) +
  geom_text(aes(label = devX), nudge_y=c(5, -5, 5, -5, 5, 5, -5))

p3
```

]

---
count: false

# The Trouble with Covariance

.pull-left[

$$Cov_{xy}=\frac{\sum_{i=1}^n(x_i-\bar{x})\color{#BF1932}{(y_i-\bar{y})}}{n-1}$$
<br>

|        $x_i - \bar{x}$          |            $y_i - \bar{y}$      |
|--------------------------------:|--------------------------------:|
| `r round(sampleDat$varX[1], 2)` | `r round(sampleDat$varY[1], 2)` |
| `r round(sampleDat$varX[2], 2)` | `r round(sampleDat$varY[2], 2)` |
| `r round(sampleDat$varX[3], 2)` | `r round(sampleDat$varY[3], 2)` |
| `r round(sampleDat$varX[4], 2)` | `r round(sampleDat$varY[4], 2)` |
| `r round(sampleDat$varX[5], 2)` | `r round(sampleDat$varY[5], 2)` |
| `r round(sampleDat$varX[6], 2)` | `r round(sampleDat$varY[6], 2)` |
| `r round(sampleDat$varX[7], 2)` | `r round(sampleDat$varY[7], 2)` |

]



.pull-right[

```{r, echo = F, fig.width=5, fig.height=2.5}
p3 
```

```{r, echo = F, fig.width=5, fig.height=2.5}
devY <- round((sampleDat$maxDistance - mean(sampleDat$maxDistance)), 2)

p4 <- p2 + geom_text(aes(label = devY), nudge_y=c(-3, -3, -3, 3, 3, 3, -3))

p4
```

]

---
count: false

# The Trouble with Covariance

.pull-left[

$$Cov_{xy}=\frac{\sum_{i=1}^n\color{#BF1932}{(x_i-\bar{x})(y_i-\bar{y})}}{n-1}$$
<br>


|        $x_i - \bar{x}$          |            $y_i - \bar{y}$      | $(x_i - \bar{x})(y_i - \bar{y})$   |
|--------------------------------:|---------------------------------------------------------------------:|
| `r round(sampleDat$varX[1], 2)` | `r round(sampleDat$varY[1], 2)` |  `r round(sampleDat$varXY[1], 2)`  |
| `r round(sampleDat$varX[2], 2)` | `r round(sampleDat$varY[2], 2)` |  `r round(sampleDat$varXY[2], 2)`  |
| `r round(sampleDat$varX[3], 2)` | `r round(sampleDat$varY[3], 2)` |  `r round(sampleDat$varXY[3], 2)`  |
| `r round(sampleDat$varX[4], 2)` | `r round(sampleDat$varY[4], 2)` |  `r round(sampleDat$varXY[4], 2)`  |
| `r round(sampleDat$varX[5], 2)` | `r round(sampleDat$varY[5], 2)` |  `r round(sampleDat$varXY[5], 2)`  |
| `r round(sampleDat$varX[6], 2)` | `r round(sampleDat$varY[6], 2)` |  `r round(sampleDat$varXY[6], 2)`  |
| `r round(sampleDat$varX[7], 2)` | `r round(sampleDat$varY[7], 2)` |  `r round(sampleDat$varXY[7], 2)`  |

]


.pull-right[

```{r, echo = F, fig.width=5, fig.height=2.5}
p3
```

```{r, echo = F, fig.width=5, fig.height=2.5}
p4
```

]

---
count: false

# The Trouble with Covariance

.pull-left[

$$Cov_{xy}=\frac{\color{#BF1932}{\sum_{i=1}^n}(x_i-\bar{x})(y_i-\bar{y})}{n-1}$$

<br>


|        $x_i - \bar{x}$          |            $y_i - \bar{y}$      | $(x_i - \bar{x})(y_i - \bar{y})$   |
|--------------------------------:|---------------------------------------------------------------------:|
| `r round(sampleDat$varX[1], 2)` | `r round(sampleDat$varY[1], 2)` |  `r round(sampleDat$varXY[1], 2)`  |
| `r round(sampleDat$varX[2], 2)` | `r round(sampleDat$varY[2], 2)` |  `r round(sampleDat$varXY[2], 2)`  |
| `r round(sampleDat$varX[3], 2)` | `r round(sampleDat$varY[3], 2)` |  `r round(sampleDat$varXY[3], 2)`  |
| `r round(sampleDat$varX[4], 2)` | `r round(sampleDat$varY[4], 2)` |  `r round(sampleDat$varXY[4], 2)`  |
| `r round(sampleDat$varX[5], 2)` | `r round(sampleDat$varY[5], 2)` |  `r round(sampleDat$varXY[5], 2)`  |
| `r round(sampleDat$varX[6], 2)` | `r round(sampleDat$varY[6], 2)` |  `r round(sampleDat$varXY[6], 2)`  |
| `r round(sampleDat$varX[7], 2)` | `r round(sampleDat$varY[7], 2)` |  `r round(sampleDat$varXY[7], 2)`  |
|                                 |                                 |  **`r round(sum(sampleDat$varXY), 2)`**|

]

.pull-right[

```{r, echo = F, fig.width=5, fig.height=2.5}
p3
```

```{r, echo = F, fig.width=5, fig.height=2.5}
p4
```

]

---

# The Trouble with Covariance

.pull-left[

$$Cov_{xy}=\frac{\sum_{i=1}^n(x_i-\bar{x})(y_i-\bar{y})}{n-1}=\frac{`r round(sum(sampleDat$varXY), 2)`}{7-1}=`r round(sum(sampleDat$varXY)/6, 2)`$$


<br>


|        $x_i - \bar{x}$          |            $y_i - \bar{y}$      | $(x_i - \bar{x})(y_i - \bar{y})$   |
|--------------------------------:|---------------------------------------------------------------------:|
| `r round(sampleDat$varX[1], 2)` | `r round(sampleDat$varY[1], 2)` |  `r round(sampleDat$varXY[1], 2)`  |
| `r round(sampleDat$varX[2], 2)` | `r round(sampleDat$varY[2], 2)` |  `r round(sampleDat$varXY[2], 2)`  |
| `r round(sampleDat$varX[3], 2)` | `r round(sampleDat$varY[3], 2)` |  `r round(sampleDat$varXY[3], 2)`  |
| `r round(sampleDat$varX[4], 2)` | `r round(sampleDat$varY[4], 2)` |  `r round(sampleDat$varXY[4], 2)`  |
| `r round(sampleDat$varX[5], 2)` | `r round(sampleDat$varY[5], 2)` |  `r round(sampleDat$varXY[5], 2)`  |
| `r round(sampleDat$varX[6], 2)` | `r round(sampleDat$varY[6], 2)` |  `r round(sampleDat$varXY[6], 2)`  |
| `r round(sampleDat$varX[7], 2)` | `r round(sampleDat$varY[7], 2)` |  `r round(sampleDat$varXY[7], 2)`  |
|                                 |                                 |  **`r round(sum(sampleDat$varXY), 2)`**|

]

.pull-right[

```{r, echo = F, fig.width=5, fig.height=2.5}
p3
```

```{r, echo = F, fig.width=5, fig.height=2.5}
p4
```

]


---

# The Trouble with Covariance

+ A value of `r round(sum(sampleDat$varXY)/6, 2)` seems high...I think. Is it?

  + Maybe? But maybe not. 
  
  + Covariance is related specifically to the scales of the variables we are analysing. 
  
  + Variables with larger scales will naturally have larger covariance values.


---

# The Trouble with Covariance

+ Consider what would happen if we converted our distance data to kilometers instead of miles.

```{r, echo=F}
sampleDat$maxDistancekm <- sampleDat$maxDistance*1.61
```


.pull-left[

.center[**Miles**]

```{r, echo = 5, fig.height=3.5, fig.width=5}
p4
```


]

.pull-right[

.center[**Kilometers**]

```{r, echo = F, fig.height=3.5, fig.width=5}
sampleDat$varYkm <- sampleDat$maxDistancekm-mean(sampleDat$maxDistancekm)
sampleDat$varXYkm <- sampleDat$varX*sampleDat$varYkm

ggplot(sampleDat, aes(Names, maxDistancekm)) + geom_point(size=2, colour = '#88B04B') +
  labs(x='Participant Name', y = 'Max Distance (km)') + 
  theme(axis.text = element_text(size=10), axis.title = element_text(size=12, face = 'bold')) +
  scale_y_continuous(breaks=seq(20, 50, by = 10), limits = c(0, 50)) + geom_hline(yintercept = mean(sampleDat$maxDistancekm)) +
  geom_segment(x = sampleDat$Names, y = sampleDat$maxDistancekm, 
               xend = sampleDat$Names, yend = mean(sampleDat$maxDistancekm), linetype = "dashed",
               colour= '#88B04B') +
  geom_text(aes(label = devX), nudge_y=c(-3, -3, -3, 3, 3, 3, -3))
```



]

---

# The Trouble with Covariance

.pull-left[

.center[**Miles**]

$$Cov_{xy}=`r round(sum(sampleDat$varXY)/6, 2)`$$


|        $x_i - \bar{x}$          |            $y_i - \bar{y}$      | $(x_i - \bar{x})(y_i - \bar{y})$   |
|--------------------------------:|---------------------------------------------------------------------:|
| `r round(sampleDat$varX[1], 2)` | `r round(sampleDat$varY[1], 2)` |  `r round(sampleDat$varXY[1], 2)`  |
| `r round(sampleDat$varX[2], 2)` | `r round(sampleDat$varY[2], 2)` |  `r round(sampleDat$varXY[2], 2)`  |
| `r round(sampleDat$varX[3], 2)` | `r round(sampleDat$varY[3], 2)` |  `r round(sampleDat$varXY[3], 2)`  |
| `r round(sampleDat$varX[4], 2)` | `r round(sampleDat$varY[4], 2)` |  `r round(sampleDat$varXY[4], 2)`  |
| `r round(sampleDat$varX[5], 2)` | `r round(sampleDat$varY[5], 2)` |  `r round(sampleDat$varXY[5], 2)`  |
| `r round(sampleDat$varX[6], 2)` | `r round(sampleDat$varY[6], 2)` |  `r round(sampleDat$varXY[6], 2)`  |
| `r round(sampleDat$varX[7], 2)` | `r round(sampleDat$varY[7], 2)` |  `r round(sampleDat$varXY[7], 2)`  |
|                                 |                                |  **`r round(sum(sampleDat$varXY), 2)`**|

]

.pull-right[

.center[**Kilometers**]

```{r, echo = F}
sampleDat$varYkm <- sampleDat$maxDistancekm-mean(sampleDat$maxDistancekm)
sampleDat$varXYkm <- sampleDat$varX*sampleDat$varYkm
```


$$Cov_{xy}=`r round(sum(sampleDat$varXYkm)/6, 2)`$$


|        $x_i - \bar{x}$          |            $y_i - \bar{y}$      | $(x_i - \bar{x})(y_i - \bar{y})$   |
|--------------------------------:|---------------------------------------------------------------------:|
| `r round(sampleDat$varX[1], 2)` | `r round(sampleDat$varYkm[1], 2)` |  `r round(sampleDat$varXYkm[1], 2)`  |
| `r round(sampleDat$varX[2], 2)` | `r round(sampleDat$varYkm[2], 2)` |  `r round(sampleDat$varXYkm[2], 2)`  |
| `r round(sampleDat$varX[3], 2)` | `r round(sampleDat$varYkm[3], 2)` |  `r round(sampleDat$varXYkm[3], 2)`  |
| `r round(sampleDat$varX[4], 2)` | `r round(sampleDat$varYkm[4], 2)` |  `r round(sampleDat$varXYkm[4], 2)`  |
| `r round(sampleDat$varX[5], 2)` | `r round(sampleDat$varYkm[5], 2)` |  `r round(sampleDat$varXYkm[5], 2)`  |
| `r round(sampleDat$varX[6], 2)` | `r round(sampleDat$varYkm[6], 2)` |  `r round(sampleDat$varXYkm[6], 2)`  |
| `r round(sampleDat$varX[7], 2)` | `r round(sampleDat$varYkm[7], 2)` |  `r round(sampleDat$varXYkm[7], 2)`  |
|                                 |                                |  **`r round(sum(sampleDat$varXYkm), 2)`**|


]

---

# Correlation

+ Correlation allows you to compare continuous variables with widely different scales without the magnitude of the variables skewing your results.

+ The correlation coefficient, $r$, is the standardised version of covariance:

.f4[

$$r=\frac{\frac{\sum_{i=1}^n(x_i-\bar{x})(y_i-\bar{y})}{n-1}}{\sqrt{\frac{\sum_{i=1}^n(x_i-\bar{x})^2}{n-1}}\sqrt{\frac{\sum_{i=1}^n(y_i-\bar{y})^2}{n-1}}}$$
]


---
count: false

# Correlation

+ Correlation allows you to compare continuous variables across different scales without the magnitude of the variables skewing your results.

+ **Pearson's product moment correlation**, $r$, is the standardised version of covariance:

.f4[

$$r=\frac{\frac{\sum_{i=1}^n(x_i-\bar{x})(y_i-\bar{y})}{n-1}}{\sqrt{\frac{\sum_{i=1}^n(x_i-\bar{x})^2}{n-1}}\sqrt{\frac{\sum_{i=1}^n(y_i-\bar{y})^2}{n-1}}} = \frac{Cov_{xy}}{s_xs_y}$$
]

---
# Correlation

.pull-left[
+ By dividing covariance by the product of the standard deviations of $x$ and $y$, we remove issues with scale differences in the original variables.

+ Because of this, you can use $r$ to investigate the relationships between continuous variables with completely different ranges. 

]


.pull-right[
.center.f3[**Miles**]

<br>

$$r=\frac{`r round(sum(sampleDat$varXY)/6, 2)`}{`r round(sd(sampleDat$daysInProgram),2)`\cdot`r round(sd(sampleDat$maxDistance),2)`}=`r round((sum(sampleDat$varXY)/6)/(sd(sampleDat$daysInProgram)*sd(sampleDat$maxDistance)), 2)`$$

<br>

.center.f3[**Kilometers**]

<br>

$$r=\frac{`r round(sum(sampleDat$varXYkm)/6, 2)`}{`r round(sd(sampleDat$daysInProgram),2)`\cdot`r round(sd(sampleDat$maxDistancekm),2)`}=`r round((sum(sampleDat$varXYkm)/6)/(sd(sampleDat$daysInProgram)*sd(sampleDat$maxDistancekm)), 2)`$$

]

---
# Correlations

+ Correlations measure the degree of association between two variables.

+ If one variable changes, does the other variable also change?

+ If so, do they rise and fall together, or does one rise as the other falls?


---
# Correlation in R

+ To run a simple correlation in R, you can use `cor()`

+ Let's compute the correlation between the number of days in the program and the max running distance for our entire sample of `r nrow(dat)`:

--

```{r}
cor(dat$daysInProgram, dat$maxDistance)
```

--

+ Now we have a correlation value. But what does it mean?

---
# Interpreting $r$

+ Values of $r$ fall between -1 and 1.

  + Values closer to 0 indicate a weaker relationship
  
  + More extreme values indicate a stronger association

  + Interpretation:

```{r, echo = F}
ESint <- tibble(Strength=c('Weak', 'Moderate', 'Strong'), Value = c('.1 < |r| < .3', '.3 < |r| < .5', '|r| > .5'))
kable(ESint, align = c('l', 'c'))
```

---
# Interpreting $r$

+ Values of $r$ fall between -1 and 1.

  + Values closer to 0 indicate a weaker relationship
  
  + More extreme values indicate a stronger association

.center[
```{r, echo = F, message = F, fig.height=4, fig.width = 9}
set.seed(504)
strCor <- rnorm_multi(n=200, mu = c(mean(dat$daysInProgram), mean(dat$maxDistance)), sd = c(sd(dat$daysInProgram),  sd(dat$maxDistance)), varnames = c('daysInProgram', 'maxDistance'), r = 0.8)
strCor$maxDistance[strCor$maxDistance<=0] <- 5
strCor$daysInProgram[strCor$daysInProgram<=0] <- 2
strCor$corrStrength <- 'Strong'

set.seed(1022)
modCor <- rnorm_multi(n=200, mu = c(mean(dat$daysInProgram), mean(dat$maxDistance)), sd = c(sd(dat$daysInProgram),  sd(dat$maxDistance)), varnames = c('daysInProgram', 'maxDistance'), r = 0.45)
modCor$corrStrength <- 'Moderate'
modCor$maxDistance[modCor$maxDistance<=0] <- 5

set.seed(86)
lowCor <- rnorm_multi(n=200, mu = c(mean(dat$daysInProgram), mean(dat$maxDistance)), sd = c(sd(dat$daysInProgram),  sd(dat$maxDistance)), varnames = c('daysInProgram', 'maxDistance'), r = 0.15)
lowCor$corrStrength <- 'Weak'

set.seed(86)
noCor <- rnorm_multi(n=200, mu = c(mean(dat$daysInProgram), mean(dat$maxDistance)), sd = c(sd(dat$daysInProgram),  sd(dat$maxDistance)), varnames = c('daysInProgram', 'maxDistance'), r = 0)
noCor$corrStrength <- 'None'

corDat <- rbind(strCor, modCor, lowCor, noCor)

corDat$corrStrength <- factor(corDat$corrStrength, levels = c('None', 'Weak', 'Moderate', 'Strong'))

corPlot <- ggplot(corDat, aes(daysInProgram, maxDistance)) + 
  geom_point(alpha = 0.7, size = 1) + 
  geom_smooth(method = 'lm', se = F, colour = baseColour) +
  labs(x = 'Days in Program', y = 'Max Distance (miles)')

ann_text <- data.frame(daysInProgram = rep(75, 4),maxDistance = rep(0, 4),
                       lab = c(paste('r =', round(cor.test(noCor$daysInProgram, noCor$maxDistance)$estimate, 2)),
                               paste('r =', round(cor.test(lowCor$daysInProgram, lowCor$maxDistance)$estimate, 2)),
                               paste('r =', round(cor.test(modCor$daysInProgram, modCor$maxDistance)$estimate, 2)),
                               paste('r =', round(cor.test(strCor$daysInProgram, strCor$maxDistance)$estimate, 2))),
                       corrStrength = c('None', 'Weak', 'Moderate', 'Strong'))

ann_text$corrStrength <- factor(ann_text$corrStrength, levels = c('None', 'Weak', 'Moderate', 'Strong'))

corPlot + facet_grid(.~corrStrength) +
  theme(strip.text = element_text(size = 10, face = 'bold')) +
  geom_text(data = ann_text, label = ann_text$lab, size = 4)

```
]

---
# Interpreting $r$

+ The sign of $r$ says nothing about the strength of the relationship, but its direction

  + Positive values indicate that the two variables rise together or fall together.
  
  + Negative values indicate that as one variable increases, the other decreases, and vice versa

.pull-left[
```{r, echo = F, fig.height=3.5, fig.width=5}
set.seed(806)
negCor <- rnorm_multi(n=200, mu = c(mean(dat$daysInProgram), mean(dat$maxDistance)), sd = c(sd(dat$daysInProgram),  sd(dat$maxDistance)), varnames = c('daysInProgram', 'maxDistance'), r = -0.5)
negCor$maxDistance[negCor$maxDistance<=0] <- 5

ggplot(negCor, aes(daysInProgram, maxDistance)) + geom_point() + geom_smooth(method = 'lm', se = F, colour = baseColour) + 
  labs(x='Days in Program', y='Max Distance (miles)') +
  scale_y_continuous(breaks = seq(0, 30, 10), limits = c(0, 35)) +
  theme(axis.text = element_text(size = 10), axis.title = element_text(size = 12, face = 'bold')) + 
  annotate('text', label = paste('r =', round(cor.test(negCor$daysInProgram, negCor$maxDistance)$estimate, 2)),
           x=150, y = 30, size = 4)
```
]

.pull-right[
```{r, echo = F, fig.height=3.5, fig.width=5}
ggplot(modCor, aes(daysInProgram, maxDistance)) + geom_point() + geom_smooth(method = 'lm', se = F, colour = baseColour) + 
  labs(x='Days in Program', y = 'Max Distance (miles)') +
  scale_y_continuous(breaks = seq(0, 30, 10), limits = c(0, 35)) +
  theme(axis.text = element_text(size = 10), axis.title = element_text(size = 12, face = 'bold')) + 
  annotate('text', label = paste('r =', round(cor.test(modCor$daysInProgram, modCor$maxDistance)$estimate, 2)), 
           x = 45, y = 30, size = 4)
```
]
  
---

class: center, middle
# Questions?

---

class: inverse, center, middle
# Part 2: Hypothesis Testing with $r$

---

# Hypotheses

+ In some cases, $r$ is considered a descriptive statistic.

  + $r$ is actually a direct measure of effect size:
  
      + It provides information about the strength of the relationship between two variables.
      
      + It is a standardized measure

+ However, there may be times that a correlation is the test of interest, and we can formulate associated hypotheses tests.

---

# Hypotheses

+ There is no real relationship between two random variables, so the null hypothesis should reflect this.

  + $H_0:r=0$

  + $H_{1\ two-tailed}:r\not=0$
  
  + $H_{1\ one-tailed}:r>0\ \lor\ r<0$

---

# Assumptions of Pearson correlation

1. Variables must be interval or ratio (continuous)
  
  + Knowledge of your data
  
  + No Likert Scales!
	

---

# Assumptions of Pearson correlation

1. Variables must be interval or ratio (continuous)
	
2. Variables must be normally distributed.

--

.pull-left[
.center[**Max Distance**]
```{r, echo = F, fig.height=2.5, fig.width=5}
shapiro.test(dat$maxDistance)
ggplot(dat, aes(maxDistance)) + geom_histogram(aes(y=..density..), bins = 18, fill = baseColour, colour = 'darkgray') +
  geom_density() + labs(x='Max Distance', y = 'Density') +
  stat_function(fun = dnorm, args = list(mean = mean(dat$maxDistance), sd = sd(dat$maxDistance)), colour = '#BF1932')
```
]

.pull-right[
.center[**Days in Program**]
```{r, echo = F, fig.height=2.5, fig.width=5}
shapiro.test(dat$daysInProgram)
ggplot(dat, aes(daysInProgram)) + geom_histogram(aes(y=..density..), bins = 18, fill = baseColour, colour = 'darkgray') +
  geom_density() + labs(x='Days in Program', y = 'Density') +
  stat_function(fun = dnorm, args = list(mean = mean(dat$daysInProgram), sd = sd(dat$daysInProgram)), colour = '#BF1932')
```
]

---

# Assumptions of Pearson correlation

1. Variables must be interval or ratio (continuous)
	
2. Variables must be normally distributed.
	
3. There must be no extreme outliers in your data.

--

.pull-left[
.center[**Ok!**]
```{r, echo = F, fig.height=3.25, fig.width=5}
datPlot <- ggplot(dat, aes(daysInProgram, maxDistance)) + geom_point(alpha = 0.5) + 
  geom_smooth(method = 'lm', se = F, colour = baseColour) +
  labs(x='Days in Program', y = 'Max Distance (miles)') +
  scale_x_continuous(breaks=seq(0, 160, 40), limits=c(0, 165)) +
  scale_y_continuous(breaks=seq(0, 30, 10), limits = c(0, 35))

datPlot
```
]

.pull-right[
.center[**Should be investigated**]
```{r, echo = F, fig.height=3.25, fig.width=5}
dat2 <- dat
dat2[nrow(dat2)+1, ] <- c(100, 49)
dat2[nrow(dat2)+1, ] <- c(250, 18)

ggplot(dat2, aes(daysInProgram, maxDistance)) + geom_point(alpha = 0.5) + geom_smooth(method = 'lm', se = F, colour = baseColour) +
  labs(x='Days in Program', y = 'Max Distance (miles)') +
  scale_x_continuous(limits=c(0, 250)) +
  scale_y_continuous(breaks=seq(0, 50, 10)) +
  annotate('point', x=100, y=49, colour="#BF1932") +
  annotate('point', x=250, y=18, colour="#BF1932")
```
]

---

# Assumptions of Pearson correlation

.pull-left[

1. Variables must be interval or ratio (continuous)
	
2. Variables must be normally distributed.
	
3. There must be no extreme outliers in your data.

4. The relationship between the two variables must be linear.

]

--

.pull-right.center[
**Ok!**
```{r, echo = F, fig.height = 2.25, fig.width = 5, message=F}
datPlot
```
<br>
**Should be investigated**
```{r, echo = F, fig.height = 2.25, fig.width = 5, message = F}
nonLinDat <- read.csv('https://uoepsy.github.io/data/anx_perf.csv')
nonLinDat <- nonLinDat[nonLinDat$anxiety<30,]
ggplot(nonLinDat, aes(anxiety, performance)) + geom_point() + geom_smooth(se=F, colour = baseColour)
```
]

---

# Assumptions of Pearson correlation

.pull-left[

1. Variables must be interval or ratio (continuous)
	
2. Variables must be normally distributed.
	
3. There must be no extreme outliers in your data.

4. The relationship between the two variables must be linear.
	
5. Homoscedasticity (homogeneity of variance)   

]

--

.pull-right.center[
**Ok!**
```{r, echo = F, fig.height = 2.25, fig.width = 5, message=F}
datPlot
```
<br>
**Should be investigated**
```{r, echo = F, fig.height = 2.25, fig.width = 5, message = F}
set.seed(113)
dat2 <- rbind(dat, rnorm_multi(n=150, mu=c(130, 18), sd = c(15, 20), varnames = c('daysInProgram', 'maxDistance')))

ggplot(dat2, aes(daysInProgram, maxDistance)) + geom_point(alpha = 0.5) + geom_smooth(method = 'lm', se = F, colour = baseColour) +
  labs(x='Days in Program', y = 'Max Distance (miles)')
```
]

---

# Significance Testing

Remember the key steps of hypothesis testing:

1. Compute a test statistic

2. Locate the test statistic on a distribution that reflects the probability of each test statistic value, given that H0 is true.

3. Determine whether the probability associated with your test statistic is lower than $\alpha$

---

# Significance Testing

**Compute a test statistic**

+ The sampling distribution for $r$ is approximately normal with a large $n$, and is $t$ distributed when $n$ is small.

  + Thus, significance is assessed using a $t$-distribution

--

+ The $t$-statistic for a correlation is calculated as:

$$t=r\sqrt{\frac{n-2}{1-r^2}}$$
+ So in our example:

```{r, echo = F}
rStat <- round(cor.test(dat$daysInProgram, dat$maxDistance)$estimate, 2)
denom <-round(1-(rStat^2),2)
tStat <- as.numeric(round(rStat*(sqrt((nrow(dat)-2)/denom)),2))
pVal <- as.numeric(round(pt(rStat*(sqrt((nrow(dat)-2)/denom)), df = nrow(dat)-2, lower.tail=F), 2))
```

$$t\ =\ `r rStat`\sqrt{\frac{`r nrow(dat)`-2}{1-`r rStat`^2}}\ =\ `r rStat`\sqrt{\frac{`r nrow(dat)-2`}{`r denom`}}\ =\ `r rStat`\sqrt{`r round((nrow(dat)-2)/denom, 2)`}\ =\ `r tStat`$$
---

# Significance Testing

**Locate the test statistic on a distribution**

.pull-left[
+  We use a $t$ distribution with $n-2$ degrees of freedom
  
  + $n-2$: we had to calculate the means of *two* variables (`daysInProgram` and `maxDistance`)

]


.pull-right[
```{r, echo=F, fig.height=3.5, fig.width=5, message = F}
corTestPlot <- ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = nrow(dat)-2), colour = baseColour, size = 1) +
  labs(x='t', y = 'Probability') + 
  annotate(geom = 'text', x = -2, y = 0.3, label = paste('df =',nrow(dat)-2), colour = baseColour, size = 8) +
  theme(axis.text = element_text(size = 12), axis.title = element_text(size = 14, face = 'bold'))

corTestPlot
```
]

---

# Significance Testing

**Locate the test statistic on a distribution**

.pull-left[
+  We use a $t$ distribution with $n-2$ degrees of freedom
  
  + $n-2$: we had to calculate the means of *two* variables (`daysInProgram` and `maxDistance`)

]


.pull-right[
```{r, echo=F, fig.height=3.5, fig.width=5, message = F}
corTestPlot + geom_vline(xintercept=tStat, linetype = 'dashed', color = baseColour)
```
]

---

# Significance Testing

**Determine whether the probability associated with your test statistic is lower than $\alpha$**

.pull-left[

+ We will use two-tailed $\alpha$ = .05

+ The probability of a test statistic at least as extreme as `r tStat` is only `r pVal`. 

+ $`r pVal`<.05$, so we conclude our results are significant.

]

.pull-right[
```{r, echo=F, fig.height=3.5, fig.width=5, message = F}
testDat <- seq(tStat, 4, by = .1)
dTest <- dt(testDat, df = nrow(dat)-2)
testDat <- data.frame(vals=testDat, prob=dTest)

corTestPlot + geom_vline(xintercept=tStat, linetype = 'dashed', color = baseColour) +
  geom_area(data = testDat, aes(x=vals,y=prob), fill = baseColour) 
```
]

---

# Significance Testing in R

+ To run a full hypothesis test on a correlation, you can use `cor.test()`

```{r}
cor.test(dat$daysInProgram, dat$maxDistance)
```

+ You'll note values are very slightly different here than on the previous slides due to rounding differences.

---

class: center, middle
# Questions?

---

class: inverse, center, middle
# Part 3: Other Types of Correlation

---

# Types of Correlation

| Variable 1  | Variable 2  | Correlation Type |
|-------------|-------------|------------------|
| Continuous  | Continuous  | Pearson          |
| Continuous  | Categorical | Polyserial       |
| Continuous  | Binary      | Biserial         |
| Categorical | Categorical | Polychoric       |
| Binary      | Binary      | Tetrachoric      |
| Rank        | Rank        | Spearman         |
| Nominal     | Nominal     | Chi-square       |


---
# Spearman's Correlation

+ Spearman's $\rho$ (or rank-order correlation) uses data on the rank-ordering of $x$, $y$ responses for each individual.
  
+ Spearman's $\rho$ is a nonparametric version of Pearson's $r$, so it doesn't require the same constraints on your data

+ When would we choose to use the Spearman correlation?
	+ If our data are naturally ranked data (e.g. a survey where the task is to rank foods and drinks in terms of preference)
	+ Our data are ordinal (e.g., Likert Scales)
	+ If the data are non-normal or skewed
	+ If the data shows evidence of non-linearity

---
# Spearman's Correlation

+ Spearman's is not testing for linear relations, it is testing for increasing monotonic relationship.

--

  + What?

--

.pull-left.center[
**Linear**
```{r, echo = F, fig.height=3.5, fig.width=5}
mono <- tibble(
  A = c(1,2,3,4,5,6,7,8,9,10),
  B = c(1,2,3,4,5,6,7,8,9,10),
  C = c(1,4,5,6,8,9,10,13,15,16)
)

lp <- mono %>%
  ggplot(., aes(x=A, y=B)) +
  geom_point() + 
  geom_line()

mp <- mono %>%
  ggplot(., aes(x=A, y=C)) +
  geom_point() + 
  geom_line()

lp
```
<br>
A perfectly linear relationship between A & B

]

.pull-right.center[
**Increasing Monotonic**
```{r, echo = F, fig.height=3.5, fig.width=5}
mp
```
<br>
A perfectly increasing monotonic relationship between A & C
]

---
# Monotonic Relationship

+ **Perfect Monotonic Relationship:** The rank position of all observations on Variable A is the same as the rank position of all observations on Variable C.

.pull-left[

```{r, echo = F}
tibble(
  A = c(1,2,3,4,5,6,7,8,9,10),
  B = c(1,2,3,4,5,6,7,8,9,10),
  C = c(1,4,5,6,8,9,10,13,15,16)
) %>%
  mutate(
    ID = paste("ID", 1:10, sep = ""),
    Rank_A = rank(A),
    Rank_C = rank(C)
  ) %>%
  dplyr::select(., ID, A, C, Rank_A, Rank_C) %>%
  kable() %>%
  kable_styling(full_width = F)

```

]

.pull-right[
```{r, echo = F, fig.height=3.5, fig.width=5}
mp
```

]

---

# Calculating Spearman's $\rho$

$$\rho=1-\frac{6\sum{d^2_i}}{n(n^2-1)}$$
+ $d_i=$ rank( $x_i$ ) $-$ rank( $y_i$ )

+ Steps
  1. Rank each variable from largest to smallest 
  2. Calculate the difference in rank for each person on the two variables
  3. Square the difference
  4. Sum the squared values

---

# Calculating Spearman's $\rho$ in R

+ You can also use `cor()` and `cor.test()` to calculate Spearman's $\rho$ in R:

+ Imagine we want to know whether the participants' ratings (on a 1-5 scale) of the program are associated with how difficult they found the program (on a 1-5 scale).

.pull-left[
<br>
```{r, echo = F}
ratings <- tibble(Names=sampleDat$Names, ProgRating=c(3, 5, 4, 2, 4, 3, 2), Difficulty=c(2, 3, 2, 5, 3, 1, 4))
kable(ratings)
```
]

--

.pull-left[
```{r}
cor(ratings$ProgRating, ratings$Difficulty,
    method='spearman')

cor.test(ratings$ProgRating, ratings$Difficulty,
         method='spearman')
```

]

---

class: center, middle
# Questions?

---

# Summary of Today

+ We reviewed the differences between variance, covariance, and correlation

+ We learned how to calculate both covariance and correlation

+ We discussed how to interpret both the correlation coefficient and the results of the associated significance test

+ We reviewed other methods for correlation and calculated Spearman's $\rho$

**Tomorrow**
+ Live R - We'll review how to run a correlation in R, generate correlation matrices and plot a correlogram.

---

class: center, middle
# Thanks for Listening!


