---
title: "Sort a dataframe with `arrange()`"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```


We'll use the `msleep` dataset that comes from `tidyverse`.
`msleep` contains some data about how much different kinds of mammals sleep.
Here's what the first few rows of `msleep` look like.

```{r}
msleep |> head()
```

## Sort values ascending

If we have a list of numbers [8, 5, 2, 13, 3], and we sort those numbers *ascending* (i.e., in ascending order), we get [2, 3, 5, 8, 13].
The smallest values come first.

Let's sort the `sleep_total` column ascending, using the tidyverse function `arrange()`.
The top of the data frame will show us the mammals who sleep the least.

The default behaviour of `arrange()` is to sort the given column ascending.

```{r}
msleep |>
  arrange(sleep_total)
```

(TIL that giraffes sleep less than two hours a day??)


## Sort values descending

If we have a list of numbers [8, 5, 2, 13, 3] and we sort those numbers *descending* (i.e., in descending order), we get [13, 8, 5, 3, 2].
The largest values come first.

Let's sort the `sleep_total` column descending.
To tell `arrange()` that we want to sort this column in descending order, we need to  first wrap the column name in the function `desc()` and then wrap that in `arrange()`, as below:

```{r}
msleep |>
  arrange(desc(sleep_total))
```

Turns out bats, opossums, and armadillos sleep a *lot*!



## Sorting works for numbers and letters

If you use `arrange()` on a numeric column, as above, it'll sort the values in numeric order.

If you use `arrange()` on an alphabetical column, it'll sort the values in alphabetical order!

```{r}
msleep |>
  arrange(name)
```


```{r}
msleep |>
  arrange(desc(name))
```