---
title: "One-sample t-test"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```

## Example RQ and data

**RQ: For remote workers, is their average score on the Work-Life Balance (WLB) Scale below the scale's midpoint of 45, indicating reasonable work–life balance?**

```{r include=F}
set.seed(1)
wlb_data <- tibble(
  wlb = rnorm(80, 35, 12) |> round()
) |>
  filter(wlb >= 15, wlb <= 75) |>
  rownames_to_column(var = 'worker_id') |>
  mutate(worker_id = paste0('worker', worker_id))
```

Here's what the first six rows of the data looks like:

```{r}
wlb_data |>
  head()
```


Work–Life Balance (`wlb`) scores can range between 15 and 75.
Higher scores represent greater *conflict* between work commitments and life commitments, so good work–life balance is reflected in *lower* scores.

Here's how we might plot this data (you'll see more plotting options in Block 3):

```{r}
wlb_data |>
  ggplot(aes(x = wlb)) +
  geom_boxplot() +
  labs(
    x = 'Work-life balance score'
  )
```


::: {.callout-caution collapse="true"}
#### How do I hide the meaningless numbers on the y axis?

Add in a couple extra lines of code, where in the `theme()` of the plot, we say "set the axis text on the y axis equal to a blank element", and the same for the "axis ticks on the y axis" (the little horizontal lines).

```{r}
wlb_data |>
  ggplot(aes(x = wlb)) +
  geom_boxplot() +
  labs(
    x = 'Work-life balance score'
  ) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

:::



## Define hypotheses

We want to know whether remote workers' work–life balance score is below the scale's midpoint of 45, on average.

Thus our null hypothesis $H_0$ is that the true average work–life balance score $\mu$ (Greek letter mu, pronounced "mew") is *equal* to 45, or formally

$$
H_0: \mu = 45
$$

And based on our RQ, we know that our alternative hypothesis $H_1$ is that the average work–life balance score $\mu$ is *lower* than 45, or formally

$$
H_0: \mu < 45
$$


## Run the test

Use the function `t.test()`, and give it four pieces of information:

- what data we want to test (usually we name a column of the data frame using dollar sign notation; see the [Subset a dataframe](subset.html){target="_blank"} flash card)
- `mu` = the mean value we want to test against
- `alternative` = `two.sided`, `less`, or `greater`; the direction of the $H_1$
- `conf.level` = 1 minus our $\alpha$ level; in Psychology, this will nearly always be .95.

```{r}
t.test(
  wlb_data$wlb,
  mu = 45,
  alternative = 'less',
  conf.level = .95
)
```


## Report the test

> To find out whether remote workers experience work–life balance that's better (that is, lower) than the midpoint of the Work–Life Balance Scale, we ran a one-sided t-test against the null hypothesis that remote workers' Work–Life Balance score is, on average, equal to 45.
We were able to reject this H0; remote workers have a significantly lower Work–Life Balance score of 37.2 on average ($t$(76) = –7, $p$ < .001, one-tailed).


## For Block 3: Check assumptions and data requirements


| Type of requirement / assumption | What to check for a one-sample t-test                                          |
| ------------------------------ | ------------------------------------------------------------------------------- |
| Data requirements              | The variable is numeric.                                                        |
| Assumption of independence     | The obtained sample data are a random sample from the population of interest.   |
| Assumption of normality        | EITHER the population follows a normal distribution<br>OR the sample size $\geq$ 30. |

### Data requirements

**The variable we're analysing should be interval, ratio, or integer data (that is, numeric).**

The `wlb` variable is indeed numeric (specifically, it's integer data!):

```{r}
wlb_data$wlb
```

### Assumption of independence

**The obtained sample data should be a random sample from the population of interest.**

This is an issue of study design more than something we can check directly.

But we can at least check that each person contributed only one data point:

```{r}
wlb_data |> head()
```


Yes, looks good!


### Assumption of normality

**EITHER the population follows a normal distribution OR the sample size is $\geq$ 30.**

#### Check sample size

If each row represents a different observation, then we can figure out the size of the sample by counting the number of rows in the dataframe.

```{r}
nrow(wlb_data)
```

We have 77 observations, which is greater than 30.
This means that, by the Central Limit Theorem, the sampling distribution of sample means is sufficiently normal.
So we can be satisfied that the assumption of normality is met.

For illustration, here's how we'd check normality if we didn't have sufficient sample size:

#### Check normality with a Q-Q plot

The closer the points are to the diagonal line, the better the variable matches an ideal normal distribution.


```{r}
wlb_data |>
  ggplot(aes(sample = wlb)) +  # notice it's "sample", not "x"!
  geom_qq() +
  geom_qq_line() +
  labs(
    x = "Theoretical quantiles",
    y = 'Sample quantiles'
  )
```

The match to the diagonal is overall not too bad here!


#### Check normality through Shapiro-Wilk test

The Shapiro-Wilk test is a hypothesis test where:

- $H_0$: The population data follow a normal distribution
- $H_1$: The population data do not follow a normal distribution

We don't want to reject this H0, so we want a non-significant result!

```{r}
shapiro.test(wlb_data$wlb)
```

To report this test, you might say:

> We performed a Shapiro-Wilk test against the null hypothesis of normality of the population data. 
Based on the sample, we could not reject the null hypothesis of normality in the population ($W$ = 1, $p$ = .6).


## For Block 3: Calculate effect size

For t-tests, the appropriate measure of standardised effect size is Cohen's D.

To compute this in R, first attach the `effectsize` package:

```{r}
library(effectsize)
```

This package gives us the function `cohens_d()`.
Give it the exact same arguments you'd give `t.test()` (see summary above):

```{r}
cohens_d(
  wlb_data$wlb,
  mu = 45,
  alternative = 'less',
  conf.level = .95
)
```

### Interpreting Cohen's D

- Small or weak effect: around 0.2 (whether positive or negative)
- Medium or moderate effect: around 0.5 (whether positive or negative)
- Large or strong effect: around 0.8 or greater (whether positive or negative)


### Reporting effect size

In addition to the information reported above, you might write something like

> With a Cohen's D of –0.80, this difference represents a large effect.