---
title: "Combine plots with `patchwork`"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```


We'll use the `msleep` dataset that comes from `tidyverse`.
`msleep` contains some data about how much different kinds of mammals sleep.
Here's what the first few rows of `msleep` look like.

```{r}
msleep |> head()
```

# Load library

You'll need to (install and) load the library `patchwork`.

```{r}
library(patchwork)
```


# Make plots

Let's say we want to combine these two plots into a single graphic.

The first plot:

```{r}
msleep |>
  ggplot(aes(y = vore)) +
  geom_bar(colour = 'darkslategray', fill = 'darkseagreen')
```

The second plot:

```{r}
msleep |>
  ggplot(aes(x = sleep_total)) +  
  geom_histogram(fill = 'cadetblue1', colour = 'cyan4')
```


# Combine plots

To patchwork two plots together, you'll first need to store each of the plots under their own name, for example:

```{r}
# the first plot
plot_vorebar <- msleep |>
  ggplot(aes(y = vore)) +
  geom_bar(colour = 'darkslategray', fill = 'darkseagreen')

# the second plot
plot_sleephist <- msleep |>
  ggplot(aes(x = sleep_total)) +  
  geom_histogram(fill = 'cadetblue1', colour = 'cyan4')
```

### Side by side

Then as long as you've got the `patchwork` library loaded, you can write the two names with a + sign in between, and you'll get the two plots placed side by side:

```{r}
# the first plot + the second plot
plot_vorebar + plot_sleephist
```

(If `patchwork` is not loaded, you'll get an error about bad use of arithmetic operator, because R won't know how to handle the + sign.)

### One above the other

You can also write the two names with a / sign in between, and you'll get one plot placed on top of the other:

```{r}
# the first plot / the second plot
plot_vorebar / plot_sleephist
```

(If `patchwork` is not loaded, you'll get an error about bad use of arithmetic operator, because R won't know how to handle the / sign.)

You can do a LOT more than this with `patchwork`, it is a very sophisticated tool!
Check out [the patchwork website for more](https://patchwork.data-imaginist.com){target="_blank"}.