---
title: "Explore a normally-distributed variable"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```

For the purposes of DAPR1, you can assume that if we convert a variable into z-scores, that new z-scored variable follows a normal distribution with mean 0 and SD 1.

A normal distribution with these parameters is called a "standard normal distribution".
Here's what it looks like.

```{r echo=F, fig.width=5, fig.height=3}
x <- seq(-4, 4, length.out = 1000)
y <- dnorm(x, mean = 0, sd = 1)
df <- data.frame(x = x, y = y)

stdnorm <- ggplot(df, aes(x = x, y = y)) +
  geom_line(linewidth = 1, color = "#0F4C81") +
  labs(x = "Z", y = "Density") +

  theme_classic() +
  NULL

stdnorm
```

## What's the probability of a value falling above/below some specific z-score?

Probability corresponds to the area under the standard normal distribution.

To find the proportion of the area under the distribution that falls above or below some given z-score, we use the function `pnorm()`.

### Probability below a given z-score

What amount of the area under the distribution—that is, what probability—falls below a z-score of 0?

```{r}
pnorm(0)
```

.5 of the total area under the curve, or a probability of 50%.

Visually:

```{r echo=F, fig.width=5, fig.height=3}
zz <- 0
df2 <- subset(df, x <= zz)

stdnorm +
  geom_area(data = df2, aes(y = y), fill = "#0F4C81", alpha = 0.5) +
  geom_vline(xintercept = zz, linetype = "dashed", color = "red", linewidth = 1) +
  annotate(
    "text", 
    x = 0, y = 0.2, 
    label = paste0(round(pnorm(zz)*100),"%"),
    size = 5, 
    fontface = "bold") +
  scale_x_continuous(breaks=c(-4:4, zz))+
  NULL
```


<br>

How much probability falls below a z-score of 1.5?

```{r}
pnorm(1.5)
```


```{r echo=F, fig.width=5, fig.height=3}
zz <- 1.5
df2 <- subset(df, x <= zz)

stdnorm +
  geom_area(data = df2, aes(y = y), fill = "#0F4C81", alpha = 0.5) +
  geom_vline(xintercept = zz, linetype = "dashed", color = "red", linewidth = 1) +
  annotate(
    "text", 
    x = 0, y = 0.2, 
    label = paste0(round(pnorm(zz)*100),"%"),
    size = 5, 
    fontface = "bold") +
  scale_x_continuous(breaks=c(-4:4, zz))+
  NULL
```


### Probability above a given z-score

By default, `pnorm()` gives us the probability below / to the left of the given z-score.

To find the probability above / to the right of that z-score, we subtract the output of `pnorm()` from 1.

For example, how much probability falls above a z-score of 1.5?
(It should be the approx 7% left over from the 93% in the plot just above.)

```{r}
1 - pnorm(1.5)
```

```{r echo=F, fig.width=5, fig.height=3}
zz <- 1.5
df2 <- subset(df, x >= zz)

stdnorm +
  geom_area(data = df2, aes(y = y), fill = "#0F4C81", alpha = 0.5) +
  geom_vline(xintercept = zz, linetype = "dashed", color = "red", linewidth = 1) +
  annotate(
    "text", 
    x = 0, y = 0.2, 
    label = paste0(round((1 - pnorm(zz))*100),"%"),
    size = 5, 
    fontface = "bold") +
  scale_x_continuous(breaks=c(-4:4, zz))+
  NULL
```


## What z-score has a specific probability of values falling above/below it?

This question is asking about the **quantile** of the probability distribution.

To find the z-score at a given quantile of the distribution—that is, where a given amount of probability falls above or below it—use `qnorm()`.

`qnorm()` takes a probability and tells you the exact z-score where the area under the standard normal curve reaches that amount.


### Z-score with given probability below it

50% of the standard normal distribution falls below / to the left of what z-score?

```{r}
qnorm(.5)
```

```{r echo=F, fig.width=5, fig.height=3}
prob <- .5
zz <- qnorm(prob)
df2 <- subset(df, x <= zz)

stdnorm +
  geom_area(data = df2, aes(y = y), fill = "#0F4C81", alpha = 0.5) +
  geom_vline(xintercept = zz, linetype = "dashed", color = "red", linewidth = 1) +
  annotate(
    "text", 
    x = 0, y = 0.2, 
    label = paste0(round(prob*100),"%"),
    size = 5, 
    fontface = "bold") +
  scale_x_continuous(breaks=c(-4:4, zz))+
  NULL
```

<br>

30% of the standard normal distribution falls below / to the left of what z-score?

```{r}
qnorm(.3)
```

```{r echo=F, fig.width=5, fig.height=3}
prob <- .3
zz <- qnorm(prob)
df2 <- subset(df, x <= zz)

stdnorm +
  geom_area(data = df2, aes(y = y), fill = "#0F4C81", alpha = 0.5) +
  geom_vline(xintercept = zz, linetype = "dashed", color = "red", linewidth = 1) +
  annotate(
    "text", 
    x = 0, y = 0.2, 
    label = paste0(round(prob*100),"%"),
    size = 5, 
    fontface = "bold") +
  scale_x_continuous(breaks=c(-4:4, zz))+
  NULL
```


### Z-score with given probability above it

By default, `qnorm()` gives us the quantile with the given probability below / to the left of it.

To find the quantile where the probability falls above / to the right of it, we take 1, subtract from that the probability we're using, and give that to `qnorm()`:

For example, what's the z-score that has 70% probability above it?
(Should be the same z-score as in the plot just above!)

```{r}
qnorm(1 - .7)
```

Or you can compute this difference in your head and input the result directly into `qnorm()`:

```{r}
qnorm(.3)
```


```{r echo=F, fig.width=5, fig.height=3}
prob <- .3
zz <- qnorm(prob)
df2 <- subset(df, x >= zz)

stdnorm +
  geom_area(data = df2, aes(y = y), fill = "#0F4C81", alpha = 0.5) +
  geom_vline(xintercept = zz, linetype = "dashed", color = "red", linewidth = 1) +
  annotate(
    "text", 
    x = 0, y = 0.2, 
    label = paste0(round((1-prob)*100),"%"),
    size = 5, 
    fontface = "bold") +
  scale_x_continuous(breaks=c(-4:4, zz))+
  NULL
```


