---
title: "Subset a dataframe"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```

There are lots of different ways to subset a dataframe, depending on the kind of output you want.

We'll illustrate using the `msleep` dataset that comes from `tidyverse`.
`msleep` contains some data about how much different kinds of mammals sleep.
Here's what the first few rows of `msleep` look like.

```{r}
msleep |> head()
```


## Extracting cells

### Square-bracket indexing

In computer programming, an "index" is a number that tells you what position some element is in.
For example, a row with index 1 is the first row in the dataframe; a column with index 3 is the third column in the dataframe.
And where those two intersect, in row 1 and column 3, there is one cell of the dataframe.

As for the "square bracket" part: you can name a dataframe (e.g., `my_df`), then write an opening square bracket `[`, write the row index you want, then a comma `,`, then the column index you want, the finish by writing the closing square bracket `]`.

For example, to get the cell in the first row and the third column of `msleep`:

```{r}
msleep[1, 3]
```


## Extracting column(s)

### Dollar-sign notation

Use this method if you want to quickly grab one single column and treat it as a vector.

In dollar-sign notation, we write the name of the dataframe (e.g., `my_df`), a dollar sign `$`, then the name of the column we want to extract(e.g., `my_column`): `my_df$my_column`.

For example, to get the column `name` from `msleep`:

```{r}
msleep$name
```


### `pull()`

Use the tidyverse function `pull()` if you already have a sequence of piped-together functions and want to finish by extracting one single column and treating it as a vector.

You can pipe any dataframe into `pull()` and name the column you'd like to extract:

```{r}
msleep |>
  pull(name)
```



### `select()`

Use the tidyverse function `select()` if you want to select one or more columns and keep them in dataframe/tibble form.

For example, to select only name and genus from `msleep`:

```{r}
msleep |>
  select(name, genus)
```

You can also use `select()` to remove specific columns!
Between brackets, you write a minus sign `-` and then the name of the column to remove.

For example, let's imagine we don't care about the animals' genus after all:

```{r}
msleep |>
  select(-genus)
```

The `genus` column is gone, we go straight from `name` to `vore`!

### Square-bracket indexing

If you only want to extract a column with a specific index, you can do that by leaving the "row" part of the square bracket `[row index, column index]` notation blank.

You still need the comma though!! 
That's how R knows whether it's the column or the row that you're after.

For example, to grab column 4:

```{r}
msleep[, 4]
```

It looks weird, but it works!


## Extract row(s) 

### `slice()`

The tidyverse function `slice()` fits well into a string of piped-together functions.

Imagine we want to keep rows 2 through 7:

```{r}
msleep |>
  slice(2:7)
```



### Square-bracket indexing

If you only want to extract a row with a specific index, you can do that by leaving the "column" part of the square bracket `[row index, column index]` notation blank.

You still need the comma though!! 
That's how R knows whether it's the column or the row that you're after.

For example, to grab row 4:

```{r}
msleep[4, ]
```

### Rows that meet a condition: `filter()`

Use the tidyverse function `filter()` if you want to keep only the rows that match some specific condition.

For example, to keep only the rows of `msleep` that contain carnivores, we want to filter `msleep` for the rows where the value of `vore` is `"carni"`.

(An important note: we write this as `vore == "carni"`, with two equals signs!
Try it yourself with one equals sign and see what happens.)

```{r}
msleep |>
  filter(vore == "carni")
```

There's a LOT more you can do with `filter()`!
[Click here for some examples from the official documentation.](https://dplyr.tidyverse.org/reference/filter.html#ref-examples){target="_blank"}