---
title: "Calculate z-scores"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```


We'll calculate z-scores using the `msleep` dataset that comes from `tidyverse`.
`msleep` contains some data about how much different kinds of mammals sleep.
Here's what the first few rows of `msleep` look like.

```{r}
msleep |> head()
```

**Let's convert each mammal's total hours of sleep (in column `sleep_total`) into z-scores.**

# What is a z-score?

A z-score is a standardised version of a variable.

What does it mean for a variable to be "standardised"?

- The mean of a z-scored variable is always zero, no matter what the variable's original mean was.
- The SD of a z-scored variable is always one, no matter what the variable's original SD was.

These properties are "standard" for all z-scores.

# How to convert a variable into z-scores

The formula for converting one observation from a given variable into a z-score:

$$
\text{z-score} = \frac{\text{observation – variable's mean}}{\text{variable's SD}}
$$

or purely in mathematical form:

$$
Z = \frac{X - M}{SD}
$$


## Method 1: Use `mutate()` from tidyverse

```{r}
msleep <- msleep |>
  mutate(
    Z_sleep_total = (sleep_total - mean(sleep_total)) / sd(sleep_total)
  )
```

<br>

Now let's check the result.

Let's have a look at just the columns `name`, `sleep_total`, and `Z_sleep_total` (for more detail on subsetting using `select()`, see the [Subset a dataframe](subset.qmd){target="_blank"} flash card).

```{r}
msleep |>
  select(name, sleep_total, Z_sleep_total)
```

- Animals with positive z-scores, like cheetahs and three-toed sloths, sleep more than this sample's average.
- Animals with negative z-scores, like cows and vesper mice, sleep less than this sample's average.

Sense check that the mean is 0 and the SD is 1:

```{r}
msleep |>
  summarise(
    mean_Z_sleep = mean(Z_sleep_total),
    sd_Z_sleep   = sd(Z_sleep_total),
  )
```

This mean value `-2.98e-17` is an abbreviated way (using scientific notation) of writing the VERY small number 0.0000000000000000298.


For technical computational reasons, R sometimes will give us very small numbers when really it means zero.

So our mission is accomplished!

## Method 2: Calculate manually using the formula

We extract the column `sleep_total` from `msleep` using the dollar sign notation.
And we create a new column, `Z_sleep_total`, in `msleep`, using the dollar sign notation too.
(for more detail on the dollar sign notation, see the [Subset a dataframe](subset.qmd){target="_blank"} flash card)

```{r}
msleep$Z_sleep_total <- (msleep$sleep_total - mean(msleep$sleep_total)) / sd(msleep$sleep_total)
```


Sense check that the mean is 0 and the SD is 1 (this time, rounding the output to the nearest whole number, so that the very small number above will be rounded to 0):

```{r}
msleep |>
  summarise(
    mean_Z_sleep = round(mean(Z_sleep_total)),
    sd_Z_sleep   = round(sd(Z_sleep_total)),
  )
```


# How to convert z-scores back onto the original scale

You'll need to know the original scale's mean and SD.

RESUME




