---
title: "The standard normal distribution and z-scores"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
```

Z-scores are a useful way to transform data.
How are they related to normal distributions?

Backstory:
The standard normal distribution, denoted $Z \sim N(0, 1)$, is a normal distribution with mean = 0 and standard deviation = 1.

The R functions `dnorm(x, mean, sd)` and `pnorm(x, mean, sd)` assume that mean = 0 and sd = 1 if not provided. In other words, they assume a standard normal distribution by default.

For example, the following two functions return the same output

```{r}
dnorm(2)
dnorm(2, mean = 0, sd = 1)
```

and similarly for `pnorm()`.


In order to transform a value $x$ from a normal distribution with mean = $\mu$ and sd = $\sigma$ to a score on the standard normal scale, we must use the z-score transformation.

The z-score measures distance in terms of the number of standard deviations from the mean:
$$
z = \frac{x - \mu}{\sigma} \qquad (\text{standard units or } z\text{-}scores)
$$

::: callout-note
__z-score__

The z-score of $x$ is the number of standard deviations $x$ is from the mean.
:::

<br>

Let $X \sim N(\mu, \sigma)$ and $Z \sim N(0, 1)$. The following relation holds:
$$
P(X < x) = P \left( \frac{X - \mu}{\sigma} < \frac{x - \mu}{\sigma} \right) = P(Z < z)
$$

This means that the probability, when computed on the original normal distribution or on the standard normal distribution, will be the same as long as you use the z-score of $x$ for the standard normal distribution.

<br>

Consider the following cases:

a. You observe $x = 4.3$ from a $N(3, 2)$ distribution. The z-score is $z = (4.3 - 3) / 2 = 0.65$
b. You observe $x = 4.3$ from a $N(3, 0.2)$ distribution. The z-score is $z = (4.3 - 3) / 0.2 = 6.5$

In case (a), your observed value 4.3 is 0.65 standard deviations away from the mean. 
In case (b), the observed value 4.3 is 6.5 standard deviations away from the mean.

Observing a value of 4.3 from a $N(3,2)$ population is very likely, as 68% of the observations are within 1 $\sigma$ of the mean.
However, observing a value of 4.3 from a $N(3,0.2)$ distribution is very unlikely, as almost all the values will be within 3 $\sigma$ of the mean. Such a high z-score provides us with an element of surprise. We might want to inspect that value further as it could be an outlier or a typing error of who has entered the data.


As you can see, z-scores are particularly useful for comparing observations across distributions having a different mean and standard deviation.
