---
title: "DAPR1"
subtitle: "2022/23"
date: ""
output:
  rmdformats::material:
    highlight: kate
    self_contained: true
    # code_folding: show
    thumbnails: true
    gallery: true
    fig_width: 4
    fig_height: 4
    df_print: kable
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{css}
.header-panel {
    background-color: #0F4C81;
}

a {
  color: #5265a2;
}

```


# Course materials {.tabset .tabset-pills}

_Please note: materials are added incrementally as this course progresses through the year._

## Semester 1

```{r echo=FALSE}
library(tidyverse)
library(kableExtra)

course <- tribble(~a, ~b, ~c, ~d, ~e)

course <- course %>%
    bind_rows(
        c(a = 'Sem. 1: Welcome Week',
          b = 'Course introduction',
          c = '[Lecture](https://uoepsy.github.io/dapr1/2223/lectures/dapr1_welcome_week.pdf)',
          d = '',
          e = '')
    ) %>%
        bind_rows(
        c(a = 'Sem. 1: Week 1',
          b = 'Research Design & Data',
          c = '[Lecture](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec1_ResearchDesign-Data.html#1)',
          d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_01_design_and_data.html)',
          e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_01.html)')
    ) %>%
        bind_rows(
        c(a = 'Sem. 1: Week 2',
          b = 'Describing Categorical Data',
          c = '[Lecture](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec2_DescribingCategoricalData.html#1)',
          d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_02_categorical_data.html)',
          e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_02.html)')
    ) %>%
        bind_rows(
        c(a = 'Sem. 1: Week 3',
          b = 'Describing Continuous Data',
          c = '[Lecture](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec3_DescribingContData.html#1)',
          d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_03_numeric_data.html)',
          e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_03.html)')
    ) %>%
        bind_rows(
        c(a = 'Sem. 1: Week 4',
          b = 'Describing Relationships',
          c = '[Lecture](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec4_DescribingRelationships.html#1)',
          d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_04_relationships.html)',
          e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_04.html)')
    ) %>%
        bind_rows(
        c(a = 'Sem. 1: Week 5',
          b = 'Functions',
          c = '[Lecture](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec5_Functions.html#1)',
          d = '',
          e = '')
    )

names(course) <- c('Week', 'Topic', 'Lecture slides', 'Lab materials', 'Readings')

course %>%
    kbl() %>%
    kable_styling()
```


## Semester 2

Coming next year.


# Course description

Data Analysis for Psychology in R 1 (DAPR1) is a course taken by 1st year students in Psychology. It introduces you to data collection, description, and visualisation using the R programming language. In semester 2, you will also learn how to use experimental data to test simple research hypotheses.


# Archive

- [Year 2021/22](https://uoepsy.github.io/dapr1/2122/index.html)
- [Year 2020/21](https://uoepsy.github.io/dapr1/2021/index.html)

