---
title: "DAPR1"
subtitle: "2022/23"
date: ""
output:
  rmdformats::material:
    highlight: kate
    self_contained: true
    # code_folding: show
    thumbnails: true
    gallery: true
    fig_width: 4
    fig_height: 4
    df_print: kable
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{css}
.header-panel {
    background-color: #0F4C81;
}

a {
  color: #5265a2;
}
```


# Course materials {.tabset .tabset-pills}

_Please note: materials are added incrementally as this course progresses through the year._

## Semester 1

```{r echo=FALSE}
library(tidyverse)
library(kableExtra)

course <- tribble(~a, ~b, ~c, ~d, ~e)

course <- course %>%
    bind_rows(
        c(a = 'Sem. 1: Welcome Week',
          b = 'Course introduction',
          c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapr1_welcome_week.pdf)',
          d = '',
          e = '')
    ) %>%
    bind_rows(
        c(a = 'Sem. 1: Week 1',
          b = 'Research Design & Data',
          c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec1_ResearchDesign-Data.html#1)',
          d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_01_design_and_data.html)',
          e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_01.html)')
    ) %>%
    bind_rows(
        c(a = 'Sem. 1: Week 2',
          b = 'Describing Categorical Data',
          c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec2_DescribingCategoricalData.html#1)',
          d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_02_categorical_data.html)',
          e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_02.html)')
    ) %>%
    bind_rows(
        c(a = 'Sem. 1: Week 3',
          b = 'Describing Continuous Data',
          c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec3_DescribingContData.html#1)',
          d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_03_numeric_data.html)',
          e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_03.html)')
    ) %>%
    bind_rows(
        c(a = 'Sem. 1: Week 4',
          b = 'Describing Relationships',
          c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec4_DescribingRelationships.html#1)',
          d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_04_relationships.html)',
          e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_04.html)')
    ) %>%
    bind_rows(
        c(a = 'Sem. 1: Week 5',
          b = 'Functions',
          c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec5_Functions.html#1)',
          d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_05_formative_report_a.html)',
          e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_05.html)')
    ) %>%
    bind_rows(
        c(a = 'Sem. 1: Week 6',
          b = 'Formative feedback week (A)',
          c = '',
          d = '',
          e = '')
    ) %>%
    bind_rows(
        c(a = 'Sem. 1: Week 7',
          b = 'Probability Theory',
          c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec6_IntroProbability.html#1)',
          d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_07_prob_theory.html)',
          e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_07.html)')
    ) %>%
    bind_rows(
        c(a = 'Sem. 1: Week 8',
          b = 'Probability Rules',
          c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec7_ProbabilityRules.html#1)',
          d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_08_prob_rules.html)',
          e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_08.html)')
    ) %>%
    bind_rows(
        c(a = 'Sem. 1: Week 9',
          b = 'Random Variables (Discrete)',
          c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec8_DiscreteProbabilityDist.html#1)',
          d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_09_discrete_dist.html)',
          e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_09.html)')
    ) %>%
    bind_rows(
        c(a = 'Sem. 1: Week 10',
          b = 'Random Variables (Continuous)',
          c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec9_ContinuousProbabilityDist.html#1)',
          d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_10_cont_dist.html)',
          e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_10.html)')
    ) %>%
    bind_rows(
        c(a = 'Sem. 1: Week 11',
          b = 'Sampling',
          c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec10_Samples-SamplingDist.html#1)',
          d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_11_sampling.html)',
          e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_11.html)')
    ) %>%
    bind_rows(
        c(a = 'Sem. 1: Week 12',
          b = 'Formative feedback week (B)',
          c = '',
          d = '',
          e = '')
    )

names(course) <- c('Week', 'Topic', 'Lecture slides', 'Lab materials', 'Readings')

course %>%
    kbl() %>%
    kable_styling()
```


## Semester 2

```{r}
course.s2 <- tribble(~a, ~b, ~c, ~d, ~e)

course.s2 <- course.s2 %>%
    bind_rows(
        c(a = 'Sem. 2: Week 1',
          b = 'Confidence intervals',
          c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapr1_2_01_confint_2223.pdf)',
          d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/2_01_confint.html)',
          e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd2_01.html)')
    )  %>%
    bind_rows(
        c(a = 'Sem. 2: Week 2',
          b = 'Hypothesis testing: p-values',
          c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapr1_2_02_ht_pvalues.pdf)',
          d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/2_02_ht_pvalues.html)',
          e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd2_02.html)')
    ) %>%
    bind_rows(
        c(a = 'Sem. 2: Week 3',
          b = 'Hypothesis testing: critical values',
          c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapr1_2_03_ht_critvalues.pdf)',
          d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/2_03_ht_critvalues.html)',
          e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd2_03.html)')
    ) %>%
bind_rows(
    c(a = 'Sem. 2: Week 4',
      b = 'Connecting Hypothesis testing and confidence intervals',
      c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapr1_2_04_htci.pdf)',
      d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/2_04_htci.html)',
      e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd2_04.html)')
) %>%
bind_rows(
    c(a = 'Sem. 2: Week 5',
      b = 'Errors, Power, Effect size, and Assumptions',
      c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapr1_2_05_errorspower.pdf)',
      d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/2_05_hterrorspower.html)',
      e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd2_05.html)')
) %>%
bind_rows(
    c(a = 'Sem. 2: Week FLW',
      b = 'Formative feedback week (C)',
      c = '',
      d = '',
      e = '')
) 
# %>%
    # bind_rows(
    #     c(a = 'Sem. 2: Week 7',
    #       b = 'Probability Theory',
    #       c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec6_IntroProbability.html#1)',
    #       d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_07_prob_theory.html)',
    #       e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_07.html)')
    # ) %>%
    # bind_rows(
    #     c(a = 'Sem. 2: Week 8',
    #       b = 'Probability Rules',
    #       c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec7_ProbabilityRules.html#1)',
    #       d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_08_prob_rules.html)',
    #       e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_08.html)')
    # ) %>%
    # bind_rows(
    #     c(a = 'Sem. 2: Week 9',
    #       b = 'Random Variables (Discrete)',
    #       c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec8_DiscreteProbabilityDist.html#1)',
    #       d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_09_discrete_dist.html)',
    #       e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_09.html)')
    # ) %>%
    # bind_rows(
    #     c(a = 'Sem. 2: Week 10',
    #       b = 'Random Variables (Continuous)',
    #       c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec9_ContinuousProbabilityDist.html#1)',
    #       d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_10_cont_dist.html)',
    #       e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_10.html)')
    # ) %>%
    # bind_rows(
    #     c(a = 'Sem. 2: Week 11',
    #       b = 'Sampling',
    #       c = '[Slides](https://uoepsy.github.io/dapr1/2223/lectures/dapR1_lec10_Samples-SamplingDist.html#1)',
    #       d = '[Lab](https://uoepsy.github.io/dapr1/2223/labs/1_11_sampling.html)',
    #       e = '[Reading](https://uoepsy.github.io/dapr1/2223/labs/rd1_11.html)')
    # )

names(course.s2) <- c('Week', 'Topic', 'Lecture slides', 'Lab materials', 'Readings')

course.s2 %>%
    kbl() %>%
    kable_styling()
```



# Course description

Data Analysis for Psychology in R 1 (DAPR1) is a course taken by 1st year students in Psychology. It introduces you to data collection, description, and visualisation using the R programming language. In semester 2, you will also learn how to use experimental data to test simple research hypotheses.


# Archive

- [Year 2021/22](https://uoepsy.github.io/dapr1/2122/index.html)
- [Year 2020/21](https://uoepsy.github.io/dapr1/2021/index.html)


# Site map

[Home](https://uoepsy.github.io/) > [DAPR1](https://uoepsy.github.io/dapr1/)
