---
title: "Connecting hypothesis testing and confidence intervals"
callout-appearance: simple
---


```{r setup}
#| include: false

source('assets/setup.R')
```


```{r include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', 
                      fig.height = 5, fig.width = 6.2, 
                      out.width = '80%')

library(tidyverse)
library(patchwork)

theme_set(theme_light(base_size = 15))

set.seed(123)
```


## Hypothesis testing

Consider the two-sided hypothesis testing case 

$$H_0 : \mu = \mu_0$$
$$H_1 : \mu \neq \mu_0$$

Where the test statistic used in order to test the above claim is:

$$
t = \frac{\bar x - \mu_0}{s / \sqrt{n}}
$$

At the 5% significance level:

- we __reject__ the null hypothesis $H_0$ whenever the observed t-statistic lies beyond the critical values:

$$t \leq -t^* \qquad \text{or} \qquad t \geq +t^*$$

```{r, echo=FALSE}
n <- 50

plt <- tibble(
    x = seq(-5, 5, 0.1),
    y = dt(x, n-1),
)

plt.tstar <- qt(c(0.025, 0.975), n-1)
tobs <- -2.99

ggplot(plt, aes(x = x, y = y)) +
    geom_line(color = 'blue') +
    geom_vline(xintercept =  plt.tstar, color = 'red') + 
    geom_vline(xintercept = tobs, color = 'darkgreen') +
    labs(x = 't-statistic', y = 'Density') +
    annotate('text', x = plt.tstar, y = c(0.02, 0.02), label = c('-t*', '+t*'),
             color = 'red', size = 6, adj = -0.5) +
    annotate('text', x = tobs, y = 0.02, label = 't',
             color = 'darkgreen', size = 6, adj = 1.5)
```


- we __do not reject__ the null hypothesis $H_0$ whenever the observed t-statistic lies within the critical values:

$$-t^* < t < +t^*$$

```{r echo=FALSE}
tobs <- 1.3

ggplot(plt, aes(x = x, y = y)) +
    geom_line(color = 'blue') +
    geom_vline(xintercept =  plt.tstar, color = 'red') + 
    geom_vline(xintercept = tobs, color = 'darkgreen') +
    labs(x = 't-statistic', y = 'Density') +
    annotate('text', x = plt.tstar, y = c(0.02, 0.02), label = c('-t*', '+t*'),
             color = 'red', size = 6, adj = -0.5) +
    annotate('text', x = tobs, y = 0.02, label = 't',
             color = 'darkgreen', size = 6, adj = 1.5)
```

## Confidence interval

A 95% confidence interval for the population mean is given by

$$\left[ \bar x - t^* \times \frac{s}{\sqrt n}, \ \ \bar x + t^* \times \frac{s}{\sqrt n} \right]$$

This is often written as

$$
\bar x \pm t^* \times \frac{s}{\sqrt n}
$$

where $\pm t^*$ are the quantiles of a t-distribution jointly cutting an overall probability of $\alpha$ in the tails.


## From HT to CI

In the hypothesis test, we do not reject the null hypothesis at the 5% significance level whenever $\mu_0$ lies inside of the 95% CI:

$$
\begin{aligned}
\textbf{Do not reject } H_0 : \mu = \mu_0 \textbf{ if} \\ \quad \\ 
-t^* &< t < +t^* \\
-t^* &< \frac{\bar x - \mu_0}{\frac{s}{\sqrt n}} < +t^* \\
-t^* \times \frac{s}{\sqrt n} &< \bar x - \mu_0 < +t^* \times \frac{s}{\sqrt n} \\
-\bar x -t^* \times \frac{s}{\sqrt n} &< - \mu_0 < -\bar x +t^* \times \frac{s}{\sqrt n} \\
\bar x + t^* \times \frac{s}{\sqrt n} &> \mu_0 > \bar x - t^* \times \frac{s}{\sqrt n} \\
\bar x - t^* \times \frac{s}{\sqrt n} &< \mu_0 < \bar x + t^* \times \frac{s}{\sqrt n} \\
\mu_0 \text{ inside of } &\left[ \bar x - t^* \times \frac{s}{\sqrt n}, \ \ \bar x + t^* \times \frac{s}{\sqrt n} \right] \\
\mu_0 \text{ inside of } &\text{95\% CI}
\end{aligned}
$$
