---
title: "Relationships between variables"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
```


Consider the dataset available at <https://uoepsy.github.io/data/RestaurantTips.csv>, containing 157 observations on the following 7 variables:

```{r, echo=FALSE}
library(tidyverse)
tribble(
    ~'Variable Name', ~'Description',
    'Bill', 'Size of the bill (in dollars)',
    'Tip', 'Size of the tip (in dollars)',
    'Credit', 'Paid with a credit card? n or y',
    'Guests', 'Number of people in the group',
    'Day', 'Day of the week: m=Monday, t=Tuesday, w=Wednesday, th=Thursday, or f=Friday',
    'Server', 'Code for specific waiter/waitress: A, B, or C',
    'PctTip', 'Tip as a percentage of the bill'
) |>
    knitr::kable()
```

These data were collected by the owner of a bistro in the US, who was interested in understanding the tipping patterns of their customers. The data are adapted from @lock2020.

```{r}
library(tidyverse)
tips <- read_csv("https://uoepsy.github.io/data/RestaurantTips.csv")
head(tips)
```

We can replace each factor level with a clearer label:

```{r}
tips$Day <- factor(tips$Day, 
                   levels = c("m", "t", "w", "th", "f"),
                   labels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))

tips$Credit <- factor(tips$Credit, 
                      levels = c("n", "y"),
                      labels = c("No", "Yes"))

tips$Server <- factor(tips$Server)
```

```{r}
summary(tips)
```

The percentage of total bill has a maximum value of 221, which seems very strange. Someone is very unlikely to tip more than their bill total. In this case 221% of their bill value seems unlikely.

Let's inspect the row where `PctTip` is greater than 100:

```{r}
tips |> 
    filter(PctTip > 100)
```

With a bill of 49.6, the tip would be 109.62 dollars:

```{r}
49.6 * 221 / 100
```

Furthermore, we also notice that the tipping amount is not available (NA). The corresponding value in the percentage of total tip seems likely an inputting error, perhaps due to double typing the leading 2 when recording the data. We will set that value to not available (NA) with the following code:

```{r}
tips$PctTip[tips$PctTip > 100] <- NA
```

Consider, for example, the relationship between bill and tip size. As these are two numerical variables, we visualise the relationship with a scatterplot:

```{r}
#| label: fig-tips-scatter
#| fig-cap: "Scatterplot displaying the relationship between bill and tip size"
ggplot(tips, aes(x = Bill, y = Tip)) +
    geom_point() +
    labs(x = "Bill size (in US dollars)",
         y = "Tip size (in US dollars)")
```

:::{.column-margin}
The code starts by setting up a blank canvas for plotting the dataset `tip`, and placing on the x axis the variable `Bill` and on the y axis the variable `Tip`:

```
ggplot(tips, aes(x = Bill, y = Tip))
```

The following line adds a geometric shape to the plot, in this case points:

```
geom_point()
```

The final line uses more informative labels for the reader, setting a label for the x and y axis respectively:

```
labs(x = "Bill size (in US dollars)", 
     y = "Tip size (in US dollars)")
```

The layers of the plot need to be added to each other with a `+` symbol at the end of each line, excluding the last one.
:::

We can numerically summarise this relationship with the covariance between the two variables:

```{r}
cov(tips$Bill, tips$Tip)
```

There are missing values, so the covariance cannot be computed if one or both of the values $X, Y$ is missing.  
To fix this, we use the option `use = "pairwise.complete.obs"` to tell R to only keep the complete pairs to compute the covariance, i.e. ignoring pairs where at least one number is NA:

```{r}
round(cov(tips$Bill, tips$Tip, use = "pairwise.complete.obs"), digits = 2)
```

To investigate the relationship between bill and tip size for those who paid by credit card and those who didn't we can create faceted scatterplots:

```{r}
ggplot(tips, aes(x = Bill, y = Tip)) +
    geom_point() +
    labs(x = "Bill size (in US dollars)",
         y = "Tip size (in US dollars)") +
    facet_wrap(~Credit)
```

::: {.column-margin}
Instead of `facet_wrap()` you can use the `facet_grid()` function. 
This allows you to facet the rows by a variable, and the columns by another variable, or both:  

- `facet_grid(rows ~ .)`  
- `facet_grid(. ~ cols)`  
- `facet_grid(rows ~ cols)`

Try replacing the last line of code with:

`facet_grid(Server ~ Credit)`
:::

We can improve the labelling by using `labeller = "label_both"`, which displays not only the group value as label, but both the variable and value:

```{r}
#| label: fig-tips-scatter-group
#| fig-cap: "Relationship between bill and tip size by paying method"
ggplot(tips, aes(x = Bill, y = Tip)) +
    geom_point() +
    labs(x = "Bill size (in US dollars)",
         y = "Tip size (in US dollars)") +
    facet_wrap(~Credit, labeller = "label_both")
```

Similarly, you can compute grouped covariances via:

```{r eval=FALSE}
library(kableExtra)
tips |>
    group_by(Credit) |>
    summarise(Cov = cov(Bill, Tip, use = "pairwise.complete.obs")) |> 
    kable(digits = 2, booktabs = TRUE,
        caption = "Relationship between bill and tip size by credit card usage")
```

```{r echo=FALSE}
#| label: tbl-tips
#| tbl-cap: "Relationship between bill and tip size by credit card usage"
library(kableExtra)
tips |>
    group_by(Credit) |>
    summarise(Cov = cov(Bill, Tip, use = "pairwise.complete.obs")) |> 
    kable(digits = 2) |>
    kable_styling()
```


:::{.callout-tip}
### Example write-up

@fig-tips-scatter highlights a strong positive relationship between bill and tip size (in US dollars). The covariance between the two variables is 25.96 squared dollars.
The relationship between bill and tip size is stronger for those who paid by credit card than those who did not, as highlighted by @fig-tips-scatter-group and @tbl-tips, where the covariance between the two variables is 37.68 for those that used a credit card and 15.38 for those that did not.
:::

