---
title: "One-sample t-tests <br> and their assumptions"
author: "Elizabeth Pankratz (elizabeth.pankratz@ed.ac.uk)"
editor_options: 
  chunk_output_type: inline
format:
  revealjs:
    smaller: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)

dapr1blue <- '#0F4C81'

source('_theme/theme_quarto.R')

theme_update(
  text = element_text(family = 'Source Sans 3'),
  panel.grid.minor = element_blank(),
  strip.background = element_blank(),
  strip.text.x = element_text(size = 24)
)
```

# Welcome back to DAPR!

## Welcome back warm-up

TODO WOO

Interactively walk through the logic of NHST somehow: <https://nullworlds.andrewheiss.com>

- thought exercise: imagine a world where the effect we're interested in is not real
- this is the world where the null hypothesis is true = null world
- in that world, what would my test results look like
- some variability here bc of the randomness in the world
- now i compare my actual test results to that
- what fraction of null world outcomes are more extreme than my actual test result
- = how improbable is my actual test result, assuming that the null world is the true world
- we'll accept a 5% chance of being wrong
- so we want our actual test result to have a less than 5% chance of occurring in the null world



## This week's learning objectives

:::dapr1callout

What kind of test do we use to compare a variable's mean to a specific value?

- one-sample t-test

:::

<!-- :::dapr1callout -->
<!-- How do we formalise the hypotheses that we're testing? -->

<!-- - h0h1 -->
<!-- - tailedness depends on directionality of hyp -->
<!-- ::: -->

:::dapr1callout
What does this test require/assume to be true about the data that it models?

(0) dependent variable is numeric;
(1) variable is normally distributed OR sample size >= 30;
(2) observations are sampled independently
:::

:::dapr1callout
How do we check whether we're satisfied with the test's requirements/assumptions?

(0) check that variable is on interval/ratio/integer scale
(1)  visually with plots (hist, dens, qq), statistically with Shapiro-Wilk
(2) check that each data point comes from a separate source (e.g., a separate person)
:::

<!-- :::dapr1callout -->
<!-- What do we do if we're not satisfied with the test's assumptions? -->

<!-- - Wilcoxon signed-rank test (beyond the scope of DAPR1, you won't be assessed on this, but we mention it so you know what to do if you're in this scenario later) -->
<!-- ::: -->



# Recap: One-sample t-test

## One-sample t-test example: College loneliness

With a one-sample t-test, we can test whether a sample's mean is significantly different from some specific value.

**Research question (RQ): Based on the 20-item UCLA Loneliness Scale, do college students report loneliness above the "low loneliness" threshold of 34?**

This example uses real data from [Russell (1996)](https://www.tandfonline.com/doi/epdf/10.1207/s15327752jpa6601_2){target="_blank"}.


```{r include = F}
set.seed(1)
N_SUBJS <- 487
lonely_data <- tibble(
  subj_ID =  c( 
    paste("student_0", 1:9, sep=""),  
    paste("student_", 10:N_SUBJS, sep="")  
  ),
  loneliness = round(rnorm(n = N_SUBJS, mean = 40.08, sd = 9.50)) # from Table 2 (p. 26).
)
```

<br>

::::{.columns}
:::{.column width="50%"}
```{r}
lonely_data |> head(12)
```

:::
:::{.column width="50%"}
```{r fig.width = 5, fig.height = 5}
#| code-fold: true
lonely_data |>
  ggplot(aes(y = loneliness)) +
  geom_boxplot() +
  # the theme() bit below removes the uninformative x axis numbering
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank()
  )
```
:::
::::


## A few ways to visualise this data

::::{.columns}
:::{.column width="33%"}
**Box plot:** 

<br>

```{r fig.width = 4, fig.height = 5}
#| code-fold: true
lonely_data |>
  ggplot(aes(y = loneliness)) +
  geom_boxplot() +
  # the theme() bit below removes the uninformative x axis numbering
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank()
  )
```
:::
:::{.column width="33%"}
**Violin plot with mean and <br> stacked data points:**

```{r fig.width = 4, fig.height = 5}
#| code-fold: true
lonely_data |>
  ggplot(aes(x = 1, y = loneliness)) +
  geom_violin() +     # add violin
  geom_point(         # add points
    alpha = 0.2,      # make points 20% opaque
  ) +
  stat_summary(fun = mean, geom = 'point', size = 5) +  # add big point for mean
  # the theme() bit below removes the uninformative x axis numbering
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank()
  )
```
:::
:::{.column width="33%"}
**Violin plot with mean and <br> jittered data points:**

```{r fig.width = 4, fig.height = 5}
#| code-fold: true
lonely_data |>
  ggplot(aes(x = 1, y = loneliness)) +
  geom_violin() +     # add violin
  geom_jitter(        # add points with jitter (random horizontal offset)
    width = 0.2,      # how big is the max horizontal offset?
    alpha = 0.2,      # make the points 20% opaque
    ) +     
  stat_summary(fun = mean, geom = 'point', size = 5) +  # add big point for mean
  # the theme() bit below removes the uninformative x axis numbering
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank()
  )
```
:::
::::

<br>

**Best practice:** make plots that show all the data points *and also* usefully summarise their distribution.

(This is why Elizabeth likes violin + points plots more than box plots.)


## From RQ to hypotheses

RQ: Based on the 20-item UCLA Loneliness Scale, do college students report loneliness above the "low loneliness" threshold of 34?

<br>

**The null hypothesis $H_0$ is always that there's no difference, no effect, nothing interesting going on:**

- There's no difference between the mean loneliness of college students ($\mu$) and 34.
- Formally:

$$
H_0 : \mu = 34
$$

<br>

**For the alternative hypothesis $H_1$, we need to decide:**

- Is our RQ **directional?** (Do we expect a specific direction to the effect?) <img style="margin-bottom:0px; width: 1em;" src="figs/codicon--thumbsup-filled.svg"> <img style="margin-bottom:-10px; width: 1em;" src="figs/codicon--thumbsdown-filled.svg">

- If our RQ is directional, which alternative hypothesis matches the direction of the effect we expect to see?

::::{.columns}
:::{.column width="50%"}
$$
H_1 : \mu > 34
$$

:::hcenter
<img style="width: 2em;" src="figs/codicon--thumbsup-filled.svg">
:::


:::
:::{.column width="50%"}
$$
H_1 : \mu < 34
$$

:::hcenter
<img style="width: 2em;" src="figs/codicon--thumbsdown-filled.svg">
:::

:::
::::


## How do we test our research question?

**RQ: Based on the 20-item UCLA Loneliness Scale, do college students report loneliness above the "low loneliness" threshold of 34?**

A hypothesis test has five parts:

::::{.columns}
:::{.column width="33%"}
**(1) The null hypothesis $H_0$**
:::
:::{.column width="66%"}
College students' mean loneliness $\mu$ is exactly equal to 34.

$$
H_0 : \mu = 34
$$
:::
::::

::::{.columns}
:::{.column width="33%"}
**(2) The alternative hypothesis $H_1$**
:::
:::{.column width="66%"}
College students' mean loneliness $\mu$ is above the threshold of 34.

$$
H_1 : \mu > 34
$$
:::
::::

::::{.columns}
:::{.column width="33%"}
**(3) The test statistic**
:::
:::{.column width="66%"}
For testing a mean, we use the t-statistic.

- The t-statistic measures how many SEs away from the hypothesised mean (34) the observed sample mean is.

- It answers the question: How much does the observed data differ from what we would expect if the H0 were true?
:::
::::

::::{.columns}
:::{.column width="33%"}
**(4) The significance level $\alpha$**
:::
:::{.column width="66%"}
What % of the time do we accept that we will incorrectly reject the H0, even if it is actually true? (Typically $\alpha = .05$.)

<!-- - In Psychology, typically $\alpha = .05$, so we'll incorrectly reject H0 one time in twenty. -->
:::
::::

::::{.columns}
:::{.column width="33%"}
**(5) The p-value**
:::
:::{.column width="66%"}
Assuming the H0 is true, how probable is it that we would obtain a test statistic at least as extreme as the one we calculated in Step 3?
:::
::::

## Run Steps 3, 4, 5 in R

```{r}
lonely_t <- t.test(
  lonely_data$loneliness,   # take the loneliness column of lonely_data
  mu = 34,                  # compare it to the reference value of 34
  alternative = "greater",  # alternative hypothesis: mu is *greater* than 34
  conf.level = .95          # confidence level is 1 minus alpha, so 1 – .05
)
```

```{r}
lonely_t
```

## A visual intuition for Steps 3, 4, 5

```{r echo=F}
cat( paste0(capture.output(
  lonely_t
), '\n')[4:12])
```


```{r echo=F, fig.width = 12, fig.height = 6}
p_overallt <- tibble(
  t = seq(-16, 16, length.out = 1000),
  density = dt(t, df = 486)
) |>
  ggplot(aes(x = t, y = density)) +
  geom_ribbon(aes(ymin = 0, ymax = density), fill = dapr1blue) +
  geom_text(
    x = -16, y = 0.35, label = 'a t-distribution\nwith 486\ndegrees of freedom', hjust = 0, size = 7,
    family = 'Source Sans 3'
    ) +
  geom_vline(xintercept = 15, colour = '#BF1932', linewidth = 2) +
  scale_x_continuous(breaks = seq(-15, 15, by=5)) 

p_zoomt <- tibble(
  t = seq(14.9, 15.2, length.out = 10),
  density = dt(t, df = 486)
) |>
  ggplot(aes(x = t, y = density)) +
  geom_ribbon(aes(ymin = 0, ymax = density), fill = dapr1blue) +
  geom_vline(xintercept = 15, colour = '#BF1932', linewidth = 2) +
  labs(
    y = ''
  )

w <- 6
layout <- c(
  area(t = 1, l = 1, b = 5, r = w),
  area(t = 3, l = w+1, b = 4.5, r = w+3)
)
p_overallt + p_zoomt +
  plot_layout(design = layout)
```


## How would we report this analysis?

<br>

> We wanted to find out whether college students score above the threshold for "low loneliness" on the UCLA Loneliness scale, which is 34.
At the 5% significance level, we performed a one-sided test of significance against the null hypothesis that college students' loneliness is equal to 34.
The results indicate that the mean loneliness for college students, 40.3, is significantly higher than the "low loneliness" threshold of 34 ($t$(486) = 15, $p$ < .001, one-sided).


# The one-sample t-test is only valid if three things are true

## The one-sample t-test is only valid if three things are true

<br>

::::{.columns}
:::{.column width="30%"}
**Data requirements:**
:::
:::{.column width="70%"}
The variable is numeric.
:::
::::

::::{.columns}
:::{.column width="30%"}
**Assumption of independence:**
:::
:::{.column width="70%"}
The obtained sample data are a random sample from the population of interest.
:::
::::

::::{.columns}
:::{.column width="30%"}
**Assumption of normality:**
:::
:::{.column width="70%"}
EITHER the population follows a normal distribution <br> OR the sample size $\geq$ 30.
:::
::::

<br>

**What if one or more of these things isn't true?**

- Then you shouldn't use a one-sample t-test to analyse your data.
- Instead, use a test called the "Wilcoxon Signed-Rank Test". (We aren't covering it in DAPR, so you don't need to know anything more about it than this!)

<br>

**How do we check each requirement/assumption?**

## Data requirements for a one-sample t-test

<br>

:::hcenter
**The variable is numeric.**
:::

<br>

Look at the data and make sure that the variable we're analysing represents **interval, ratio, or integer data.**

```{r}
lonely_data
```

<br>

`loneliness` is integer data: it consists of whole numbers &nbsp; ✅


## Assumption of independence for a <br> one-sample t-test

<br>

:::hcenter
**The obtained sample data are a random sample from the population of interest.**
:::

<br>

Before generalising the conclusion of our test to some population, we need to first make sure that our sample was randomly selected from that population.

This is more of a study design issue, not something we can test directly.

- Was each observation sampled independently of the others? (e.g., does each observation come from a different person?)
- Is the sample representative of the population? (so that we avoid sampling bias)

**Let's assume that our data was sampled randomly from the population of all students, and that each observation is independent of each other one ✅**


## Assumption of normality for a one-sample t-test

<br>

:::hcenter
**EITHER the population follows a normal distribution OR the sample size $\geq$ 30.**
:::

<br>

Backstory: All our formulas for confidence intervals and hypothesis tests are built around the assumption that **the sampling distribution of the sample mean** is normally distributed.

<br>

When is the sample mean's sampling distribution normally distributed?

<br>

When either one of these is true:

- **The population follows a normal distribution** (no matter what size the sample is), or
- **Our sample size is greater than 30** (no matter how the population data is distributed)
  - Why 30? 
  - Think back to the Central Limit Theorem in Semester 1: after we've sampled about 30 observations, the sampling distribution of sample means will become sufficiently normal


## Check the sample size

<br>

Check the sample size:

```{r}
nrow(lonely_data)
```

<br>

**So the assumption of normality of the sampling distribution of sample means is met, because we have a large enough sample &nbsp; ✅ **

<br>

But for illustration, let's look at how we'd check on the normality of the population.



## One way to check the population distribution: Plot the sample (1)

We don't know how the population is distributed, **but if our sample is approximately normally distributed, then the population probably is too.**

::::{.columns}
:::{.column width="50%"}
**Histogram:**

```{r fig.width = 6, fig.height = 6}
lonely_data |>
  ggplot(aes(x = loneliness)) +
  geom_histogram()
```

:::
:::{.column width="50%"}
**Density plot:**

```{r fig.width = 6, fig.height = 6}
lonely_data |>
  ggplot(aes(x = loneliness)) +
  geom_density()
```
:::
::::


## One way to check the population distribution: Plot the sample (2)

A plot specifically designed to check how closely a variable matches a normal distribution: <br> a **Quantile-Quantile plot** aka **Q-Q plot.**

The closer the points are to the diagonal line, the more normally-distributed the variable is.


::::{.columns}
:::{.column width="50%"}
<br>

```{r eval=F}
lonely_data |>
  ggplot(aes(sample = loneliness)) +
  geom_qq() +
  geom_qq_line() +
  labs(
    x = "Theoretical quantiles",
    y = 'Sample quantiles'
  )
```
:::
:::{.column width="50%"}
```{r echo=F, fig.width = 5, fig.height = 5}
lonely_data |>
  ggplot(aes(sample = loneliness)) +
  geom_qq() +
  geom_qq_line() +
  labs(
    x = "Theoretical quantiles",
    y = 'Sample quantiles'
  )
```
:::
::::


## Examples of worrisome Q-Q plots

```{r echo=F, fig.width = 9, fig.height = 4}
df1 <- data.frame(y = rexp(100))
df2 <- data.frame(y = runif(100, 2, 10))
plt1 <- ggplot(df1, aes(sample = y)) + 
    geom_qq() +
    geom_qq_line() +
    labs(x = "Theoretical quantiles", 
         y = "Sample quantiles")
plt2 <- ggplot(df2, aes(sample = y)) + 
    geom_qq() +
    geom_qq_line() +
    labs(x = "Theoretical quantiles", 
         y = "Sample quantiles")
plt1 | plt2
```



## Another way to check normality of the population distribution: Shapiro-Wilk test

The Shapiro-Wilk test is a hypothesis test where:

- $H_0$: The population data follow a normal distribution
- $H_1$: The population data do not follow a normal distribution

We don't want to reject this H0, so **we want a non-significant result!**

<br>

```{r}
shapiro.test(lonely_data$loneliness)
```

<br>

> We performed a Shapiro-Wilk test against the null hypothesis of normality of the population data.
Based on the sample, we could not reject the null hypothesis of normality in the population ($W$ = 1, $p$ = .3).


## Recap: Assumption of normality for a <br> one-sample t-test

<br>

:::hcenter
**EITHER the population follows a normal distribution OR the sample size $\geq$ 30.**
:::

<br>

- To check if the sample size is greater than 30:
  - Count the observations in your dataset
- To check if the population follows a normal distribution:
  - Plot the sample:
    - Histogram or density plot
    - Q-Q plot
  - Run a Shapiro-Wilk test (you want a non-significant result!)



# What's the rest of Block 3 about?

## What's the rest of Block 3 about?

<br>
 
**Over the next several weeks, we'll use the tools you've gained so far to explore how psychologists answer different kinds of research questions.**

<br>

So far in DAPR1, you've been looking at just one variable at a time.

But in psychological research, most RQs are about the **associations between variables.**

- So that's our focus for the next five weeks: 
  - Common ways to statistically test whether the relationship between two variables is different from what we would expect by chance.
  - The assumptions/requirements of each test and how to check them.
  

[fragment: show all the kinds of research qs (no tests)]


<!-- TODO WOO: -->

<!-- - could practice with diff kinds of RQs, identifying the kinds of variables that they are asking about -->


# Back matter

## Revisiting this week's learning objectives

TODO


