---
title: "Independent samples t-test"
author: "Elizabeth Pankratz (elizabeth.pankratz@ed.ac.uk), <br> Emma Waterston (Emma.Waterston@ed.ac.uk)"
editor_options: 
  chunk_output_type: inline
format:
  revealjs:
    smaller: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)
library(kableExtra)
library(simglm)

dapr1blue <- '#0F4C81'

source('_theme/theme_quarto.R')

theme_update(
  text = element_text(family = 'Source Sans 3')
)
```

# Welcome back!

## This week's learning objectives

What kind of test do we use to compare the means of two independent groups?

- independent samples t-test

How do we formalise the hypotheses that we're testing?

- h0h1
- tailedness depends on directionality of hyp

What does this test assume is true about the data that it models?

(0) dependent variable is numeric;
(1) each group's variable is normally distributed OR each group's size >= 30;
(2) observations are sampled independently, both within and across groups;
(3) population variance is the same in both groups

How do we check whether we're satisfied with the test's assumptions?

(0) check that variable is on interval/ratio/integer scale
(1) descriptively with measures of skew, visually with plots (hist, dens, qq), statistically with Shapiro-Wilks for each group separately
(2) check that each individual belongs to only one group and has only one obs in group they belong to;
(3) test statistically with F-test (if not met, use Welch's t-test instead)

What do we do if we're not satisfied with the test's assumptions?

- Welch's t-test (beyond the scope of DAPR1, you won't be assessed on this, but we mention it so you know what to do if you're in this scenario later)

Imagine we have paired data, where e.g. we measure the same people at two different time points. How do we analyse that data using the one-sample t-test machinery?

- compute pairwise diffs
- compare them to 0

## Where we are in the Block 3 concept map

[rstack fragment: show map up to indep t]

[fragment: add in indep t]


# This week's data

## This week's data: Relationship quality and stress

<br>

::::{.columns}
:::{.column width="30%"}

:::bluebox
Researchers asked people to label their current or most recent partner relationship as "happy" or "unhappy"
:::
:::
:::{.column width="5%"}
:::{.hcenter style="font-size:200%;"}
→
:::
:::
:::{.column width="30%"}
:::bluebox
People spent five minutes vividly imagining or remembering their relationship
:::
:::
:::{.column width="5%"}
:::{.hcenter style="font-size:200%;"}
→
:::
:::
:::{.column width="30%"}
:::bluebox
Researchers measured the cortisol (a stress marker) in people's saliva
:::
:::
::::

<br>

**Our research question (RQ): Do people with happy relationships have lower cortisol than people in unhappy ones?**

<br>

<!-- beware: if we looked at people's stress levels before and after, then our samples are not independent anymore! -->


This example is simplified, but it's based on real findings from [Berry & Worthington (2001)](https://psycnet.apa.org/fulltext/2001-11480-010.pdf){target="_blank"}.

```{r include=F}
# Based on the descriptives in Table 1, p. 451 of BerryWorthington2001
set.seed(6)

n_ppts <- 39

stress_data <- tibble(
  subj_ID =  c( 
    paste("subj0", 1:9, sep=""),  
    paste("subj", 10:n_ppts, sep="")  
  ),
  relationship = c(
    rep('Happy', 19),
    rep('Unhappy', 20)
  ),
  cortisol = c(
    rnorm(n = 19, mean = 0.14, sd = 0.08),
    rnorm(n = 20, mean = 0.18, sd = 0.08)
  )
)
```

##  Do people with happy relationships have lower cortisol than people in unhappy ones? 

<br>

::::{.columns}
:::{.column width="40%"}

```{r}
stress_data |> head()
```


:::
:::{.column width="60%"}

```{r fig.width = 7, fig.height = 6}
stress_data |>
  ggplot(aes(x = relationship, y = cortisol)) +
  geom_violin() +
  geom_jitter(width = 0.2) +
  stat_summary(fun = mean, geom = 'point', size = 5)
```

:::
::::

## From RQ to hypotheses

RQ: Do people with happy relationships have lower cortisol than people in unhappy ones?

<br>

**The null hypothesis $H_0$ is always that there's no difference, no effect, nothing interesting going on:**

- There's no difference between the mean cortisol of people in happy relationships ($\mu_\text{Happy}$) and the mean cortisol of people in unhappy relationships ($\mu_\text{Unhappy}$)
- Formally:

$$
H_0 : \mu_\text{Happy} = \mu_\text{Unhappy}
$$

<br>

**For the alternative hypothesis $H_1$, we need to decide:**

- Is our RQ **directional?** (Do we expect a specific direction to the effect?) <img style="margin-bottom:0px; width: 1em;" src="figs/codicon--thumbsup-filled.svg"> <img style="margin-bottom:-10px; width: 1em;" src="figs/codicon--thumbsdown-filled.svg">

- If our RQ is directional, which alternative hypothesis matches the direction of the effect we expect to see?

::::{.columns}
:::{.column width="50%"}
$$
H_1 : \mu_\text{Happy} > \mu_\text{Unhappy}
$$

:::hcenter
<img style="width: 2em;" src="figs/codicon--thumbsup-filled.svg">
:::


:::
:::{.column width="50%"}
$$
H_1 : \mu_\text{Happy} < \mu_\text{Unhappy}
$$

:::hcenter
<img style="width: 2em;" src="figs/codicon--thumbsdown-filled.svg">
:::

:::
::::



## How do we test our research question?

**RQ: Based on the 20-item UCLA Loneliness Scale, do college students report loneliness above the "low loneliness" threshold of 34?**

A hypothesis test has five parts:

::::{.columns}
:::{.column width="33%"}
**(1) The null hypothesis $H_0$**
:::
:::{.column width="66%"}
People with happy relationships will have the same average cortisol as people with unhappy relationships.

$$
H_1 : \mu_\text{Happy} = \mu_\text{Unhappy}
$$
:::
::::

::::{.columns}
:::{.column width="33%"}
**(2) The alternative hypothesis $H_1$**
:::
:::{.column width="66%"}
People with happy relationships will have lower average cortisol than people with unhappy relationships.

$$
H_1 : \mu_\text{Happy} < \mu_\text{Unhappy}
$$
:::
::::

::::{.columns}
:::{.column width="33%"}
**(3) The test statistic**
:::
:::{.column width="66%"}
For testing differences between means, we can still use the t-statistic.

- The t-statistic measures how many SEs away from zero the observed difference between means is.

- It answers the question: How much does the observed difference between means differ from what we would expect if the H0 were true?
:::
::::

::::{.columns}
:::{.column width="33%"}
**(4) The significance level $\alpha$**
:::
:::{.column width="66%"}
What % of the time do we accept that we will incorrectly reject the H0, even if it is actually true? (Typically $\alpha = .05$.)
:::
::::

::::{.columns}
:::{.column width="33%"}
**(5) The p-value**
:::
:::{.column width="66%"}
Assuming the H0 is true, how probable is it that we would obtain a test statistic at least as extreme as the one we calculated in Step 3?
:::
::::


## Independent-samples t-test in R

Runs Steps 3, 4, 5:

```{r}
stress_t <- t.test(
  
  stress_data$cortisol ~ stress_data$relationship,  
  # asks: how is `cortisol` different for each value of `relationship`
  
  alternative = "less",   
  # the direction of H1: mean(happy) is less than mean(unhappy)
  
  mu = 0,   
  # the difference we'd expect between groups if H0 were true
  
  var.equal = TRUE,   
  # related to assumptions – we'll come back to this!
  
  conf.level = .95   
  # .95 because our alpha level is .05, and confidence level is 1 – alpha
  
)
```


## Interpreting the test results

```{r}
stress_t
```

<br>

```{r echo=F, fig.width = 10, fig.height = 4}
tibble(
  t = seq(-4, 4, length.out = 100),
  density = dt(t, df = 37)
) |>
  ggplot(aes(x = t, y = density)) +
  geom_ribbon(aes(ymin = 0, ymax = density), fill = dapr1blue) +
  geom_text(
    x = 1.5, y = 0.35, label = 'a t-distribution\nwith 37 df', hjust = 0, size = 7,
    family = 'Source Sans 3'
    ) +
  geom_vline(xintercept = -2, colour = '#BF1932', linewidth = 2) +
  theme(
    panel.grid.minor = element_blank()
  )
```


## Note about how categories are ordered

<br>

You might have noticed that we wrote our H1 as 

$$
\mu_\text{Happy} < \mu_\text{Unhappy}
$$

and not as

$$
\mu_\text{Unhappy} > \mu_\text{Happy}
$$

Why?

<br>

Because by default, R will compare categories **in alphabetical order.** 

**So the category that comes first alphabetically should also come first in our H0 and H1.**

<br>

This guarantees that we'll choose the right direction for our one-sided tests!

Only `alternative = "less"` runs the test we actually want.

# What the independent-samples t-test assumes about the data

## What the independent-samples t-test assumes about the data


## How to check each assumption


# A final note: how to analyse non-independent samples

## A final note: how to analyse non-independent samples

before and after cortisol



(this is more like what the researchers in this cortisol study actually did: they compared the *difference* in cortisol between relationship groups)


# Back matter

## Revisiting this week's learning objectives

TODO


