---
title: "Independent samples t-test"
author: "Elizabeth Pankratz (elizabeth.pankratz@ed.ac.uk)"
editor_options: 
  chunk_output_type: inline
format:
  revealjs:
    smaller: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)

dapr1blue <- '#0F4C81'

source('_theme/theme_quarto.R')

theme_update(
  text = element_text(family = 'Source Sans 3'),
  panel.grid.minor = element_blank(),
  strip.background = element_blank(),
  strip.text.x = element_text(size = 24)
)
```

# Welcome back!

## This week's learning objectives

:::dapr1callout
What kind of test do we use to compare the means of two independent groups?

- independent samples t-test
:::

<!-- :::dapr1callout -->
<!-- How do we formalise the hypotheses that we're testing? -->

<!-- - h0h1 -->
<!-- - tailedness depends on directionality of hyp -->
<!-- ::: -->

:::dapr1callout
What does this test require/assume to be true about the data that it models?

(0) dependent variable is numeric;
(1) each group's variable is normally distributed OR each group's size >= 30;
(2) observations are sampled independently, both within and across groups;
(3) population variance is the same in both groups
:::

:::dapr1callout
How do we check whether we're satisfied with the test's assumptions?

(0) check that variable is on interval/ratio/integer scale
(1) visually with plots (hist, dens, qq), statistically with Shapiro-Wilks for each group separately
(2) check that each individual belongs to only one group and has only one obs in group they belong to;
(3) test statistically with F-test (if not met, use Welch's t-test instead)
:::

<!-- :::dapr1callout -->
<!-- What do we do if we're not satisfied with the test's assumptions? -->

<!-- - Welch's t-test (beyond the scope of DAPR1, you won't be assessed on this, but we mention it so you know what to do if you're in this scenario later) -->
<!-- ::: -->

:::dapr1callout
Imagine we have paired data, where e.g. we measure the same people at two different time points. How do we analyse that data using the one-sample t-test machinery?

- compute pairwise diffs
- compare them to 0
:::


## Where we are in the Block 3 concept map

[TODO figure]

# This week's data

## This week's data: Relationship quality and stress

<br>

::::{.columns}
:::{.column width="30%"}

:::bluebox
Researchers asked people to label their current or most recent partner relationship as "happy" or "unhappy"
:::
:::
:::{.column width="5%"}
:::{.hcenter style="font-size:200%;"}
→
:::
:::
:::{.column width="30%"}
:::bluebox
People spent five minutes vividly imagining or remembering their relationship
:::
:::
:::{.column width="5%"}
:::{.hcenter style="font-size:200%;"}
→
:::
:::
:::{.column width="30%"}
:::bluebox
Researchers measured the cortisol (a stress marker) in people's saliva
:::
:::
::::

<br>

**Our research question (RQ): Do people with happy relationships have lower cortisol than people in unhappy ones?**

<br>

<!-- beware: if we looked at people's stress levels before and after, then our samples are not independent anymore! -->


This example is simplified, but it uses real data from [Berry & Worthington (2001)](https://psycnet.apa.org/fulltext/2001-11480-010.pdf){target="_blank"}.

```{r include=F}
# Based on the descriptives in Table 1, p. 451 of BerryWorthington2001
set.seed(6)

n_ppts <- 39

stress_data <- tibble(
  subj_ID =  c( 
    paste("subj0", 1:9, sep=""),  
    paste("subj", 10:n_ppts, sep="")  
  ),
  relationship = c(
    rep('Happy', 19),
    rep('Unhappy', 20)
  ),
  cortisol = c(
    rnorm(n = 19, mean = 0.14, sd = 0.08),
    rnorm(n = 20, mean = 0.18, sd = 0.08)
  )
)
```

##  Do people with happy relationships have lower cortisol than people in unhappy ones? 

<br>

::::{.columns}
:::{.column width="40%"}

```{r}
stress_data |> head(12)
```


:::
:::{.column width="60%"}

```{r fig.width = 6, fig.height = 5}
#| code-fold: true
stress_data |>
  ggplot(aes(x = relationship, y = cortisol)) +
  geom_violin() +
  geom_jitter(width = 0.2, alpha = 0.5, size = 3) +
  stat_summary(fun = mean, geom = 'point', size = 8)
```

:::
::::

## From RQ to hypotheses

<br>

RQ: Do people with happy relationships have lower cortisol than people in unhappy ones?

<br>

**The null hypothesis $H_0$ is always that there's no difference, no effect, nothing interesting going on:**

- There's no difference between the mean cortisol of people in happy relationships ($\mu_\text{Happy}$) and the mean cortisol of people in unhappy relationships ($\mu_\text{Unhappy}$)
- Formally:

$$
H_0 : \mu_\text{Happy} = \mu_\text{Unhappy}
$$

<br>

**For the alternative hypothesis $H_1$, we need to decide:**

- Is our RQ **directional?** (Do we expect a specific direction to the effect?) <img style="margin-bottom:0px; width: 1em;" src="figs/codicon--thumbsup-filled.svg"> <img style="margin-bottom:-10px; width: 1em;" src="figs/codicon--thumbsdown-filled.svg">

- If our RQ is directional, which alternative hypothesis matches the direction of the effect we expect to see?

::::{.columns}
:::{.column width="50%"}
$$
H_1 : \mu_\text{Happy} > \mu_\text{Unhappy}
$$

:::hcenter
<img style="width: 2em;" src="figs/codicon--thumbsup-filled.svg">
:::


:::
:::{.column width="50%"}
$$
H_1 : \mu_\text{Happy} < \mu_\text{Unhappy}
$$

:::hcenter
<img style="width: 2em;" src="figs/codicon--thumbsdown-filled.svg">
:::

:::
::::



## How do we test our research question?

<br>

RQ: Do people with happy relationships have lower cortisol than people in unhappy ones?

<br>

A hypothesis test has five parts:

::::{.columns}
:::{.column width="33%"}
**(1) The null hypothesis $H_0$**
:::
:::{.column width="66%"}
<!-- People with happy relationships will have the same average cortisol as people with unhappy relationships. -->

$$
H_0 : \mu_\text{Happy} = \mu_\text{Unhappy}
$$
:::
::::

::::{.columns}
:::{.column width="33%"}
**(2) The alternative hypothesis $H_1$**
:::
:::{.column width="66%"}
<!-- People with happy relationships will have lower average cortisol than people with unhappy relationships. -->

$$
H_1 : \mu_\text{Happy} < \mu_\text{Unhappy}
$$
:::
::::

::::{.columns}
:::{.column width="33%"}
**(3) The test statistic**
:::
:::{.column width="66%"}
For testing differences between means, we can still use the t-statistic.

- The t-statistic measures how many SEs away from zero the observed difference between means is.

- It answers the question: How much does the observed difference between means differ from what we would expect if the H0 were true?
:::
::::

::::{.columns}
:::{.column width="33%"}
**(4) The significance level $\alpha$**
:::
:::{.column width="66%"}
What % of the time do we accept that we will incorrectly reject the H0, even if it is actually true? (Typically $\alpha = .05$.)
:::
::::

::::{.columns}
:::{.column width="33%"}
**(5) The p-value**
:::
:::{.column width="66%"}
Assuming the H0 is true, how probable is it that we would obtain a test statistic at least as extreme as the one we calculated in Step 3?
:::
::::


## Independent-samples t-test in R

Runs Steps 3, 4, 5:

```{r}
stress_t <- t.test(
  
  stress_data$cortisol ~ stress_data$relationship,  
  # asks: how is `cortisol` different for each value of `relationship`
  # cortisol is the "dependent variable"
  
  alternative = "less",   
  # the direction of H1: mean(happy) is less than mean(unhappy)
  
  mu = 0,   
  # the difference we'd expect between groups if H0 were true
  
  var.equal = TRUE,   
  # related to assumptions – we'll come back to this!
  
  conf.level = .95   
  # .95 because our alpha level is .05, and confidence level is 1 – alpha
  
)
```


## Interpreting the test results

```{r}
stress_t
```

<br>

```{r echo=F, fig.width = 10, fig.height = 4}
tibble(
  t = seq(-4, 4, length.out = 100),
  density = dt(t, df = 37)
) |>
  ggplot(aes(x = t, y = density)) +
  geom_ribbon(aes(ymin = 0, ymax = density), fill = dapr1blue) +
  geom_text(
    x = 1.5, y = 0.35, label = 'a t-distribution\nwith 37 df', hjust = 0, size = 7,
    family = 'Source Sans 3'
    ) +
  geom_vline(xintercept = -2, colour = '#BF1932', linewidth = 2) +
  theme(
    panel.grid.minor = element_blank()
  )
```


## Note about how categories are ordered

<br>

You might have noticed that we wrote our H1 as 

$$
\mu_\text{Happy} < \mu_\text{Unhappy}
$$

and not as

$$
\mu_\text{Unhappy} > \mu_\text{Happy}
$$

Why?

<br>

Because by default, R will compare categories **in alphabetical order.** 

**So the category that comes first alphabetically should also come first in our H0 and H1.**

<br>

This guarantees that we'll choose the right direction for our one-sided tests!

Only `alternative = "less"` runs the test we actually want.

# What the independent-samples t-test requires / assumes about the data

## What the independent-samples t-test requires / assumes about the data

<br>

::::{.columns}
:::{.column width="30%"}
**Data requirements:**
:::
:::{.column width="70%"}
The dependent variable is numeric.
:::
::::

::::{.columns}
:::{.column width="30%"}
**Assumption of independence:**
:::
:::{.column width="70%"}
The obtained sample data are a random sample from the population of interest, both within and across groups.
:::
::::

::::{.columns}
:::{.column width="30%"}
**Assumption of normality:**
:::
:::{.column width="70%"}
EITHER the population for each group follows a normal distribution <br> OR each group's sample size $\geq$ 30.
:::
::::

::::{.columns}
:::{.column width="30%"}
**Assumption of equal variance:**
:::
:::{.column width="70%"}
The population variance is the same for both groups.
:::
::::

<br>

**What if one or more of these things isn't true?**

- Then you shouldn't use an independent-samples t-test to analyse your data.
- A common alternative that makes fewer assumptions is the "Welch's t-test". (We aren't covering it in DAPR, so you don't need to know anything more about it than this!)

<br>

**How do we check each requirement/assumption?**

## Data requirements for the <br> independent-samples t-test

:::hcenter
**The dependent variable is numeric.**
:::

<br>

Look at the data and make sure that the variable we're analysing represents **interval, ratio, or integer data.**

```{r}
stress_data
```

<br>

`cortisol` is ratio data: it is positive-only with a true zero point &nbsp; ✅

## Independence assumption for the <br> independent-samples t-test

:::hcenter
**The obtained sample data are a random sample from the population of interest, <br> both within and across groups.**
:::

<br>

This is more of a study design issue, not something we can test directly.

- Does each individual who contributed data belong to only one group?
- Does each individual have only one observation in the group they belong to?
- Within groups, is each sample representative of the population?
- Were people allocated to each group at random?
  - In this case, they weren't! 
  - People described their own relationships as "Happy" or "Unhappy", so in a way, they allocated themselves to groups.
  - This might limit our conclusions: maybe people with higher cortisol tend to be in unhappy relationships?
  - (Note: the actual study by Berry & Worthington did account for this, but our example does not!)


## Normality assumption for the <br> independent-samples t-test

:::hcenter
**EITHER the population for each group follows a normal distribution OR each group’s sample size ≥ 30.**
:::

<br>

The independent-samples t-test is designed on the premise that the sampling distributions of each group's sample means are normally distributed.

The sampling distribution of both group's sample means is normal as long as either one of the conditions above is true!

<br>

Check the size of each group:

```{r}
stress_data |>
  group_by(relationship) |>
  count()
```


The groups both contain less than 30 observations, so we can't support the assumption of normality this way.


## One way to check the population distributions: Plot the sample

We don't know how the population is distributed, **but if our sample is approximately normally distributed, then the population probably is too.**

<br>

::::{.columns}
:::{.column width="42%"}
```{r eval = F}
stress_data |>
  ggplot(aes(x = cortisol)) +
  geom_density() +
  facet_wrap(~ relationship)
```
:::
:::{.column width="58%"}
```{r echo=F, fig.width = 7, fig.height = 3}
stress_data |>
  ggplot(aes(x = cortisol)) +
  geom_density() +
  facet_wrap(~ relationship)
```
:::
::::


::::{.columns}
:::{.column width="42%"}
```{r eval=F}
stress_data |>
  ggplot(aes(sample = cortisol)) +
  geom_qq() +
  geom_qq_line() +
  facet_wrap(~ relationship) +
  labs(
    x = "Theoretical quantiles",
    y = 'Sample quantiles'
  ) 
```

:::
:::{.column width="58%"}
```{r echo=F, fig.width = 7, fig.height = 3}
stress_data |>
  ggplot(aes(sample = cortisol)) +
  geom_qq() +
  geom_qq_line() +
  facet_wrap(~ relationship) +
  labs(
    x = "Theoretical quantiles",
    y = 'Sample quantiles'
  ) 
```
:::
::::


## Another way to check the population distributions: Shapiro-Wilk tests

Use `filter()` then `pull()` to extract the `cortisol` data from each group separately.
For example:

```{r}
stress_data |>
  filter(relationship == 'Happy') |>
  pull(cortisol)
```


<br>

Pipe that data directly into `shapiro.test()`:

::::{.columns}
:::{.column width="50%"}
```{r}
stress_data |>
  filter(relationship == 'Happy') |>
  pull(cortisol) |>
  shapiro.test()
```
:::
:::{.column width="50%"}
```{r}
stress_data |>
  filter(relationship == 'Unhappy') |>
  pull(cortisol) |>
  shapiro.test()
```
:::
::::

<br>

We fail to reject the Shapiro-Wilk test's H0 that the population data is normally distributed &nbsp; ✅ 



## Equal variance assumption for the <br> independent-samples t-test

:::hcenter
**The population variance is the same for both groups.**
:::

<br>

We test whether the population variances in the two groups are equal using something called the F-test.

<br>

The F-test is a hypothesis test where:

- $H_0$: The population variances are equal
- $H_1$: The population variances are not equal

<br>

We don't want to reject this H0, so **we want a non-significant result!**


## F-test for equal variance


```{r}
var.test(
  stress_data$cortisol ~ stress_data$relationship, 
  ratio = 1
)
```

**Why `ratio = 1`?** If variances are equal (say they're both 25), then the ratio of variances equals 1:

$$
\frac{\text{variance}_\text{Happy}}{\text{variance}_\text{Unhappy}} = \frac{25}{25} = 1
$$

Take-home message: &nbsp; ✅

> We ran an F-test against the null hypothesis that the populations that each group was sampled from have equal variance.
Based on the sample, we could not reject the null hypothesis of equal variances in the population ($F$(18, 19) = 0.6, $p$ = .3).


# A final note: How to analyse non-independent samples

## One example of non-independent samples


::::{.columns}
:::{.column width="30%"}

:::bluebox
Researchers measured the cortisol in people's saliva before the experiment
:::
:::
:::{.column width="5%"}
:::{.hcenter style="font-size:200%;"}
→
:::
:::
:::{.column width="30%"}
:::bluebox
People spent five minutes vividly imagining or remembering a happy relationship
:::
:::
:::{.column width="5%"}
:::{.hcenter style="font-size:200%;"}
→
:::
:::
:::{.column width="30%"}
:::bluebox
Researchers measured the cortisol in people's saliva again
:::
:::
::::


**RQ: Does cortisol decrease after imagining a happy relationship?**

```{r include=F}
# Based on the descriptives in Table 1, p. 451 of BerryWorthington2001
set.seed(3)

baseline_data <- tibble(
  subj_ID =  c( 
    paste("subj0", 1:9, sep=""),  
    paste("subj", 10:n_ppts, sep="")  
  ),
  relationship = c(
    rep('Happy', 19),
    rep('Unhappy', 20)
  ),
  cortisol = c(
    rnorm(n = 19, mean = 0.19, sd = 0.11),
    rnorm(n = 20, mean = 0.20, sd = 0.10)
  )
)

stress_data2 <- bind_rows(
  t1 = baseline_data,
  t2 = stress_data,
  .id = 'time'
) |>
  filter(relationship == 'Happy') |>
  select(-relationship) |>
  arrange(subj_ID, time)
```


::::{.columns}
:::{.column width="50%"}
<br>

```{r}
stress_data2 |> head(10)
```
:::
:::{.column width="50%"}

```{r fig.width = 6, fig.height = 5}
#| code-fold: true
stress_data2 |>
  ggplot(aes(x = time, y = cortisol)) +
  geom_violin() +
  geom_jitter(width = 0.2, alpha = 0.5, size = 3) +
  stat_summary(fun = mean, geom = 'point', size = 8) +
  labs(
    x = 'time'
  )
```
:::
::::

## These two samples are paired

Paired groups are a specific type of non-independence where one observation in each group is matched with an observation in the other group.

Here, each subject has contributed data to both groups:

```{r echo=F, fig.width = 12, fig.height= 7}
set.seed(1)
stress_data2 |>
  rowwise() |>
  mutate(
    jittertime = ifelse(time == 't1', 1, 2) + rnorm(n = 1, sd = 0.1),
  ) |>
  ggplot(aes(x = jittertime, y = cortisol)) +
  geom_violin(aes(group = time)) +
  geom_point(aes(colour = subj_ID), alpha = 0.5, size = 5) +
  geom_line(aes(group = subj_ID, colour = subj_ID)) +
  scale_x_continuous(breaks = c(1, 2), limits = c(0.7, 2.3)) +
  labs(
    x = 'time'
  ) 
```


## Here's the trick: Paired samples can become <br> a one-sample t-test!

<br>

If you notice that your data contains paired samples, then you can analyse it as follows:

1. Make a new variable that represents the **difference** between the two groups
2. Run a one-sample t-test on that variable, comparing it to 0.

<br>

For Step 1, we'll need to do some data wrangling.

::::{.columns}

:::{.column width="50%"}

**What we've got:**

```{r}
stress_data2 |> head()
```

:::

:::{.column width="50%"}

**What we want:**

```{r include=F}
stress_data2_wide <- stress_data2 |>
  pivot_wider(
    names_from = time,
    values_from = cortisol
  ) |>
  mutate(
    cortisol_diff = t2 - t1
  )
```

```{r}
stress_data2_wide |> head()
```

:::
::::


## Rearrange the data with `pivot_wider()`

```{r}
stress_data2 |>
  pivot_wider(
    names_from = time,      # values of `time` become new column names
    values_from = cortisol  # values of `cortisol` become the column contents
  ) |> 
  head()
```


## Subtract `t1` from `t2` with `mutate()`

```{r}
stress_data2 |>
  pivot_wider(
    names_from = time,      # values of `time` become new column names
    values_from = cortisol  # values of `cortisol` become the column contents
  ) |>
  mutate(
    cortisol_diff = t2 - t1
  ) |> 
  head()
```


## Run a one-sample t-test on `cortisol_diff`

**RQ: Does cortisol decrease after imagining a happy relationship?**

- For H0: If there's no change in cortisol, then the difference will be equal to 0.
- For H1: The RQ predicts that cortisol is higher at `t1` and lower at `t2`.
  - `t2 - t1` = a small number minus a big number = a number less than zero.

$$
\begin{align}
H_0 &: \mu = 0 \\
H_1 &: \mu < 0\\
\end{align}
$$

::::{.columns}
:::{.column width="60%"}
```{r}
t.test(
  stress_data2_wide$cortisol_diff,
  mu = 0,
  alternative = 'less',
  conf.level = .95
)
```
:::
:::{.column width="5%"}
:::
:::{.column width="35%"}

<br>
<br>
<br>
<br>
<br>

>  We fail to reject the H0 that there's no difference in cortisol before and after imagining a happy relationship.
In other words, we find no evidence that thinking about a happy relationship reduces salivary cortisol ($t$(18) = –0.7, $p$ = 0.2, one-sided).

:::
::::





# Back matter

## Revisiting this week's learning objectives

TODO


