---
title: "Chi-squared test of independence"
author: "Elizabeth Pankratz (elizabeth.pankratz@ed.ac.uk)"
editor_options: 
  chunk_output_type: inline
format:
  revealjs:
    smaller: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)

dapr1blue <- '#0F4C81'

source('_theme/theme_quarto.R')

theme_update(
  text = element_text(family = 'Source Sans 3'),
  panel.grid.minor = element_blank(),
  strip.background = element_blank(),
  strip.text.x = element_text(size = 24)
)
```

# Welcome back!

## This week's learning objectives

- TODO

## Where we are in the Block 3 concept map

[TODO figure]


# This week's data

## This week's data: Co-occurrence of ASD and ADHD in children

**ASD** = Autism Spectrum Disorder; **ADHD** = Attention Deficit Hyperactive Disorder

<br>

**Do ASD and ADHD co-occur in children more often than we would expect, based on the prevalence of each neurotype alone?**

```{r include=F}
# counts from Table 1, p. 3
federico <- tribble(
  ~sex, ~yesADHD_noASD, ~noADHD_yesASD, ~yesADHD_yesASD, ~noADHD_noASD,
  'female', 2437, 224, 233, 31385,
  'male', 4971, 883, 906, 29874
) |>
  pivot_longer(
    cols = yesADHD_noASD:noADHD_noASD,
    names_to = 'diagnosis',
    values_to = 'count'
  )

# aggregating over sex for this lecture
neurotype_data <- federico |>
  group_by(diagnosis) |>
  summarise(count = sum(count)) |>
  separate(diagnosis, into = c('ADHD', 'ASD')) |>
  # mutate(
  #   ADHD = gsub('ADHD', '', ADHD),
  #   ASD = gsub('ASD', '', ASD),
  # ) |>
  uncount(count)
```


::::{.columns}
:::{.column width="35%"}

```{r}
neurotype_data
```

:::
:::{.column width="65%"}

```{r fig.width = 7, fig.height = 5}
#| code-fold: true

p_neuro_obs <- neurotype_data |>
  ggplot(aes(x = ADHD, fill = ADHD)) +
  geom_bar() +
  facet_wrap(~ ASD) +
  theme(
    legend.position = 'none'
  ) +
  labs(
    x = ''
  )
p_neuro_obs
```

:::
::::


This example uses real data from [Federico et al. (2024)](https://linkinghub.elsevier.com/retrieve/pii/S1936657423000845){target="_blank"}.



## How can we find out whether ASD+ADHD <br> co-occur unexpectedly often?

Same idea as always: **We take the data we observed and compare it to what we would expect to find, if the effect/difference we care about didn't exist.** 

::::{.columns}
:::{.column width="50%"}

:::hcenter
**Observed:**
:::

```{r}
#| code-fold: true
library(stats)
xtabs(~ ADHD + ASD, data = neurotype_data)
```


:::
:::{.column width="50%"}

:::hcenter
**Expected:**
:::

```{r}
#| code-fold: true
neurotype_chisq <- chisq.test(neurotype_data$ADHD, neurotype_data$ASD)
neurotype_chisq$expected
```

:::
::::

<br>


::::{.columns}
:::{.column width="50%"}

```{r echo=F, fig.width = 6.5, fig.height = 5}
p_neuro_obs
```

:::
:::{.column width="50%"}

```{r echo=F, fig.width = 6.5, fig.height = 5}
neurotype_chisq$expected |>
  as.data.frame() |>
  rownames_to_column(var = 'ADHD') |>
  pivot_longer(cols = ends_with('ASD'), names_to = 'ASD', values_to = 'count') |>
  ggplot(aes(x = ADHD, y = count, fill = ADHD)) +
  geom_col() +
  facet_wrap(~ ASD) +
  theme(
    legend.position = 'none'
  ) +
  labs(
    x = ''
  ) +
  scale_fill_manual(values = c('#888', '#222'))
```



:::
::::



## Where do the expected counts come from?

**Expected counts are essentially fancy averages, calculated on a table of numbers, not a list of numbers.**

You do not need to memorise how to calculate them! The working-out is here only in case you're curious.

<br>

<!-- I include this slide to help demystify the process, only in case you're curious. -->

$$
\text{expected counts} ~=~ \frac{\text{total counts in row} \times \text{total counts in column}}{\text{sample size}}
$$

<br>

For example, the expected count for `yesADHD` and `yesASD`, based on the counts observed:

::::{.columns}
:::{.column width="50%"}
<br>

```{r echo=F}
xtabs(~ ADHD + ASD, data = neurotype_data)
```
:::
:::{.column width="50%"}

```{r include=F}
((7408 + 1139) * (1107 + 1139)) / 70913
```


$$
\begin{align}
\text{expected counts} &= \frac{(7408 + 1139) \times (1107 + 1139)}{70913} \\[.5em]
  &= \frac{8547 \times 2246}{70,913} \\[.5em]
  &= \frac{19,196,562}{70,913} \\[.5em]
  &= 271
\end{align}
$$

:::
::::

If the co-occurrence between ADHD and ASD were _only_ based on how common each neurotype is individually, we would expect them to co-occur in 271 cases.

Instead we observed a co-occurrence in 1139 cases—more than we'd expect. But is it **significantly more?**


## How do we compare observed and expected counts?

We use a chi-squared ($\chi^2$, pronounced "kye") test! 

The idea is really similar to the t-tests you're already familiar with:

::::{.columns}
:::{.column width="48%"}
**In a t-test:**

The difference between **two group means** is measured using a **t-statistic.**
:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
**In a $\chi^2$ test:**

The difference between **observed and expected counts** is measured using a **$\chi^2$ statistic.**
:::
::::

::::{.columns}
:::{.column width="48%"}
To find out if **group means** differ significantly, we compare the **t-statistic** to a **t-distribution** that represents the world in which **groups** do not differ.
:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
To find out if **observed and expected counts** differ significantly, we compare the **$\chi^2$ statistic** to a **$\chi^2$ distribution** that represents the world in which **counts** do not differ.
:::
::::

::::{.columns}
:::{.column width="48%"}
If the **t-statistic** is far enough into the tails of that distribution, we conclude that our observed data is very unlikely **if the true difference between group means were zero;** we reject H0.
:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
If the **$\chi^2$-statistic** is far enough into the tails of that distribution, we conclude that our observed data is very unlikely **if the expected counts were the true counts;** we reject H0.
:::
::::

## The chi-squared statistic



## The chi-squared distribution




## The formal hypotheses of a chi-squared test


## Chi-squared test in R


```{r}
neurotype_test <- chisq.test(neurotype_data$ADHD, neurotype_data$ASD)
neurotype_test$residuals
neurotype_test$expected
```


## Interpreting the test results


# What does the chi-squared test require / assume about the data?

## What does the chi-squared test require / assume about the data?


<br>

::::{.columns}
:::{.column width="30%"}
**Data requirements:**
:::
:::{.column width="70%"}
x
:::
::::

::::{.columns}
:::{.column width="30%"}
**Assumption of independence:**
:::
:::{.column width="70%"}
x
:::
::::



<br>

**What if one or more of these things isn't true?**

- x

<br>

**How do we check each requirement/assumption?**

# Back matter

## Revisiting this week's learning objectives

TODO


# Appendix {.appendix}
