---
title: "Chi-squared test of independence"
author: "Elizabeth Pankratz (elizabeth.pankratz@ed.ac.uk)"
editor_options: 
  chunk_output_type: inline
format:
  revealjs:
    smaller: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)

dapr1blue <- '#0F4C81'

source('_theme/theme_quarto.R')

theme_update(
  text = element_text(family = 'Source Sans 3'),
  panel.grid.minor = element_blank(),
  strip.background = element_blank(),
  strip.text.x = element_text(size = 24)
)
```

# Welcome back!

## This week's learning objectives

- TODO

What kind of test do we use to analyse whether combinations of two categorical variables are more frequent than we'd expect?




## Where we are in the Block 3 concept map

[TODO figure]


# This week's data

## This week's data: Co-occurrence of ASD and ADHD in children

**ASD** = Autism Spectrum Disorder; **ADHD** = Attention Deficit Hyperactive Disorder

<br>

**Do ASD and ADHD co-occur in children more often than we would expect, based on the prevalence of each neurotype alone?**

```{r include=F}
# counts from Table 1, p. 3
federico <- tribble(
  ~sex, ~yesADHD_noASD, ~noADHD_yesASD, ~yesADHD_yesASD, ~noADHD_noASD,
  'female', 2437, 224, 233, 31385,
  'male', 4971, 883, 906, 29874
) |>
  pivot_longer(
    cols = yesADHD_noASD:noADHD_noASD,
    names_to = 'diagnosis',
    values_to = 'count'
  )

# aggregating over sex for this lecture
neurotype_data <- federico |>
  group_by(diagnosis) |>
  summarise(count = sum(count)) |>
  separate(diagnosis, into = c('ADHD', 'ASD')) |>
  # mutate(
  #   ADHD = gsub('ADHD', '', ADHD),
  #   ASD = gsub('ASD', '', ASD),
  # ) |>
  uncount(count)
```


::::{.columns}
:::{.column width="35%"}

```{r}
neurotype_data
```

:::
:::{.column width="65%"}

```{r fig.width = 7, fig.height = 5}
#| code-fold: true

p_neuro_obs <- neurotype_data |>
  ggplot(aes(x = ADHD, fill = ADHD)) +
  geom_bar() +
  facet_wrap(~ ASD) +
  theme(
    legend.position = 'none'
  ) +
  labs(
    x = ''
  )
p_neuro_obs
```

:::
::::


This example uses real data from [Federico et al. (2024)](https://linkinghub.elsevier.com/retrieve/pii/S1936657423000845){target="_blank"}.



## How can we find out whether ASD+ADHD <br> co-occur unexpectedly often?

Same idea as always: **We take the data we observed and compare it to what we would expect to find, if the effect/difference we care about didn't exist.** 

::::{.columns}
:::{.column width="50%"}

:::hcenter
**Observed contingency table:**
:::

```{r}
#| code-fold: true
library(stats)
xtabs(~ ADHD + ASD, data = neurotype_data)
```


:::
:::{.column width="50%"}

:::hcenter
**Expected contingency table:**
:::

```{r}
#| code-fold: true
neurotype_chisq <- chisq.test(neurotype_data$ADHD, neurotype_data$ASD)
neurotype_chisq$expected
```

:::
::::

<br>


::::{.columns}
:::{.column width="50%"}

```{r echo=F, fig.width = 6.5, fig.height = 5}
p_neuro_obs
```

:::
:::{.column width="50%"}

```{r echo=F, fig.width = 6.5, fig.height = 5}
neurotype_chisq$expected |>
  as.data.frame() |>
  rownames_to_column(var = 'ADHD') |>
  pivot_longer(cols = ends_with('ASD'), names_to = 'ASD', values_to = 'count') |>
  ggplot(aes(x = ADHD, y = count, fill = ADHD)) +
  geom_col() +
  facet_wrap(~ ASD) +
  theme(
    legend.position = 'none'
  ) +
  labs(
    x = ''
  ) +
  scale_fill_manual(values = c('#888', '#222'))
```



:::
::::



## Where do the expected counts come from?

**Expected counts are essentially fancy averages, calculated on a table of numbers, not a list of numbers.**

You do not need to memorise how to calculate them! The working-out is here only in case you're curious.

<br>

<!-- I include this slide to help demystify the process, only in case you're curious. -->

$$
\text{expected counts} ~=~ \frac{\text{total counts in row} \times \text{total counts in column}}{\text{sample size}}
$$

<br>

For example, the expected count for `yesADHD` and `yesASD`, based on the counts observed:

::::{.columns}
:::{.column width="50%"}
<br>

```{r echo=F}
xtabs(~ ADHD + ASD, data = neurotype_data)
```
:::
:::{.column width="50%"}

```{r include=F}
((7408 + 1139) * (1107 + 1139)) / 70913
```


$$
\begin{align}
\text{expected counts} &= \frac{(7408 + 1139) \times (1107 + 1139)}{70913} \\[.5em]
  &= \frac{8547 \times 2246}{70,913} \\[.5em]
  &= \frac{19,196,562}{70,913} \\[.5em]
  &= 271
\end{align}
$$

:::
::::

If the co-occurrence between ADHD and ASD were _only_ based on how common each neurotype is individually, we would expect them to co-occur in 271 cases.

Instead we observed a co-occurrence in 1139 cases—more than we'd expect. But is it **significantly more?**

# How do we compare observed and expected counts?

## How do we compare observed and expected counts?

We use a chi-squared ($\chi^2$, pronounced "kye") test! 

Specifically: a **$\chi^2$ test of independence,** since we're essentially testing **if ASD and ADHD occur independently from one another.**

<br>

**The null and alternative hypotheses that we're testing:**

- $H_0$: ASD is independent of ADHD
- $H_1$: ASD is not independent of ADHD

(Don't worry about mathematical notation for these hypotheses!)


## The logic of a chi-squared test of independence

<br>

The idea is really similar to the t-tests you're already familiar with!

::::{.columns}
:::{.column width="48%"}
**In a t-test:**

The difference between **two group means** is measured using a **t-statistic.**
:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
**In a $\chi^2$ test:**

The difference between **observed and expected counts** is measured using a **$\chi^2$ statistic.**
:::
::::

::::{.columns}
:::{.column width="48%"}
To find out if **group means** differ significantly, we compare the **t-statistic** to a **t-distribution** that represents the world in which **the true difference between groups means is zero.**
:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
To find out if **observed and expected counts** differ significantly, we compare the **$\chi^2$ statistic** to a **$\chi^2$ distribution** that represents the world in which **the true counts are the expected  (independent) counts.**
:::
::::

::::{.columns}
:::{.column width="48%"}
If the **t-statistic** is far enough into the tails of that distribution, we conclude that our observed data is very unlikely **if the true difference between group means were zero;** we reject H0.
:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
If the **$\chi^2$-statistic** is far enough into the tails of that distribution, we conclude that our observed data is very unlikely **if the true counts were the expected (independent) counts;** we reject H0.
:::
::::

## The chi-squared distribution

How $\chi^2$ statistics would be distributed if the true counts were the expected counts <br> (i.e., if the H0 of independence is true):

```{r}
#| echo: false

tibble(x = c(0, 15)) |>
  ggplot(aes(x = x)) +
  stat_function(fun = dchisq, args = list(df = 2), colour = '#002060', linewidth = 2) +
  stat_function(fun = dchisq, args = list(df = 4), colour = '#BF1932', linewidth = 2) +
  stat_function(fun = dchisq, args = list(df = 6), colour = '#88B04B', linewidth = 2) +
  stat_function(fun = dchisq, args = list(df = 8), colour = '#FCBB06', linewidth = 2) +
  labs(x=expression(chi^{2}), y = 'Probability density') +
  annotate(geom = 'text', x = 12, y = 0.45, label = 'df = 2', colour = '#002060', size = 8) +
  annotate(geom = 'text', x = 12, y = 0.4,  label = 'df = 4', colour = '#BF1932', size = 8) +
  annotate(geom = 'text', x = 12, y = 0.35, label = 'df = 6', colour = '#88B04B', size = 8) +
  annotate(geom = 'text', x = 12, y = 0.3,  label = 'df = 8', colour = '#FCBB06', size = 8) +
  NULL
```

Like the t-distribution, the exact shape of the $\chi^2$ distribution depends on degrees of freedom (df).

For $\chi^2$, degrees of freedom are based on the size of your contingency table:

$$(\text{number of rows} - 1) \times (\text{number of columns} - 1)$$



<!-- - Shape of the distribution depends on the degrees of freedom $(df)$  -->
<!--   - For $\chi^2$ distribution, $df$ is based on the number of groups $(g)$ within your data -->

<!-- - As the number of comparison groups increases, the distribution curve flattens -->
<!--   - Larger $\chi^2$ values become more probable -->
<!--   - A wider range of $\chi^2$ values become more likely -->

<!-- - The $\chi^2$ distribution begins at 0 -->
<!--   - Categorical variables don't have direction -->
<!--   - $p$-value is computed only in one direction (right-tail) as the Probability of observing a $\chi^2$ statistic as big or bigger than the one obtained -->

## Finding p-values

- We take our observed $\chi^2$ statistic
- We compare it to the $\chi^2$ distribution with appropriate df
- We see how probable it is to observe a $\chi^2$ statistic as extreme or more extreme than the one we observed
  - That probability = the area under the distribution to the right of our observed $\chi^2$ statistic 
  - That probability = **our test's p-value**

<br>

```{r echo=F, warning=F, fig.width = 9, fig.height = 5}
nx <- seq(0.1, 6, length.out = 1000)
p_chisq <- tibble(
  x = nx,
  chisq_dens = dchisq(x = nx, df=1)
) |>
  ggplot(aes(x, y = chisq_dens)) +
  geom_line(linewidth = 2) +
  annotate(geom = 'text', x = 5.5, y = 1.15, label = 'df = 1', colour = '#000', size = 8) +
  labs(y = 'Probability density', x = expression(chi^{2})) +
  NULL
p_chisq
```


## Finding p-values {visibility="uncounted"}

- We take our observed $\chi^2$ statistic
- We compare it to the $\chi^2$ distribution with appropriate df
- We see how probable it is to observe a $\chi^2$ statistic as extreme or more extreme than the one we observed
  - That probability = the area under the distribution to the right of our observed $\chi^2$ statistic 
  - That probability = **our test's p-value**

<br>

```{r echo=F, warning=F, fig.width = 9, fig.height = 5}
chisq_stat <- 1

p_chisq +
  geom_ribbon( aes( x = ifelse(x >= chisq_stat, x, NA), ymin = 0, ymax = chisq_dens ), alpha = 0.3, fill = dapr1blue) +
  geom_vline(xintercept = chisq_stat, linewidth = 1, colour = dapr1blue) +
  annotate( geom = 'text', x = chisq_stat + 0.1, y = 0.88, hjust = 0, label = paste( 'chi-squared\nstatistic:', chisq_stat ), colour = dapr1blue, size = 8) +
  annotate( geom = 'text', x = 5.5, y = 0.14,  label = paste( 'p-value:', round(pchisq(chisq_stat, df = 1, lower.tail=FALSE), 2) ),  colour = dapr1blue, size = 8) +
  NULL
```

## Finding p-values {visibility="uncounted"}

- We take our observed $\chi^2$ statistic
- We compare it to the $\chi^2$ distribution with appropriate df
- We see how probable it is to observe a $\chi^2$ statistic as extreme or more extreme than the one we observed
  - That probability = the area under the distribution to the right of our observed $\chi^2$ statistic 
  - That probability = **our test's p-value**

<br>

```{r echo=F, warning=F, fig.width = 9, fig.height = 5}
chisq_stat <- 3.5

p_chisq +
  geom_ribbon( aes( x = ifelse(x >= chisq_stat, x, NA), ymin = 0, ymax = chisq_dens ), alpha = 0.3, fill = dapr1blue) +
  geom_vline(xintercept = chisq_stat, linewidth = 1, colour = dapr1blue) +
  annotate( geom = 'text', x = chisq_stat + 0.1, y = 0.88, hjust = 0, label = paste( 'chi-squared\nstatistic:', chisq_stat ), colour = dapr1blue, size = 8) +
  annotate( geom = 'text', x = 5.5, y = 0.14,  label = paste( 'p-value:', round(pchisq(chisq_stat, df = 1, lower.tail=FALSE), 2) ),  colour = dapr1blue, size = 8) +
  NULL
```

## Finding p-values {visibility="uncounted"}

- We take our observed $\chi^2$ statistic
- We compare it to the $\chi^2$ distribution with appropriate df
- We see how probable it is to observe a $\chi^2$ statistic as extreme or more extreme than the one we observed
  - That probability = the area under the distribution to the right of our observed $\chi^2$ statistic 
  - That probability = **our test's p-value**

<br>

```{r echo=F, warning=F, fig.width = 9, fig.height = 5}
chisq_stat <- 4.1

p_chisq +
  geom_ribbon( aes( x = ifelse(x >= chisq_stat, x, NA), ymin = 0, ymax = chisq_dens ), alpha = 0.3, fill = dapr1blue) +
  geom_vline(xintercept = chisq_stat, linewidth = 1, colour = dapr1blue) +
  annotate( geom = 'text', x = chisq_stat + 0.1, y = 0.88, hjust = 0, label = paste( 'chi-squared\nstatistic:', chisq_stat ), colour = dapr1blue, size = 8) +
  annotate( geom = 'text', x = 5.5, y = 0.14,  label = paste( 'p-value:', round(pchisq(chisq_stat, df = 1, lower.tail=FALSE), 2) ),  colour = dapr1blue, size = 8) +
  NULL
```




## Run the chi-squared test in R

```{r}
neurotype_chisq <- chisq.test(
  neurotype_data$ADHD,   # name one variable
  neurotype_data$ASD,    # name the other variable (order doesn't matter)
  correct = FALSE        # run the basic version of the test, not "corrected"
)
neurotype_chisq
```

<br>


```{r echo=F, warning=F, fig.width = 9, fig.height = 5}
chisq_stat <- 3267

nx <- seq(0.1, 4000, length.out = 1000)
tibble(
  x = nx,
  chisq_dens = dchisq(x = nx, df=1)
) |>
  ggplot(aes(x, y = chisq_dens)) +
  geom_line(linewidth = 2) +
  labs(y = 'Probability density', x = expression(chi^{2})) +
  geom_ribbon( aes( x = ifelse(x >= chisq_stat, x, NA), ymin = 0, ymax = chisq_dens ), alpha = 0.3, fill = dapr1blue) +
  geom_vline(xintercept = chisq_stat, linewidth = 1, colour = dapr1blue) +
  NULL
```


## Explore results in more detail

Based on the significance test, we can reject the H0 that ASD and ADHD are independent.

But that's very general!
How do we find out specifically if ASD and ADHD co-occur **more than we would expect?**

<br>

<!-- We can look at *how* different the observed data is from the expected data. -->

We can find this out from the $\chi^2$ test's **residual:**


```{r}
neurotype_chisq$residuals
```

<br>

- **Negative** residuals = counts were **lower than expected** in that cell.
- **Positive** residuals = counts were **higher than expected** in that cell.

<br>

For the mathy among you:

$$
\chi^2 ~ \text{residuals} ~=~ \frac{\text{observed} - \text{expected}}{\text{square root(expected)}}
$$

# What does the chi-squared test require / assume about the data?

## What does the chi-squared test require / assume about the data?


<br>

::::{.columns}
:::{.column width="30%"}
**Data requirements:**
:::
:::{.column width="70%"}
x
:::
::::

::::{.columns}
:::{.column width="30%"}
**Assumption of independence:**
:::
:::{.column width="70%"}
x
:::
::::



<br>

**What if one or more of these things isn't true?**

- If expected values are less than 5: Fisher's Exact test (not covered in DAPR, you don't need to know anything more about it).


<br>

**How do we check each requirement/assumption?**

# Back matter

## Revisiting this week's learning objectives

TODO


# Appendix {.appendix}

## Calculating the chi-squared statistic (1)

For each cell in the contingency table, calculate:

$$
\frac{\text{(observed minus expected)}^\text{squared}}{\text{expected}}
$$

then add all those numbers up.
The result is the $\chi^2$ statistic.


::::{.columns}
:::{.column width="50%"}
**Observed:**
```{r echo=F}
obs <- xtabs(~ ADHD + ASD, data = neurotype_data) |> unname()
obs
```

:::
:::{.column width="50%"}
**Expected:**
```{r echo=F}
exp <- neurotype_chisq$expected |> unname()
exp
```

:::
::::

<br>

**So:** (calculation continued on next slide)

|         | noASD                                  | yesASD                         |
| ------- | :------------------------------------: | :----------------------------: |
| <b>noADHD</b>  | $$\frac{(61,259 - 60,391)^2}{60,391}$$ | $$\frac{(1107-1975)^2}{1975}$$ |
| <b>yesADHD</b> | $$\frac{(7408-8276)^2}{8276}$$         | $$\frac{(1139-271)^2}{271}$$   |
||||

## Calculating the chi-squared statistic (2)



**Subtract terms in the numerator to get:** (cool how the same number comes out everywhere!)

|         | noASD                                  | yesASD                         |
| ------- | :------------------------------------: | :----------------------------: |
| <b>noADHD</b>  | $$\frac{868^2}{60,391}$$ | $$\frac{-868^2}{1975}$$ |
| <b>yesADHD</b> | $$\frac{-868^2}{8276}$$         | $$\frac{868^2}{271}$$   |
||||

<br>

**Square the numerator to get:**

|         | noASD                                  | yesASD                         |
| ------- | :------------------------------------: | :----------------------------: |
| <b>noADHD</b>  | $$\frac{753,935}{60,391}$$ | $$\frac{753,935}{1975}$$ |
| <b>yesADHD</b> | $$\frac{753,935}{8276}$$         | $$\frac{753,935}{271}$$   |
||||

<br>

(continued on next slide)


## Calculating the chi-squared statistic (3)

**Divide numerator by denominator to get:**

|         | noASD                                  | yesASD                         |
| ------- | :------------------------------------: | :----------------------------: |
| <b>noADHD</b>  | $$12.5$$ | $$381.7$$  |
| <b>yesADHD</b> | $$91.1$$ | $$2785.1$$ |
||||

<br>

<!-- **(Sense check in R)** -->

<!-- ```{r} -->
<!-- (obs - exp)^2 / exp -->
<!-- ``` -->

**Add up all those numbers:**

$$
12.5 + 381.7 + 91.1 + 2785.1 = 3270 = \chi^2 \text{ statistic}
$$