---
title: "T-Test: Paired Samples"
subtitle: "By Hand Calculations"
author: "Emma Waterston"
format: html
---


In this document we are going to work through the hand calculations for a paired samples $t$-test. 

## Packages

For this example we will only need the `tidyverse` and `kableExtra` packages.

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(kableExtra)
```


## Data

Below is code to create the data we used in lecture:

```{r}
set.seed(007)

exam <- tibble(
  ID = c(rep(paste("ID", 1:50, sep=""),2)),
  stress = round(c(rnorm(50, 9.2, 2.2), rnorm(50, 7.5, 2.8))),
  time = as_factor(c(rep("t1", 50), rep("t2", 50)))
)

head(exam)
```

The `glimpse()` and `str()` functions allow us to look at the structure of our dataset:

```{r}
glimpse(exam)
```

```{r}
str(exam)
```


We can see that we have 100 rows of data, and have three variables -  `ID`, `stress`, and `time`. In terms of structure, `ID` is being recognised as a character (`chr`) vector, `stress` as numeric (`num`/`dbl`), and `time` as a factor (`fct`/`factor`) with two levels - t1 and t2. Each participant has two entries of data (i.e., each `ID` is repeated twice), where they have two stress scores - one for t1 and one for t2. 

## Research Question & Hypotheses

We are interested in whether stress levels are significantly different pre- and post-time management course

- I elect to use a two-tailed test with alpha $(\alpha)$ of .01

- I specify the following hypotheses:

$$
\begin{matrix}
H_0: \mu_{d} = \mu_{d_0}\\
H_1: \mu_{d} \neq \mu_{d_0}\\
\end{matrix}
$$

## Calculation

We can calculate a paired samples $t$-test using the following:

$$
t = \frac{\overline{d} - \mu_{d_0}}{SE_\bar{d}} \qquad \text{where} \qquad {SE_{\bar d}} = \frac{s_d}{\sqrt n}
$$

- Where
  - $\bar{d}$ = mean of the individual difference scores $(d_i)$ 
      - where $d_i = x_{i1} - x_{i2}$
  - $\mu_{d_0}$ is the hypothesised population mean difference in the null hypothesis
      - usually assumed to be 0
  - $SE_\bar{d}$ = standard error of mean difference $(d_i)$
      - $s_{d}$ = standard deviation of the difference scores $(d_i)$
      - $n$ = sample size = number of matched pairs

<br>

- Sampling distribution is a $t$-distribution with $n-1$ degrees of freedom
  - $n-1$ degrees of freedom as we only have 1 group of participants

### Steps in Calculations

Let's break down what we need to do into more simple steps: 

1. Calculate the difference scores for individuals $d_i$ 
2. Calculate the mean of the difference scores $\bar{d}$
3. Calculate the $s_{d}$ of the difference scores
4. Check our $n$
5. Calculate the standard error of mean difference $(SE_\bar{d})$


## Data Organisation

Before we get started with our calculations, we need to make sure our data are in the correct format. It's currently in *long* format (i.e., two rows for each ID), and we need it to be in *wide* (i.e., one row for each ID). To convert, we can use the `pivot_wider()` function:

```{r, echo = TRUE}
exam_wide <- exam |>
  pivot_wider(id_cols =  ID, 
              names_from = time, 
              values_from = stress)
head(exam_wide)
```


## Calculations

The first thing we need to do is calculate the difference between t1 and t2 (i.e., our difference scores for all individuals ($d_i$)):

```{r}
exam_wide <- exam_wide |>  
  mutate(
    dif = t1 - t2)
head(exam_wide)
```


Now we have our difference scores, we can use `R` to calculate the values we need. Recall that we need to calculate the following - the mean of the difference scores $(\bar{d})$, the $(s_{d})$ of the difference scores, and our sample size $(n)$. 

```{r}
exam_wide |>  
  summarise(
    dbar = mean(dif), #step 2
    Sd = sd(dif), #step 3
    mu_d0 = 0,
    n = n()) |> #step 4
  kable(digits = 2) |>
  kable_styling(full_width = FALSE)
```


From the above, we can plug the values directly into the formula: 

$$
t \quad = \quad \frac{\bar{d} - \mu_{d_0}}{SE_\bar{d}} \quad  = \quad \frac{2.1-0}{\frac{3.55}{\sqrt{50}}} \quad = \quad  \frac{2.1}{0.5} \quad = \quad 4.20
$$

Or alternatively, we can ask `R` to do the hard work for us by adding in a couple of extra sums to the above code to calculate the standard error of mean difference $(SE_\bar{d})$ and our $t$-statistic:

```{r}
exam_wide |>  
  summarise(
    dbar = mean(dif), #step 2
    Sd = sd(dif), #step 3
    mu_d0 = 0,
    n = n()) |> #step 4
  mutate(
    SEd = (Sd /sqrt(n)), #step 5
    t = ((dbar-mu_d0)/SEd) #bonus step: calculate t-stat
    ) |>
  kable(digits = 2) |>
  kable_styling(full_width = FALSE)
```


<br>

So in our example $t= 4.20$ (or $t= 4.19$ - when doing hand calculations, there might be a small amount of rounding error when we compare to $t$ calculated in `R` - given that `R` is likely more accurate due to rounding it might be safest to stick with what it tells us!). 


## Is our Test Significant?

We *almost* have all the pieces of information we need to be able to answer that question:

- $t = 4.19$ 
- $df = n-1 = 50 - 1 = 49$
- Hypothesis to test is $two-tailed$
- $\alpha = .01$

But we either need to be told what the corresponding critical values and exact $p$-value is from the associated $t$-distribution in order to make our decision, **or** we can calculate via `R` (i.e., this bit we can't do using by hand calculations).

To calculate the critical values, we can use the `qt()` function; and for the $p$-value, the `pt()` function: 
  
- `qt()`: since using $\alpha = .01$, computes the $t$-statistic corresponding to the 0.5th percentile (lower critical value) and 99.5th percentile (upper critical value) of the $t$-distribution with 49 degrees of freedom    
-  `pt()`: computes the cumulative probability of a $t$-statistic as extreme as 4.19 with 49 degrees of freedom. Since we are conducting a two-tailed test, it multiplies the one-tailed probability by 2 to get the exact $p$-value for both tails    
- `round()`: we are rounding our values to two decimal places for our critical values, and to five decimal places for our $p$-value (note only rounding to so many decimal places since its such a small number! When reporting, use 3 d.p. in line with APA guidelines)  


```{r}
#| echo: true

tibble(
  LowerCrit = round(qt(0.005, 49),2),
  UpperCrit = round(qt(0.995, 49),2),
  Exactp = round(2*(1-pt(4.19, 49)),5)
)
```

If we were to visualise our results, we would see that our sample leads to an observed $t$-statistic that has a relatively low probability. In other words, the observed $t$-statistic was unlikely to be obtained from random samples if the null were true. 

```{r}
#| echo: false

ggplot(tibble(x = c(-6,6)), aes(x = x)) +
  stat_function(fun=dt,
                geom = "line",
                args = list(df=49)) +
  stat_function(fun = dt, 
                geom = "area",
                xlim = c(qt(0.005, 49), -6),
                alpha=.25,
                fill = "blue",
                args = list(df=49)) +
    stat_function(fun = dt, 
                geom = "area",
                xlim = c(qt(0.995, 49), 6),
                alpha=.25,
                fill = "blue",
                args = list(df=49)) +
  geom_vline(xintercept = 4.19, col="red") +
  xlab("\n t") +
  ylab("") +
  ggtitle("t-distribution (df=49); t-statistic = 4.19 (red line)")
```

Linking back to the output of our significance testing, we would say that the critical value is 2.68, and our $t$-statistic (4.19) is larger than this; and we found that $p < .001$, which is $< \alpha$. Therefore, we **reject the null hypothesis**. 


## Effect Size

To calculate the effect size for a paired samples $t$-test, we use:

$$
D = \frac{\bar{d} - \mu_{d_0}}{s_{d}}
$$

- where
  - $\bar{d}$ = mean of the difference scores $(d_i)$  
  - $\mu_{d_0}$ = the hypothesised population difference in means in the null hypothesis
  -	$s_{d}$ = standard deviation of the difference scores $(d_i)$  
      
<br>

In our example, we know that:  
  
  - $\bar{d}$ = 2.1  
  - $\mu_{d_0}$ = 0  
  -	$s_{d}$ = 3.55  
  
We can plug these values into the above formula:

$$
D = \frac{\bar{d} - \mu_{d_0}}{s_{d}} = \frac{2.1 - 0}{3.55} = \frac{2.1}{3.55} = 0.59
$$








