---
title: "T-Test: Independent Samples"
author: "Emma Waterston (Emma.Waterston@ed.ac.uk)"
editor_options: 
  chunk_output_type: console
format:
  revealjs:
    smaller: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)
library(kableExtra)
library(simglm)

source('_theme/theme_quarto.R')

```


# Course Overview 

<br>

:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block1_name = "Exploratory Data Analysis"
block1_lecs = c("W1",
                "W2",
                "W3",
                "W4",
                "W5")
block2_name = "Inference"
block2_lecs = c("W7",
                "W8",
                "W9",
                "W10",
                "W11")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")
course_table(block1_name,block2_name,block1_lecs,block2_lecs, week = 0)
```

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block3_name = "Basic Tests"
block3_lecs = c("W1",
                "W2",
                "W3",
                "W4",
                "W5")
block4_name = "Intro Linear Models"
block4_lecs = c("W6",
                "W7",
                "W8",
                "W9",
                "W10")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")
course_table(block3_name,block4_name,block3_lecs,block4_lecs,week=2)
```

:::

::::


## This Week's Learning Objectives

<br>


::: {style="font-size: 125%;"}

::: {.fragment}
::: {.dapr3callout}
Understand when to use an independent samples $t$-test
:::
:::


::: {.fragment}
::: {.dapr3callout}
Understand how to calculate the test statistic
:::
:::

::: {.fragment}
::: {.dapr3callout}
Know how to conduct the test in `R`
:::
:::

::: {.fragment}
::: {.dapr3callout}
Understand the data requirements and assumptions for a independent samples $t$-test
:::
:::

:::


# T-Test: Independent Samples


## Purpose
- Used when we want to test whether the mean of one group is statistically different from/greater than/less than the mean of another group

<br>

- Independent (i.e., distinct, mutually exclusive) groups in different study designs:
    - Is the average reaction time different between participants who do and do not drink caffeinated drinks?
    - Do participants who identify as 'early birds' have a higher average rating of sleep quality than those who identify as 'night owls'?
    - Do teens who use social media daily have lower self-esteem levels than teens who abstain from social media use? 

## t-statistic

$$
t \quad = \quad \frac{(\bar{x}_1 - \bar{x}_2) - \delta_0}{SE_{(\bar{x}_1 - \bar{x}_2)}} \quad = \quad \frac{\text{difference between sample means} - \text{hypothesised difference in means}}{\text{variation in sample mean differences due to random sampling}}
$$


<br>

- Sampling distribution is a $t$-distribution with $n-2$ degrees of freedom
  - $n-2$ degrees of freedom as we have 2 group of participants, where $n$ = $n_1 + n_2$

<br>
  
- TO DO: LINK TO FLASHCARDS/HANDOUT FOR HOW TO DO BY HAND


## Hypotheses {.smaller}

:::: {.columns}

::: {.column width="50%"}

**Two-tailed (`alternative = "two.sided"`)**

$$
\begin{aligned}
H_0: \mu_1 = \mu_2 
\qquad \text{vs} \qquad 
H_1: \mu_1 \neq \mu_2
\end{aligned}
$$

<br>


**One-tailed (`alternative = "less" | "greater"`)**

$$
\begin{aligned}
H_0: \mu_1 = \mu_2 
\qquad \text{vs} \qquad 
H_1: \mu_1 < \mu_2 
\end{aligned}
$$

<br>

$$
\begin{aligned}
H_0: \mu_1 = \mu_2
\qquad \text{vs} \qquad 
H_1: \mu_1 > \mu_2
\end{aligned}
$$

::: 

::: {.column width="50%"}

<br>

$$
\begin{aligned}
H_0: \mu_1 -\mu_2 = 0
\qquad \text{vs} \qquad 
H_1: \mu_1 - \mu_2 \neq 0 
\end{aligned}
$$

<br>


<br>


$$
\begin{aligned}
H_0: \mu_1 -\mu_2 = 0
\qquad \text{vs} \qquad 
H_1: \mu_1 - \mu_2 < 0 
\end{aligned}
$$

<br>

$$
\begin{aligned}
H_0: \mu_1 -\mu_2 = 0 
\qquad \text{vs} \qquad 
H_1: \mu_1 - \mu_2 > 0
\end{aligned}
$$

::: 

::::


# Example

## Stereotype Threat

- Example taken from Howell, D.C. (2010). *Statistical Methods for Psychology, 7th Edition*. Belmont, CA: Wadsworth Cengage Learning.

- Data from Aronson, Lustina , Good, Keough, Steele and Brown (1998). Experiment on stereotype threat.

	- Two independent groups college students ($n$=12 control; $n$=11 threat condition) 
	- Both samples excel in maths
	- Threat group told certain students usually do better in the test


## Hypotheses

- I elect to use a one-tailed test with alpha $(\alpha)$ of .05

<br>

- I specify the following hypotheses:

$$
\begin{matrix}
H_0: \mu_1 = \mu_2\\
H_1: \mu_1 < \mu_2\\
\end{matrix}
$$


## Data

```{r}
#| echo: false

threat <- tibble(
  Group = as_factor(c(rep("Threat", 11), rep("Control", 12))),
  Score = c(7,5,6,5,6,5,4,7,4,3,6,10,11,8,9,12,11,10,9,8,7,11,9)
)
```

```{r}
#| echo: false

threat
```

## Descriptive Statistics

```{r}
#| echo: true
#| code-line-numbers: "2"

threat |>
  group_by(Group) |>
  summarise(
    Mean = mean(Score),
    SD = sd(Score),
    n = n()
  ) |>
  kable(digits = 2) |>
  kable_styling(full_width = FALSE)
```


## Data Visualisation

```{r}
#| echo: true
#| code-line-numbers: "1"

ggplot(data = threat, aes(x = Group, y = Score, fill = Group)) +
    geom_violin() + 
    geom_jitter(alpha = 0.5) + 
    labs(x = "Group", y = "Score")
```


## Independent Samples T-Test **in R**

:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: true

res <- t.test(threat$Score ~ threat$Group, 
       alternative = "less",
       mu = 0,
       var.equal = TRUE,
       conf.level = 0.95)
res
```

:::

::: {.column width="50%"}

To get **missing** CI - need to do a **two-sided** test

```{r}
#| eval: false
#| echo: true

t.test(threat$Score ~ threat$Group, 
       alternative = "two.sided",
       mu = 0,
       var.equal = TRUE,
       conf.level = 0.95)
```

:::

::::

## Independent Samples T-Test Interpretation TO DO!! 

[INSERT WOOCLAP]

QUESTION:

Do the threat group perform more significantly worse than the control group? Do we reject or fail to reject the null hypothesis?

 ANSWER:

- Yes! Difference estimated to be ~4, and this is statistically significant since p < .001. Thus, reject the null. 

## Write Up

```{r}
#| echo: false

calc <- threat |>
  group_by(Group) |> #<<
  summarise(
    Mean = round(mean(Score),2),
    SD = round(sd(Score),2),
    n = n()
    )
```



An independent samples $t$-test was used to determine whether the average maths score of the stereotype threat group $(n = `r calc[1,4]`)$ was significantly lower $(\alpha = .05)$ than the control group $(n = `r calc[2,4]`)$. There was a significant difference in test score between the control $(M = `r calc[2,2]`, SD=`r calc[2,3]`)$ and threat $(M = `r calc[1,2]`, SD=`r calc[1,3]`)$ groups, where the scores were significantly lower in the threat group $(t(`r res$parameter`)=`r round(res$statistic,2)`, p < .001, one-tailed)$. Therefore, we can reject the null hypothesis. The direction of difference supports our directional hypothesis and indicates that the threat group performed more poorly than the control group.


# Data Requirements & Assumptions


## Data Requirements

- A numeric variable

- A binary variable denoting groups


## Assumption Checks Summary 

```{r tbl7}
#| echo: false

tbl7 <- tibble::tribble(
~` `, ~`Description`, ~`One-Sample t-test`, ~`Independent Samples t-test`, ~`Paired Samples t-test`,
"Normality","Numeric variable (or difference) is normally distributed OR sample size is sufficiently large.","Yes (variable). Sample size guideline: n ≥ 30","Yes (variable in each group).  Sample size guideline: n1 ≥ 30 and n2 ≥ 30","Yes (difference).    Sample size guideline: number of pairs ≥ 30",
"Tests:","Descriptive Statistics and Plots; QQ-Plot; Shapiro-Wilks Test"," "," "," ",
"Independence","Observations are sampled independently.","Yes","Yes (within and across groups)","Yes (across pairs)",
"Tests:","None. Design issue."," "," "," ",
"Homogeneity of variance","Population standard deviation is the same in both groups.","NA","Yes[note]","NA",
"Tests:","F-test"," "," "," ",
"Matched Pairs in data","For paired sample, each observation must have matched pair.","NA","NA","Yes",
"Tests:","None. Data structure issue."," "," "," ",
" "," "," "," "," "
)


kable(tbl7) |>
  kable_styling(bootstrap_options = "striped", full_width = F, font_size = 18) |>
   add_footnote(c("Welch t-test is available if this is not met"), notation = "symbol")

```

## Data Requirements & Assumptions: How to Check/Test

- DV is numeric
  - The dependent variable should be measured on a interval/ratio/integer scale
  
- Normality **within groups**
  - Can be checked with descriptive statistics, visually with plots, and with a Shapiro-Wilks test for each group separately 

- Independence of observations **within and across groups**
  - More of a study design issue, and cannot directly test
  - Need to make sure that each individual only belongs to one group, and only has one observation in the group they belong to 

  
- Homogeneity of variance **across groups**
  - Can be checked using an $F$-test


##  Normality: Skew
- Skew is a descriptive statistic informing us of both the direction and magnitude of asymmetry    

- No strict cut offs for skew - these are loose guidelines on how to interpret skew   
	  
	
| Verbal label         | Magnitude of skew in absolute value |
|:--------------------:|:-----------------:|
| Small (or weak)      | $|\text{Skew}|  < 1$  |
| Medium (or moderate) | $1 \leq  |\text{Skew}|  < 2$     |
| Large (or strong)    | $|\text{Skew}| \geq 2$       |


## Skew **in R**

```{r}
#| echo: true
#| code-line-numbers: "3"

library(psych)
threat |>
  group_by(Group) |>
    summarise(
      skew = round(skew(Score),2)
  )
```


## Normality: Visual Assessment 

We can visually assess normality by plotting the distribution of our outcome variable in both groups separately

- Histograms
    - The count (or frequency) of data points that fall within specified intervals/bins
        
- Density Plots
    - The probability density (or proportion of values) of data points at each value of the observed variable  
      
- QQ-Plots (Quantile-Quantile plot)  
    - Plots the sorted quantiles of one data set (distribution) against sorted quantiles of data set (distribution)
	  - Quantile = the percent of points falling below a given value
	  - For a normality check, we can compare our own data to data drawn from a normal distribution


## Visual Assessment **in R**

```{r}
#| echo: true
#| layout-nrow: 1
#| code-line-numbers: "4,9,13"

#Histogram
ggplot(data = threat, aes(x=Score)) +
  geom_histogram() +
  facet_wrap(~ Group) + 
  labs(title = "Histogram")

#Density Plot
ggplot(data = threat, aes(x=Score)) +
  geom_density() +
  facet_wrap(~ Group) +
  labs(title = "Density")

#QQPlot
ggplot(data = threat, 
       aes(sample = Score, colour = Group)) +
  geom_qq() +
  geom_qq_line() +
      labs(title="QQ-plot", 
       x = "Theoretical quantiles",
       y = "Sample quantiles")
```


##  Normality: Shapiro-Wilks Test 

- Purpose:
  - Statistical test of normality  
  - Checks properties of the observed data against properties we would expected from normally distributed data  

<br>

- Hypotheses:
	- $H_0$: data = normal distribution
	- $H_1$: data = non-normal distribution

<br> 

- Interpretation: 
  - If $p$-value $< \alpha$, then reject the null as data are not normally distributed

<br>

- Note: 
  - Sensitive to $n$ as all $p$-values will be
  - In very large $n$, normality should also be checked with QQ-plots alongside statistical test


##  Shapiro-Wilks Test **in R**

Check outcome variable in both groups separately:

<br>

:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: true
#| code-line-numbers: "2"

threat |> 
  filter(Group == "Control") |> 
  pull(Score) |>
shapiro.test()
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| code-line-numbers: "2"

thr <- threat |> 
  filter(Group == "Threat") |>
  select(Score)
shapiro.test(thr$Score)
```


:::

::::

## Normality Interpretation TO DO!! 

[INSERT WOOCLAP]

QUESTION:

What do we conclude about normality based on skew, plots, and shapiro wilk test? Is the assumption of normality met?
 
ANSWER:

- Skew: low (< 1), so we would conclude that it is not problematic
- Plots: no concerns in histogram or density plots for either group, qqplot reasonable in both 
- Shapiro-wilk: Fail to reject the null since $p > \alpha$
- Conclusion: Normality assumption is met


##  Homogeneity of Variance: F-Test

- Purpose:
    - The $F$-test is a test that compares the variances of two groups

<br> 

- Hypotheses:
	- $H_0$: Population variances are equal
	- $H_1$: Population variances are **not** equal

<br> 

- Interpretation: 
  - If $p$-value $< \alpha$, then reject the null as the variances differ across groups
  
  
##  F-test **in R**
```{r}
#| echo: true

var.test(threat$Score ~ threat$Group, ratio = 1)
```

<br>

#### Why `ratio = 1`?

$$
\begin{matrix}
H_0: \quad \sigma_1^2 = \sigma_2^2 \quad \text{which is equivalent to} \quad \frac{\sigma_1^2}{\sigma_2^2} = 1 \\
H_1: \quad \sigma_1^2 \neq \sigma_2^2 \quad \text{which is equivalent to} \quad \frac{\sigma_1^2}{\sigma_2^2} \neq 1
\end{matrix}
$$


##  F-test Interpretation TO DO!! 

[INSERT WOOCLAP]

QUESTION:

What do we conclude? Do we reject or fail to reject the null hypothesis? Recall:

	- $H_0$: Population variances are equal
 - $p$-value $< \alpha$ = reject the null, variances differ across groups
 
 ANSWER:

- Fail to reject the null since $p > \alpha$
- Equality of variances assumption is met


##  Violation of Homogeneity of Variance 

- If the variances differ, we can use a Welch test

- Conceptually very similar, but we do not use a pooled standard deviation
	- As such our estimate of the SE of the difference changes
	- As do our degrees of freedom


##  Welch Test

- Test statistic:

$$
t = \frac{(\bar{x}_1 - \bar{x}_2) - \delta_0}{SE_{(\bar{x}_1 - \bar{x}_2)}}
$$

- SE calculation:

$$
SE_{(\bar{x}_1 - \bar{x}_2)} = \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}
$$

- Degrees of freedom:

$$
df = \frac{(\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2})^2}{\frac{(\frac{s_1^2}{n_1})^2}{n_1 -1} + \frac{(\frac{s_2^2}{n_2})^2}{n_2 -1}}
$$

## Welch Test **in R**

```{r}
#| echo: true
#| code-line-numbers: "4"

t.test(threat$Score ~ threat$Group, 
       alternative = "less",
       mu = 0,
       var.equal = FALSE, #default, only here to highlight difference
       conf.level = 0.95)
```


# Effect Size


##  Cohen's D: Independent T-Test

- If you **do** have equality of variances ^[if you do not have equality of variances, need to calculate via `cohens_d()` function from `effectsize` package - do not calculate by hand]:

$$
D \quad = \quad \frac{(\bar{x}_1 - \bar{x}_2) - \delta_0}{s_p} \quad = \quad \frac{\text{difference between sample means} - \text{hypothesised difference in means}}{\text{pooled standard deviation}}
$$


<br> 

Recall the common 'cut-offs' for $D$ scores:

| Verbal label         | Magnitude of $D$ in absolute value |
|:--------------------:|:-----------------:|
| Small (or weak)      | $\leq 0.20$       |
| Medium (or moderate) | $\approx 0.50$    |
| Large (or strong)    | $\geq 0.80$       |


## Cohen's D: Independent Samples T-Test **in R**


:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: true
#| warning: false

library(effectsize)

cohens_d(threat$Score ~ threat$Group, 
         mu = 0, 
         alternative = "less", 
         var.equal = TRUE, 
         ci = 0.95)
```

:::

::: {.column width="50%"}

To get **missing** CI - need to do a **two-sided** test:

```{r}
#| echo: true
#| warning: false

library(effectsize)

cohens_d(threat$Score ~ threat$Group, 
       alternative = "two.sided",
       mu = 0,
       var.equal = TRUE,
       ci = 0.95)
```

:::

::::

## Cohen's D Interpretation TO DO!! 

[INSERT WOOCLAP]

QUESTION:

What do we conclude? How would we verbally describe the effect size?
 
ANSWER:

- large

## Write Up: Data Requirements, Assumptions, & Effect Size

The DV of our study, Score, was measured on a continuous scale, and data were independent (participants belonged to one of two groups - Control or Threat). The assumption of normality was visually assessed (via histograms, density plots, and a QQplot) as well as statistically via a Shapiro-Wilks test. The QQplots did not show much deviation from the diagonal line in either group, and the Shapiro-Wilks test for both the Control $(W = 0.96, p = .716)$ and Threat $(W = 0.94, p = .518)$ conditions suggested that the samples came from a population that was normally distributed. This was inline with the histogram and density plots for each group, which suggested that Score was normally distributed (and where $skew < 1$). Based on the results of our $F$-test, there was no significant difference between the two population variances $(F(10,11) = 0.71, p = .604)$. The size of the effect was found to be large $D = -3.08~[-4.30,-2.02]$. 


## Revisiting This Week's Learning Objectives

<br>

::: {}
::: {.dapr3callout}
**When to use an independent samples t-test?**

- Used to test whether the difference in means between two groups is significantly different from the hypothesised population difference in means under the null (usually assumed to be 0)  

:::
:::

::: {}
::: {.dapr3callout}
**How to calculate the test statistic?**


$$
t = \frac{(\bar{x}_1 - \bar{x}_2) - \delta_0}{SE_{(\bar{x}_1 - \bar{x}_2)}}
$$


:::
:::

## Revisiting This Week's Learning Objectives

<br>

::: {}
::: {.dapr3callout}
**How to conduct the test in R?**

```{r}
#| eval: false

t.test(dataframe$DV ~ dataframe$IV, #numeric outcome (DV) ~ binary grouping variable (IV)
       mu = ?, #hypothesised population mean difference under the null
       alternative = "", #options: "two.sided", "less", "greater"
       var.equal = , #options: TRUE or FALSE
       conf.level = 0.??) #depends on alpha specified e.g., if alpha = .05, then 0.95
```

  
:::
:::

::: {}
::: {.dapr3callout}
**What are the data requirements & assumptions of an independent samples t-test?**

- Data are continuous
- The data are normally distributed OR the sample size is sufficiently large (rule-of-thumb $n \geq$ 30 in each group)
- The data are independent (i.e. the value of a data point does not depend on the value of another data point in any way)
- The variance is equal across groups (if not, use Welch test)

:::
:::


## This Week 

