---
title: "T-Test: Paired Samples"
author: "Emma Waterston (Emma.Waterston@ed.ac.uk)"
editor_options: 
  chunk_output_type: console
format:
  revealjs:
    smaller: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)
library(kableExtra)
library(simglm)

source('_theme/theme_quarto.R')

```


# Course Overview 

<br>

:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block1_name = "Exploratory Data Analysis"
block1_lecs = c("W1",
                "W2",
                "W3",
                "W4",
                "W5")
block2_name = "Inference"
block2_lecs = c("W7",
                "W8",
                "W9",
                "W10",
                "W11")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")
course_table(block1_name,block2_name,block1_lecs,block2_lecs, week = 0)
```

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block3_name = "Basic Tests"
block3_lecs = c("W1",
                "W2",
                "W3",
                "W4",
                "W5")
block4_name = "Intro Linear Models"
block4_lecs = c("W6",
                "W7",
                "W8",
                "W9",
                "W10")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")
course_table(block3_name,block4_name,block3_lecs,block4_lecs,week=1)
```

:::

::::


## This Week's Learning Objectives

<br>


::: {style="font-size: 125%;"}

::: {.fragment}
::: {.dapr3callout}
Understand when to use an paired samples $t$-test
:::
:::


::: {.fragment}
::: {.dapr3callout}
Understand how to calculate the test statistic
:::
:::

::: {.fragment}
::: {.dapr3callout}
Know how to conduct the test in `R`
:::
:::

::: {.fragment}
::: {.dapr3callout}
Understand the data requirements and assumptions for a paired samples $t$-test
:::
:::

:::


# T-Test: Paired Samples


## Purpose
- Used when we want to test the difference in mean scores for a sample comprising matched (or naturally related) pairs 

<br>

- Matched pairs in different study designs:
    - Longitudinal: The same participants provide responses to the same measure on two occasions  
        - Pre-test and post-test score with an intervention administered between the time points  
    - Paired measures: The same participants provide responses to two different measures  
        - A participant experiences two experimental conditions (e.g., caffeine and placebo)    
    - Paired sample: Participants in the two groups are different but inherently related. Consequently, their responses are assumed to be correlated  
        - Participants are pairs such as couples (e.g., husband and wife), siblings, twins, etc.  

## t-statistic

$$
t \quad = \quad  \frac{\overline{d} - \mu_{d_0}}{SE_\bar{d}} \quad = \quad  \frac{\text{mean difference between pairs} - \text{hypothesised mean difference}}{\text{variation in sample mean differences due to random sampling}}
$$


<br>

- Sampling distribution is a $t$-distribution with $n-1$ degrees of freedom
  - $n-1$ degrees of freedom as we have 1 group of participants

<br> 

- TO DO: LINK TO FLASHCARDS/HANDOUT FOR HOW TO DO BY HAND


## Hypotheses {.smaller}

:::: {.columns}

::: {.column width="50%"}

**Two-tailed (`alternative = "two.sided"`)**

$$
\begin{aligned}
H_0: \mu_{d} = \mu_{d_0}
\qquad \text{vs} \qquad 
H_1:\mu_{d} \neq \mu_{d_0}
\end{aligned}
$$

<br>


**One-tailed (`alternative = "less" | "greater"`)**

$$
\begin{aligned}
H_0: \mu_{d} = \mu_{d_0}
\qquad \text{vs} \qquad 
H_1: \mu_{d} < \mu_{d_0}
\end{aligned}
$$

<br>

$$
\begin{aligned}
H_0: \mu_{d} = \mu_{d_0}
\qquad \text{vs} \qquad 
H_1: \mu_{d} > \mu_{d_0}
\end{aligned}
$$

::: 

::: {.column width="50%"}

<br>

$$
\begin{aligned}
H_0: \mu_{d} - \mu_{d_0} = 0 
\qquad \text{vs} \qquad 
H_1: \mu_{d} - \mu_{d_0}  \neq 0
\end{aligned}
$$

<br>


<br>


$$
\begin{aligned}
H_0: \mu_{d} - \mu_{d_0}  = 0  
\qquad \text{vs} \qquad 
H_1: \mu_{d} - \mu_{d_0}  < 0
\end{aligned}
$$

<br>

$$
\begin{aligned}
H_0: \mu_{d} - \mu_{d_0}  = 0  
\qquad \text{vs} \qquad 
H_1: \mu_{d} - \mu_{d_0}  > 0
\end{aligned}
$$

::: 

::::


# Example

## Time Management & Stress

- I want to assess whether a time-management course influenced levels of exam stress in students

- I ask 50 students to take a self-report stress measure during their winter exams (end Semester 1)

- At the beginning of semester 2 they take a time management course 

- I then assess their self-report stress in the summer exam block (end Semester 2)
	- Let's assume for the sake of this example that I have been able to control the volume and difficulty of the exams the students take in each block


## Hypotheses

- I elect to use a two-tailed test with alpha $(\alpha)$ of .01

<br>

- I specify the following hypotheses:

$$
\begin{matrix}
H_0: \mu_{d} = \mu_{d_0}\\
H_1: \mu_{d} \neq \mu_{d_0}\\
\end{matrix}
$$


## Data

```{r}
#| echo: false

set.seed(007)
exam <- tibble(
  ID = c(rep(paste("ID", 1:50, sep=""),2)),
  stress = round(c(rnorm(50, 9.2, 2.2), rnorm(50, 7.5, 2.8))),
  time = as_factor(c(rep("t1", 50), rep("t2", 50)))
)
```

```{r}
#| echo: false

exam
```

## Descriptive Statistics

```{r}
#| echo: true
#| code-line-numbers: "2"

exam |>
  group_by(time) |>
  summarise(
    Mean = mean(stress),
    SD = sd(stress),
    n = n()
  ) |>
  kable(digits = 2) |>
  kable_styling(full_width = FALSE)
```


## Data Visualisation

```{r}
#| echo: true

ggplot(data = exam, aes(x = time, y = stress, group = ID)) +
  geom_line(color = "#0F4C81") +
  geom_point() +
  theme_minimal() +
  labs(title = "Stress Scores by Time", 
       x = "Time", 
       y = "Stress")
```


## Data Organisation

- Our data is currently in what is referred to as **long** format
  - All the scores are in one column, with two entries per participant

- To calculate the $d_i$ values, we will convert this to **wide** format
  - Where there are two columns representing the score at time 1 and time 2
  - And a single row per person
  

## Data Organisation

```{r}
#| echo: true

exam_wide <- exam |>
  pivot_wider(id_cols =  ID, 
              names_from = time, 
              values_from = stress)
head(exam_wide)
```


## Adding the Difference Scores

- We are concerned with the **difference scores**

- We can add them to `exam_wide` for ease:

```{r}
#| echo: true

exam_wide <- exam_wide |>  
  mutate(
    dif = t1 - t2)

head(exam_wide)
```


## Descriptive Statistics

```{r}
#| echo: true
#| code-line-numbers: "2"

exam_wide |>
  summarise(
    Mean = mean(dif),
    SD = sd(dif),
    n = n()
  ) |>
  kable(digits = 2) |>
  kable_styling(full_width = FALSE)
```


## Data Visualisation

```{r}
#| echo: true

ggplot(data = exam_wide, aes(x = dif)) +
  geom_histogram(fill ="#0F4C81") +
  labs(title = "Difference in Stress Scores", 
       x = "Difference in Scores between Time 1 and Time 2", 
       y = "Frequency")
```

## Paired Samples T-Test **in R**

```{r}
#| echo: true

res_wide <- t.test(exam_wide$t1, exam_wide$t2, 
       paired = TRUE, 
       mu = 0,
       alternative = "two.sided",
       conf.level = 0.99)
res_wide
```


## Paired Samples T-Test as a One Sample T-Test in Disguise **in R**

- Instead of using our two numeric columns for our t1 and t2 stress scores, we can perform a one sample $t$-test on the mean of those difference scores!

```{r}
#| echo: true

res_one <- t.test(exam_wide$dif, 
       mu = 0,
       alternative = "two.sided",
       conf.level = 0.99)
res_one
```


## Paired Samples T-Test Interpretation TO DO!! 

[INSERT WOOCLAP]

QUESTION:

Is there a statistically significant difference in stress scores? Do we reject or fail to reject the null hypothesis?

 ANSWER:

- Yes! Difference estimated to be 2.1, and this is statistically significant since p < .001. Thus, reject the null. 

## Write Up

A paired samples $t$-test was conducted in order to determine a if a statistically significant $(\alpha = .01)$ mean difference in self-reported stress was present, pre- and post-time management intervention in a sample of 50 undergraduate students. The pre-intervention mean score was higher $(M=`r mean(exam_wide$t1)`, SD = `r round(sd(exam_wide$t1),2)`)$ than the post intervention score $(M = `r mean(exam_wide$t2)`, SD = `r round(sd(exam_wide$t2),2)`)$. The difference was statistically significant $(t(`r res_wide$parameter`)= `r round(res_wide$statistic,2)`, p < . 001, two-tailed)$. We are 99% confident that post-intervention stress scores were between 0.76 and 3.44 points lower than pre-intervention stress scores. Thus, we reject the null hypothesis of no difference.


# Data Requirements & Assumptions


## Data Requirements

- A numeric variable


## Assumption Checks Summary 

```{r tbl7}
#| echo: false

tbl7 <- tibble::tribble(
~` `, ~`Description`, ~`One-Sample t-test`, ~`Independent Samples t-test`, ~`Paired Samples t-test`,
"Normality","Numeric variable (or difference) is normally distributed OR sample size is sufficiently large.","Yes (variable). Sample size guideline: n ≥ 30","Yes (variable in each group).  Sample size guideline: n1 ≥ 30 and n2 ≥ 30","Yes (difference).    Sample size guideline: number of pairs ≥ 30",
"Tests:","Descriptive Statistics and Plots; QQ-Plot; Shapiro-Wilks Test"," "," "," ",
"Independence","Observations are sampled independently.","Yes","Yes (within and across groups)","Yes (across pairs)",
"Tests:","None. Design issue."," "," "," ",
"Homogeneity of variance","Population standard deviation is the same in both groups.","NA","Yes[note]","NA",
"Tests:","F-test"," "," "," ",
"Matched Pairs in data","For paired sample, each observation must have matched pair.","NA","NA","Yes",
"Tests:","None. Data structure issue."," "," "," ",
" "," "," "," "," "
)


kable(tbl7) |>
  kable_styling(bootstrap_options = "striped", full_width = F, font_size = 18) |>
   add_footnote(c("Welch t-test is available if this is not met"), notation = "symbol")

```

## Data Requirements & Assumptions: How to Check/Test

- DV is numeric
  - The dependent variable should be measured on a interval/ratio/integer scale
  
- Normality of the difference scores $(d_i)$
  - Can be checked with descriptive statistics, visually with plots, and with a Shapiro-Wilks test for each group separately 

- Independence of observations **across pairs**
  - More of a study design issue, and cannot directly test

- Data are matched pairs 
  - More of a study design issue, and cannot directly test


##  Normality: Skew
- Skew is a descriptive statistic informing us of both the direction and magnitude of asymmetry    

- No strict cut offs for skew - these are loose guidelines on how to interpret skew   
	  
	
| Verbal label         | Magnitude of skew in absolute value |
|:--------------------:|:-----------------:|
| Small (or weak)      | $|\text{Skew}|  < 1$  |
| Medium (or moderate) | $1 \leq  |\text{Skew}|  < 2$     |
| Large (or strong)    | $|\text{Skew}| \geq 2$       |


## Skew **in R**

```{r}
#| echo: true

library(psych)
exam_wide |>
  summarise(
    skew = round(skew(dif), 2)
  )
```


## Normality: Visual Assessment 

We can visually assess normality by plotting the distribution of the difference scores $(d_i)$

- Histograms
    - The count (or frequency) of data points that fall within specified intervals/bins
        
- Density Plots
    - The probability density (or proportion of values) of data points at each value of the observed variable  
      
- QQ-Plots (Quantile-Quantile plot)  
    - Plots the sorted quantiles of one data set (distribution) against sorted quantiles of data set (distribution)
	  - Quantile = the percent of points falling below a given value
	  - For a normality check, we can compare our own data to data drawn from a normal distribution


## Visual Assessment **in R**

```{r}
#| echo: true
#| layout-nrow: 1

#Histogram
ggplot(exam_wide, aes(x=dif)) +
  geom_histogram() + 
  labs(title = "Histogram")

#Density Plot
ggplot(exam_wide, aes(x=dif)) +
  geom_density() + 
  labs(title = "Density")

#QQPlot
ggplot(exam_wide, aes(sample = dif)) +
  geom_qq() +
  geom_qq_line() +
  labs(title="QQ-plot",
       x = "Theoretical quantiles",
       y = "Sample quantiles")
```


##  Normality: Shapiro-Wilks Test 

- Purpose:
  - Statistical test of normality  
  - Checks properties of the observed data against properties we would expected from normally distributed data  

<br>

- Hypotheses:
	- $H_0$: data = normal distribution
	- $H_1$: data = non-normal distribution

<br> 

- Interpretation: 
  - If $p$-value $< \alpha$, then reject the null as data are not normally distributed

<br>

- Note: 
  - Sensitive to $n$ as all $p$-values will be
  - In very large $n$, normality should also be checked with QQ-plots alongside statistical test


##  Shapiro-Wilks Test **in R**

```{r}
#| echo: true

shapiro.test(exam_wide$dif)
```


## Normality Interpretation TO DO!! 

[INSERT WOOCLAP]

QUESTION:

What do we conclude about normality based on skew, plots, and shapiro wilk test? Is the assumption of normality met?
 
ANSWER:

- Skew: low (< 1), so we would conclude that it is not problematic
- Plots: Histogram and density plot looks fairly normally distributed. QQplot - we want our points to be as close to this line as possible, and there doesn't appear to be much deviation
- Shapiro-wilk: Fail to reject the null since $p > \alpha$


# Effect Size


##  Cohen's D: Paired T-Test

$$
D \quad = \quad  \frac{\bar{d} - \mu_{d_0}}{s_{d}} \quad = \quad \frac{\text{mean difference between pairs} - \text{hypothesised mean difference}}{\text{standard deviation of the difference scores}}
$$


<br> 

Recall the common 'cut-offs' for $D$ scores:

| Verbal label         | Magnitude of $D$ in absolute value |
|:--------------------:|:-----------------:|
| Small (or weak)      | $\leq 0.20$       |
| Medium (or moderate) | $\approx 0.50$    |
| Large (or strong)    | $\geq 0.80$       |


## Cohen's D: Paired Samples T-Test **in R**

```{r}
#| echo: true
#| warning: false

library(effectsize)

cohens_d(exam_wide$t1, exam_wide$t2, 
       paired = TRUE, 
       mu = 0,
       alternative = "two.sided",
       ci = 0.99)
```


## Cohen's D Interpretation TO DO!! 

[INSERT WOOCLAP]

QUESTION:

What do we conclude? How would we verbally describe the effect size?
 
ANSWER:

- medium-large

## Write Up: Data Requirements, Assumptions, & Effect Size

The DV of our study, Stress, was measured on a continuous scale. Independence of observations can be assumed based on the study design. Data comprised matched pairs of observations as participants were assessed twice, pre- and post- time management course. The assumption of normality was visually assessed (via histograms, density plots, and a QQplot) as well as statistically via a Shapiro-Wilks test. The QQplot did not show much deviation from the diagonal line, and the Shapiro-Wilks test suggested that the difference scores were normally distributed $(W = 0.97, p = .264)$. This was inline with the histogram and density plots, which suggested that the difference in scores between the two assessment times was normally distributed (and where $skew < 1$). The size of the effect was found to be medium-large $(D = 0.59~[0.19, 0.99])$.


## Revisiting This Week's Learning Objectives

<br>

::: {}
::: {.dapr3callout}
**When to use a paired samples t-test?**

- Used to test whether the mean difference between two sets of paired observations (e.g., from longitudinal, paired measures, or paired samples study designs) is significantly different from the hypothesised population mean difference under the null (usually assumed to be 0)  

:::
:::

::: {}
::: {.dapr3callout}
**How to calculate the test statistic?**


$$
t = \frac{\overline{d} - \mu_{d_0}}{SE_\bar{d}} \qquad \text{where} \qquad {SE_{\bar d}} = \frac{s_d}{\sqrt n}
$$


:::
:::

## Revisiting This Week's Learning Objectives

<br>

::: {}
::: {.dapr3callout}
**How to conduct the test in R?**

```{r}
#| eval: false

t.test(dataframe$var1, dataframe$var2, #must have two numeric columns
       paired = TRUE, #tells R data is comprised matched pairs
       mu = ?, #hypothesised population mean difference under the null
       alternative = "", #options: "two.sided", "less", "greater"
       conf.level = 0.??) #depends on alpha specified e.g., if alpha = .05, then 0.95
```

  
:::
:::

::: {}
::: {.dapr3callout}
**What are the data requirements & assumptions of a paired samples t-test?**

- Data are continuous
- Data must be comprised of matched pairs
- The difference scores are normally distributed OR the sample size is large (rule-of-thumb $n_{pairs} \geq$ 30)
- The difference scores are independent (i.e. the value of a the difference for one pair does not depend on the values of another pair in any way)

:::
:::


## This Week 

