---
title: "Correlations"
author: "Dr. Elizabeth Pankratz (elizabeth.pankratz@ed.ac.uk)"
editor_options: 
  chunk_output_type: inline
format:
  revealjs:
    smaller: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)

dapr1blue <- '#0F4C81'

source('_theme/theme_quarto.R')

theme_update(
  text = element_text(family = 'Source Sans 3'),
  panel.grid.minor = element_blank(),
  strip.background = element_blank(),
  strip.text.x = element_text(size = 24)
)
```

# Welcome back!

## This week's learning objectives

:::{.dapr1callout}
What kind of test do we use to analyse whether, as the value of one continuous variable changes, the value of another continuous variable changes with it?

- Pearson's correlation
:::


:::{.dapr1callout}
What kind of test do we use to analyse the same question but with ordinal/ranking variables instead?

- Spearman's correlation
:::


:::dapr1callout
What does each test require/assume to be true about the data that it models?

- Pearson:
  (0) Both variables are continuous
  (1) Data contains no extreme outliers
  (2) Association between variables is sufficiently linear
  (3) Both variables are normally distributed
  (4) Each variable's variance is equal across the range of the other variable.


- Spearman:
  (0) Both variables represent rankings/orderings
:::

:::dapr1callout
How do we check whether we're satisfied with each test's assumptions?

- Pearson:
  (0) check that variable is on interval/ratio scale;
  (1) visually inspect scatterplot for outliers
  (2) compare loess line to lm line
  (3) Shapiro-Wilk;
  (4) visually inspect scatterplot for fan/funnel shape

- Spearman:
  (0) check that variable represents rankings/orderings
:::



## Where we are in the Block 3 concept map

[TODO figure]


# What are correlations?

## What are correlations?

If two variables are **correlated,** then as the value of one variable changes, the value of the other variable changes too.

Put differently: you can make a reasonable guess about the value of one variable based on the value of the other one.


<br>

::::{.columns}
:::{.column width="30%"}
**Not correlated:**

```{r echo=F, fig.height = 4, fig.width = 4}
set.seed(3)
df1 <- MASS::mvrnorm(
  n = 100,
  mu = c(0, 0),
  Sigma = matrix(c(100, 0, 0, 100), nrow = 2)
)
colnames(df1) <- c('x', 'y')

df1 |>
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 3)
```


For any given value of `x`, you can't really predict what kind of value `y` is likely to have.

:::
:::{.column width="5%"}
:::
:::{.column width="30%"}
**Positively correlated:**

```{r echo=F, fig.height = 4, fig.width = 4}
set.seed(2)
cov <- 80
df <- MASS::mvrnorm(
  n = 100,
  mu = c(0, 0),
  Sigma = Matrix::nearPD(matrix(c(100, cov, cov, 100), nrow = 2))$mat
)
colnames(df) <- c('x', 'y')

df |>
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 3)
```

If `x` increases, then `y` increases.

If `x` decreases, then `y` decreases.

:::
:::{.column width="5%"}
:::
:::{.column width="30%"}

**Negatively correlated:**

```{r echo=F, fig.height = 4, fig.width = 4}
set.seed(1)
cov <- -80
df <- MASS::mvrnorm(
  n = 100,
  mu = c(0, 0),
  Sigma = Matrix::nearPD(matrix(c(100, cov, cov, 100), nrow = 2))$mat
)
colnames(df) <- c('x', 'y')

df |>
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 3)
```


If `x` increases, then `y` decreases.

If `x` decreases, then `y` increases.

:::
::::


## We measure correlation using a <br> "correlation coefficient" between –1 and 1

<br>


::::{.columns}
:::{.column width="30%"}
**Not correlated:**

```{r echo=F, fig.height = 4, fig.width = 4}
set.seed(3)
df1 <- MASS::mvrnorm(
  n = 100,
  mu = c(0, 0),
  Sigma = matrix(c(100, 0, 0, 100), nrow = 2)
)
colnames(df1) <- c('x', 'y')

df1 |>
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 3)
```

For uncorrelated variables, the correlation coefficient will be **close to zero.**


:::
:::{.column width="5%"}
:::
:::{.column width="30%"}
**Positively correlated:**

```{r echo=F, fig.height = 4, fig.width = 4}
set.seed(2)
cov <- 80
df <- MASS::mvrnorm(
  n = 100,
  mu = c(0, 0),
  Sigma = Matrix::nearPD(matrix(c(100, cov, cov, 100), nrow = 2))$mat
)
colnames(df) <- c('x', 'y')

df |>
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 3)
```

For positively correlated variables, the correlation coefficient will be **between 0 and 1** (that is, positive).


:::
:::{.column width="5%"}
:::
:::{.column width="30%"}

**Negatively correlated:**

```{r echo=F, fig.height = 4, fig.width = 4}
set.seed(1)
cov <- -80
df <- MASS::mvrnorm(
  n = 100,
  mu = c(0, 0),
  Sigma = Matrix::nearPD(matrix(c(100, cov, cov, 100), nrow = 2))$mat
)
colnames(df) <- c('x', 'y')

df |>
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 3)
```

For negatively correlated variables, the correlation coefficient will be **between –1 and 0** (that is, negative).

:::
::::


# This week's data: <br> Socio-economic status (SES) and test scores

## Socio-economic status (SES) and test scores

**RQ: Do students from higher-income families achieve higher scores on the reading comprehension section of the SAT ("SAT Verbal")?** 
(The SAT is an American college admission test.)

This example is based on real data from [Zwick and Greif Green (2007)](https://onlinelibrary.wiley.com/doi/full/10.1111/j.1745-3984.2007.00025.x){target="_blank"}.

<br>

```{r include=F}
# sample size from Table 3 on p. 32
zwick_n <- 98391

# correlation from Table 7 on p. 36 (total within-school correlation between income and SAT verbal score)
zwick_cor <- 0.219

# mean score on SAT verbal from Table 3 on p. 33, arbitrary SD of 100, thanks google
zwick_SAT_mean <- 504.5
zwick_SAT_sd <- 100

# income mean and SD approximated from income distribution from Table 3 on p. 32
zwick_income_dist <- c(
  rep(5000, 41),
  rep(15000, 81),
  rep(25000, 99),
  rep(35000, 111),
  rep(45000, 92),
  rep(55000, 96),
  rep(65000, 87),
  rep(75000, 85),
  rep(90000, 122),
  rep(120000, 187)
)
zwick_income_mean <- mean(zwick_income_dist)
zwick_income_sd <- sd(zwick_income_dist)

# create vcov mtx
zwick_cov  <- zwick_cor * zwick_income_sd * zwick_SAT_sd
zwick_vcov <- matrix(c(zwick_income_sd^2, zwick_cov, zwick_cov, zwick_SAT_sd^2), nrow = 2)

set.seed(1)
ses_data <- MASS::mvrnorm(
  n = 800,
  mu = c(zwick_income_mean, zwick_SAT_mean),
  Sigma = zwick_vcov
)
colnames(ses_data) <- c('Income', 'Score')
ses_data <- ses_data |>
  as_tibble() |>
  # filter for realism
  filter(Income > 0, Score >= 200, Score <= 800)

cor(ses_data$Income, ses_data$Score)
cor.test(ses_data$Income, ses_data$Score)
```

::::{.columns}
:::{.column width="30%"}


```{r}
ses_data |> head(12)
```

:::
:::{.column width="70%"}

```{r fig.align="center", fig.width = 7, fig.height = 6}
#| code-fold: true

ses_data |>
  ggplot(aes(x = Income, y = Score)) +
  geom_point(size = 3) +
  labs(
    x = "Parents' income (USD)",
    y = 'Score on SAT Verbal'
  )
```

:::
::::


## Plotting continuous/continuous correlations

::::{.columns}
:::{.column width="50%"}
```{r eval=F, fig.align="center", fig.width = 7, fig.height = 6}
ses_data |>
  ggplot(aes(x = Income, y = Score)) +
  geom_point() +
  labs(
    x = "Parents' income (USD)",
    y = 'Score on SAT Verbal'
  )
```
:::
:::{.column width="50%"}
```{r echo=F, fig.align="center", fig.width = 7, fig.height = 6}
ses_data |>
  ggplot(aes(x = Income, y = Score)) +
  geom_point(size = 3) +
  labs(
    x = "Parents' income (USD)",
    y = 'Score on SAT Verbal'
  )
```
:::
::::




## Plotting continuous/continuous correlations {visibility="uncounted"}


::::{.columns}
:::{.column width="50%"}
```{r eval=F, fig.align="center", fig.width = 7, fig.height = 6}
ses_data |>
  ggplot(aes(x = Income, y = Score)) +
  geom_point() +
  labs(
    x = "Parents' income (USD)",
    y = 'Score on SAT Verbal'
  ) +
  
  # add a line that fits the data
  geom_smooth(      
    
    # what kind of line? 
    # lm = Linear Model, 
    # a straight line through the data
    method = 'lm',  
    
    # show standard error ribbon around 
    # line? FALSE = don't show
    se = FALSE
  )
```
:::
:::{.column width="50%"}
```{r echo=F, fig.align="center", fig.width = 7, fig.height = 6}
ses_data |>
  ggplot(aes(x = Income, y = Score)) +
  geom_point(size = 3) +
  labs(
    x = "Parents' income (USD)",
    y = 'Score on SAT Verbal'
  ) +
  geom_smooth(method = 'lm', se = F, linewidth = 2)
```
:::
::::



## What's the correlation coefficient for this data?

<br>

We find out the correlation coefficient using the function `cor()`, giving it each column of our data frame.

The format to use in general:

```{r eval=F}
cor(DATA$COLUMN1, DATA$COLUMN2)
```

**For our income/test score data:**

```{r}
cor(ses_data$Income, ses_data$Score)
```

<br>

- Positive correlation coefficient $\rightarrow$ as income increases, so does SAT score.

- You can interpret the size of a correlation coefficient kind of like a percentage: 21.6% is kind of small, but definitely not nothing.

  - **Weak correlations:** between .1 and .3 (whether positive or negative)
  - **Moderate correlations:** between .3 and .5 (whether positive or negative)
  - **Strong correlations:** greater than .5 (whether positive or negative)
  

<br>

The specific kind of correlation coefficient that we get from `cor()` by default is **Pearson's correlation coefficient.**

We represent this coefficient as $r$.


## Testing Pearson's correlation

<br>

What we often really want to know: how likely is it that we would get a correlation coefficient like this one if there were *no* association between income and test scores?

**Is the correlation of 0.216 stronger than we would expect if, in reality, the two variables were uncorrelated?**

<br>

We define our null hypothesis, H0: **The Pearson correlation coefficient $r$ is equal to 0 (that is, the variables are uncorrelated).**

$$H_0 : r = 0$$

We also define the alternative hypothesis, H1: **The Pearson correlation coefficient $r$ is not equal to zero (that is, the variables *are* correlated somehow).**


$$H_1 : r \neq 0$$

## Run the test

<br>

Instead of `cor()`, use `cor.test()`:

```{r}
cor.test(ses_data$Income, ses_data$Score)
```

<br>

:::{.dapr1callout .hcenter}

Can we reject the H0 that income and test score are uncorrelated?

<img style="margin-bottom:-10px; width: 2em;" src="figs/codicon--thumbsup-filled.svg"> <img style="margin-bottom:-20px; width: 2em;" src="figs/codicon--thumbsdown-filled.svg">

:::

## Reporting Pearson's correlation

<br>

> We found a weak positive correlation between a student's score on the SAT Verbal test and their parents' income (Pearson's $r$ = .22, $t$(755) = 6, $p$ < .001).
This result suggests that being from a higher-income family is associated with higher test scores, while being from a lower-income family is associated with lower test scores.

<br>

APA notation format for reporting the key inferential values:

Pearson's $r$ = [correlation coefficient], $t$([degrees of freedom]) = [t-statistic], $p$ = [p-value]

<br>

**However, before writing our report and calling it a day, we need to check the test's assumptions.**

# What does Pearson's correlation require / assume about the data?

## What does Pearson's correlation require / assume about the data?


<br>

::::{.columns}
:::{.column width="30%"}
**Data requirements:**
:::
:::{.column width="70%"}
Both variables are continuous.

Data contains no extreme outliers.
:::
::::


::::{.columns}
:::{.column width="30%"}
**Assumption of linearity:**
:::
:::{.column width="70%"}
The association between the variables is sufficiently linear.
:::
::::

::::{.columns}
:::{.column width="30%"}
**Assumption of independence:**
:::
:::{.column width="70%"}
Both variables are normally distributed.
:::
::::

::::{.columns}
:::{.column width="30%"}
**Assumption of independence:**
:::
:::{.column width="70%"}
Each variable's variance is equal across the range of the other variable.
:::
::::

<br>

**What if one or more of these things isn't true?**

- Then you shouldn't use Pearson's correlation to analyse your data.

- Instead, you can almost always use Spearman's correlation, which we'll talk about in a bit.

<br>

**How do we check each requirement/assumption?**




## Data requirements for Pearson's correlation (1)

:::hcenter
**Both variables are continuous.**
:::

<br>

Review your dataset:

```{r}
glimpse(ses_data)
```

<br>

:::{.dapr1callout .hcenter}

Are you satisfied that this requirement is met?

<img style="margin-bottom:-10px; width: 2em;" src="figs/codicon--thumbsup-filled.svg"> <img style="margin-bottom:-20px; width: 2em;" src="figs/codicon--thumbsdown-filled.svg">

:::


## Data requirements for Pearson's correlation (2)

:::hcenter
**Data contains no extreme outliers.**
:::

<br>

Plot the data and look for any data points that are far away from the others.

::::{.columns}
:::{.column width="50%"}
```{r fig.width=7, fig.height=6}
#| code-fold: true

ses_data |>
  ggplot(aes(x = Income, y = Score)) +
  geom_point(size = 3) +
  labs(
    x = "Parents' income (USD)",
    y = 'Score on SAT Verbal'
  )
```
:::
:::{.column width="50%"}

<br>
<br>
<br>
<br>
<br>
<br>


:::{.dapr1callout .hcenter}

Are you satisfied that this requirement is met?

<img style="margin-bottom:-10px; width: 2em;" src="figs/codicon--thumbsup-filled.svg"> <img style="margin-bottom:-20px; width: 2em;" src="figs/codicon--thumbsdown-filled.svg">

:::


:::
::::







## Assumption of linearity for Pearson's correlation

:::hcenter
**The association between the variables is sufficiently linear.**
:::

Plot the data with two `geom_smooth()`s, and you want them to closely match: 

- one with `method = 'lm'` (the straight line)
- one with `method = 'loess'` (LOcally Estimated Scatterplot Smoothing, a wobbly line)

::::{.columns}
:::{.column width="50%"}
```{r eval=F}
ses_data |>
  ggplot(aes(x = Income, y = Score)) +
  geom_point) +
  labs(
    x = "Parents' income (USD)",
    y = 'Score on SAT Verbal'
  ) +
  geom_smooth(
    method = 'lm', 
    se = FALSE
  ) +
  geom_smooth(
    method = 'loess', 
    se = FALSE, 
    colour = 'red'
  )
```

:::
:::{.column width="50%"}
```{r echo=F, fig.width=7, fig.height=6}
ses_data |>
  ggplot(aes(x = Income, y = Score)) +
  geom_point(size = 3) +
  labs(
    x = "Parents' income (USD)",
    y = 'Score on SAT Verbal'
  ) +
  geom_smooth(method = 'lm', se = FALSE, linewidth = 2) +
  geom_smooth(method = 'loess', se = FALSE, colour = 'red', linewidth = 2)
```
:::
::::



:::{.dapr1callout .hcenter}

Are you satisfied that this assumption is met?

<img style="margin-bottom:-10px; width: 2em;" src="figs/codicon--thumbsup-filled.svg"> <img style="margin-bottom:-20px; width: 2em;" src="figs/codicon--thumbsdown-filled.svg">

:::




## Assumption of normality for <br> Pearson's correlation (1): `Score`

:::hcenter
**Both variables are normally distributed.**
:::

First let's look at `Score`:

::::{.columns}
:::{.column width="50%"}
```{r fig.width=5, fig.height=5}
#| code-fold: true

ses_data |>
  ggplot(aes(sample = Score)) +
  geom_qq(size = 3) +
  geom_qq_line() +
  labs(
    x = "Theoretical quantiles",
    y = 'Sample quantiles'
  )
```

:::
:::{.column width="50%"}

<br>

```{r}
shapiro.test(ses_data$Score)
```
:::
::::

:::{.dapr1callout .hcenter}

Are you satisfied that this assumption is met for `Score`?

<img style="margin-bottom:-10px; width: 2em;" src="figs/codicon--thumbsup-filled.svg"> <img style="margin-bottom:-20px; width: 2em;" src="figs/codicon--thumbsdown-filled.svg">

:::


## Assumption of normality for <br> Pearson's correlation (2): `Income`

:::hcenter
**Both variables are normally distributed.**
:::

Now let's look at `Income`:

::::{.columns}
:::{.column width="50%"}
```{r fig.width=5, fig.height=5}
#| code-fold: true

ses_data |>
  ggplot(aes(sample = Income)) +
  geom_qq(size = 3) +
  geom_qq_line() +
  labs(
    x = "Theoretical quantiles",
    y = 'Sample quantiles'
  )
```

:::
:::{.column width="50%"}

<br>

```{r}
shapiro.test(ses_data$Income)
```
:::
::::

:::{.dapr1callout .hcenter}

Are you satisfied that this assumption is met for `Income`?

<img style="margin-bottom:-10px; width: 2em;" src="figs/codicon--thumbsup-filled.svg"> <img style="margin-bottom:-20px; width: 2em;" src="figs/codicon--thumbsdown-filled.svg">

:::


## Assumption of equal variance for <br> Pearson's correlation

:::hcenter
**Each variable's variance is equal across the range of the other variable.**
:::

Plot the data and look for a fairly uniform cloud of data points.

::::{.columns}
:::{.column width="50%"}
**Made-up data with very unequal variance:**

```{r fig.width=7, fig.height=5}
#| code-fold: true

n_obs <- 250
set.seed(1)
tibble(
  x = 1:n_obs,
  error = 1/2*x,
  y = rnorm(n = n_obs, mean = 0 + x, sd = error)
) |>
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 3)
```


:::
:::{.column width="50%"}

**Our `ses_data`:**

```{r fig.width=7, fig.height=5}
#| code-fold: true

ses_data |>
  ggplot(aes(x = Income, y = Score)) +
  geom_point(size = 3) +
  labs(
    x = "Parents' income (USD)",
    y = 'Score on SAT Verbal'
  )
```

:::
::::

:::{.dapr1callout .hcenter}

Are you satisfied that this assumption is met?

<img style="margin-bottom:-10px; width: 2em;" src="figs/codicon--thumbsup-filled.svg"> <img style="margin-bottom:-20px; width: 2em;" src="figs/codicon--thumbsdown-filled.svg">

:::


# Because assumptions <br> weren't all met: <br> Spearman's correlation

## What is Spearman's correlation?

A more flexible kind of correlation that's able to handle any variables that represent rankings or orderings.

<br>

::::{.columns}
:::{.column width="50%"}

**Pearson's correlation**

- can only be a **straight line.**
- assumes that `y` changes **by the same amount** for every increment of `x`.

```{r fig.width=5, fig.height=5}
#| code-fold: true
tibble(
  x = -3:3,
  y = x
) |>
  ggplot(aes(x=x, y=y)) +
  geom_point(size = 5) +
  geom_path(linewidth = 1.5) +
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank()
  )
```


:::
:::{.column width="50%"}
**Spearman's correlation**

- does **not** have to be a straight line.
- only cares about the **direction of change,** not the amount.

```{r fig.width=5, fig.height=5}
#| code-fold: true
set.seed(1)
tibble(
  x = -3:3,
  y = x + rnorm(7, mean = 0, sd = 0.5)
) |>
  ggplot(aes(x=x, y=y)) +
  geom_point(size = 5) +
  geom_path(linewidth = 1.5) +
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank()
  )
```

:::
::::

## Spearman's correlation example

RQ: **On RateMyProfessor.com, when students rate an instructor as more difficult, do they tend to also give that instructor lower star ratings?**

<br>

::::{.columns}
:::{.column width="45%"}

```{r include=F}
rmp_data <- read_csv('https://uoepsy.github.io/data/ratemyprof_stars.csv')

cor.test(
  rmp_data$student_difficult,
  rmp_data$student_star,
  method = 'spearman'
)

# raises warning bc of "tied ranks" but nbd
```

```{r}
rmp_data |> head(10)
```


[(data from He, 2020)](https://data.mendeley.com/datasets/fvtfjyvw7d/2){target="_blank"}
:::
:::{.column width="55%"}

```{r echo=F, fig.width = 7, fig.height = 5}
rmp_data |>
  mutate(student_star = factor(student_star)) |>
  ggplot(aes(x = student_difficult, fill = student_star)) +
  geom_bar()
```

:::
::::

You'll play with this data (and try to recreate this plot yourselves) in this week's workshop!

But for now: **Because Spearman's correlation is more flexible, it can also handle cases where Pearson's assumptions weren't met.**

## Testing our SES data with <br> Spearman's correlation

<br>

We saw that Pearson's correlation coefficient is represented by $r$.

**Spearman's correlation coefficient is represented by the Greek letter rho,  $\rho$.**

Otherwise, hypotheses are defined just the same way!

<br>

$$
\begin{align}
H_0 &: \rho = 0 \\
H_1 &: \rho \neq 0\\
\end{align}
$$

<!-- could woo here: put these hyps into words? -->

## Run the test in R

We can use `cor.test()` again.
We'll just change the type of correlation by adding `method = 'spearman'`:

```{r}
cor.test(
  ses_data$Income,
  ses_data$Score,
  method = 'spearman'
)
```


<br>

:::{.dapr1callout .hcenter}

Can we reject the H0 that Income and SAT score are uncorrelated?

<img style="margin-bottom:-10px; width: 2em;" src="figs/codicon--thumbsup-filled.svg"> <img style="margin-bottom:-20px; width: 2em;" src="figs/codicon--thumbsdown-filled.svg">

:::


## Reporting Spearman's correlation (and this whole analysis journey)

<br>

> We wanted to find out whether students from higher-income families tend to achieve higher test scores than students from lower-income families.
However, a Shapiro-Wilk test indicated that our income variable deviated significantly from a normal distribution ($W$ = 1, $p$ < .001).
We therefore used Spearman's $\rho$ instead of Pearson's $r$ to test the correlation between income and test score.
>
> We found a weak positive correlation (Spearman's $\rho$ = .2, $p$ < .001).
This result suggests that higher family income is significantly associated with higher test scores, and lower family income is associated with lower test scores.

<br>

APA notation format for reporting the key inferential values:

Spearman's $\rho$ = [correlation coefficient], p = [p-value]





# What does Spearman's correlation require / assume about the data?

## What does Spearman's correlation require / assume about the data?


<br>

::::{.columns}
:::{.column width="30%"}
**Data requirements:**
:::
:::{.column width="70%"}
Both variables represent rankings/orderings.
:::
::::


<br>

**What if this isn't true?**

- Then you shouldn't use Spearman's correlation to analyse your data.

<br>

**How do we check this requirement?**




## Data requirements for Spearman's correlation

:::hcenter
**Both variables represent rankings/orderings.**
:::

<br>

Review your dataset.

**Can each variable be sorted into a sensible order?**

<br>

::::{.columns}
:::{.column width="50%"}
```{r}
ses_data |> head(12)
```

:::
:::{.column width="50%"}

<br>
<br>
<br>
<br>


:::{.dapr1callout .hcenter}

Are you satisfied that this requirement is met?

<img style="margin-bottom:-10px; width: 2em;" src="figs/codicon--thumbsup-filled.svg"> <img style="margin-bottom:-20px; width: 2em;" src="figs/codicon--thumbsdown-filled.svg">

:::

:::
::::



# Back matter

## Revisiting this week's learning objectives

TODO


# Appendix {.appendix}
