---
title: "Lecture 20: Alternative forms"
subtitle: Data Analysis for Psychology in R 1
author: Tom Booth
---
```{r premable, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(kableExtra)
library(patchwork)
library(moderndive)
library(polycor)
library(psych)
```

## Today
- Briefly consider different forms of correlation and calculation in R

## Learning objectives
- Understand when to use different forms of correlation.
- Be able to run correct forms of correlation in R.

## Types of correlation

| X           | Y           | Correlation Type |
|-------------|-------------|------------------|
| Continuous  | Continuous  | Pearson          |
| Continuous  | Categorical | Polyserial       |
| Continuous  | Binary      | Biserial         |
| Categorical | Categorical | Polychoric       |
| Binary      | Binary      | Tetrachoric      |
| Rank        | Rank        | Spearman         |
| Nominal     | Nominal     | Chi-square       |

## Spearman correlation
- Spearman's $\rho$ (or rank-order correlation) is a *non-parametric* estimate of the association between two variables.
  - As name suggests, it uses data on the rank-ordering of X,Y responses for each individual.
- When would we choose to use the Spearman correlation?
	- If our data are naturally ranked data (e.g. imaging an experiment where the task is to rank foods and drinks in terms of preference).
	- If the data are non-normal or skewed.
	- If the data shows evidence of non-linearity.

## Spearman correlation
- Spearman's is not testing for linear relations, it is testing for increasing monotonic relationship.
  - Huh?
  
## Linear vs. monotonic
```{r, echo=FALSE}
mono <- tibble(
  A = c(1,2,3,4,5,6,7,8,9,10),
  B = c(1,2,3,4,5,6,7,8,9,10),
  C = c(1,4,5,6,8,9,10,13,15,16)
)

p1 <- mono %>%
  ggplot(., aes(x=A, y=B)) +
  geom_point() + 
  geom_line() +
  ggtitle("Linear")

p2 <- mono %>%
  ggplot(., aes(x=A, y=C)) +
  geom_point() + 
  geom_line() +
  ggtitle("Increasing Monotonic")

p1 + p2

```

## Steps in Spearman's
$$
\rho = 1 - \frac{6\Sigma{d^2_i}}{n(n^2-1)}
$$

- Calculation steps:
	- Rank each variable from largest to smallest.
	- If there are ties in ranks, assign the average of the rankings to each case.
	- Calculate the difference in rank for each person on the two variables.
	- Tabulate data
	- Square the difference.
	- Sum the squared values.
	
## Quick example

```{r}
rank <- tibble(
  ID = paste("ID", 1:6, sep = ""),
  RT =c(.264, .311, .265, .291, .350, .500),
  Caff = c(210,280,150,90,200,450)
)
rank
```

## Calculation
```{r}
rank_calc <- rank %>%
  mutate(
    RT_rank = rank(RT),
    Caff_rank = rank(Caff),
    di = RT_rank - Caff_rank,
    di2 = di^2
  )
rank_calc
```

## Calculation
```{r}
rank_calc
```

$$
\rho = 1 - \frac{6\Sigma{d^2_i}}{n(n^2-1)} = 1 - \frac{6*18}{6(6^2-1)} = 1 - \frac{108}{210} = 1 - 0.514 = 0.486
$$

## In R
```{r}
cor(rank$RT, rank$Caff, method = "spearman")
```

## Other forms
- General principle (simplified a little) of the other forms of correlation is roughly the same.
- We assume that the categorical variable is a crude measurement of an underlying normal variable.
- Aiming to provide an estimate of the association between these underlying variables.

## Underlying continuous variable

```{r, echo=FALSE}
dat <- tibble(
  x = factor(sample(c("Disagree", "Neither Agree/Disagree", "Agree"), 
                       500, replace = T, prob = c(0.3, 0.5, 0.2)), 
             levels = c("Disagree", "Neither Agree/Disagree", "Agree")
             )
  )
p3 <- dat %>% 
  ggplot(., aes(x=x)) +
  geom_bar() 
p4 <- ggplot(NULL, aes(x = c(-5,5))) +
  stat_function(fun=dnorm,
                geom = "line") +
  theme_void()

p3+p4
```

## In R
- Estimating correlation is straight forward.
- All we need to do is make sure R knows the type of data we have, then use `hetcor`

```{r}
pers_items <- bfi[,c(1:5)]
pers_items <- pers_items %>%
  mutate(
    A1 = as_factor(A1)
  )
pers_cors <- hetcor(pers_items)
```

## In R

```{r}
pers_cors$correlations
```


```{r}
pers_cors$type
```


## Tasks for this week...
- Keep safe and healthy.

