---
title: "Probability Theory"
subtitle: "Semester 1 - Week 7"
callout-appearance: simple
---


```{r setup}
#| include: false

source('assets/setup.R')
library(tidyverse)
movies <- read_csv("https://uoepsy.github.io/data/hollywood_movies_subset.csv")
```


## Formative report B

[Download the template Rmd file and start your work there](https://uoepsy.github.io/dapr1/2425/useful_files/template_formative_report_b.Rmd). Save your work regularly by clicking File > Save.

::: {.callout-important collapse="true"}
### Instructions - Don't Skip! {-}

Formative Report B covers the labs from weeks 7-11 of the DAPR1 course. You’ll need to create a PDF report using RMarkdown, which will be submitted by 12 noon on Friday, 29th November 2024. Remember, these submissions are group-based, so there are no extensions. Expect written formative feedback in week 12.

Your report should be tailored for a reader with basic statistical knowledge and should not include any references to R code or functions in the main report write-up. Instead, keep the main report focused on text, figures, and tables. All R code should be included in the compulsory Appendix B for reproducibility, which is automatically created for you in the template Rmd file. 
If you need to add extra tables or figures that don’t fit in the main part of the report, you can use an optional Appendix A. Remember, the main report should be a PDF file and should not exceed six sides of A4 paper, though appendices at the end don’t count towards this limit.

Ensure to use the default settings for font and page margins in your RMarkdown file. Also, make sure your report title includes your group name: Group NAME.LETTER, and list the exam numbers of all group members in the author section.

Don't forget to register for your lab table group on LEARN by navigating to the course LEARN page, clicking on Groups, selecting Labs_1_2_3, finding your group, and clicking Join. 

:::

::: {.callout-tip collapse="true"}
### Formatting resources {-}

At [this page](zf_formatting_resources.html){target="_blank"} you can find resources to help you with your report formatting.

:::


### Data

**Hollywood Movies**. At the link <https://uoepsy.github.io/data/hollywood_movies_subset.csv> you will find data on Hollywood movies released between 2012 and 2018 from the top 5 lead studios and top 10 genres. Among all variables, the following will be used for this report:

<!-- - `Movie`: Title of the movie  -->
- `LeadStudio`: Primary U.S. distributor of the movie
<!-- - `RottenTomatoes`: Rotten Tomatoes rating (critics)  -->
- `AudienceScore`: Audience rating (via Rotten Tomatoes) 
- `Genre`: One of Action Adventure, Black Comedy, Comedy, Concert, Documentary, Drama, Horror, Musical, Romantic Comedy, Thriller, or Western 
<!-- - `TheatersOpenWeek`: Number of screens for opening weekend  -->
<!-- - `OpeningWeekend`: Opening weekend gross (in millions)  -->
<!-- - `BOAvgOpenWeekend`: Average box office income per theater, opening weekend  -->
<!-- - `Budget`: Production budget (in millions)  -->
<!-- - `DomesticGross`: Gross income for domestic (U.S.) viewers (in millions)  -->
<!-- - `WorldGross`: Gross income for all viewers (in millions)  -->
<!-- - `ForeignGross`: Gross income for foreign viewers (in millions)  -->
<!-- - `Profitability`: WorldGross as a percentage of Budget  -->
<!-- - `OpenProfit`: Percentage of budget recovered on opening weekend  -->
<!-- - `Year`: Year the movie was released -->
- `IQ1`-`IQ50`: IQ score of each of 50 audience raters (every movie had different raters)
- `Snacks`: How many of the 50 audience raters bought snacks
- `PrivateTransport`: How many of the 50 audience raters reached the cinema via private transportation



### Tasks

For formative report B, you will be asked to perform the following tasks, each related to a week of teaching in this course.  
This week's task is highlighted in bold below. Please only focus on completing that task this week. In the next section, you will also find the guided sub-steps that you need to consider to complete this week's task.

> **B1) Create a new categorical variable, Rating, taking the value 'Good' if the audience score is > 50, and 'Bad' otherwise. Inspect and describe the joint probability distribution of movie genre and rating.**  
> B2) Investigate if a movie receiving a good rating is independent of the genre.   
> B3) Computing and plotting probabilities with a binomial distribution.  
> B4) Computing and plotting probabilities with a normal distribution.  
> B5) Plot standard error of the mean, and finish the report write-up (i.e., knit to PDF, and submit the PDF for formative feedback).  


### B1 sub-tasks

__This week you will only focus on task B1.__ Below there are sub-steps you need to consider to complete task B1.

:::{.column-margin}
:::{.callout-tip}
#### Tip

To see the hints, hover your cursor on the superscript numbers.
:::
:::

- Read the Hollywood movies data into R and give it a helpful name. Data link: <https://uoepsy.github.io/data/hollywood_movies_subset.csv>

- Find the 3 most frequent movie genres. Next, filter the data to only keep movies from the 3 most frequent genres. This is the dataset you will be working with for the rest of the report.[^top3-filter]

[^top3-filter]: Hint: use `sort(table(DATA$COLUMN))` to find top 3 movie genres, and `filter()` to select the specific 3 genres. 

    Example: For the `starwars` dataset, we can filter to include only the two most frequent species, Humans and Droids, via the following code:
             
    ```{r, results='hide'}
    # sort the frequency table
    sort(table(starwars$species))
    
    # or, for decreasing order, you can use:
    sort(table(starwars$species), decreasing = TRUE)
    
    # keep only the rows where species is Human or Droid
    # in R, or is the vertical bar |
    starwars2 <- starwars |> 
        filter(species == "Human" | species ==  "Droid")
    ```

- Create a variable called Rating where the AudienceScore variable is recoded so that the movies scoring less than or equal to 50 are coded as "Bad" and those scoring over 50 (i.e., >50) are "Good".[^func-if-else]

[^func-if-else]: Hint: use `ifelse()` and `mutate()` functions to create a new 'Rating' variable capturing whether movies are good or bad. 
    
    Example: For the `starwars` dataset, we can create a variable to capture whether characters were short (<180cm) or tall (>180cm) based on their recorded height using the following code:

    ```{r}
    starwars2 <- starwars2 |>
        mutate(size = ifelse(height < 180, 's', 't'))
    ```
    
    In the code above, `mutate()` changes to data to include
    a column named `size`, computed with the computation that follows the equal sign.  
    The `ifelse` function checks a condition (is `height < 180`?) and uses the value `'s'` if the condition is TRUE, and `'t'` if the condition is FALSE.

- Ensure Rating and Genre are coded as factors.[^func-factor]

[^func-factor]: Hint: use the `factor()` function.

    Example: For the starwars data, if I wanted to give better labels for my size variable, I could use the following code. Since the names are fine as is for species, I will just specify `factor()`.

    ```{r}
    starwars2$size <- factor(starwars2$size,
                             levels = c("s", "t"),
                             labels = c("short", "tall"))
    
    starwars2$species <- factor(starwars2$species)
    ```


- Create a contingency table of Rating by Genre.[^func-table]

[^func-table]: Hint: use the `table()` function, passing to it the two columns that you want to be included in your contingency table.

    Example: For the `starwars` dataset, if I wanted a contingency table displaying how many humans and droids were short and tall, I could use the following code: 
               
    ```{r}
    swars_freq_tbl <- table(starwars2$species, starwars2$size)
    swars_freq_tbl
    ```
             
- Transform the table of counts to a relative frequency table.[^func-prop-tbl]

[^func-prop-tbl]: Hint: use the `prop.table()` function. You will want to pass your contingency table created above to this. 

    Example: For the `starwars` dataset, if I wanted a relative frequency table, I could use the following code: 

    ```{r}
    swars_rel_freq_tbl <- swars_freq_tbl |>
        prop.table()
    swars_rel_freq_tbl
    ```

- Do the numbers in the table satisfy the requirements of probabilities?[^prob-01]

[^prob-01]: Hint: Recall the requirements of probabilities - are the proportions >= 0 or <= 1?; and that the values in the relative frequency table should sum to 1.  The `all()` and `sum()` functions could be useful here. 

    Example: For the `starwars` dataset, I could check these requirements using the following code: 

    ```{r}
    all(swars_rel_freq_tbl >= 0)
    all(swars_rel_freq_tbl <= 1)
    sum(swars_rel_freq_tbl)
    ```

- Instead of checking the probability requirements manually, add row and column totals to the relative frequency table.[^func-addmargins]

[^func-addmargins]: Hint: Use the `addmargins()` function.

    Example: For the `starwars` dataset, I could add this using the following code:

    ```{r}
    swars_rel_freq_sum <- swars_freq_tbl |>
        prop.table() |>
        addmargins()
    swars_rel_freq_sum
    ```

- Visualise the relative frequency table as a mosaic plot, making sure to add a main title and clear axis titles.[^func-mosaic]

[^func-mosaic]: Hint: Make sure to load the `ggmosaic()` package so that you can specify `geom_mosaic()` when building your plot. To add a title, as well as x- and y-axis titles, specify `labs(title = , x = , y = )`.

    Example: For the `starwars` dataset, I create a mosaic plot using the following code:

    ```{r}
    #| label: fig-swars1
    #| fig-cap: "Starwars Mosaic Plot Example Title"
    library(ggmosaic)
    m_plot <- ggplot(starwars2) +
        geom_mosaic(aes(x = product(species, size), fill = species)) +
        labs(title = "Starwars Mosaic Plot Example Title", 
             x = "Size", 
             y = "Species",
             fill = "Species")
    m_plot
    ```

- Does your plot have any NAs? If yes, you can drop the NAs before re-plotting, see hint. If there are no NAs, no action is needed but read over the hint anyway.[^func-drop] 

[^func-drop]: Hint: Here you could make sense of the `drop_na()` function.

    Example: For the `starwars` dataset, I do have `NA`s in my 'size' variable, so I need to remove these before re-plotting:

    ```{r}
    starwars2 <- starwars2 |>
        drop_na(size) |>
        mutate(size = factor(size))
    ```

    ```{r}
    #| label: fig-swars2
    #| fig-cap: "Starwars Mosaic Plot Example Title - No NAs"
    library(ggmosaic)
    m_plot <- ggplot(starwars2) +
        geom_mosaic(aes(x = product(species, size), fill = species)) +
        labs(title = "Starwars Mosaic Plot Example Title", 
             x = "Size", 
             y = "Species",
             fill = "Species")
    m_plot
    ```

- In the introduction section of your report, write up a small introduction to the data and variables.

    - What are the data that you are analysing and where can these be found?
    - Which variables are you using and what do they represent?
    - What is the type of these variables?
    - Are there any missing values in these variables?
    - In the next weeks, update the introduction based on the new variables that you will consider.

- In the analysis section of your report, write up a summary of what you have reported above, using proper rounding to 2 decimal places and avoiding any reference to R code or functions.


## Worked Example

Consider the dataset available at <https://uoepsy.github.io/data/RestaurantTips.csv>, containing 157 observations on the following 7 variables:

```{r, echo=FALSE}
library(tidyverse)
library(kableExtra)

tribble(
    ~'Variable Name', ~'Description',
    'Bill', 'Size of the bill (in dollars)',
    'Tip', 'Size of the tip (in dollars)',
    'Credit', 'Paid with a credit card? n or y',
    'Guests', 'Number of people in the group',
    'Day', 'Day of the week: m=Monday, t=Tuesday, w=Wednesday, th=Thursday, or f=Friday',
    'Server', 'Code for specific waiter/waitress: A, B, or C',
    'PctTip', 'Tip as a percentage of the bill'
) |>
    kbl() |>
    kable_styling(full_width = FALSE)
```

These data were collected by the owner of a bistro in the US, who was interested in understanding the tipping patterns of their customers. The data are adapted from @lock2020.

```{r}
library(tidyverse)  # we use read_csv and glimpse from tidyverse
tips <- read_csv("https://uoepsy.github.io/data/RestaurantTips.csv")
head(tips)
glimpse(tips)
```

We can filter our data to only include rows of data from Servers A and B, and save this filtered data to a new dataset called "tips2". 

Because we want to include servers A and B, but **not** C, we can use the `!=` (or does not equal) operator. 

:::{.column-margin}

Some common operators include: 

| Operator        | Description |
|:----------------|:--------------|
| <               | less than |
| >               | greater than |
| <=              | less than or equal to |
| >=              | greater than or equal to |
| ==              | equal to |
| !=              | not equal to |
| LEFT %in% RIGHT | is LEFT a member of RIGHT? |

:::

```{r}
tips2 <- tips |> 
    filter(Server != "C")
```

Consider if, for example, we wanted to recode the Tip variable so that those tipping less than or equal to 15% were coded as "Below" average, and those tipping over 15% were coded as "Above" average (15% being used as below/above average tips cut-off in relation to usual US tipping rates). To do so, we could create a new variable called "Tip_Avg".

```{r}
tips2 <- tips2 |>
    mutate(Tip_Avg = ifelse(PctTip <= 15, 'Below', 'Above'))

table(tips2$Tip_Avg)
```

Now that we have the variables we want, it would be a good point to make these both factors:

```{r}
tips2$Server <- factor(tips2$Server)
tips2$Tip_Avg <- factor(tips2$Tip_Avg)
```

To visually represent the distribution of how many customers were served by each server and if they left a below or above average tip, we could Create a contingency table:

```{r}
freq_tbl <- table(tips2$Server, tips2$Tip_Avg)
freq_tbl
```

We could then transform the table of counts above to instead represent a relative frequency table: 

```{r}
rel_freq_tbl <- freq_tbl |>
    prop.table()
rel_freq_tbl
```

Before we interpret our results, we must ensure that the numbers above satisfy the requirements of probabilities. We can do this two ways:

::: {.panel-tabset}

#### Manually

Check that all values in the proportions table are greater than or equal 0, all values are less than or equal to 1, and they all sum to 1:

```{r}
all(rel_freq_tbl >= 0)
```

```{r}
all(rel_freq_tbl <= 1)
```

```{r}
sum(rel_freq_tbl)
```


#### Using `addmargins()`

Instead of checking manually, we can use the function `addmargins()` to check that the probabilities sum to 1:

```{r}
rel_freq_tbl <- freq_tbl |>
    prop.table() |>
    addmargins()
rel_freq_tbl
```

:::

In order to visualise our results in a figure, we could use a mosaic plot:

```{r}
#| label: fig-mos
#| fig-cap: "Association between Servers and Tips"
library(ggmosaic)
mos_plot <- ggplot(tips2) +
    geom_mosaic(aes(x = product(Server, Tip_Avg), fill = Server)) +  
    labs(title = "Association between Servers and Tips", x = "Tip", y = "Server")
mos_plot
```
 
:::{.callout-tip}
#### Example writeup

More customers tipped above (64%) than below (36%) average. Both Server A and Server B received an equal distribution of tips above average (32%), but server B had a higher proportion of tips below average (20%) in comparison to server A (16%). These associations are visually represented in @fig-mos.

:::


## Student Glossary

To conclude the lab, add the new functions to the glossary of `R` functions. 

| Function   | Use and package |
|:-----------|:--------------|
| `filter` | ? |
| `mutate` | ? |
| `drop_na` | ? |
| `factor` | ? |
| `table` | ? |
| `ifelse` | ? |
| `prop.table` | ? |
| `addmargins` | ? |
| `all` | ? |
| `|` | ? |
| `geom_mosaic` | ? |

