---
title: "Random Variables (Discrete)"
subtitle: "Semester 1 - Week 9"
callout-appearance: simple
editor_options: 
  chunk_output_type: console
---

```{r setup}
#| include: false

set.seed(532)
source('assets/setup.R')
library(tidyverse)

movies <- read_csv("https://uoepsy.github.io/data/hollywood_movies_subset.csv")

movies2 <- movies %>%
    filter(Genre %in% c("Drama", "Action", "Comedy")) %>%
    mutate(Rating = ifelse(AudienceScore <= 50, 'Bad', 'Good')) %>%
    mutate(Genre = factor(Genre),
           Rating = factor(Rating))

movies_prop <- table(movies2$Genre, movies2$Rating) %>%
    prop.table() %>%
    addmargins()
movies_prop

movies_prop[, 2] / movies_prop[, 3]

library(ggmosaic)
m_plot <- ggplot(movies2) +
    geom_mosaic(aes(x = product(Rating), fill = Rating, conds = product(Genre)))
m_plot
```


## Formative Report B

::: {.callout-important}
Instructions and data were released in [week 7](1_07_prob_theory.html).
:::


### Tasks

For formative report B, you will be asked to perform the following tasks, each related to a week of teaching in this course.  
This week's task is highlighted in bold below. Please only focus on completing that task this week. In the next section, you will also find the guided sub-steps that you need to consider to complete this week's task.

> B1) Create a new categorical variable, Rating, taking the value 'Good' if the audience score is > 50, and 'Bad' otherwise. Inspect and describe the joint probability distribution of movie genre and rating.  
> B2) Investigate if a movie receiving a good rating is independent of the genre.  
> **B3) Computing and plotting probabilities with a binomial distribution.**  
> B4) Computing and plotting probabilities with a normal distribution.  
> B5) Plot standard error of the mean, and finish the report write-up (i.e., knit to PDF, and submit the PDF for formative feedback).  


## B3 sub-tasks

__This week you will only focus on task B3.__ Below there are sub-steps you need to consider to complete task B3.

:::{.column-margin}
:::{.callout-tip}
#### Tip

To see the hints, hover your cursor on the superscript numbers.
:::
:::

::: {.callout-important}
#### Context
A new movie theatre is opening in Texas, and the management team are reviewing the requirements for snack stalls and their car parking capacity. They are interested in determining whether to have only one or multiple snack stalls selling popular food and drink items; and are considering whether to expand their car park size or introduce a bus stop on cinema grounds.

In this lab, you will need to consider both the `Snacks` and `Private Transport` variables from the Hollywood movies dataset when answering the questions below. 
:::

- Reopen last week's Rmd file, and continue building on last week's work. Make sure you are still using the movies dataset filtered to only include the top 3 genres.[^share-file]

[^share-file]: Hint: Ask last week's driver for the Rmd file, they should share it with the group via the group discussion space or email. To download the file from the server, go to the RStudio Files pane, tick the box next to the Rmd file, and select More > Export.

- Consider the `Snacks` and `Private Transport` variables, are they discrete or continuous? 

- Plot separately the sample frequency distribution of both the `Snacks` and `Private Transport`variables with barplots.[^plot-freq]

[^plot-freq]: Hint: 

    First, compute the frequency distribution of each variable, using `count()` and `mutate()`. 
    
    Next, plot each frequency distribution as a barplot. Since you already have the bar heights, use `geom_col()` to create the bars. 
    
    **Warning**: The function `geom_bar()`, instead, does the counting for you and requires a data set instead of a frequency table.

- What kind of distribution do these follow? Estimate the parameters of your two distributions from the sample data.[^est-params]

[^est-params]: Hint: A Binomial distribution has two parameters, the _number of trials_ or _size_ ($n$) and the _probability of success_ ($p$).

    - To estimate the _number of trials_, $n$, use the study design information that you can find in the data description.

    - To estimate the _probability of success_, $p$, use the following fact. The expected value $E(X)$ of a Binomial distribution (i.e., the mean) is $E(X) = n * p$ where $n$ = number of trials, and $p$ is the probability of success. From this, we have that $p = E(X) / n$.
    
    In the movies data, the number of trials (size) is $n = 50$. This is because the variables `Snacks` and `PrivateTransport` record how many, out of the 50 audience movie raters, bought snacks or travelled to the cinema by private transportation, respectively.

- Plot the fitted Binomial distribution on top of the sample frequency distribution for each variable. Is the Binomial distribution a good model for each variable?[^plt-binom]

[^plt-binom]: First, create a new tibble with two columns: (1) the possible values of the Binomial distribution, and (2) the probabilities of each outcome.

    Next, take the barplots previously created and use `geom_point(data = ?, aes(x = ?, y = ?))` to add the Binomial probabilities on top of each barplot with points.
    
    If the plot looks weird, are your bar heights probabilities or counts? The Binomial probabilities are, as the words says, probabilities so make sure you plot two relatable measures.


<!-- - What is the probability that exactly half of the audience raters for each movie (i.e., 25 viewers) bought snacks? What is the probability that exactly half traveled via private transport?[^half-dbinom] -->

<!-- [^half-dbinom]: Hint: Here we need to use the function `dbinom()`, as we are interested in the probability of getting one specific result. -->

<!--     The function `dbinom()` will require you to state three arguments: (1) x - the value(s) you want to compute the probability for, (2) size - the number of trials, and (3) prob - the probability of success. -->

<!--     ``` -->
<!--     dbinom(x = ?, size = ?, prob = ?) -->
<!--     ``` -->

- What is the probability that **more than** half of the audience for each movie (i.e., >25 viewers) bought snacks? What is the probability that **less than** half of the audience for each movie (i.e., <25 viewers) traveled via private transportation?[^snack-pbinom]

[^snack-pbinom]: Hint: Here we need to use the function `pbinom()`, as we are interested in the cumulative probability of getting a specific result:
    
    ```
    pbinom(q = ?, size = ?, prob = ?, lower.tail = ?)
    ```
    
    The function `pbinom()` will require you to state four arguments:
    
    1. `q`: the number of successful outcomes
    2. `size`: the number of trials (i.e., the Binomial parameter $n$)
    3. `prob`: the probability of success (i.e., the Binomial parameter $p$)
    4. `lower.tail`: if TRUE probabilities are $P(X \leq q)$, if FALSE probabilities are $P(X > q)$. This is TRUE by default, change to FALSE if you require $P(X > q)$.


- Based on the probabilities you have reported above, do you think that the new movie theatre should (1) invest in multiple snack stations, or just one; and (2) increase car parking capacity or add a bus stop? Justify your answer. 

- A [recent survey](https://web.archive.org/web/20230325235010/http://fj.webedia.us/features/snack-statistics-surveyme-findings-reveal-moviegoers-concession-preferences) suggested that 49% of movie viewers always buy some form of snack item (i.e., popcorn, drinks, sweets) when watching a movie, and another suggested that 70% of movie viewers travel to the cinema via private transport. Are your two probability estimates consistent with the survey-reported values?[^value-consistent]

[^value-consistent]: Hint: Compare your estimated probability of buying snacks to that reported in the above survey (which suggested that 49% of movie viewers always buy some form of snack item (i.e., popcorn, drinks, sweets) when watching a movie); and compare your estimated probability of travelling via private transportation to that reported in the above survey (which suggested that 70% of movie viewers viewers travel to the cinema via private transport). 

- In the analysis section of your report, write up a summary of what you have reported above, using proper rounding to 2 decimal places and avoiding any reference to R code or functions. 


## Worked Example

The dataset available at <https://uoepsy.github.io/data/RestaurantTips2.csv> was collected by the owner of a US bistro, and contains 99 observations on 10 variables. 
It is a subset of the RestaurantTips.csv data presented in the past weeks, _focusing only on parties of 2 people_.^[Data adapted from @lock2020.]

The bistro owner is interested in coffee sales, and whether they should consider introducing a 2 for 1 coffee deal to entice customers to purchase one of their Christmas coffees. Another option they are considering is starting a loyalty scheme for customers to be rewarded for every coffee purchase. Your job is to advise them on which scheme they should run to benefit most customers. 

For context, Americans drink *a lot* of coffee, but slightly less than Norwegians (89.4% of Norwegians drink at least one coffee per day!^[Tverdal, A., Hjellvik, V. & Selmer, R. Coffee intake and oral–oesophageal cancer: follow-up of 389,624 Norwegian men and women 40–45 years. Br J Cancer 105, 157–161 (2011).]). We are interested in estimating the probability that an adult American will drink coffee. 

```{r, echo=FALSE}
library(tidyverse)
library(kableExtra)

tribble(
    ~'Variable Name', ~'Description',
    'Bill', 'Size of the bill (in dollars)',
    'Tip', 'Size of the tip (in dollars)',
    'Credit', 'Paid with a credit card? n or y',
    'Guests', 'Number of people in the group',
    'Day', 'Day of the week: m=Monday, t=Tuesday, w=Wednesday, th=Thursday, or f=Friday',
    'Server', 'Code for specific waiter/waitress: A, B, or C',
    'PctTip', 'Tip as a percentage of the bill',
    'HadCoffee', 'Number of guests in the group who had coffee',
    'IQ1', 'Score on IQ test for guest 1',
    'IQ2', 'Score on IQ test for guest 2'
) %>%
    kable(format = 'simple')
```

First, let's read the new dataset into R:

```{r}
library(tidyverse)
tips2 <- read_csv("https://uoepsy.github.io/data/RestaurantTips2.csv")
head(tips2)
```

As mentioned in the data description, the variable `HadCoffee` records how many, out of the 2 people in each party, had coffee. Clearly this can only take the values 0, 1, or 2 if none, only one, or both of the people in the party ordered coffee.

- If we were asked to describe what kind of variable `HadCoffee` is, and to comment on the kind of probability distribution it may follow, we could say:

The number of coffee-consuming guests out of parties of size 2 is a discrete random variable that could be modeled by a Binomial probability distribution.

- We can plot the frequency distribution of the `HadCoffee` variable in the sample as following:

Compute the frequency distribution and relative frequency:

```{r}
# freq_distr data contains the sample frequencies
freq_distr <- tips2 %>%
    count(HadCoffee) %>%
    mutate(rel_freq = n / sum(n)) %>%
    rename(freq = n)    # the code rename(new_name = current_name)
                        # renames column 'n' created by count() to 'freq'
freq_distr
```

Plot the frequency distribution of the discrete variable `HadCoffee` using a barplot.
As we already have the x values (`HadCoffee`) and the y values (the height of the bars, `rel_freq`), we use `geom_col()`.

::: column-margin
`geom_col()` is used to create a barplot when you have already computed the bar heights, i.e. the frequency table. This draws columns at `x` of height given by `y` as specified inside `aes(x = ?, y = ?)`  
<br>
`geom_bar()` takes the original data, and does the counting for you. Because of this, it only requires you to specify `x` into `aes(x = ?)`. The height is computed for you internally by counting the unique values in `x`.
:::

```{r}
ggplot(freq_distr, aes(x = HadCoffee, y = rel_freq)) +
    geom_col() +
    labs(x = "Coffee-drinking customers per party",
         y = "Relative frequency")
```


:::{.callout-tip collapse="true"}
### Fitting a distribution

Fitting a Binomial distribution to data involves estimating the parameters of the distribution from the data. In other words, we want to find values for $n$ and $p$ from the variable `HadCoffee` in the our data.
:::


- We can now fit a Binomial distribution to the variable. To do so, we need to start by estimating the parameters of the Binomial distribution:

    + $n$, the number of trials (or size)
    + $p$, the probability of success

For this dataset, $n$ represents the size of each party, i.e. $n = 2$. The discrete variable `HadCoffee` represents how many guests had coffee, out of the 2 possible guests per party. 

::: column-margin
In a Binomial distribution, the number of trials ($n$) should not be confused with the __sample size__. The sample size would be the total number of parties in the dataset, i.e. `r nrow(tips2)`. It's just unfortunate that both use the same symbol $n$, but which one is the correct one should be clear from the context.
:::

The event "had coffee" represents our "success", and $p$ denotes the probability of success. In other words, $p$ represents the probability of an individual having coffee.

:::{.callout-tip collapse="true"}
### Estimating the probability of success

The expected value $E(X)$ of a Binomial distribution (i.e., the mean) is $E(X) = n * p$ where $n$ = number of trials = size, and $p$ is the probability of success. From this, we have that $p = E(X) / n$.  
We typically denote the estimated probability of success from the sample data with a hat on top, $\hat{p}$, written in text as `$\hat{p}$`.
:::

```{r}
# The mean of the discrete random variable
EX <- mean(tips2$HadCoffee)  

# Each party size (n in the formula)
size <- 2

# Estimated probability of having coffee: E(X) / n
p_hat <- EX / size

#check value
p_hat
```


- We can then compare the sample frequency distribution to the Binomial distribution, and comment on whether the Binomial fit is good:

We can create a new tibble having two columns. The first has the possible values of the Binomial distribution (0, 1, or 2 guests ordering coffee out of the 2). The second column has the theoretical probabilities, for each of the possible values, predicted by the Binomial distribution with parameters $n = 2$ and $p = 0.76$.

```{r}
# binom_distr data contains the binomial probabilities
binom_distr <- tibble(
    HadCoffee = 0:2,
    binom_prob = dbinom(x = HadCoffee, size = 2, prob = p_hat)
)
binom_distr
```

We can plot the sample frequency distribution as a bar plot and put on top the fitted Binomial probability distribution as dots (see Option 1), and even add segments to show the Binomial distribution (see Option 2).

In both cases, you need to specify different data for each `geom_`. The bars will use the sample frequency distribution computed using `count()` before. The points will use the Binomial probabilities computed using `dbinom()`.

::: {.panel-tabset}

## Option 1

```{r}
ggplot() +
    geom_col(data = freq_distr, aes(x = HadCoffee, y = rel_freq)) + # uses freq_distr data
    geom_point(data = binom_distr,                                  # uses binom_distr data
               aes(x = HadCoffee, y = binom_prob),  
               colour = 'red', size = 3)
```

## Option 2

```{r}
#| label: fig-coffee
#| fig-cap: "Probability of guests drinking coffee."
ggplot() +
    geom_col(data = freq_distr, aes(x = HadCoffee, y = rel_freq)) +
    geom_point(data = binom_distr, 
               aes(x = HadCoffee, y = binom_prob),  # adds points
               colour = 'red', size = 3) +
    geom_segment(data = binom_distr, 
                 aes(x = HadCoffee, y = binom_prob, # adds line (optional) 
                     xend = HadCoffee, yend = 0),   # from (x,y) to (xend,yend)
                 colour = 'red')
```

:::

The Binomial distribution seems to be a good fit for the sample distribution, as the probabilities tend to agree. Among all parties of two guests, the highest probability is that both guests had coffee, and no one having coffee has the lowest probability.

:::{.callout-tip collapse="true"}
### Probability Mass Function

- $P(X = x)$ = `dbinom(x, size, prob)`

The probability mass function computes $P(X = x)$ for a Binomial distribution where number of trials is `size` and probability of success is `prob`.
:::

- To calculate the probability that 1 person in the party orders coffee, we can compute the following:

The probability P(X = 1) for a Binomial with $x = 1$, $size = 2$, and $p = 0.76$ is:

```{r}
dbinom(1, size = 2, prob = p_hat)
```

- To calculate the probability that 2 people in the party order coffee,  we can compute the following:

The probability P(X = 2) for a Binomial with $x = 2$, $size = 2$, and $p = 0.76$ is:

```{r}
dbinom(2, size = 2, prob = p_hat)
```

:::{.callout-tip collapse="true"}
### Cumulative Distribution Function

- $P(X \leq q)$ = `pbinom(q, size, prob)`

The cumulative distribution function $P(X \leq q)$ gives the probability of having less than or equal to $q$ successes. 

From the total probability rule, the probability of a value being greater than $q$ is computed as $P(X > q) = 1 - P(X \leq q)$:

```
# option 1: P(X > q) = 1 - P(X <= q)
1 - pbinom(q, size, prob)
```

```
# option 2: P(X > q) directly
pbinom(q, size, prob, lower.tail = FALSE)
```
:::

- To calculate the probability that 1 person or less orders coffee,  we can compute the following:

We do this with the `pbinom(q, size, prob)` function, with `q = 1` to have $P(X \leq 1) = P(X = 0) + P(X = 1)$:

```{r}
pbinom(1, size = 2, prob = p_hat)
```

You can also see this by creating a new column in the binomial distribution which has the cumulative sums of the probabilities, i.e. the values P(X = 0), followed by P(X = 0) + P(X = 1), and finally by P(X = 0) + P(X = 1) + P(X = 2):

```{r}
binom_distr %>% 
    mutate(
        cumul_prob = cumsum(binom_prob)
    )
```

As you can see, P(X = 0) + P(X = 1) is 0.42 in the second row, which agrees with the result computed using `pbinom` above.

- To calculate the probability that at least one person from the group orders coffee, we can compute one of the below four equivalent calucaltions:

::: {.panel-tabset}

## 1 - P(X <= 0)

```{r}
1 - pbinom(0, size = 2, prob = p_hat)
```

## P(X > 0)

```{r}
pbinom(0, size = 2, prob = p_hat, lower.tail = FALSE)
```


## P(X = 1) + P(X = 2)

```{r}
# sum up the probability of that 1 person orders coffee and the probability that both persons order coffee
# P(1 guest has coffee) + P(2 guests has coffee)
dbinom(1, size = 2, prob = p_hat) + dbinom(2, size = 2, prob = p_hat)
```

## 1 - P(X = 0)

```{r}
# 1 - P(X = 0)
# As the total probability is 1, we can get it as 1 - P(0 guests have coffee). 
# This follows from the total probability rule
1 - dbinom(0, size = 2, prob = p_hat)
```

:::

- To check whether our estimated probability of drinking coffee consistent with the one reported by a recent [YouGov survey](https://today.yougov.com/topics/society/articles-reports/2022/09/29/how-do-people-us-take-their-coffee), which reported that three quarters of adult Americans drink coffee, we can take a look at our estimated $\hat{p}$. We can also then comment on whether adult Americans are more or less likely to drink coffee than Norwegians?

```{r}
# YouGov reported probability
p_survey <- 3/4
p_survey
```

```{r}
# Our estimate
p_hat
```

The estimated probability of drinking coffee, based on the data from a US bistro, is `r round(p_hat, 2)`. This is relatively close to the YouGov reported result of 0.75. The small deviation may be due to sampling variability, due to having chosen a different sample of people from the one that were considered in this bistro. 

According to a recent survey, 89.4% of Norwegians drink at least one coffee per day^[Tverdal, A., Hjellvik, V. & Selmer, R. Coffee intake and oral–oesophageal cancer: follow-up of 389,624 Norwegian men and women 40–45 years. Br J Cancer 105, 157–161 (2011).] As such, while adult Americans drink lots of coffee, an adult American is  less likely to drink coffee than a Norwegian.

:::{.callout-tip}

#### Example writeup

@fig-coffee displays the distribution of coffee-drinking customers in two-party tables, with a Binomial fit superimposed as red dots. Both party guests are more likely to both drink coffee (0.58) than not (0.06). The probability of only one guest drinking coffee is 0.36. As such, the probability of at least one guest drinking coffee is 0.94.
The owner of the bistro should not consider running a 2 for 1 offer on Christmas coffees, as they would lose out on income. It is most likely that groups of two customers will already buy two coffees when they visit the bistro (0.58). Instead, the owner might want to offer a loyalty scheme to reward customers for purchasing coffees, regardless of the quantity, as this will also reward parties of two where only one person purchases a coffee. 
:::

## Student Glossary

To conclude the lab, add the new functions to the glossary of `R` functions. 

| Function   | Use and package |
|:-----------|:--------------|
| `rename() `     | ? |
| `dbinom() `     | ? |
| `pbinom()`      | ? |
| `geom_col() `     | ? |
| `geom_point() `     | ? |
| `geom_segment() `     | ? |
| `cumsum()`      | ? |

