---
title: "Hypothesis testing: critical values"
subtitle: "Semester 2 - Week 3"
callout-appearance: simple
editor_options: 
  chunk_output_type: inline
---


```{r setup}
#| include: false

source('assets/setup.R')

library(tidyverse)
library(patchwork)
library(kableExtra)

grad <- read_csv("https://uoepsy.github.io/data/dataset-ipeds-2012-subset2.csv")
head(grad)
glimpse(grad)

plt.grad.h <- ggplot(grad, aes(x = gradratew)) +
    geom_histogram(color = 'white')

plt.grad.d <- ggplot(grad, aes(x = gradratew)) +
    geom_density()

plt.grad.b <- ggplot(grad, aes(x = gradratew)) +
    geom_boxplot()

plt.grad.h / plt.grad.b

stats.grad <- grad %>%
    summarise(n = length(gradratew),
              Min = min(gradratew),
              Max = max(gradratew),
              M = mean(gradratew),
              SD = sd(gradratew))

stats.grad %>%
    kbl(digits = 2) %>%
    kable_styling(full_width = FALSE)

psych::describe(grad$gradratew)

xbar <- mean(grad$gradratew)
n <- nrow(grad)
tstar <- qt(c(0.025, 0.975), n - 1)
se <- sd(grad$gradratew) / sqrt(n)
xbar + tstar * se

stats.grad$M + tstar * stats.grad$SD / sqrt(stats.grad$n)

# week 2 code ====

xbar <- mean(grad$gradratew)
n <- nrow(grad)
s <- sd(grad$gradratew)
se <- s / sqrt(n)

tobs <- (xbar - 50) / se
tobs

pvalue <- 2 * pt(abs(tobs), df = n - 1, lower.tail = FALSE)
pvalue

tstar <- qt(c(0.025, 0.975), n - 1)
xbar + tstar * se

t.test(grad$gradratew, mu = 50)


shapiro.test(grad$gradratew)

qqnorm(grad$gradratew)
qqline(grad$gradratew)

dim(grad)


# week 3 code

tstar <- qt(c(0.025, 0.975), df = n - 1)
tstar

tobs

tobs >= tstar[2]
tobs <= tstar[1]
```




## Tasks

The data __dataset-ipeds-2012-subset2__, available at <https://uoepsy.github.io/data/dataset-ipeds-2012-subset2.csv>, is a subset of data derived from the Integrated Postsecondary Education Data System (IPEDS) at the National Center for Education Statistics, 2012. The variables include:

```{r}
#| echo: false
library(tidyverse)
library(kableExtra)

cb <- tibble(
    Variable = 
        c("type", "region", "gradrate", "gradratem", "gradratew"),
    Description = 
        c("College Type: −1 = Not reported; 1 = Public; 2 = Private for-profit; 3 = Private not-for-profit (no religious affiliation); 4 = Private not-for-profit (religious affiliation)", "Region: 0 = US Service schools; 1 = New England; 2 = Mid East; 3 = Great Lakes; 4 = Plains; 5 = Southeast; 6 = Southwest; 7 = Rocky Mountains; 8 = Far West; 9 = Outlying areas", "Graduation Rate – All", "Graduation Rate – Men", "Graduation Rate – Women")
)

cb %>%
    kable() %>%
    kable_styling(full_width = FALSE)
```


In formative report C, you will investigate the mean graduation rate for female students at colleges and universities in the United States. Specifically, you are asked to perform the following tasks, each related to a week of teaching in this course.  

__This week you will only focus on task C3.__

C1) Read the data into R, describe the variable of interest both visually and numerically, and provide a 95% CI for the mean graduation rate of female students at colleges and universities in the United States.  
C2) At the 5% significance level and using the p-value method, test whether the mean graduation rate for female students at colleges and universities in the United States is significantly different from a rate of 50 percent.  

:::{.callout-note}
### This week's task
C3) At the 5% significance level and using the critical value method, test whether the mean graduation rate for female students at colleges and universities in the United States is significantly different from a rate of 50 percent.  
:::

C4) Tidy up your report so far, making sure to have 3 sections: introduction, analysis and discussion.  
C5) Compute and report the effect size, check if the assumptions underlying the t-test are violated.


 


## C3 sub-tasks

In this section you will find some guided sub-steps you may want to consider to complete task C3.

::: {.callout-tip}
#### Tip

To see the hints, hover your cursor on the superscript numbers.
:::

- Reopen last week's Rmd file, as you will continue last week's work (i.e., the top 3 most frequent movie genres that were rated as 'good' and 'bad'), and build on it.[^share-file]

[^share-file]: Hint: Ask last week's driver for the Rmd file, they should share it with the group via email or Teams. To download the file from the server, go to the RStudio Files pane, tick the box next to the Rmd file, and select More > Export.

- State the null and alternative hypotheses.[^hypotheses]

[^hypotheses]: Hint: Identify the hypothesised value of the mean, $\mu_0$, and replace that value in the following equations:

    $$H_{0} : \mu = \mu_{0}$$
    $$H_{1} : \mu \neq \mu_{0}$$

    The equations can be written with the following code, where the text within curly braces after an underscore is rendered as a subscript:

    ```
    $$ H_{0} : \mu = \mu_{0} $$
    $$ H_{1} : \mu \neq \mu_{0} $$
    ```

- Compute the value of the t-statistic from the sample mean graduation rate of female students.[^t-stat]

[^t-stat]: Hint: Use the following formula for the t-statistic:

    $$t = \frac{\bar{x} - \mu_0}{SE_{\bar{x}}} \qquad \text{where} \qquad SE_{\bar{x}} = \frac{s}{\sqrt{n}}$$
    
    In the above:
    
    - $\bar{x}$ is the sample mean
    - $\mu_0$ is the hypothesised value for the population mean
    - $s$ is the sample standard deviation
    - $n$ is the sample size

- Identify the null distribution.^[Hint: the t-statistic follows a $t(n-1)$ distribution where $n-1$ is the degrees of freedom.  
The question asks you: what are the degrees of freedom to use in the t distribution?]

- Compute the critical values of the null distribution using the appropriate significance level.^[Hint: The `qt()` function may be useful]

- Make a decision on whether or not to reject the null hypothesis.

- Provide a write up of your results in the context of the research question.

::: {.callout-note}
__Reporting with critical values__
When you use the critical value method, you don't have a computed p-value, so the reporting will only say whether p < .05 or p > .05 depending on whether the observed t-statistic is beyond or not beyond the critical value(s).

- If $t$ is in the rejection region:

    + p < .05 (if you use a different $\alpha$, change accordingly)

- If $t$ is not in the rejection region:

    + p > .05 (if you use a different $\alpha$, change accordingly)
:::


## Worked Example

The Procrastination Assessment Scale for Students (PASS) was designed to assess how individuals approach decision situations, specifically the tendency of individuals to postpone decisions ([Solomon & Rothblum, 1984](https://psycnet.apa.org/doiLanding?doi=10.1037%2F0022-0167.31.4.503)).

The PASS assesses the prevalence of procrastination in six areas: writing a paper; studying for an exam; keeping up with reading; administrative tasks; attending meetings; and performing general tasks. For a measure of total endorsement of procrastination, responses to 18 questions (each measured on a 1-5 scale) are summed together, providing a single score for each participant (range 0 to 90). The mean score from [Solomon & Rothblum, 1984](https://psycnet.apa.org/doiLanding?doi=10.1037%2F0022-0167.31.4.503) was 33.

> __Research question:__
> 
> Does the mean procrastination score of Edinburgh University students differ from the Solomon & Rothblum average of 33?

To answer this question, we will use data collected for a random sample of students from the University of Edinburgh: <https://uoepsy.github.io/data/pass_scores.csv>

```{r, echo=FALSE}
library(tidyverse)
library(kableExtra)

tribble(
    ~'Variable Name', ~'Description',
    'sid', 'Subject identifier',
    'school', 'School each subject belonged to',
    'PASS', 'Total endorsement of procrastination score'
) %>%
    kbl() %>%
    kable_styling(full_width = FALSE)
```

Necessary packages:

::: {.column-margin}
- tidyverse for using `read_csv()`, using `summarise()` and `ggplot()`.
- patchwork for arranging plots side by side or underneath
- kableExtra for creating user-friendly tables
:::


```{r}
library(tidyverse)
library(patchwork)
library(kableExtra)
```

Read the data into R:

```{r}
pass_scores <- read_csv("https://uoepsy.github.io/data/pass_scores.csv")
dim(pass_scores)
```

To inspect the data:

::: {.panel-tabset}

## head()
```{r}
head(pass_scores)
```

## glimpse()
```{r}
glimpse(pass_scores)
```

## summary()
```{r}
summary(pass_scores)
```

:::

Visualise the distribution of `PASS` scores:

::: {.column-margin}
__Note__

The boxplot highlights an outlier (40). However, this value is well within the plausible range of the scale (0 -- 90), hence it is of no concern and the point can be kept for the analysis.
:::

```{r}
plt_hist <- ggplot(pass_scores, aes(x = PASS)) + 
    geom_histogram(color = 'white')

plt_box <- ggplot(pass_scores, aes(x = PASS)) + 
    geom_boxplot()

plt_hist / plt_box
```

Descriptive statistics:

```{r}
stats <- pass_scores %>%
    summarise(n = n(),
              Min = min(PASS),
              Max = max(PASS),
              M = mean(PASS),
              SD = sd(PASS))
```

```{r, eval=FALSE}
stats %>%
    kbl(booktabs = TRUE, digits = 2, 
        caption = "Descriptive statistics for PASS scores")
```

```{r, echo=FALSE}
#| label: pass-stats
stats %>%
    kbl(digits = 2, caption = "Descriptive statistics for PASS scores") %>%
    kable_styling(full_width = FALSE)
```

**Step 1: Identify the null and alternative hypotheses.**  

First we need to write the null and alternative hypothesis, which take the form $H_0 : \mu = \mu_0$ vs $H_1: \mu \neq \mu_0$. From the research question, we identify the hypothesised value $\mu_0$ to be 33, hence:

::: {.column-margin}
These are written as:

```
$$H_{0}: \mu = 33$$
$$H_{1}: \mu \neq 33$$
```

In H_{0} and H_{1} the 0 and 1 within curly braces are written as subscripts. The curly braces delimit what goes in the subscript. The symbol `\mu` denotes the greek letter "mu" that stands for the population mean (a parameter). The symbol `\neq` means not equal.
:::


$$H_0: \mu = 33$$
$$H_1: \mu \neq 33$$

**Step 2: Compute the t-statistic**  

Next, we compute the t-statistics, which compares the difference between the sample and hypothesised mean ($\bar{x} - \mu_0$) to the variation due to random sampling ($SE_{\bar{x}}$).

To test the hypothesis, we need to compute the t-statistic, 

$$
t = \frac{\bar{x} - \mu_0}{SE_{\bar{x}}} \qquad \text{where} \qquad SE_{\bar{x}} = \frac{s}{\sqrt{n}}
$$

```{r}
# Sample mean
xbar <- stats$M

# Standard error
s <- stats$SD
n <- stats$n
se <- s / sqrt(n)

# Observed t-statistic
tobs <- (xbar - 33) / se
tobs
```

**Step 3: Identify the null distribution**, i.e. the distribution of the t-statistic assuming the null to be true.  

As the sample size is $n = $`r n`, if the null hypothesis is true the t-statistic will follow a t(`r n - 1`) distribution.

**Step 4: Compute the critical values for $\alpha = 0.05$**  

As the alternative hypothesis is two-sided (or two-tailed), we have two critical values that cut an area of 0.05 beyond them. That area must be equally split in both tails, so 0.025 to the left and 0.025 to the right

```{r}
tstar <- qt(c(0.025, 0.975), df = n - 1)
tstar
```

**Step 5: Make a decision**  

Compare the observe statistic to the critical values:

```{r}
tobs
tstar
```

Is the observed t-statistic within the critical values (hence we do not reject) or beyond (hence we reject)?

```{r}
tobs <= tstar[1]
tobs >= tstar[2]
```

Using the critical-value method, we compare the observed t-statistic with the critical values for the appropriate significance level ($\alpha = 0.05$ in this case). 
As 0.005800318 < 0.05, we reject $H_0$. Please note this is just an explanation and not how you would write up the result!


A potential way to report the t-test results:

:::{.callout-tip}
#### Example hypothesis test write-up
At the 5% significance level, the sample data provide evidence against the null hypothesis and in favour of the alternative one that the mean procrastination score of Edinburgh University students is different from the Solomon & Rothblum reported average of 33: $t(19) = -3.11, p < .05$, two-sided.
:::


## Student Glossary

To conclude the lab, add the new functions to the glossary of `R` functions. 

| Function   | Use and package |
|:-----------|:--------------|
| `geom_histogram` | ? |
| `geom_boxplot` | ? |
| `summarise` | ? |
| `n()` | ? |
| `mean` | ? |
| `sd` | ? |
| `abs` | ? |
| `qt` | ? |
