---
title: 'dapR1: Hypothesis Testing & P Values Lec Activities'
author: "Department of Psychology<br>University of Edinburgh"
date: "AY 2020-2021"
output:
  html_document:
    theme: flatly
---

# Activity 1  

Please watch this video, from Tom Stoppard's play "Rosencrantz and Guildenstern Are Dead".   
(It's tempting to skip ahead in the video because it gets a bit repetitive, but don't be impatient - humour needs suspense). 

<iframe width="560" height="315" src="https://www.youtube.com/embed/gOwLEVQGbrM" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

### After watching

Read the first 2 sections of [this article for daily.jstor](https://daily.jstor.org/statistics-of-coin-tosses-theater-geeks/#:~:text=In%20Stoppard's%20scene%2C%20the%20bit,it%20lands%20as%20heads%20again) (you don't have to read beyond "a short history of coin flips").  

Think about the idea here in terms of what Rosencrantz and Guildenstern's null hypothesis is. 

We can actually calculate their p-value! 
They get 78 Heads in 78 Coin Tosses. What's the probability of this happening **with a fair coin** (one that lands on heads/tails evenly)?  

```{r echo=FALSE}
heads = 0:78
plot(heads,dbinom(heads,size=78,prob=.5),
     type='h',
     main='Binomial Distribution where n = 78 and p=0.5',
     ylab='Probability',
     xlab ='# Heads',
     lwd=3)
```

(for a little refresh on the binomial, see the lecture from [Sem 1 Week 9](https://uoepsy.github.io/dapr1/lectures/dapR1_lec8_DiscreteProbabilityDist.html), especially [slide 16](https://uoepsy.github.io/dapr1/lectures/dapR1_lec8_DiscreteProbabilityDist.html#28)).  

We can actually work this out using R, with the `pbinom()` function. 
Run this in your console (bottom left window of RStudio).
What do you get?  
```{r}
# we have the "1-" because we want the right hand side. 
# we're asking "what's the probability of getting as many or more than 78 heads out of 78 coin tosses, IF the coin is unbiased?"
1 - pbinom(78, size = 78, prob = 0.5)
```
Is this a valid answer that R has given you?  
Think about the possible values a probability can take.  

# Activity 2  

Have you seen sometimes in some of R's output we get things like `2.26e-17`, or `1.742926e-10` or `1.34e+81` and wondered what on earth they mean?  

$$
10^6 = 1,000,000 \\
\qquad \\
1.64\text{e}6 \qquad = \qquad 1.64 \times 10^6 \qquad  = \qquad 164 \times 1,000,000 \qquad  = \qquad 1,640,000
$$
```{r}
1.64e6
```
$$
10^{-6} = 0.000006 \\
\qquad \\
\text{2.4e-6}\qquad  = \qquad ????
$$


<br>
<br>

---  

*Scroll down for answers*  
  
---

<div style="margin-bottom:750px;">

</div>

## Activity 1 

P-values are probabilistic statements, and are always between 0 and 1, and they are never exactly 0 or 1.   
$0 < p < 1$  - YES!  
$0 \leq p \leq 1$ - NO!  
  
p = 0 would mean an **impossible** event.  
p = 1 would mean a **certain** event.  


## Activity 2

$$
\text{2.4e-6}\qquad  = \qquad 2.4 \times 10^{-6} \qquad = \qquad 2.4 \times 0.000006 \qquad = \qquad 0.0000144
$$
