---
title: "Block 4 Week 5: Fitness & Health Report"
output: bookdown::pdf_document2
toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{r include = FALSE}
######## LOAD LIBRARIES & DATA ######## 
library(tidyverse)
library(kableExtra)
library(psych)
library(patchwork)
library(kableExtra)


dat <- read_csv("https://uoepsy.github.io/data/fitnessProgram.csv")

###### DATA CHECKS ###### 

str(dat)
summary(dat)


###################### RQ1 ########################

####### ASSUMPTION CHECKS #######

# Days in programme
shapiro.test(dat$daysInProgram)

hd_viz_time <- ggplot(dat, aes(x = daysInProgram, y = after_stat(density))) + 
  geom_histogram() + 
  geom_density()
hd_viz_time

q_viz_time <- ggplot(dat, aes(sample = daysInProgram)) + 
  geom_qq() + 
  geom_qq_line() + 
  labs(title = "QQPlot - Days in Program")
q_viz_time

#CV health
shapiro.test(dat$cvHealth)

hd_viz_cv <- ggplot(dat, aes(x = cvHealth, y = after_stat(density))) + 
  geom_histogram() + 
  geom_density()
hd_viz_cv 

q_viz_cv <- ggplot(dat, aes(sample = cvHealth)) + 
  geom_qq() + 
  geom_qq_line() + 
  labs(title = "QQPlot - CV Health")
q_viz_cv

####### VISUALISATION #######

#Because we are interested in running a hypothesis test on the correlation between `daysInProgram` and `cvHealth`, we can produce a scatterplot to check for linearity, outliers, and heteroscedasticity. Can also use this to visulaise the association between our two variables:

viz_time_cv <- ggplot(dat, aes(x = daysInProgram, y = cvHealth)) + 
  geom_point() + 
  geom_smooth(method = 'lm', colour = 'red', se = F) +
  labs(x='Time in Program (days)', y = 'Cardiovascular Health', title = "Association Between Time in Program and Cardiovascular Health") 

####### CORRELATION #######

cor(dat$daysInProgram, dat$cvHealth)

cor.test(dat$daysInProgram, dat$cvHealth,
         alternative = "two.sided")

###################### RQ2 ########################

####### VISUALISATION #######

# we can see that a scatterplot is not informative for Likert scale data:
ggplot(dat, aes(x = Difficulty, y = Satisfaction)) + 
  geom_point() + 
  geom_smooth(method = 'lm', colour = 'red', se = F)

#boxplots or barplots may be more informative 

ggplot(dat, aes(x = factor(Difficulty), y = Satisfaction)) + 
  geom_boxplot()

viz_diff_sat <- ggplot(dat, aes(x = factor(Difficulty), fill = factor(Satisfaction))) + 
  geom_bar(position = "fill") + 
  labs(x = "Difficulty Rating", y = "Satisfaction Rating", fill = "Satisfaction Rating")
viz_diff_sat

####### CORRELATION #######

cor.test(dat$Difficulty, dat$Satisfaction, 
         alternative = "two.sided",
         method = "spearman")

#To suppress warning message, use the following `exact = FALSE` argument witin cor.test

###################### BONUS CONTENT ########################


#There are a few ways that you can plot correlation data. When you are looking at many correlation values, it may be helpful to plot a correlogram. There are lots of packages that allow you to do this. We're going to use the `corrplot` package. 

#Instead of running separate correlations for each variable pair, it's much similar to create a correlation matrix to look at the correlations between all continuous variables. 

corDat <- round(cor(dat[1:5]),2)
corDat

####### CORR PLOT #######

library(corrplot)

#You can make a correlogram with the numeric correlation values:
corrplot(corDat, method = 'number')

#or with representative colors:
corrplot(corDat, method = 'color')

#You can mix numbers and colors with the `corrplot.mixed` function. I've also added the `tl.col` argument to make the text black:  
corrplot.mixed(corDat, lower='number', upper='color', tl.col='black')

#the text is cut off in the diagonal, so I've added the `tl.pos` argument to set the text position to the left axis and the top:  
corrplot.mixed(corDat, lower='number', upper='color', tl.col='black', tl.pos = 'lt')

####### CORRR #######

library(corrr)
# Need to use dataset - so using columns 1-5 from dat dataset in this example

#Network plot - variables that are more highly correlated appear closer together and are joined by stronger paths. Blue = positive, red = nagative correlations
dat[1:5] |>
  correlate() |>
  network_plot()

#shave() retain only one part of correlation matrix - upper or lower 
#rplot() the correlations with shapes in place of the values - bigger shapes = larger associations
dat[1:5] |>
  correlate() |>
  shave() |>
  rplot()


# present in well formatted table
dat[1:5] |>
  correlate() |>
  shave() |>
  fashion() |>
  kable(caption = "Correlations", digits = 2) |>
  kable_styling()

```


# Introduction

The data available at https://uoepsy.github.io/data/fitnessProgram.csv were provided by a Personal Trainer who specializes in fitness programs, specifically running programs. Their dataset contained information on 8 variables - the number of days their clients have been enrolled on the program (`daysInProgram`), the maximum distance they can run (in miles; `maxDistance`), their base fitness level (`baseFitness`), their cardiovascular health rating (`cvHealth`), their average resting pulse rate (`pulse`), their age (in years; `age`), their difficulty rating of the program (on a scale of 1-5; `Difficulty`), and their overall rating of satisfaction with the program (on a scale of 1-5; `Satisfaction`). 

## Research Questions

+ RQ1: Is there a significant association between the number of days one spends in the program and one's cardiovascular health?
+ RQ2: Is there a significant association between satisfaction and difficulty ratings?

# Analysis 

## Research Question 1

Total time enrolled in the program was moderately positively associated with cardiovascular health, and this association was statistically significant $(r= .46, t(498) = 11.64, p < .001)$. This association is visually presented in Figure \@ref(fig:viz-rq1). These results suggested that a greater number of days in the program was positively associated with better cardiovascular health.

```{r viz-rq1, fig.cap = "Association Between Time in Program and Cardiovascular Health", fig.height = 2}
viz_time_cv
```

Both days in program ($W = 0.998, p = .725$) and cardiovascular health ($W = 0.996, p = .163$) were normally distributed. From Figure \@ref(fig:viz-rq1), we can see that there are no extreme outliers, the association between the two variables is linear, and there is no evidence of heteroscedasticity. 

## Research Question 2

We used Spearman's $\rho$ to determine whether there was a significant association between difficulty and satisfaction ratings. We found a significant weak negative association between difficulty and satisfaction $(\rho = -.21, p < .001)$, such that those who found the program more difficult reported lower satisfaction with the program overall.

```{r viz-rq2, fig.cap = "Association Between Difficulty and Satisfaction", fig.height = 2}
viz_diff_sat
```

# Appendix

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```

