---
title: "Models"
subtitle: Data Analysis for Psychology in R 1
author: Tom Booth
---

```{r premable, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(kableExtra)
```

## Today
- Introduce the idea of models in statistics
- Think through models as functions
- Consider how models, plots and functions relate.
- **Admin**: I want to talk through the lab tests

## Learning objectives
- Understand the general concept of a model.
- Understand how a model relates to a function.
- Understand how a model provides predictions about the world

## What is a model?
- Pretty much all statistics is about models.
- A model is a formal representation of a system.
- Put another way, a model is an idea about the way the world is.

## Models in psychology
- Rescorla-Wagner Model: classical conditioning

$$
\Delta V_{X}^{n+1} = \alpha_X \beta(\lambda - V_{tot})
$$

- $\Delta V_{X}^{n+1}$ = change in strength of an association.
- $\alpha_x$ = Salience of X (CS)
- $\beta$ = salience of US
- $\lambda$ = maximum possible conditioning for US
- $V_{tot}$ = strength between CS and US for all possible stimuli

## Models in Psychology

$$
\Delta V_{X}^{n+1} = \alpha_X \beta(\lambda - V_{tot})
$$

- As a function:

$$
f(\alpha, \beta, \lambda, z = [V_{tot}]) = \alpha_X \beta(\lambda - V_{tot})
$$

## A model as a function
- We tend to represent mathematical models as functions.
  - which can be very helpful.
- It allows for the precise specification about what is important (arguments) and what those things do (operations)
  - This leads to predictions
  - And these predictions can be tested.

## An Example
- To think through these relations, we can use a simpler example.
- Suppose I have a model for growth of babies.

$$
Length = 55 + 4*month
$$

## Visualizing a model

```{r, warning=FALSE, message=FALSE}
fun1 <- function(x) 55 + (4*x)
m1 <- ggplot(data = data.frame(x=0), aes(x=x)) +
  stat_function(fun = fun1) +
  xlim(0,24) +
  ylim(0,150) +
  ylab("Length (cm)") +
  xlab("Age (months)")
```

## Visualizing a model
```{r, warning=FALSE, message=FALSE}
m1
```

## Models as "a state of the world"
- Let's suppose my model is true.
  - That is, it is a perfect representation of how babies grow.
- What are the implications of this?

## Models and predictions
- My models creates predictions.
- **IF** my model is a true representation of the world, **THEN** data from the world should closely match my predictions.
  - Sometimes this is referred to as my model matching the *data generation process*


## Predictions and data

```{r, eval=FALSE}
tibble(
  Age = seq(10, 12, 0.25)
) %>%
  mutate(
    Prediction = 55 + (Age*4)
  ) %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```


## Predictions and data

```{r, echo=FALSE}
tibble(
  Age = seq(10, 12, 0.25)
) %>%
  mutate(
    Prediction = 55 + (Age*4)
  ) %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

- Our predictions are points which fall on our line (representing the model, as a function)


## Predictions and data

```{r, echo=FALSE, warning=FALSE, message=FALSE}
m1
```

- Consider the predictions when the children get a lot older...

## Predictions and data

```{r, eval=FALSE}
tibble(
  Age = seq(216,300, 12)
) %>%
  mutate(
    Year = Age/12,
    Prediction = 55 + (Age*4),
    Predction_M = Prediction/100
  ) %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```


## Predictions and data

```{r, echo=FALSE}
tibble(
  Age = seq(216,300, 12)
) %>%
  mutate(
    Year = Age/12,
    Prediction = 55 + (Age*4),
    Predction_M = Prediction/100
  ) %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

- What do you think this would mean for our actual data?
- Will the data fall on the line?

## How good is my model?
- How might we judge how good our model is?

1. Model is represented as a function
2. We see that as a line (or surface if we have more things to consider)
3. That yields predictions (or values we expect if our model is true)
4. We can collect data
5. If the predictions do not match the data (points deviate from our line), that says something about our model.


## Models and Statistics
- In statistics we (roughly) follow this process.
- We define a model that represents one state of the world (probabilistically)
- We then collect data to compare to it.
- These comparisons lead us to make inferences about how the world actually is, by comparison to a world that we specify by our model. 

## Length & Age is non-linear

```{r, eval=FALSE, warning=FALSE, message=FALSE}
df <- read_csv("length_age.csv", col_names = c("Month", "Mean", "SD", "Lower", "Upper"))
m1 + 
  geom_line(data = df, aes(x=Month, y = Mean), colour = "red") + 
  geom_line(data = df, aes(x=Month, y = Lower), colour = "red", linetype = "dashed") +
  geom_line(data = df, aes(x=Month, y = Upper), colour = "red", linetype = "dashed")
```


## Length & Age is non-linear

```{r, echo=FALSE, warning=FALSE, message=FALSE}
df <- read_csv("length_age.csv", col_names = c("Month", "Mean", "SD", "Lower", "Upper"))
m1 + 
  geom_line(data = df, aes(x=Month, y = Mean), colour = "red") + 
  geom_line(data = df, aes(x=Month, y = Lower), colour = "red", linetype = "dashed") +
  geom_line(data = df, aes(x=Month, y = Upper), colour = "red", linetype = "dashed")
```

- Our red line is plotted based on the mean length for different ages [real data](https://www.cdc.gov/growthcharts/who/boys_length_weight.htm)

# Lab tests

## What to expect...
- You will be assigned to either the first or second hour of your normal lab.
- You will come in, have 10 minutes to settle, open computers etc.
- You will open an .rmd file that will contain questions and empty code chunks
- You will need to write some R-code to answer the question
  - You will knit your .rmd file
  - And hand in the .rmd and the knitted file via LEARN

## Types of question
- Question will draw on the content of the course up to the point where the test.
- They will focus mainly on recent material, but may have some things repeated (like sorting data)
- Examples for test 1:
  - Make a plot
  - Produce summary statistics for variable X
  - How many subjects are there in the data that meet conditions A, B and C
- There will be 4 or 5 questions in each lab test.
  - The last of which will be hard.

## Final things
- If you have adjustments, you will have extra time as if this is an exam.
- You can have all your notes and the internet
- The admin for the test in week 10 will happen next week
- There will be a practice submission during the lab before the first test.

## Tasks for this week...
1. Finish any existing tasks from labs.
2. Reading: See LEARN
3. Quiz 7 on functions
    - **This quiz counts**
    - Live now (as of Monday at 09:00).
    - Closes Sunday at 17:00

## Recommendations of the week
- Podcast: [Another In Our Time](https://www.bbc.co.uk/programmes/b00qj2nq)
- Recipe: [Breakfast Egg Muffins](https://cafedelites.com/breakfast-egg-muffins-3-ways/)
- Book: [SPQR, Mary Beard](https://www.goodreads.com/book/show/28789711-spqr)
- Thing to do: [St. Giles Cathedral](https://stgilescathedral.org.uk/music/)
- Coffee/food Place: [Miros Portobello Beach](http://www.mirosportobello.co.uk/)
