---
title: "<b>Week 5: Functions </b>"
subtitle: "Data Analysis for Psychology in R 1<br><br> "
author: "Patrick Sturt"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    self_contained: true
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
    base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(kableExtra)
knitr::opts_chunk$set(fig.asp=.9)
```

# Course Overview

.pull-left[

```{r echo = FALSE, results='asis'}
block1_name = "Exploratory Data Analysis"
block1_lecs = c("Research design and data",
                "Describing categorical data",
                "Describing continuous data",
                "Describing relationships",
                "Functions")
block2_name = "Probability"
block2_lecs = c("Probability theory",
                "Probability rules",
                "Random variables (discrete)",
                "Random variables (continuous)",
                "Sampling")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")
course_table(block1_name,block2_name,block1_lecs,block2_lecs,week=5)
```


]

.pull-right[


```{r echo = FALSE, results='asis'}
block3_name = "Foundations of inference"
block3_lecs = c("Confidence intervals",
                "Hypothesis testing (p-values)",
                "Hypothesis testing (critical values)",
                "Hypothesis testing and confidence intervals",
                "Errors, power, effect size, assumptions")
block4_name = "Common hypothesis tests"
block4_lecs = c("One sample t-test",
                "Independent samples t-test",
                "Paired samples t-test",
                "Chi-square tests",
                "Correlation")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")
course_table(block3_name,block4_name,block3_lecs,block4_lecs,week=0)
```

]

---


# Weeks Learning Objectives
1. Understand the basic principles of functions.

2. Understand concept of data transformations.

3. Understand the calculation of z-scores.

---
# Topics for today
+ What is a function?

--

+ Linear and non-linear functions

--

+ How do we use functions in statistics?

--

+ An example of z-scores
  
---
# What is a function?
+ A function takes an **input**, **does something**, and provides an **output.**

--

+ **Input**

$$
x = 
\begin{bmatrix}
1 \\ 
2 \\ 
3 \\ 
\end{bmatrix}
$$

--

+ **Doing something**

$$
f(x) = x-2
$$

--

+ **An output **

$$
y = 
\begin{bmatrix}
-1 \\ 
0 \\ 
1 \\ 
\end{bmatrix}
$$

???
+ Functions can become as complex as we want them to be.
+ Why $y$?


---
# Functions and relations
+ It is important to think of the function as showing the *relationship* between input and output.

+ We can link this to the idea of relationships from week 4.

+ The function links an input (predictors, $x$), to an output (outcome, $y$)

--

+ So we can write

$$y = f(x) = x-2$$


---
# Visualising Functions
+ An important tool in understanding functions is to plot them.

+ So let's look at the following:


$$
y = f(x) = 10 + 2x
$$

???
+ This helps us both understand plots
+ And gain intuition about functions.
  
---
# Visualising Functions
+ Our input $x$ is a vector of numbers:

$$
x = 
\begin{bmatrix}
1 \\ 
2 \\ 
3 \\
4 \\
5 \\ 
6 \\ 
7 \\
8 \\
\end{bmatrix}
$$


???
+ As well as the functions getting more complex, so can the inputs.
+ We are sticking with small examples to help visualize what is happening and get an intuition
    + But this could be 10,000 elements long
    + could contain values like 1.7875453
    + Computers can deal with all this for us quite easily!


---

# Visualising Simple Functions

.pull-left[
```{r, echo=FALSE}
func_x <- tibble(
  x = c(1,2,3,4,5,6,7,8), 
  y = 10 + (2*x) 
)
  
kable(func_x) %>%
  kable_styling(., full_width = F)
  
```
]

.pull-right[

```{r, eval=FALSE}
func_x <- tibble(
  x = c(1,2,3,4,5,6,7,8), 
  y = 10 + (2*x) #<<
)
```

+ `tibble` is used to create a data set

+ `x` is our original data entered as a vector of numbers using `c()`

+ `y` is the output of the function `f(x) = 10+(2*x)`

]


---

# Visualising Simple Functions

.pull-left[
```{r, echo=FALSE}
func_x <- tibble(
  x = c(1,2,3,4,5,6,7,8), 
  y = 10 + (2*x) 
)
  
kable(func_x) %>%
  kable_styling(., full_width = F)
  
```
]

.pull-right[
$$
y = f(x) = 10 + 2x
$$

+ Example row 1:

$$
10 + (2*1) = 12
$$

+ Example row 5:

$$
10 + (2*5) = 20
$$

]


---
# Visualising Functions
.pull-left[
```{r, echo=FALSE}
ggplot(func_x[1,], aes(x, y)) +
  scale_x_continuous(name = "x", breaks = c(0:10), labels = c(0:10), limits = c(0,10)) +
  scale_y_continuous(name = "y", breaks = c(seq(0,30,5)), labels = c(seq(0,30,5)), limits = c(0,30)) 
```
]

.pull-right[

**Our Data**

```{r, echo=FALSE}
kable(func_x) %>%
  kable_styling(., full_width = F)
```
]

---
# Visualising Functions

.pull-left[
```{r, echo=FALSE}
ggplot(func_x[1,], aes(x, y)) + 
  scale_x_continuous(name = "x", breaks = c(0:10), labels = c(0:10), limits = c(0,10)) +
  scale_y_continuous(name = "y", breaks = c(seq(0,30,5)), labels = c(seq(0,30,5)), limits = c(0,30))  +
  geom_point(colour = "red", size = 3) +
  geom_segment(aes(x = x, y = y, xend = x, yend = 0), 
               arrow=arrow(type = "closed", length = unit(0.25, "cm"))) +
  geom_segment(aes(x = x, y = y, xend = 0, yend = y), 
               arrow=arrow(type = "closed", length = unit(0.25, "cm")))
```
]

.pull-right[

**Our Data**

```{r, echo=FALSE}
kable(func_x) %>%
  kable_styling(., full_width = F)
```
]

---
# Visualising Functions

.pull-left[
```{r, echo=FALSE}
ggplot(func_x, aes(x, y)) + 
  scale_x_continuous(name = "x", breaks = c(0:10), labels = c(0:10), limits = c(0,10)) +
  scale_y_continuous(name = "y", breaks = c(seq(0,30,5)), labels = c(seq(0,30,5)), limits = c(0,30)) +
  geom_point(colour = "red", size = 2)
```
]

.pull-right[

**Our Data**

```{r, echo=FALSE}
kable(func_x) %>%
  kable_styling(., full_width = F)
```
]

---
# Visualising Functions

.pull-left[
```{r, echo=FALSE}
fun1 <- function(x) 10+(2*x)

ggplot(func_x, aes(x, y)) + 
  geom_point(colour = "red", size = 2) +
  stat_function(fun = fun1) +
  scale_x_continuous(name = "x", breaks = c(0:10), labels = c(0:10), limits = c(0,10)) +
  scale_y_continuous(name = "y", breaks = c(seq(0,30,5)), labels = c(seq(0,30,5)), limits = c(0,30)) 
```
]

.pull-right[

**Our Data**

```{r, echo=FALSE}
kable(func_x) %>%
  kable_styling(., full_width = F)
```
]


---
# Visualising Functions (R-code)

.pull-left[
```{r, echo=FALSE}
fun1 <- function(x) 10+(2*x)

ggplot(func_x, aes(x, y)) + 
  geom_point(colour = "red", size = 2) +
  stat_function(fun = fun1) +
  scale_x_continuous(name = "x", breaks = c(0:10), labels = c(0:10), limits = c(0,10)) +
  scale_y_continuous(name = "y", breaks = c(seq(0,30,5)), labels = c(seq(0,30,5)), limits = c(0,30)) 
```
]

.pull-right[

**R-code**

```{r, eval=FALSE}
fun1 <- function(x) 10+(2*x) #<<

ggplot(func_x, aes(x, y)) + 
  geom_point(colour = "red", size = 2) +  #<<
  stat_function(fun = fun1) +             #<<
  scale_x_continuous(name = "x", 
                     breaks = c(0:10), 
                     labels = c(0:10), 
                     limits = c(0,10)) +
  scale_y_continuous(name = "y", 
                     breaks = c(seq(0,30,5)), 
                     labels = c(seq(0,30,5)), 
                     limits = c(0,30)) 
```
]


---
# Multiple arguments
+ Functions can take multiple arguments. Consider:

$$
y = f(x,z) = 10 + (x*z)
$$

+ Where:

$$
x = 
\begin{bmatrix}
1 \\ 
2 \\ 
3 \\ 
\end{bmatrix}
$$



$$
z = 
\begin{bmatrix}
1 \\ 
2 \\ 
3 \\ 
\end{bmatrix}
$$



---

# Multiple arguments

.pull-left[
```{r,echo=FALSE}
func_xz <- tibble(
  x = c(1,1,1,2,2,2,3,3,3),
  z = c(1,2,3,1,2,3,1,2,3)
)

kable(func_xz) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```
]

.pull-right[

+ Notice that when we have multiple inputs, our rows correspond to pairs of inputs.

+ So $x$ = 1, pairs with:
  + $z$ = 1
  + $z$ = 2
  + $z$ = 3
  
+ and so on for all values of $x$

]

---

# Multiple arguments

.pull-left[
```{r,echo=FALSE}
func_xz2 <- tibble(
  x = c(1,1,1,2,2,2,3,3,3),
  z = c(1,2,3,1,2,3,1,2,3),
  "f(x,z)" = 10 + (x*z)
)

kable(func_xz2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```
]

.pull-right[

$$
y = f(x,z) = 10 + (x*z)
$$

+ Example 1, row 2

$$
10 + (1*2) = 12
$$

+ Example, row 8

$$
10 + (3*2) = 16
$$

]

---
# Linear vs non-linear functions
+ Each of the examples so far have been linear functions.
    + If we plot them, we get a straight line (or flat surface)
    
+ Can also have non-linear functions:
    + A non-linear function would contain powers or roots
    

---
# Non-linear functions

.pull-left[
```{r, echo=FALSE}
fun2 <- function(x) 5+(x^2)
func_x2 <- tibble(
  x = c(1,2,3,4,5,6,7,8), 
  y = 5 + (x^2) 
)
ggplot(func_x2, aes(x, y)) + 
  geom_point(colour = "red", size = 2) +
  stat_function(fun = fun2) +
  xlim(0, 10) +
  ylab("y") +
  ylim(0, 125)
```

]

.pull-right[

**Example of non-linear function**

$$
y = f(x) = 5 + x^2
$$

]

---
# Why are functions important?
+ There are going to be lots of examples of functions in action.

--

+ Two primary examples are:
    + **Data transformations**
    + **Describing formal models**

--

+ We will start with transformations, and come back to models at the end of the course.


---
# z-scores
+ One of the most common transformations in data analysis is standardizing variables.

+ What is standardizing?

--

  + It is putting all variables onto the same scale so they can be compared.

--

+ We refer to standardized variables as $z$-scores (the reason we will explain later)

--

+ $z$-score:

$$z = \frac{x - \mu}{\sigma}$$

---
# Z-score for measured variable
+ $z$-score for $x$:

$$z_{x_i} = \frac{x_i - \bar{x}}{s_x}$$

+ Where
  + $x_i$ = individual score on $x$
  + $\bar{x}$ = mean of $x$
  + $s_x$ = standard deviation of $x$


---
# z-scores
+ A $z$-score will have a mean = 0, and a SD = 1.

--

+ What this means is there is a standard way to interpret $z$-scores.

--

  + $z$-score = 1.5, means a respondent is 1.5 SD above the mean.

--

  + $z$-score = -2, means a respondent is 2 SD below the mean.


---
# Summary of today
+ Functions take input, do something, and produce an output.

+ Functions can have multiple arguments, be linear or non-linear

+ Typically we will visualize functions

+ We use functions frequently in statistics. 

+ In fact almost everything we are going to see involves functions.


---
# This week

<script src="https://cdn.jsdelivr.net/npm/iconify-icon@2.1.0/dist/iconify-icon.min.js"></script>

.pull-left[
<iconify-icon icon="clarity:tasks-solid" width="64" height="64"  style="color: #0F4C81"></iconify-icon>

## Tasks

- Attend lectures and lab

- Complete the weekly quiz
    + Opens Monday at 9am
    + Closes Sunday at 5pm
    
- Submit Formative Report A by 12 noon on Friday 17th Oct 2025 (see instructions in the lab materials)
 
- The Thursday 10am lab will be in room 40 George Square LG.07 next week, rather than Appleton Tower M2 (where all the other labs will be).

<!-- - Submit Formative Report B by 12 noon on Friday the 29th of November 2024 -->
<!-- - Submit Formative Report C by 12 noon on Friday the 14th of February 2025 -->
<!-- - Submit the Assessed Report by 12 noon on Friday the 28th of March 2025 -->
]


.pull-right[
<iconify-icon icon="raphael:help" width="64" height="64"  style="color: #0F4C81"></iconify-icon>

## Support

- **Office hours**: for one-to-one support on course materials or assessments<br>(see LEARN > Course information > Course contacts)

- **Piazza**: help each other on this peer-to-peer discussion forum

- **Student Adviser**: for general support while you are at university<br>(find your student adviser on MyEd/Euclid)
]


 