---
title: "Numeric data"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
```


Consider the dataset available at <https://uoepsy.github.io/data/RestaurantTips.csv>, containing 157 observations on the following 7 variables:

```{r, echo=FALSE}
library(tidyverse)
library(kableExtra)

tribble(
    ~'Variable Name', ~'Description',
    'Bill', 'Size of the bill (in dollars)',
    'Tip', 'Size of the tip (in dollars)',
    'Credit', 'Paid with a credit card? n or y',
    'Guests', 'Number of people in the group',
    'Day', 'Day of the week: m=Monday, t=Tuesday, w=Wednesday, th=Thursday, or f=Friday',
    'Server', 'Code for specific waiter/waitress: A, B, or C',
    'PctTip', 'Tip as a percentage of the bill'
) |>
    kbl() |>
    kable_styling(full_width = FALSE)
```

These data were collected by the owner of a bistro in the US, who was interested in understanding the tipping patterns of their customers. The data are adapted from @lock2020.

```{r}
library(tidyverse)
tips <- read_csv("https://uoepsy.github.io/data/RestaurantTips.csv")
head(tips)
glimpse(tips)
```

```{r}
tips$Day <- factor(tips$Day, 
                   levels = c("m", "t", "w", "th", "f"),
                   labels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))

tips$Credit <- factor(tips$Credit, 
                      levels = c("n", "y"),
                      labels = c("No", "Yes"))

tips$Server <- factor(tips$Server)

glimpse(tips)
```

```{r}
tips <- tips |>
    mutate(
        PctTip = ifelse(PctTip > 100, NA, PctTip)
    )
```

We can create a histogram of tips via:

```{r}
ggplot(tips, aes(x = Tip)) + 
    geom_histogram(color = 'white') + 
    labs(x = "Size of the tip (US dollars)")
```

We can create a single figure with the distribution of all numeric variables by using the patchwork package:

::: {.column-margin}
If you save each plot into an object with a name, e.g. `p1`, `p2`, `p3`, `p4`, you can arrange the four plots into 2 by 2 panels as follows:

```
library(patchwork)
(p1 | p2) / (p3 | p4)
```

To do 1 row of 4 plots:

```
p1 | p2 | p3 | p4
```
:::

```{r}
#| column: page
#| fig-height: 6
#| fig-width: 8

library(patchwork)

pltBill <- ggplot(tips, aes(x = Bill)) +
    geom_histogram(color = 'white', fill = 'lightblue') +
    labs(x = "Size of the bill (US dollars)")

pltTip <- ggplot(tips, aes(x = Tip)) +
    geom_histogram(color = 'white', fill = 'lightblue') +
    labs(x = "Size of the tip (US dollars)")

pltGuests <- ggplot(tips, aes(x = Guests)) +
    geom_bar(fill = 'lightblue') +
    labs(x = "Number of people in the group")

pltPctTip <- ggplot(tips, aes(x = PctTip)) +
    geom_histogram(color = 'white', fill = 'lightblue') +
    labs(x = "Tip as a percentage of the bill")

(pltBill | pltTip) / (pltGuests | pltPctTip)
```

To summarise one numeric variable, you can use the `summarise` function from tidyverse, which takes the data and computes a numeric summary. The syntax is:

```
data |>
    summarise(
        write_the_column_name = computation
    )
```

This computes the mean and SD of tip size (in US dollars), and calls the column storing the mean `M`, and the column storing the standard deviation `SD`:

```{r eval=F}
tips |>
    summarise(
        M = mean(Tip, na.rm = TRUE),
        SD = sd(Tip, na.rm = TRUE)
    ) |>
    kbl(digits = 2, booktabs = TRUE)
```

```{r echo=F}
tips |>
    summarise(
        M = mean(Tip, na.rm = TRUE),
        SD = sd(Tip, na.rm = TRUE)
    ) |>
    kbl(digits = 2) |>
    kable_styling(full_width = FALSE)
```

To summarise all of the numeric variables into a single table of descriptive statistics you can use the `describe` function from the psych package:

```{r eval=F}
library(kableExtra)
library(psych)

tips |>
    select(Bill, Tip, Guests, PctTip) |>
    describe() |>
    kbl(booktabs = TRUE, digits = 2)
```

```{r echo=F}
library(kableExtra)
library(psych)

tips |>
    select(Bill, Tip, Guests, PctTip) |>
    describe() |>
    kbl(digits = 2) |>
    kable_styling(full_width = FALSE)
```

To only keep the sample size, mean, SD, median, min, max, we use the `select` function from tidyverse:

```{r eval=F}
library(kableExtra)
library(psych)

tips |>
    select(Bill, Tip, Guests, PctTip) |>
    describe() |>
    select(n, mean, sd, median, min, max) |>
    kbl(booktabs = TRUE, digits = 2)
```

```{r echo=F}
library(kableExtra)
library(psych)

tips |>
    select(Bill, Tip, Guests, PctTip) |>
    describe() |>
    select(n, mean, sd, median, min, max) |>
    kbl(digits = 2) |>
    kable_styling(full_width = FALSE)
```


:::{.callout-warning}
#### Use the appropriate summary for each variable type

Ensure that you summarise variables correctly:

- For categorical variables use frequency tables
- For continuous variables use a table of descriptive statistics (mean, SD, Median, etc.)

You should not summarise categorical variables with the mean, SD, and this is why it's important to use `select()` before `describe()` to only keep the variables that are continuous.

:::


:::{.callout-tip}
#### Example writeup

The distributions of bill size, tip size, and group size are skewed to the right. The distribution of tips, as a percentage of the total bill, appears to be approximately bell shaped, with three outliers in the right tail of the distribution.  
The average bill was about \$22.73, with a SD of \$12.16. The average tip was \$3.81, with a SD of \$2.37, corresponding to an average tip as a percentage of the total bill of \$16.59, with a SD of \$4.39. The average party size comprised 2 guests, with a SD of roughly 1 person.
:::

