---
title: "Power"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
```


To compute power, you need to know the distribution of the sample mean 

- when $H_0$ is true
- when $H_1$ is true

This seldom happens in practice, so these exercises will be more conceptual.

Before going ahead, though, remember that if the population has mean $\mu$ the sampling distribution of the mean is $N(\mu, \frac{\sigma}{\sqrt n})$, where $\sigma$ is the population SD and $n$ the size of a sample.

Suppose you are testing

$$H_0 : \mu = 0$$
$$H_1 : \mu > 0$$



What is the power of the test if:

- the population has a mean $\mu = 5$
- the population has a standard deviation $\sigma = 3$
- you will take a sample of size 30


The distribution of the sample mean in the population is 

$$\bar X \sim N(5, \frac{3}{\sqrt{30}})$$


If $H_0$ is true, however, the distribution would be:

$$\bar X \sim N(0, \frac{3}{\sqrt{30}})$$

```{r echo=FALSE}
SE <- 3 / sqrt(30)
par(mfrow = c(2,1), mai = c(1,.3,.2,.5) * 1)
curve(dnorm(x, 0, SE), -3, 3, xlim = c(-4, 10), col = 'dodgerblue', lwd = 2,
      frame.plot = F, axes = F, ylab = "", xlab = "")
text(x = 2, y = 0.1, label = "H0 true")
axis(1)
curve(dnorm(x, 5, SE), 2, 8, xlim = c(-4, 10), col = 'darkorange', lwd = 2,
      frame.plot = F, axes = F, ylab = "", xlab = "")
text(x = 7, y = 0.1, label = "H1 true")
axis(1)
```


You would reject the null hypothesis, at the 5% significance level, for values larger the 0.95 quantile of a $N(0, \frac{3}{\sqrt{30}})$ distribution:

```{r}
sigma <- 3
n <- 30
SE <- 3 / sqrt(30)
```

```{r}
tstar <- qnorm(0.95, mean = 0, sd = SE)
tstar
```


```{r echo=FALSE}
par(mfrow = c(2,1), mai = c(1,.3,.2,.5) * 1)
curve(dnorm(x, 0, SE), -3, 3, xlim = c(-4, 10), col = 'dodgerblue', lwd = 2,
      frame.plot = F, axes = F, ylab = "", xlab = "")
text(x = 2, y = 0.1, label = "H0 true", col = 'dodgerblue')
axis(1)
abline(v = 0.90, col = 'red')

curve(dnorm(x, 5, SE), 2, 8, xlim = c(-4, 10), col = 'darkorange', lwd = 2,
      frame.plot = F, axes = F, ylab = "", xlab = "")
text(x = 7, y = 0.1, label = "H1 true", col = 'darkorange')
axis(1)
abline(v = 0.90, col = 'red')
text(x = 0.90, y = 0.7, label = '0.90', col = 'red', adj = -0.2)
```

The power is the probability of rejecting the null hypothesis when the null is false.

So, if the null is false, the mean follows the distribution $N(5, \frac{3}{\sqrt{30}})$. The probability of correctly rejecting the null would be the probability under that distribution beyond -1.07

```{r}
Pow <- 1 - pnorm(0.90, mean = 5, sd = SE)
Pow
```

The power is 1.


::: {.callout-note}
__Question 9.__ Assume now that the true population mean is $\mu = 2$. What is the power of the test?
:::

```{r echo=FALSE}
par(mfrow = c(2,1), mai = c(1,.3,.2,.5) * 1)
curve(dnorm(x, 0, SE), -3, 3, xlim = c(-4, 10), col = 'dodgerblue', lwd = 2,
      frame.plot = F, axes = F, ylab = "", xlab = "")
text(x = 2, y = 0.1, label = "H0 true", col = 'dodgerblue')
axis(1)
abline(v = 0.90, col = 'red')

curve(dnorm(x, 2, SE), -1, 5, xlim = c(-4, 10), col = 'darkorange', lwd = 2,
      frame.plot = F, axes = F, ylab = "", xlab = "")
text(x = 4, y = 0.1, label = "H1 true", col = 'darkorange')
axis(1)
abline(v = 0.90, col = 'red')
text(x = 0.90, y = 0.7, label = '0.90', col = 'red', adj = -0.2)
```

The distribution of the mean when $H_0$ is false is then

$$\bar X \sim N(2, \frac{3}{\sqrt{30}})$$

We need to compute the probability in a $N(2, \frac{3}{\sqrt{30}})$ distribution __to the right__ of 0.90.

```{r}
Pow <- pnorm(0.90, mean = 2, sd = SE, lower.tail = FALSE)
Pow
```

or

```{r}
Pow <- 1 - pnorm(0.90, mean = 2, sd = SE)
Pow
```

The power, i.e. the probability that the test rejects the null hypothesis when it is false, is 0.98.

::: {.callout-note}
__Question 10.__ Assume now that the true population mean is $\mu = 2$. What is the probability of a Type II error?
:::

The easiest way to compute it is as $\beta = 1 - \text{Power}$ because $\text{Power} = 1 - \beta$.

```{r}
beta <- 1 - Pow
beta
```

The probability of not rejecting the null when it is false is 0.02.

If you want to compute it from scratch you can compute it as the probability __to the left__ of 0.90. This is because values below 0.90 lead to not rejecting the null hypothesis.

```{r}
beta <- pnorm(0.90, mean = 2, sd = SE)
beta
```

