---
title: "<b>Semester 2 welcome</b>"
subtitle: "<small>Data Analysis for Psychology in R 1<br>Semester 2, Week 1</small>"
author: "<br>"
institute: "Department of Psychology<br/>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: jk_libs/libs
    css: 
      - un-xaringan-themer.css
      - jk_libs/tweaks.css
    nature:
      beforeInit: "jk_libs/macros.js"
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(digits=4,scipen=2)
options(knitr.table.format="html")
xaringanExtra::use_xaringan_extra(c("tile_view","animate_css","tachyons"))
xaringanExtra::use_extra_styles(
  mute_unhighlighted_code = FALSE
)
library(knitr)
library(tidyverse)
library(ggplot2)
knitr::opts_chunk$set(
  dev = "svg",
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
knitr::opts_chunk$set(fig.asp=.9)
#source('R/myfuncs.R')
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro"),
  outfile = "un-xaringan-themer.css"
)
```

class: inverse, center, middle
# Semester 1 recap  

---

.pull-left[

## S1 weeks 1-5  

- Lecturer: Dr Patrick Sturt

- **Block 1: Exploratory Data Analysis**  

    - Collecting Data & Types of Data  

    - Describing and Visualising Categorical Variables   

    - Describing and Visualising Numeric Variables  

    - Describing and Visualising Relationships  

    - Types of Relations (functions)  

- Labs: **Formative report A**

- Week 6: Formative Report A solutions & feedback

    - You should have studied the example Rmd and PDF solutions
]


.pull-right[
## S1 weeks 7-11  

- Lecturer: Dr Alex Doumas

- **Block 2: Probability**  

    - Probability Theory  

    - Probability Rules  

    - Discrete Random Variables   

    - Continuous Random Variables  

    - Sampling Distributions  

- Labs: **Formative report B**

- Week 12: Formative Report B solutions & feedback

    - You should have studied the example Rmd and PDF solutions
]


---

class: inverse, center, middle

# A map of semester 2

---

.pull-left[
## S2 weeks 1-5

- Lecturer: Dr Umberto Noè, FHEA

- **Block 3: Inference**. How do we use data from a sample (i.e. a subset of the population) to draw conclusions for the wider population (for which we don't have the full data)?

    - Confidence Intervals

    - Hypothesis Testing: P-values  

    - Hypothesis Testing: Critical Values  

    - Hypothesis Testing & Confidence Intervals

    - Making Decisions (Effect Sizes, Power, Errors) 

- Labs: **Formative Report C**

- FLW: Formative Report C solutions & feedback

]

.pull-right[

## S2 weeks 6-10

- Lecturer: Dr Emma Waterston, FHEA

- **Block 4: Common hypothesis tests**  

    - One Sample t-test  

    - Independent Samples t-test  

    - Paired Samples t-test

    - Chi-square tests

    - Covariance and correlation

- Labs: **Assessed Report**

- Week 11: Online exam prep session, will be recorded. Details and link will be sent in an announcement!
]


---

# Formative report C (S2 Weeks 1-5)

- Group-based report, as in the previous practice reports

- This is the last practice report

- **Due: 12 noon, Friday 13th February 2026**

- You must have joined a group by the end of week 3. No group changes allowed after week 3.

---

# Assessed report (S2 Weeks 6-10)

- The assessed report is worth 30% of the final DAPR1 grade

- To be done in groups (same groups as Formative Report C)

- Only one person per group submits on behalf of the entire group

- No extensions allowed as the extra time will be mainstreamed

- **Due: 12 noon, Friday 27th March 2026**

- Once the report has been graded, **each student's grade will be adjusted (up or down) based on their contribution as rated by their peers.** 


---

# Exam information (60% of final grade)

- Main Exam Diet (April/May). Exact date/location will be announced by the timetabling unit, instructors don't know in advance.

- Format:

    + In-person written exam in an exam hall, closed-book
    + 2 hours long
    + 60 multiple choice questions (MCQs)

- Bring (in a transparent bag):

    - Student ID card for identification.
    - A calculator (more info will be distributed later in the course)
    - Pencils
    - Eraser
    - Pencil sharpener

- You will be provided with:

    - The exam questions paper with an equation sheet
    - An MCQ answer sheet
    - Graph paper for rough work

- You **cannot** take any of the papers above outside the room.


---

# Final Exam Information

**In Flexible Learning Week, I will release a folder on LEARN called "Revision Resources".** By the end of the course, this will contain:

- Revision versions of the weekly quizzes for your to re-take as many times as you want

- Final Exam Information:

    - Equation sheet
        
        - Familiarise yourself with the equation sheet, you will be provided with a copy in the exam hall

    - Mock exam

        - A mock exam paper will be added to this folder during the online exam preparation session of week 11

    - Calculators information
    
        - Guidance on which calculators you can take into the exam

---

class: inverse, center, middle

