---
title: "<b>Semester 2 Welcome</b>"
subtitle: "<small>Data Analysis for Psychology in R 1<br>Semester 2, Week 1</small>"
author: "<br>"
institute: "Department of Psychology<br/>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: jk_libs/libs
    css: 
      - xaringan-themer.css
      - jk_libs/tweaks.css
    nature:
      beforeInit: "jk_libs/macros.js"
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(digits=4,scipen=2)
options(knitr.table.format="html")
xaringanExtra::use_xaringan_extra(c("tile_view","animate_css","tachyons"))
xaringanExtra::use_extra_styles(
  mute_unhighlighted_code = FALSE
)
library(knitr)
library(tidyverse)
library(ggplot2)
knitr::opts_chunk$set(
  dev = "svg",
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
knitr::opts_chunk$set(fig.asp=.9)
#source('R/myfuncs.R')
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

class: inverse, center, middle
# Semester 1 Recap  

---

.pull-left[

## S1 Weeks 1 - 5  

- Collecting Data & Types of Data  

- Describing and Visualising Categorical Variables   

- Describing and Visualising Numeric Variables  

- Describing and Visualising Relationships  

- Types of Relations (functions)  

]

--

.pull-right[
## S1 Weeks 7 - 11  

- Probability Theory  

- Probability Rules  

- Discrete Random Variables   

- Continuous Random Variables  

- Sampling Distributions  


]

---

# Things we hope you did last semester

.pull-left[

### Engaging with course materials

- Complete the weekly quiz

- Watch lectures and complete the lecture activities

- Attempt the weekly exercises

- Compare your attempt with the solutions each time

- Attend labs and office hours 

- Write your questions on the discussion board

] 
.pull-right[

### RStudio skills

- Basic arithmetic operations   

- Rstudio functionality 

- Data handling in R 

- RMarkdown 

- ggplot!

]

---

class: inverse, center, middle, animated, rotateInDownLeft

# A Map of Semester 2

---

.pull-left[

## S2 Weeks 1 - 5

- Confidence Intervals and Bootstrapping

- Hypothesis Testing: P-values  

- Hypothesis Testing: Critical Values  

- Hypothesis Testing & Confidence Intervals

- Making Decisions (Effect Sizes, Power, Errors) 

]
.pull-right[

## S2 Weeks 7 - 11

- One Sample t-Test  

- Independent Samples t-test  

- Paired Samples t-test

- Chi-square test

- Covariance and correlation


]


---

# Things that will be the same

- Quizzes will be delivered in the same mode as Semester 1

- Lecture videos will be delivered online in the same mode as Semester 1

- Live R session will be delivered online in the same mode as Semester 1.

- Office hours will be held online in GatherTown (see link on the left-hand side menu of Learn)
  
  + Change in time:  
  __Tuesdays 15:00-17:00__ with Alex Doumas  
  __Thursdays 13:30-15:30__ with senior teaching coordinators


---

# Changes

- From now on, __all labs__ will be held __in person__.

- You must attend only the lab session that is in your personal timetable. 

- Do __not__ attend lab sessions you were not allocated to.

- You can put in a formal request to change allocation on Learn, but you must wait until this change is confirmed before going to the new lab.


---


class: inverse, center, middle, animated, rotateInDownLeft


