---
title: "Formatting plots using ggplot2"
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
```

## Introduction

In psychology and, more generally, scientific research, a plot is not
just for decoration. It represents the evidence for your hypotheses and
conclusions. A good plot should be easy to read and the style should be
consistent across your dissertation, report, or paper.

In this reading, you will learn how to control the most useful styling
options in `ggplot2`:

- colour and fill
- themes
- font size
- labels and legends
- colour palettes that are accessible

::: callout-tip
While we go through these settings, keep the ggplot2 package cheatsheet
handy: <https://rstudio.github.io/cheatsheets/data-visualization.pdf>
:::

### Recap: the "recipe" for a ggplot

Every ggplot plot is built by adding layers together with the `+`
symbol. You will see this same recipe in every example below:

1.  `ggplot(data, aes(...))`: specifies the data and which variables to
    plot
2.  `geom_??()`: which type of plot (bar, histogram, density, boxplot)
3.  `labs()`: title and x/y axis labels.
4.  `theme_??()`: the overall look, and font size.

Changing the style of a plot almost always means adding to, or editing,
one of these layers, rather than rewriting the whole thing.

### Choosing the right plot for your variables

Before you style a plot, make sure you have picked an appropriate plot
type for your variables.

- **Categorical variable** (e.g. treatment group): use a bar plot
  (`geom_bar()`) to show counts in each category.
- **Numeric variable, one group** (e.g. all scores together): use a
  histogram (`geom_histogram()`) or a density plot (`geom_density()`) to
  show the shape of the distribution.
- **Numeric variable, comparing groups**: use a boxplot
  (`geom_boxplot()`) to compare medians and spread across groups at a
  glance. You can also use histograms or density plots with separate
  colours for each group.

## Load packages and data

We will use the built-in `sleep` dataset that ships with the basic R
installation. This is a classic dataset from a small clinical trial: 20
patients were given either a placebo or a sleep drug, and researchers
recorded how much extra sleep (in hours) each patient got. It has just
three variables, which keeps things simple while we focus on styling:

- `extra`: a numeric variable, the increase in hours of sleep.
- `group`: a categorical variable coded 1 and 2 indicating,
  respectively, if the patient received the `Placebo` or the `Drug`.
- `ID`: a numeric variable, the patient ID (1-10 for each group).

```{r}
library(tidyverse)

glimpse(sleep)

# Recode the group variable to a factor with better labels
# Store the resulting data in a new object called sleep_data
sleep_data <- sleep |>
  mutate(
    group = factor(group, 
                   levels = c(1, 2), 
                   labels = c("Placebo", "Drug"))
  )

glimpse(sleep_data)
```

## Core idea: colour vs fill

This is the most important styling rule to remember:

- use `colour` for lines and points
- use `fill` for inside areas (bars in barplots or histograms, areas
  under density plots, interior of boxplots)

If you mix these up, your plot may not change in the way you expect.

::: callout-tip
Bookmark the following page, as it lists the name of all colours in R:
<https://r-charts.com/colors>
:::

### Specifying inside vs outside `aes()`

You will see `colour` and `fill` used in two different ways, and it is
easy to confuse them:

- **Inside `aes()`**, e.g. `aes(fill = group)`: this maps a variable in
  your data to colour, so each group automatically gets a different
  colour, and a legend appears.
- **Outside `aes()`**, e.g. `geom_bar(fill = "tomato4")`: this assigns
  every bar to the same fixed colour, regardless of group. No legend is
  created, because there is no variability in colours.

As a rule of thumb: specify inside `aes()` when colour should depend on
your data, and specify outside when you just want one fixed colour for
the whole plot.

## 1. Bar plots

Bar plots are for categorical variables, like our `group` variable. The
height of each bar shows how many observations fall into that category.

### Basic bar plot

```{r}
ggplot(sleep_data, aes(x = group)) +
  geom_bar()
```

We can improve the clarity of the plot by adding a title and axis
labels. The `labs()` function is used to change the labels of the title
and x/y axes. The labels must be specified within quotation signs
`"..."` as they represent a short text sentence.

```{r}
ggplot(sleep_data, aes(x = group)) +
  geom_bar() +
  labs(
    title = "Participants per Treatment Group",
    x = "Treatment group",
    y = "Count"
  )
```

### Change fill and border colour

The option `fill` changes the colour of the bars (i.e. an area). In this
case, it is set to cornflower blue.

```{r}
ggplot(sleep_data, aes(x = group)) +
  geom_bar(fill = "cornflowerblue") +
  labs(
    title = "Participants per Treatment Group",
    x = "Treatment group",
    y = "Count"
  )
```

The `colour` option, instead, sets the line colour to black. In this
case, the line is just the border of the bars.

```{r}
ggplot(sleep_data, aes(x = group)) +
  geom_bar(fill = "cornflowerblue", colour = "black") +
  labs(
    title = "Participants per Treatment Group",
    x = "Treatment group",
    y = "Count"
  )
```

### Change line thickness

The `linewidth` option controls the thickness of a line, and it is set
here to 1. The higher the value, the thicker the line will be.

```{r}
ggplot(sleep_data, aes(x = group)) +
  geom_bar(fill = "cornflowerblue", colour = "black", linewidth = 1) +
  labs(
    title = "Participants per Treatment Group",
    x = "Treatment group",
    y = "Count"
  )
```

### Map fill to group

We can also tell ggplot to change the colour or fill depending on the
values within a variable in the data. In this case we tell R to change
the colours of the bars (`fill`) depending on the values of the `group`
variable. This is done by putting `fill = group` inside the `aes()`
function. ggplot will automatically assign a different colour to each
group and add a legend to the plot.

```{r}
ggplot(sleep_data, aes(x = group, fill = group)) +
  geom_bar(colour = "black") +
  labs(
    title = "Participants per Treatment Group",
    x = "Treatment group",
    y = "Count",
    fill = "Treatment group"
  )
```

If you don't like the default colours, you can change them using a
different colour palette. The `scale_fill_brewer()` function allows you
to choose from a range of colour palettes - see here for all the
available ones:
<https://ggplot2-book.org/scales-colour.html#brewer-scales>

For example, we can use the "Set2" palette by specifying the palette
name in the argument `palette = ...` of the `scale_fill_brewer()`
function.

```{r}
ggplot(sleep_data, aes(x = group, fill = group)) +
  geom_bar(colour = "black") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Participants per Treatment Group",
    x = "Treatment group",
    y = "Count",
    fill = "Treatment group"
  )
```

It's worth noting that in this case the different colours don't add any
extra information and are not needed. The two groups are already clearly
distinguishable from the x-axis labels "Placebo" and "Drug" underneath
each bar.

## 2. Histograms

Histograms are for a single numeric variable. They group values into
bins and show how many observations fall into each bin, giving you a
picture of the overall shape of the distribution.

The basic histogram is:

```{r}
ggplot(sleep_data, aes(x = extra)) +
  geom_histogram()
```

Important styling decisions for histograms include:

- `bins` controls how smooth or rough the picture looks. If you are
  unsure how many bins to use, leave this out as it will use the default
  `bins = 30` which works most of the time.
- Border colour (in `colour`) can make bars easier to distinguish.
- Keep a sensible contrast between fill and background.

We can improve the previous plot by changing the number of bins from the
default (30) to 10, as there are too many gaps. Next, we can change the
colour of the bars to purple and the border to black. Finally, we
improve the axes labels to be more user-friendly.

```{r}
ggplot(sleep_data, aes(x = extra)) +
  geom_histogram(bins = 10, fill = "purple", colour = 'black') +
  labs(
    title = "Distribution of change in sleep time",
    x = "Change in sleep time (hours)",
    y = "Frequency"
  )
```

## 3. Density plots

A density plot is a smoothed version of a histogram.

### Line-only density plot

A density plot is mainly based on a line, so you can control its colour
using the `colour` argument. You can also control the thickness of the
line using the `linewidth` argument.

```{r}
ggplot(sleep_data, aes(x = extra)) +
  geom_density(colour = "darkgreen", linewidth = 2) +
  labs(
    title = "Density of change in sleep time",
    x = "Change in sleep time (hours)",
    y = "Density"
  )
```

### Filled density by group

The area under the curve can be coloured using the `fill` argument.

```{r}
ggplot(sleep_data, aes(x = extra)) +
  geom_density(colour = "darkgreen", fill = 'lightgreen', linewidth = 2) +
  labs(
    title = "Density of change in sleep time",
    x = "Change in sleep time (hours)",
    y = "Density"
  )
```

### Filled density by group

As you add more plots to the same figure, e.g. if you compare density
curves for two groups, you can make the curves easier to distinguish by
making them slightly transparent.

You control the transparency using the `alpha` argument, which takes a
value between 0 (fully transparent) and 1 (fully opaque).

```{r}
ggplot(sleep_data, aes(x = extra, fill = group, colour = group)) +
  geom_density(alpha = 0.5, linewidth = 2) +
  labs(
    title = "Change in sleep time density by group",
    x = "Change in sleep time (hours)",
    y = "Density",
    fill = "Treatment group",
    colour = "Treatment group"
  )
```

## 4. Boxplots

Boxplots show the median (thicker line in the middle), the central 50%
of scores (the box), and the range of typical values (the lines
extending on either side of the box). Here, colour affects the lines,
while fill affects the inside of the box.

Useful options in `geom_boxplot()`:

- `fill` sets the interior colour of the box.
- `colour` sets the box outline colour.
- `width` controls box width. If unsure, you can leave this out and
  ggplot will use a default width that works most of the time.

```{r}
ggplot(sleep_data, aes(x = group, y = extra, fill = group)) +
  geom_boxplot(fill = 'lightpink', colour = 'purple4', width = 0.2) +
  labs(
    title = "Change in sleep time by treatment group",
    x = "Treatment group",
    y = "Change in sleep time (hours)"
  )
```

## Themes and font sizes

Themes control the overall look (background, grid lines, text style).
The default is `theme_grey()`, but you can select any of the available
options. Try them:

- `theme_grey()` is the default
- `theme_bw()`
- `theme_light()`
- `theme_dark()`
- `theme_classic()`
- `theme_minimal()`

```{r}
ggplot(sleep_data, aes(x = group, y = extra, fill = group)) +
  geom_boxplot(fill = 'lightpink', colour = 'purple4', width = 0.2) +
  labs(
    title = "Change in sleep time by treatment group",
    x = "Treatment group",
    y = "Change in sleep time (hours)"
  ) +
  theme_minimal(base_size = 14)
```

Inside of each of the themes, you can change the font size using the
`base_size` argument. For example, `theme_minimal(base_size = 14)` sets
the font size to 14 points. This is important to ensure your figures are
readable when printed or published online.

For your dissertation, paper, or report, you want to only use one theme
consistently across all figures.

## Custom theme edits

You can customise individual parts of the plot using `theme()`.

```{r}
ggplot(sleep_data, aes(x = group, y = extra, fill = group)) +
  geom_boxplot(fill = 'lightpink', colour = 'purple4', width = 0.2) +
  labs(
    title = "Change in sleep time by treatment group",
    x = "Treatment group",
    y = "Change in sleep time (hours)"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 14),
    legend.position = "none"
  )
```

Commonly used options include:

- `plot.title`: controls the title font size and style.
- `axis.title`: controls the axis label font size and style.
- `axis.text`: controls the axis tick label font size and style.
- `legend.title`: controls the legend title font size and style.
- `legend.text`: controls the legend text font size and style.
- `legend.position`: controls where the legend appears ("none", "left",
  "right", "bottom", "top", "inside").

## Accessibility and good practice

For your dissertation, paper, or report:

1.  Do not rely on colour alone to communicate key differences.
2.  Use colour palettes with good contrast.
3.  Keep text large enough to read when printed or shared online.
4.  Keep styling consistent across all plots.
5.  Avoid overly bright or decorative colours.

A helpful option is the viridis palette family:

```{r}
ggplot(sleep_data, aes(x = group, y = extra, fill = group)) +
  geom_boxplot() +
  scale_fill_viridis_d(option = "plasma", end = 0.9) +
  labs(
    title = "Change in sleep time by treatment group",
    x = "Treatment group",
    y = "Change in sleep time (hours)"
  ) +
  theme_light(base_size = 13)
```

`scale_fill_viridis_d()` applies a discrete Viridis colour scale (the
`d` stands for discrete), so each category gets a distinct colour from a
palette designed to be perceptually uniform and colourblind-friendly.

`option = "plasma"` chooses the "plasma" variant (one of the Viridis
palette options), giving a specific sequence of colours while keeping
good contrast and accessibility.

## Save your plots

Use `ggsave()` to export high-quality figures for reports.

```{r}
#| eval: false
p <- ggplot(sleep_data, aes(x = group, y = extra, fill = group)) +
  geom_boxplot() + 
  labs(
    x = "Treatment group", 
    y = "Change in sleep time (hours)"
  ) +
  theme_minimal(base_size = 12)

ggsave("images/sleep-boxplot.png", plot = p, width = 7, height = 5, dpi = 300)
```

- Save into your `images` folder to keep your project organised.
- `plot = p` specifies which plot to save (if you have multiple plots in
  the same R session).
- `width` and `height` control the size of the saved image in inches,
  play around with the values to get the best fit for your publication.
- `dpi = 300` is the resolution and 300 is appropriate for most
  publication-ready images.

## Quick checklist

- Is the plot appropriate for the variable type?
- Are axis labels clear and informative?
- Is text readable at normal zoom/print size?
- Are colours accessible and consistent across figures?
- Is the theme consistent with your other plots?
