---
title: "Central Tendency"
subtitle: Data Analysis for Psychology in R 1
author: Tom Booth
---

```{r premable, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(kableExtra)
```

## Today
- Introduce frequency distribution
- Discuss measures of central tendency
    - Mode
    - Median
    - Mean

## Learning objectives
- Understand visualizations of frequency distributions.
- Be able to define the mode, median and mean.
- Understand how to calculate these values by hand and in R.
- Understand which statistic to use for different types of data.

## Frequency Distribution

*"a mathematical function showing the number of instances in which a variable takes each of its possible values."*

## Simple example

```{r, echo=FALSE}
set.seed(1066)
df <- tibble(
    dat = factor(rbinom(10, 1, 0.5))
)
bp <- df %>%
    ggplot(., aes(x=dat)) +
    geom_bar() +
    labs(x = "Coin toss result", y = "Count") +
    scale_x_discrete(labels = c("Heads", "Tails"))


bp
```

- I toss a coin 10 times, and I plot the number of instances of head and tails.

## Graphs 
- There are many types of graph, all of which are intended to summarise different types of data, or the relations between data.
- We are going to spend a lot of time dealing with plots this semester.
- Let's pause for a moment and consider the basic structure

## Structure of graphics
- Data
- Axes 
    - (aesthetics in `ggplot` language - also includes size, colours etc.)
- Axes have scales
- Geometric objects 
    - (`geoms` in `ggplot` language; i.e. bars, points, lines)
    
## Bar plot
```{r, echo = FALSE}
bp
```

- Bar plots are used to show frequency distributions of categorical variables.

## Bar plot
- Our example:
    - Data = result of our 10 coin tosses.
    - X-axis: values of variable (here "Heads" or "Tails)
        - Scale = two discrete values
    - Y-axis: Frequency
        - Scale = count 

## Our example: Bar plot
```{r, echo = FALSE}
bp
```

## Histogram

```{r, echo=FALSE}
set.seed(1984)
df2 <- tibble(
    dat = rnorm(225, 155, 15)
)
hp <- df2 %>%
    ggplot(., aes(x=dat)) +
    geom_histogram(bins = 20) +
    labs(x = "Height (cm)", y = "Count") +
    scale_x_continuous(breaks = c(120, 130, 140, 150, 160, 170, 180, 190, 200))

hp
```


- A histogram is used to show the frequency distribution for continuous variables

## Histogram
- Our example:
    - Data = heights of the class
    - X-axis: possible values of height.
        - Commonly presented in "bins" 
        - A bin represents a range of scores (plot can look very different dependent on this)
        - Scale = dependent on the form of measurement, here centimetres 
    - Y-axis: frequency of a given value or values within "bins"

## Describing in numbers
- So how do we describe a frequency distribution numerically?
- We typically want to know something about the;
    - central point, and
    - the spread (or dispersion around centre point)
    
## Central tendency
- Central tendency measures are a set of statistics that describe the central point of data.
    - Or the central point of our frequency distributions.
- Measures of central tendency are typically referred to as "averages"
    - But this can have many meanings.

## Some data...
```{r}
ex1 <- tibble(
    ID = c(paste("ID", 101:250, sep = "")),
    Degree = factor(sample(c("Psych", "Ling", "Phil", "Joint"), 
                           150, replace=T, prob=c(.3,.3,.3,.1))),
    StatsScore = round(rnorm(150, 66, 7),0)
)
```

## Some data...
```{r}
ex1[1:10,]
```

## Some data...
```{r, echo=FALSE}
kableExtra::kable_styling(knitr::kable(ex1[1:10,]), font_size = 22)
```


##  Central Tendency: Mode 
- One way to consider the average value is the **mode.**
- **Mode**: The most commonly observed value in the data set.
- It is of course possible that we observe two (or more) values with equal frequency.
- We then have a **multi-modal** distribution.

##  Central Tendency: Mode 

```{r}
table(ex1$Degree)
```

##  Central Tendency: Median 
- A second measure of average observation is the **median.**
- **Median:** is the value for which half the data falls above, and half below the given value.

##  Central Tendency: Median 
- To calculate the median:
    - Rank order the data.
    - Find the middle value.

## Hand calculation
- When we have an odd number of observations:


$$
\frac{1}{2}(N+1)
$$

- [10,40,30,25,15,5,65]


## Hand calculation
- When we have an even number of observations:


$$
\frac{\frac{1}{2}N + (\frac{1}{2}N+1)}{2}
$$

- [10,40,25,15,5,30]


##  Median in R 

```{r}
ex1 %>%
    summarise(
        median = median(StatsScore)
    )
```


##  Central Tendency: Mean 
- A third measure of average observation is the **arithmetic mean.**
- **Mean:** is the sum of all values, divided by the total number of observations.
    - I.e. this is the average as most people think about the average.

$$
\bar{x} = \frac{\sum_{i=1}^{N}{x_i}}{N}
$$

##  Central Tendency: Mean 

$$
\bar{x} = \frac{\sum_{i=1}^{N}{x_i}}{N}
$$

- $\bar{x}$ = estimate of mean of variable x
- $x_i$ = individual values of x
- $N$ = sample size

## Hand calculation

$$
\bar{x} = \frac{\sum_{i=1}^{N}{x_i}}{N}
$$

- [10,40,30,25,15,5]

##  Central Tendency: Arithmetic Mean 

```{r}
ex1 %>%
    summarise(
        median = median(StatsScore),
        mean = mean(StatsScore)
    )
```

##  Strengths and weaknesses 

```{r, echo = FALSE}
tbl16 <- tibble::tribble(
~`Measure`, ~`Strength`, ~`Weakness`,
"Mode","Actually occurs in our data","Not algebraically calculable",
" ","Unaffected by extreme values","Probably does not exist for true continuous data (think reaction time)",
"Median","No assumptions about interval value of data","Not relatable to measures of dispersion (see next week)",
" ","Unaffected by extreme values"," ",
"Mean","Algebraically tractable","Sensitive to extreme values",
" ","Related to measures of dispersion (see next week)","Assumes data are interval or better",
" "," ","Possible no case in your data takes the value of the mean"
)

kableExtra::kable_styling(knitr::kable(tbl16), font_size = 22)
```

##  Which measure should we use? 

```{r, echo = FALSE}
tbl18 <- tibble::tribble(
~`Variable Type`, ~`Central Tendency`,
"Categorical (Nominal)","Mode",
"Categorical (Ordered)","Mode/Median",
"Continuous","Mean (any in fact)",
"Count","Mode (mean)"
)

kableExtra::kable_styling(knitr::kable(tbl18), font_size = 26)
```

- Depends on the level of measurement.

## Tasks for this week...
1. Finish any existing tasks from lab 3.
2. Reading: See LEARN
3. Quiz 4
    - **This quiz counts**
    - Live now (as of Monday at 09:00).
    - Closes Sunday at 17:00

## Recommendations of the week
- Podcast: [History of Rome](https://thehistoryofrome.typepad.com/) (find it on usual podcast site)
- Recipe: [Cullen Skink](https://www.bbcgoodfood.com/recipes/899633/cullen-skink) 
- Book: *Lucky Jim* by Kingsley Amis (In Main Library)
- Coffee/food Place: [Bearded Barista](https://twitter.com/beardedbarista_?lang=en)

