---
title: "Variability"
subtitle: Data Analysis for Psychology in R 1
author: Tom Booth
---

```{r premable, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(kableExtra)
library(cowplot)
library(modes)
```

## Today
- Last lecture we described central tendency. 
- Today we discuss variability or dispersion.
  	- Range (minimum, maximum, inter-quartile, decile)
  	- Summed deviations
  	- Absolute deviations
  	- Variance
  	- Standard deviation

## Learning objectives
- Be able to describe the 5 measures of variability.
- Understand to what data we can apply each measure of variability.
- Understand the pro's and con's of each measure.
- Develop an initial understanding of how to calculate in R.

## Data (n=150)

```{r, echo=FALSE}
set.seed(111019)
ex1 <- tibble(
    ID = c(paste("ID", 101:250, sep = "")),
    Degree = factor(sample(c("Psych", "Ling", "Phil", "Joint"), 
                           150, replace=T, prob=c(.3,.3,.3,.1))),
    StatsScore = round(rnorm(150, 66, 7),0)
)

kable(ex1[1:10,]) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

## Statistic Score Distribution

```{r echo=FALSE}
ex1 %>%
  ggplot(., aes(x=StatsScore)) +
  geom_histogram(bins = 20, color = "darkgrey", fill = "White")+
  xlab("Statistics Test Score") +
  ylab("Count \n")
```

## Range 
- The **range** of the data is simply the value between two points.
  - We can define these points in different ways.
- Range is generally used to describe the total range (max - min) of the data.


## Range 

```{r, eval = FALSE}
ex1 %>%
  select_if(is.numeric) %>%
  summarise(
    Variable = names(.),
    Sample = n(),
    Minimum = min(.),
    Maximum = max(.)) %>% 
  mutate(Range = Maximum - Minimum)
```

```{r echo=FALSE}
ex1 %>%
  select_if(is.numeric) %>%
  summarise(
    Variable = names(.),
    Sample = n(),
    Minimum = min(.),
    Maximum = max(.)) %>% 
  mutate(Range = Maximum - Minimum) %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, font_size = 20)
```

- Problem: Very sensitive to outliers.

##  Inter-quartile range 
- The **inter-quartile range (IQR)** is the difference between the 1st and 3rd quartile.
  - Rank the data
  - Split data into four equal blocks.
  - Quartiles are the points which divide these blocks.
  - They fall at 25%, 50% and 75% of rank ordered data.
  - IQR is the difference between 25% and 75%

##  Inter-quartile range 
```{r eval=FALSE}
iqr <- quantile(ex1$StatsScore)
res <- tibble(
  "0%" = iqr[[1]],
  "25%" = iqr[[2]],
  "50%" = iqr[[3]],
  "75%" = iqr[[4]],
  "100%" = iqr[[5]]
)
```


```{r, echo=FALSE}
iqr <- quantile(ex1$StatsScore)
res <- tibble(
  "0%" = iqr[[1]],
  "25%" = iqr[[2]],
  "50%" = iqr[[3]],
  "75%" = iqr[[4]],
  "100%" = iqr[[5]]
)
  
kable(res) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, font_size = 20)
```

- So our IQR = `r res[,4] - res[,2]`

##  Inter-quartile range 
- We can calculate various quantiles using the `quantile()` function, and the IQR directly using the function `IQR()`
- Problem:   
  - Still somewhat sensitive to outliers.
  - Ignores half the data.

##  Use of range 
- In principle, we can define a range using any two values.
- Commonly researchers will also use deciles (blocks of 10%) to define a range.
    - E.g. Range between 10% and 90%


##  Use of range 

```{r eval=FALSE}
iqr2 <- quantile(ex1$StatsScore, c(.1, .9))
res2 <- tibble(
  "10%" = iqr2[[1]],
  "80%" = iqr2[[2]],
)
```


```{r, echo=FALSE}
iqr2 <- quantile(ex1$StatsScore, c(.1, .9))
res2 <- tibble(
  "10%" = iqr2[[1]],
  "80%" = iqr2[[2]],
)
  
kable(res2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, font_size = 20)

```


- So the range across 80% of the ranked data is `r res2[,2] - res2[,1]`.

##  Use of range 
- Ranges rely on **rank-ordered data** .
  - Hence they are most often used alongside the median as a measure of central tendency.


## Variation around the mean

```{r, echo=FALSE}
mean_vec <- rep(mean(ex1$StatsScore), 150)
x <- 1:150
dev <- ex1 %>%
  ggplot(., aes(x=factor(ID), y=StatsScore)) +
  geom_point(size=1.5) + 
  xlab("ID") +
  ylab("Statistics Score \n") +
  theme(axis.text.x = element_blank(), 
        axis.ticks.x=element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())

dev
```

## Variation around the mean

```{r, echo=FALSE}
dev +
  geom_hline(yintercept = mean(ex1$StatsScore), size = 1, col = "darkgrey") +
  geom_segment(x = x, y = ex1$StatsScore , xend = x, yend = mean_vec, linetype = "dashed")

```

## Sum of deviations 
- We could just add up the amount by which each observation differs from the mean.
- This is called the **sum of deviations.**

$$
S = \sum_{i=1}^{N}{(x_i - \bar{x})}
$$

- $x_i$ = individual observations
- $\bar{x}$ = mean of $x$

## Calculation

```{r, echo=FALSE}
ex1 %>%
  select(ID, StatsScore) %>%
  slice(., 1:10) %>%
  mutate(
    StatsScore = StatsScore,
    Mean = rep(mean(StatsScore), 10),
    Deviance = StatsScore - mean(StatsScore)
  ) %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

##  Problem: Sum of deviations 

```{r eval=FALSE}
ex1 %>%
  summarise(
    Variable = "Statistics Test Score",
    "Sum Deviation" = round(sum(StatsScore - mean(StatsScore)),2)
  )
```


```{r echo=FALSE}
ex1 %>%
  summarise(
    Variable = "Statistics Test Score",
    "Sum Deviation" = round(sum(StatsScore - mean(StatsScore)),2)
  ) %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, font_size = 20)
```

- The positive and negative values cancel.
- This means S will always be 0 when calculated around the mean.


##  Sum of absolute deviations 

- Suppose we just ignored the direction of the difference, and just considered the magnitude of the difference?
- This is the sum of ***absolute*** deviations.

$$
S_{abs} = \sum_{i=1}^{N}|{(x_i - \bar{x})}|
$$

- Here the || are read as absolute, which means that we remove the sign.

## Calculation

```{r, echo=FALSE}
ex1 %>%
  select(ID, StatsScore) %>%
  slice(., 1:10) %>%
  mutate(
    StatsScore = StatsScore,
    Mean = rep(mean(StatsScore), 10),
    Deviance = StatsScore - mean(StatsScore),
    Absolute = abs(Deviance)
  ) %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

##  Sum of absolute deviations 

```{r eval=FALSE}
ex1 %>%
  summarise(
    Variable = "Statistics Test Score",
    "Sum Deviation" = round(sum(StatsScore - mean(StatsScore)),2),
    "Abs Sum Deviation" =  round(sum(abs(StatsScore - mean(StatsScore))),2)
  )
```


```{r echo=FALSE}
ex1 %>%
  summarise(
    Variable = "Statistics Test Score",
    "Sum Deviation" = round(sum(StatsScore - mean(StatsScore)),2),
    "Abs Sum Deviation" =  round(sum(abs(StatsScore - mean(StatsScore))),2)
  )  %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, font_size = 20)
```

- Problem: 
  - As sample gets bigger, the sum of deviations continues to grow.

##  Mean sum of absolute deviations 
- To resolve, we could scale it by sample size. 
- Thus we get the ***mean sum of absolute deviations.***

$$
\bar{S_{abs}} = \frac{\sum_{i=1}^{N}|{(x_i - \bar{x})}|}{N}
$$

- Tells us how far *on average* all points are from the centre of the values.


##  Mean sum of absolute deviations 

```{r eval=FALSE}
ex1 %>%
  summarise(
    Variable = "Statistics Test Score",
    "Sum Deviation" = round(sum(StatsScore - mean(StatsScore)),2),
    "Abs Sum" =  round(sum(abs(StatsScore - mean(StatsScore))),2),
    "Mean Abs Sum" = round(sum(abs(StatsScore - mean(StatsScore)))/length(StatsScore),2)
  )
```


```{r echo=FALSE}
ex1 %>%
  summarise(
    Variable = "Statistics Test Score",
    "Sum Deviation" = round(sum(StatsScore - mean(StatsScore)),2),
    "Abs Sum" =  round(sum(abs(StatsScore - mean(StatsScore))),2),
    "Mean Abs Sum" = round(mean(abs(StatsScore - mean(StatsScore))),2)
  ) %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, font_size = 20)
```

- Divide the absolute sum by N.

##  Variance 
- Instead of using absolute values to deal with negatives, we could instead square the differences.
- This is called the ***variance*** .

$$
\sigma^2 = \frac{\sum_{i=1}^{N}{(x_i - \bar{x})}^2}{N}
$$

- Variance is the mean squared deviation from the mean.
- $\sigma^2$ = variance (Greek letter lower case sigma)

## Calculation

```{r, echo=FALSE}
ex1 %>%
  select(ID, StatsScore) %>%
  slice(., 1:10) %>%
  mutate(
    StatsScore = StatsScore,
    Mean = rep(mean(StatsScore), 10),
    Deviance = StatsScore - mean(StatsScore),
    Deviance_sq = Deviance^2
  ) %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```


##  Variance 

```{r eval=FALSE}
ex1 %>%
  summarise(
    Variable = "Statistics Test Score",
    "Sum Deviation" = round(sum(StatsScore - mean(StatsScore)),2),
    "Abs Sum" =  round(sum(abs(StatsScore - mean(StatsScore))),2),
    "Mean Abs Sum" = round(sum(abs(StatsScore - mean(StatsScore)))/length(StatsScore),2),
    Variance = round((sum((StatsScore - mean(StatsScore))^2))/length(StatsScore),2)
  )
```

```{r echo=FALSE}
ex1 %>%
  summarise(
    Variable = "Statistics Test Score",
    "Sum Deviation" = round(sum(StatsScore - mean(StatsScore)),2),
    "Abs Sum" =  round(sum(abs(StatsScore - mean(StatsScore))),2),
    "Mean Abs Sum" = round(mean(abs(StatsScore - mean(StatsScore))),2),
    Variance = round((sum((StatsScore - mean(StatsScore))^2))/length(StatsScore),2)
  )   %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, font_size = 20)
```

- Problem: 
  - Our units here are not quite right.
  - Variance is the mean **squared** deviation from the mean.

##  Standard deviation 
- What about a measure of variation in the same units as the mean/variable?
- The ***standard deviation.***
- The standard deviation is the square root of the variance.
    - This fixes our unit/scaling problem.

$$
\sigma = \sqrt{\frac{\sum_{i=1}^{N}{(x_i - \bar{x})}^2}{N}}
$$

##  Standard deviation 

```{r eval=FALSE}
ex1 %>%
  summarise(
    Variable = "Statistics Test Score",
    "Sum Deviation" = round(sum(StatsScore - mean(StatsScore)),2),
    "Abs Sum" =  round(sum(abs(StatsScore - mean(StatsScore))),2),
    "Mean Abs Sum" = round(mean(abs(StatsScore - mean(StatsScore))),2),
    Variance = round((sum((StatsScore - mean(StatsScore))^2))/length(StatsScore),2),
    SD = round(sqrt((sum((StatsScore - mean(StatsScore))^2))/length(StatsScore)),2)
  ) 
```


```{r echo=FALSE}
ex1 %>%
  summarise(
    Variable = "Statistics Test Score",
    "Sum Deviation" = round(sum(StatsScore - mean(StatsScore)),2),
    "Abs Sum" =  round(sum(abs(StatsScore - mean(StatsScore))),2),
    "Mean Abs Sum" = round(mean(abs(StatsScore - mean(StatsScore))),2),
    Variance = round((sum((StatsScore - mean(StatsScore))^2))/length(StatsScore),2),
    SD = round(sqrt((sum((StatsScore - mean(StatsScore))^2))/length(StatsScore)),2)
  )   %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, font_size = 20)
```

## Easier options

```{r eval=FALSE}
ex1 %>%
  summarise(
    Variable = "Statistics Test Score",
    "Sum Deviation" = round(sum(StatsScore - mean(StatsScore)),2),
    "Abs Sum" =  round(sum(abs(StatsScore - mean(StatsScore))),2),
    "Mean Abs Sum" = round(mean(abs(StatsScore - mean(StatsScore))),2),
    Variance = round(var(StatsScore),2),
    SD = round(sd(StatsScore),2)
  )
```


##  Which measure should we use? 

```{r tbl23, echo = FALSE}
tbl23 <- tibble::tribble(
~`Variable Type`, ~`Central Tendency`, ~`Dispersion`,
"Categorical (Nominal)","Mode","Frequency Table",
"Categorical (Ordered)","Mode/Median","Range",
"Continuous","Mean (any in fact)","Variance & Standard Deviation",
"Count","Mode (mean)","Range (Variance & SD)"
)

kableExtra::kable_styling(knitr::kable(tbl23), font_size = 22)
```

- Depends on the level of measurement.

##  A few extra bits?

- You may come across the mathematical language of *moments.*
- Moments describe the shape of a set of points
  - Mean
  - Variance
  - Skew
  - Kurtosis

## Skew 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
tibble(
  Value = c(rbeta(10000, 5, 2), rbeta(10000, 5, 5), rbeta(10000, 2,5)),
  Skew = c(rep("Negative", 10000), rep("None",10000), rep("Positive",10000))
) %>%
  ggplot(., aes(x=Value, fill = Skew, color = Skew)) +
  geom_density(alpha = .5) +
  ylab("Density \n")
```

- Is a measure of asymmetry of a distribution.

##  Kurtosis

```{r, echo=FALSE, warning=FALSE, message=FALSE}
tibble(
  Value = c(rnorm(10000, 0, 0.5), rnorm(10000, 0, 1), rnorm(10000, 0, 2)),
  Kurtosis = c(rep("Leptokurtic", 10000), rep("None",10000), rep("Platykurtic",10000))
) %>%
  ggplot(., aes(x=Value, fill = Kurtosis, color = Kurtosis)) +
  geom_density(alpha = .5) +
  ylab("Density \n")
```

- Kurtosis is a measure of the flatness of the peak and the fatness of the tails of the distribution.

##  Do they matter? 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
set.seed(07021984)
df <- tibble(Data = rsn(1000, 5, 2, 5)) 

df %>%
  ggplot(., aes(x=Data)) +
  geom_histogram(colour = "darkgrey", fill = "white") +
  geom_vline(xintercept = mean(df$Data), col = "red", size = 2) +
  geom_vline(xintercept = median(df$Data), col = "blue", size = 2) +
  geom_vline(xintercept = modes(df$Data, type = 2, digits = 2, nmore = 1)[1,1], col = "green", size = 2) +
  ylab("Count \n")
```

- It can make a difference in how we describe data.
- Both skew and kurtosis impact the **normality** of the distribution of the data.

## Descriptive statistics can deceive
```{r, echo=FALSE}
anscombe %>%
  gather(., key= set, value = score) %>%
  group_by(set) %>%
  summarise(
    mean = round(mean(score),2), 
    sd = round(sd(score),2)
  ) %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```



## Always visualize data
```{r, echo=FALSE}
a <- ggplot(anscombe, aes(x=x1, y=y1)) +
  geom_point(size = 2) +
  xlab("X1") +
  ylab("Y1")

b <- ggplot(anscombe, aes(x=x2, y=y2)) +
  geom_point(size = 2) +
  xlab("X2") +
  ylab("Y2")

c <- ggplot(anscombe, aes(x=x3, y=y3)) +
  geom_point(size = 2) +
  xlab("X3") +
  ylab("Y3")

d <- ggplot(anscombe, aes(x=x4, y=y4)) +
  geom_point(size = 2) +
  xlab("X4") +
  ylab("Y4")

plot_grid(a, b, c, d)
```


## Tasks for this week...
1. Finish any existing tasks from lab 4.
2. Reading: Linked at the top of lab 5.
3. Quiz 5: Central tendency
    - **This quiz counts**
    - Live now (as of Monday at 09:00).
    - Closes Sunday at 17:00

## Recommendations of the week
- Podcast: [Ineos159](https://www.youtube.com/watch?v=7XW63dM_1d0) Preview documentary
- Recipe: [Roasted Veggie Lasagna](https://www.bbcgoodfood.com/recipes/roasted-vegetable-lasagne) 
- Book: [Stoner](https://www.theguardian.com/books/2013/dec/13/stoner-john-williams-julian-barnes), John Williams
- Thing to do: [Portobello Beach](https://www.visitscotland.com/info/towns-villages/portobello-beach-p730811)
- Coffee/food Place: [Mary's Milk Bar](http://www.marysmilkbar.com/) 
