---
title: 'dapR1 (week 5): Functions Activity2'
author: "Tom Booth & Alex Doumas"
date: ""
output: html_document
---

# Problem 1

The below code is from the lecture and, you'll recall, creates a function called fun1, and then plots that function. 

```{r, echo=FALSE, warning=F,message=F}
library(tidyverse)
fun1 <- function(x) 10+(2*x)

func_x <- tibble(
  x = c(1,2,3,4,5,6,7,8), 
  y = 10 + (2*x) 
)

ggplot(func_x, aes(x, y)) + 
  geom_point(colour = "red", size = 2) +
  stat_function(fun = fun1) +
  scale_x_continuous(name = "x", breaks = c(0:10), labels = c(0:10), limits = c(0,10)) +
  scale_y_continuous(name = "y", breaks = c(seq(0,30,5)), labels = c(seq(0,30,5)), limits = c(0,30)) 
```

Please open up the slides associated with this activity (linked on the learn page), and copy and paste the code above into the editor in rstudio. Then mess with the code to create and plot the following functions (hint: all you need to change is the definition of the function and the data set created by func_x). 

1. $f(x) = x$
2. $f(x) = \frac{x}{2}$
3. $f(x) = \frac{2}{x}$
4. $f(x) = 4x^2$ (hint: remember order of operations for this function.)

Now, mess with the plot a little. 

5. Remove the code:
```{r, eval=FALSE}
geom_point(colour = "red", size = 2)
```
6. Plot red points from 1-20 instead of just 1-8.



# Answers
1. fun2 <- function(x) x
2. fun3 <- function(x) x/2
3. fun4 <- function(x) 2/x
4. fun5 <- function(x) 4*(x^2)

5. Delete the line of code and the first + from the ggplot function call. 
6. Edit func_x to the following:
```{r, eval=FALSE}
func_x <- tibble(
  x = c(1:20), 
  y = 10 + (2*x) 
)
```
And then edit the scale_x_continuous and scale_y_continuous in the ggplot function to make the x axis go from 1-20 instead of 1-10. 



