---
title: "<b>Q&A</b>"
subtitle: "Data Analysis for Psychology in R 1<br><br> "
author: ""
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

# YOUR SETTINGS

**PLEASE MAKE SURE YOUR MICORPHONES ARE ON MUTE AND YOUR CAMERAS OFF**

---
# AUDIO CHECK

**Can you all hear me talking?**

Please type yes or no into the text chat.

**And can you hear Jo?**

**And Umberto?**

---
# ASKING QUESTIONS
+ Please type all questions into the text chat.

+ Please wait to ask questions until asked to do so. 
  + This makes it easier for us to keep track.
  + I will stop and give lots of chances.

+ When I do, please type them into the chat.
  + Keep an eye in case the question has been asked already.
  
---
# RECORDING
This session will now be recorded. Any further information that you provide during a session is optional and in doing so you give us consent to process this information.

These sessions will be stored by the University of Edinburgh for one year and published on our website during that time.

By taking part in a session you give us your consent to process any information you provide during it.

---
# RECORDING

**I AM NOW TURNING THE RECORDING ON**

---
# CONCERNS
+ As we move through, things highlighted in blue are summary topics/questions from the Padlet and survey.

+ **Programming and stats**
  + A little history on course
  + Slow and steady
  + Data analysis vs stats

---
# ONLINE LEARNING
+ **MISSING MATERIAL**
  + We will do a LEARN tour later.
  + Everything will be in one file for the week.
  + If anything happens centrally to impact access (e.g. yesterday!) we will take account of that.
  
---
# ONLINE LEARNING
+ **How will on-line teaching work?**
+ **It's all on-line so less structured**
+ **Not having access to resources due to being online**
  + This may be the most structured we have ever been!
  
+ Lectures
  + Recorded, with exercises.

+ Labs
  + Self-directed learning with supported drop-ins
  + Lecturer office hours
  

---
# DROP-IN'S AND PEER SUPPORT
+ **Not being able to ask questions as we work though**
  + We hope we have made enough spaces that you will be able to.
  
+ 4 drop-in's per week.

+ In principle, you can attend them all
  + But, we need to have a bit of a trust/honour system.
  + If you have attended already in the week, and the session gets very full, allow others to ask questions

+ Open course room
  + Come and study any time.
  + If it is always full, we may have capacity issues, but I dont think this will happen.
  + Arrange with others to study togther here.
  + Discussion forum thread to find study partners at times that work for you all.

---
# FINDING HELP
+ Drop-in's

+ Office hours

+ Discussion boards

---
class: center, middle
# Questions


---
# TIMETABLE
+ **Is there a timetable?**
  + No.
  + There are no compulsory synchronous sessions on the course.

+ Suggestion:
  + Quiz (Monday)
  + Lectures & exercises (Monday-Tuesday)
  + Lab (Wednesday-Thursday)
  + Consolidation, notes, reading (Friday)

+ Total time ~5-8hrs.

---
class: center, middle
# Questions


---
# ASSESSMENT
+ **When do we have tests, how often?**
  + Every week.
  + Quizzes or lab tests will appear at the top of each "Weekly Materials" folder.
  
+ Quizzes
  + Quizzes will be the first 4 weeks of each 5 week teaching block. 
  + Weeks 1 and 2 are practice and do not count towards your assessment.
  + There are 16 assessed quizzes. 
  + Always released on a Monday at 9am, close Sunday at 5pm GMT.
  + You have 1 hr to complete each test. 
  +~10 MCQ questions each time (sometimes fewer)
  + Best 12/16 scores are averaged to be your quiz mark.
  + Worth 20% overall grade.

---
# ASSESSMENT
+ Lab tests
  + Every 5th week
  + Week 5 semester 1 = practice and does not count
  + Remaining 3 are 10% each (30% total).
  + We will give you a skeleton Rmd file, you fill in the code chunks.

---
# ASSESSMENT
+ **What happens if we miss a test?**
  + You can miss 4, or performing worse than you would have liked, and it has no impact (14/18)
  + If you miss more, and have reasons of suitable magnitude you can submit special circumstances
  + Otherwise it lowers your grade.
  
---
# ASSESSMENT
+ **When are the report deadlines?**
  + On LEARN dashboard there is a button for assessment. 
  + We will put all information on reports here.
  + We will add the deadlines this week
  + And let you know as and when we add details of the assessments.

+ Report (50%) semester 2

---
class: center, middle
# Questions


---
class: center, middle
# Tour of LEARN

---
class: center, middle
# Open Questions
