---
title: "Measure central tendency"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```

We'll illustrate measures of central tendency with the `msleep` dataset that comes from `tidyverse`.
`msleep` contains some data about how much different kinds of mammals sleep.
Here's what the first few rows of `msleep` look like.

```{r}
msleep |> head()
```


## Mean

The mean is a variable's average value.

It can be computed in two ways:

- by summing all observations together using `sum()`, then dividing by the total number of observations `n()`
- by using the `mean()` function

For example, to find the mean hours of sleep in `msleep`:

```{r}
msleep |>
  summarise(
    mean_by_hand     = sum(sleep_total) / n(),
    mean_by_function = mean(sleep_total)
  )
```

Same number!

### Means by group

To find the mean hours of sleep for each `genus` in `msleep`, add `group_by()` for the variable you want to group by, then pipe that into `summarise()`:

```{r}
msleep |>
  group_by(genus) |>
  summarise(
    mean_sleep = mean(sleep_total)
  )
```


### I'm just seeing "NA"

`mean()` worked for `sleep_total`, but if we try it with `sleep_rem` (hours of REM sleep), we get "NA" instead of a number.

```{r}
msleep |>
  summarise(
    mean_REM = mean(sleep_rem)
  )
```

This is because the data has not recorded the REM sleep for every animal.
Where the hours of REM sleep are unknown, the data contains the value NA.
NA stands for "not available".
And any NAs in this variable will make the `mean()` function return NA instead of an actual number.

To force `mean()` to overlook the NAs and return the mean of all the values we *do* have, use the code `na.rm = TRUE` (`na.rm` stands for "NA remove").

```{r}
msleep |>
  summarise(
    mean_REM = mean(sleep_rem, na.rm = TRUE)
  )
```




## Median

The median is the value for which 50% of observations a lower and 50% are higher.
Think of it as the mid-point of the list of sorted values.

```{r}
msleep |>
  summarise(
    median_sleep = median(sleep_total)
  )
```


### Medians by group

```{r}
msleep |>
  group_by(genus) |>
  summarise(
    median_sleep = median(sleep_total)
  )
```

### I'm just seeing "NA"

`na.rm` works for `median()` too.

Attempting to find the median of a variable containing NAs:

```{r}
msleep |>
  summarise(
    median_REM = median(sleep_rem)
  )
```

With `na.rm = TRUE`:

```{r}
msleep |>
  summarise(
    median_REM = median(sleep_rem, na.rm = TRUE)
  )
```



## Mode

The mode of a variable is that variable's most frequent value, or in other words, the value that occurs the most often.

There's no `mode()` function in R.
So we find the mode by generating a frequency table and looking at it to find which value has the greatest frequency.

```{r}
msleep |>
  count(vore)
```

The mode of `vore` is "herbi".

::: {.callout-caution collapse="true"}
#### Advanced: Sort the table by frequency

The `arrange()` function from `tidyverse` will sort whatever column you name in ascending order (i.e., smallest values first).

```{r}
msleep |>
  count(vore) |>
  arrange(n)
```

To sort in descending order, use `arrange(desc())`.

```{r}
msleep |>
  count(vore) |>
  arrange(desc(n))
```


The value in the first row of this table is the mode.

:::

## Which measures for which variable types

|   |  Mode |  Median |  Mean |
|--:|--:|--:|--:| 
|  Nominal (unordered categorical) | &#10004; | &#10008; | &#10008; | 
|  Ordinal (ordered categorical) | &#10004; | &#10004; | **?**<br>(you may see it for certain types of ordinal data; people disagree) |
|  Numeric Continuous  | &#10004; | &#10004; | &#10004; |
