---
title: "Manage factor variables"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```

`factor()` is a function in base R (that is, not in any additional package) that encodes categorical variables in the most R-friendly way.

We'll illustrate the use of `factor()` with the `msleep` dataset that comes from `tidyverse`.
`msleep` contains some data about how much different kinds of mammals sleep.
Here's what the first few rows of `msleep` look like.

```{r}
msleep |> head()
```


## To convert a categorical variable to a factor

This code uses `mutate()`; for more detail, see [the Mutate columns flash card](mutate.html){target="_blank"}.

```{r}
msleep <- msleep |>
   mutate(
    vore = factor(vore)
  ) 

msleep
```

To make sure it's worked, check the data type that `str()` gives us for `vore`:

```{r}
str(msleep)
```

`vore` is stored as a "Factor w/ 4 levels", and those levels are labelled "carni", "herbi", "insecti", and "omni".

## To change the order of levels in a factor

By default, factor levels use the labels' alphabetical order.
To change the factor levels, say to put "insecti" first and "carni" at the end, we can use `levels` argument of `factor()`:

```{r}
msleep <- msleep |>
   mutate(
    vore = factor(
      vore,
      levels = c('insecti', 'herbi', 'omni', 'carni')  
      # make sure the labels you type match the existing labels exactly!
    )
  ) 

msleep
```

The preview doesn't look any different, but we can see the factor levels have changed order in `str()`, because now "insecti" is listed first:

```{r}
str(msleep)
```


## To change the labels shown for factor levels

To relabel "carni" as "carnivore", "insecti" as "insectivore", and so on, we can also use the `labels` argument of `factor()`.

```{r}
msleep <- msleep |>
   mutate(
    vore = factor(
      vore,
      labels = c('insectivore', 'herbivore', 'omnivore', 'carnivore')
      # make sure the labels match the order of the levels, else you might relabel carnivores as insectivores etc.!
    )
  ) 

msleep
```

This change is more apparent in the `vore` column!

And we can also see it using `str()`:

```{r}
str(msleep)
```


