---
title: "Mutate columns"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```

`mutate()` is a function from `tidyverse` that allows us to add on new columns or modify existing ones.

We'll illustrate the use of `mutate()` with the `msleep` dataset that comes from `tidyverse`.
`msleep` contains some data about how much different kinds of mammals sleep.
Here's what the first few rows of `msleep` look like.

```{r}
msleep |> head()
```


<br>

To convert the categorical `vore` variable into a factor:

```{r}
msleep |>
  mutate(
    vore = factor(vore)
  )
```


This code will not modify `msleep` in R's memory.
It just outputs the modified dataframe so that we can look at it.

To add this column to `msleep` in R's memory, we need to re-assign the modified dataframe to the same variable name `msleep`:

```{r}
msleep <- msleep |>
  mutate(
    vore = factor(vore)
  )
```


<br>

To add together numbers in two existing columns, `sleep_total` and `awake`, and create a new column called `total_hours` with the sum:

```{r}
msleep |>
  mutate(
    total_hours = sleep_total + awake 
  )
```

This code will not modify `msleep` in R's memory.
It just outputs the modified dataframe so that we can look at it.

To add this column to `msleep` in R's memory, we need to re-assign the modified dataframe to the same variable name `msleep`:

```{r}
msleep <- msleep |>
  mutate(
    total_hours = sleep_total + awake 
  )
```


<br>


For lots more detail on `mutate()`, check out the [R for Data Science online textbook](https://r4ds.hadley.nz/data-transform.html#sec-mutate){target="_blank"}.