---
title: "Read in and inspect data"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
```

## To read in data

To read data from an online source into R, use the following code:

```{r}
library(tidyverse)

tips2 <- read_csv("https://uoepsy.github.io/data/RestaurantTips2.csv")
```

`read_csv()` comes from the `tidyverse` package, so you'll need to load the `tidyverse` package if you haven't already.


## To inspect data

**To look at the first few rows, use `head()`:**

::::{.columns}
:::{.column width="48%"}
```{r}
head(tips2)
```
:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
```{r}
tips2 |>
  head()
```

:::
::::

Note: Both versions of this code accomplish the same thing.
One places `tips2` inside the `head()` function, and the other one uses the [pipe](pipe.html){target="_blank"} to feed `tips2` into `head()`.

<br>

**To see how many rows and columns are in the data, use `dim()`:**

::::{.columns}
:::{.column width="48%"}
```{r}
dim(tips2)
```
:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
```{r}
tips2 |>
  dim()
```

:::
::::

<br>

**To see how many rows are in the data, use `nrow()`:**

::::{.columns}
:::{.column width="48%"}
```{r}
nrow(tips2)
```
:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
```{r}
tips2 |>
  nrow()
```

:::
::::

<br>



**To see how many columns are in the data, use `ncol()`:**

::::{.columns}
:::{.column width="48%"}
```{r}
ncol(tips2)
```
:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
```{r}
tips2 |>
  ncol()
```

:::
::::

<br>

**To see the structure of the data, use `str()`:**

::::{.columns}
:::{.column width="48%"}
```{r}
str(tips2)
```
:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
```{r}
tips2 |>
  str()
```

:::
::::

<br>


**To take a glimpse at the data types and the first few values of each column of the data, use `glimpse()`:**


::::{.columns}
:::{.column width="48%"}
```{r}
glimpse(tips2)
```
:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
```{r}
tips2 |>
  glimpse()
```

:::
::::
