---
title: "Store data with vectors"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
```


To store a sequence of numbers into R, we combine the values using the combine function `c()` and give the sequence a name. A sequence of elements all of the same type is called a **vector**.
To view the stored content, simply type the name of the vector.

```{r}
myfirstvector <- c(1, 5, 3, 7)
myfirstvector
```

We can perform arithmetic operations on each value of the vector. For example, to add five to each entry:
```{r}
myfirstvector + 5
```

<br>

Vectors can hold things other than numbers.
They can also hold words, for example, which are of data type "character".
Words need to be written inside quotation marks, e.g. "enquoted".

```{r}
wordsvector <- c("cat", "dog", "parrot", "peppapig")
wordsvector
```

You can use either double-quote or single-quote symbols:

```{r}
c("cat", "dog", "parrot", "peppapig")
c('cat', 'dog', 'parrot', 'peppapig')
```

The function `class()` will tell you the type of the object.
In this case, it is a character vector.

```{r}
class(wordsvector)
```

Some operations like addition and multiplication (which only make sense for numbers) will make R upset if you try them on character vectors.
For example, if we try to add 5 to each word in `wordsvector`:

```{r eval=FALSE}
wordsvector + 5
```

```
Error in wordsvector + 5 : non-numeric argument to binary operator
```

<br>

Finally, R requires that vectors contain elements that are all of the same type.
A practical consequence of this is that, if you try to combine numbers and characters in a single vector, then R will transform all of those different things to be of the same type.

Typically, R chooses the most general or easily-convertible type between the two. 
In the case below, it makes everything a character—you can tell because all the elements, even the numbers, are within quotation marks.

```{r}
mysecondvector <- c(4, 5, 6, "cat")
mysecondvector
```
