---
title: "Formative Report A (Group <NUMBER>.<LETTER>)"
author: "<insert exam numbers here, e.g. B001, B002, B003, B004, B005>"
date: "<insert date here>"
output: bookdown::pdf_document2
toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
# Write your code here.
```


# Introduction

Write here your introduction. Text is written outside of a code chunk, just like this. If you want to write something in italic, write it like _this_ or *this*. If you want to write something in bold, write it like __this__ or **this**.


# Analysis

Present here your analysis and results.


# Discussion

Write here the key take-home messages from the results and analyses you have presented.


# Appendices

## Appendix A: Additional figures and tables

Include here additional tables and figures, with captions, and properly referenced. These should be used somewhere in the text, do not include tables or figures which are not referenced anywhere in your writing.

## Appendix B: R code

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```

