---
title: "<b>Semester 2 Welcome</b>"
subtitle: "Data Analysis for Psychology in R 1"
author: ""
institute: "Department of Psychology<br/>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: jk_libs/libs
    css: 
      - xaringan-themer.css
      - jk_libs/tweaks.css
    nature:
      beforeInit: "jk_libs/macros.js"
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(digits=4,scipen=2)
options(knitr.table.format="html")
xaringanExtra::use_xaringan_extra(c("tile_view","animate_css","tachyons"))
xaringanExtra::use_extra_styles(
  mute_unhighlighted_code = FALSE
)
library(knitr)
library(tidyverse)
library(ggplot2)
knitr::opts_chunk$set(
  dev = "svg",
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
#source('R/myfuncs.R')
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

class: inverse, center, middle
# Semester 1 Recap  

---

.pull-left[

## Weeks 1 - 5  

- Collecting Data & Types of Data  

- Describing and Visualising Categorical Variables   

- Describing and Visualising Numeric Variables  

- Describing and Visualising Relationships  

- Types of Relations (functions)  

]

--

.pull-right[
## Weeks 7 - 11  

- Probability Theory  

- Probability Rules  

- Discrete Random Variables   

- Continuous Random Variables  

- Sampling Distributions  


]

---

# Things we hope you did last semester

.pull-left[

## Engaging with course materials

- Watched lectures and attempted the weekly exercises
- Checked the solutions each time
- Attended drop-ins and office hours 
- Written on the discussion board
- Checked in with study group

] 
.pull-right[

## RStudio skills

- Basic arithmetic operations   
- Rstudio functionality 
- Data handling in R 
- RMarkdown 
- ggplot!

]

---

# Things we hope you did last semester

.pull-left[

## Engaging with course materials

- Watched lectures and attempted the weekly exercises
- Checked the solutions each time
- attended drop-ins and office hours 
- written on the discussion board
- checked in with study group

] 
.pull-right[

## RStudio skills

- Basic arithmetic operations   
  - [sem 1 week 0 reading](https://uoepsy.github.io/dapr1/labs/intro_r_rstudio_year1.html#Glossary)
- Rstudio functionality 
  - [tour of rstudio videos](https://media.ed.ac.uk/playlist/dedicated/181985631/1_8iq22lv2/)
  - [sem 1 week 0 reading](https://uoepsy.github.io/dapr1/labs/intro_r_rstudio_year1.html)  
- Data handling in R 
  - [sem 1 week 1 exercsies](https://uoepsy.github.io/dapr1/labs/01_data_types.html)
- RMarkdown 
  - [tour of rstudio videos](https://media.ed.ac.uk/playlist/dedicated/181985631/1_8iq22lv2/)
  - [sem 1 week 6 extra materials (short)](https://www.learn.ed.ac.uk/webapps/blackboard/content/listContentEditable.jsp?content_id=_5290495_1&course_id=_81753_1)
- ggplot!

]
---

class: inverse, center, middle, animated, rotateInDownLeft

# Looking ahead to Semester 2

---

.pull-left[

## Weeks 1 - 5

- Bootstrap & Confidence Intervals  

- Hypothesis Testing: P-values  

- Hypothesis Testing: Critical Regions  

- Hypothesis Testing & Confidence Intervals

- Making Decisions (Effect Sizes, Power, Errors) 

]
.pull-right[

## Weeks 7 - 11

- One Sample t-Test  

- Independent Samples t-test  

- Paired Samples t-test

- Chi-square test

- Correlation


]

---

class: inverse, center, middle, animated, rotateInDownLeft

# Changes

---

## New classroom!  

- Drop-in sessions are moving to **gather.town** (it's a bit more interactive).  
  - It will be the same link for all sessions, and it will always be open should you want to arrange to meet up with other students there. 
  
---

## Q&A >> live coding

- The Q&A sessions which were run in the previous semester are going to now be used for live-coding.  
    - Watch us attempt previous weeks' materials in RStudio, watch us mess up and make lots of typos, listen to us get confused and angry at R, but learn from watching someone fix their mistakes and talk through what they are doing.  
    - Sessions will be recorded, **and will continue to take place on Blackboard Collaborate** 


---


class: inverse, center, middle, animated, rotateInDownLeft


