---
title: "Paired t-test"
subtitle: "Data Analysis for Psychology in R 1<br><br> "
author: "dapR1 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: TRUE
      countIncrementalSlides: FALSE
      
---
```{r premable, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(kableExtra)
library(patchwork)
library(moderndive)
```


# Learning Objectives
- Understand when to use an paired sample $t$-test

- Understand the null hypothesis for an paired sample $t$-test

- Understand how to calculate the test statistic

- Know how to conduct the test in R

- Know how to calculate Cohen's $D$ for each form of $t$-test


---
# Topics for today

- Recording 1: Conceptual background and introduction to our example

--

- Recording 2: Calculations and R-functions

--

- Recording 3: Assumptions and effect size
	

---
# Purpose & Data
- The paired sample $t$-test is used when we want to test the difference in mean scores for a sample measured at two points in time. 
  - Thus this is a first example of a repeated measures design.

- Data Requirements
  - A continuously measured variable.
  - A binary variable denoting time.

---
# Example
- I want to assess whether a time-management course helps reduce exam stress in students. 

- I ask 50 students to take a self-report stress measure during their winter exams. 

- At the beginning of semester 2 they take a time management course. 

- I then assess their self-report stress in the summer exam block.
	- Let's assume for the sake of this example that I have been able to control the volume and difficulty of the exams the students take in each block.


---
# Data
```{r, echo = FALSE}
set.seed(007)
exam <- tibble(
  ID = c(rep(paste("ID", 1:50, sep=""),2)),
  stress = round(c(rnorm(50, 9.2, 2.2), rnorm(50, 7.5, 2.8))),
  time = as_factor(c(rep("t1", 50), rep("t2", 50)))
)
```

```{r, echo=FALSE}
head(exam)
```


---
# Calculating difference
- In the paired $t$-test, we specifically calculate and analyse the difference in scores at time 1 and time 2 per participant.

$$d_i = x_{i1} - x_{i2}$$

---
# Test statistic
- The resulting test statistic:

$$
t = \frac{\bar{d}}{s_{d} / \sqrt{n}}
$$

- where:
  - $\bar{d}$ = mean of the individual difference scores ( $d_i$ )
  - $s_{d}$ = standard deviation of the difference scores ( $d_i$ )
  - $n$ = sample size

- The associated sampling distribution is a $t$-distributon with $n-1$ degrees of freedom.
  - Note, this is just essentially a one sample test on the difference scores.


---
# Hypotheses
- Two-tailed:

$$
\begin{matrix}
H_0: \mu_{d} = 0 \\
H_1: \mu_{d} \neq 0
\end{matrix}
$$

- One-tailed

$$
\begin{matrix}
H_1: \mu_{d} < 0 \\
H_1: \mu_{d} > 0
\end{matrix}
$$
---
class: center, middle
# Time for a break

---
class: center, middle
# Welcome Back!

**Let's calculate a paried-sample t-test!**



---
# Our Example
- I elect to use a two-tailed test with alpha of .01

- I want to be quite sure the intervention has worked and stress levels have changed.

- So my hypotheses are:

$$
\begin{matrix}
H_0: \mu_{d} = 0 \\
H_1: \mu_{d} \neq 0
\end{matrix}
$$


---
# Calculation
- Steps in my calculations:
  - Calculate the difference scores for individuals.
  - Calculate the mean of the difference scores.
  - Calculate the SD of the difference scores.
  - Check I know my N.
  - Calculate the standard error of the mean difference.
  - Use all this to calculate $t$
  - Calculate my degrees of freedom


---
# Data organisation
- Our data is currently in what is referred to as long format.
  - All the scores are in one column, with two entries per participant.

- To calcuate the $d_i$ values, we will convert this to wide format.
  - Where there are two columns representing the score at time 1 and time 2
  - And a single row per person
  

---
# Data organisation
```{r}
exam_wide <- exam %>%
  pivot_wider(id = ID, 
              names_from = time, 
              values_from = stress)
```

```{r, echo=FALSE}
head(exam_wide)
```


---
# Calculation
```{r}
calc <- exam_wide %>%  
  mutate(
    dif = t1 - t2) %>%
  
  summarise(
    D = mean(dif),
    SDd = round(sd(dif),2),
    N = n()) %>%
  
  mutate(
    SEd = round(SDd /sqrt(N),2),
    t = round(D/SEd,2)
  ) 
```

```{r, echo=FALSE}
calc
```

---
# Is my test significant?
- So we have all the pieces we need:
	- $t$ = `r calc[[5]]` 
	- $df$ = $n-1$ = 49
	- Hypothesis to test (two-tailed)
	- $\alpha = 0.01$

- So now all we need is the critical value from the associated $t$-distribution in order to make our decision .


---
# Is my test significant?

.pull-left[
```{r, echo=FALSE}
ggplot(NULL, aes(x = c(-6,6))) +
  stat_function(fun=dt,
                geom = "line",
                args = list(df=49)) +
  stat_function(fun = dt, 
                geom = "area",
                xlim = c(qt(0.005, 49), -6),
                alpha=.25,
                fill = "blue",
                args = list(df=49)) +
    stat_function(fun = dt, 
                geom = "area",
                xlim = c(qt(0.995, 49), 6),
                alpha=.25,
                fill = "blue",
                args = list(df=49)) +
  geom_vline(xintercept = calc[[5]], col="red") +
  xlab("\n t") +
  ylab("") +
  ggtitle("t-distribution (df=49)")
```
]

.pull-right[
```{r}
tibble(
  LowerCrit = round(qt(0.005, 49),2),
  UpperCrit = round(qt(0.995, 49),2),
  Exactp = round(2*(1-pt(calc[[5]], 49)),5)
)
```
]


---
# In R
```{r}
res <- t.test(exam_wide$t1, exam_wide$t2, 
       paired = TRUE, 
       alternative = "two.sided")
```

```{r, echo=FALSE}
res
```

- Again, slight rounding differences.

---
# Write-up
A paired-sample $t$-test was conducted in order to determine a if a statistically significant ( $\alpha$ = .01) mean difference in self-report stress was present, pre- and post-time management intervention in a sample of 50 undergraduate students. The pre-intervention mean score was higher (Mean=`r mean(exam_wide$t1)`) than the post intervention score (Mean = `r mean(exam_wide$t2)`). The difference was statistically significant ( $t$(`r res$parameter`)= `r round(res$statistic,2)`, $p$ < . 01, two-tailed). Thus, we reject the null hypothesis of no difference.


---
class: center, middle
# Time for a break

---
class: center, middle
# Welcome Back!

**Now to check assumptions for a paired t-test and calculate Cohen's D**


---

# Assumption checks summary 

```{r tbl7, echo = FALSE}
tbl7 <- tibble::tribble(
~` `, ~`Description`, ~`One-Sample  t-test`, ~`Independent Sample t-test`, ~`Paired Sample t-test`,
"Normality","Continuous variable (and difference) is normally distributed.","Yes (Population)","Yes (Both groups/ Difference)","Yes (Both groups/ Difference)",
"Tests:","Descriptive Statistics; Shapiro-Wilks Test; QQ-plot"," "," "," ",
"Independence","Observations are sampled independently.","Yes","Yes (within and across groups)","Yes (within groups)",
"Tests:","None. Design issue."," "," "," ",
"Homogeneity of variance","Population level standard deviation is the same in both groups.","NA","Yes","Yes",
"Tests:","F-test"," "," "," ",
"Matched Pairs in data","For paired sample, each observation must have matched pair.","NA","NA","Yes",
"Tests:","None. Data structure issue."," "," "," "
)


kable(tbl7) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, font_size = 18)
```

---
# Assumptions
1. Normality of the difference scores ( $d_i$ )
2. Independence of observations **within** group/time
3. Data are matched pairs (design)

- We will briefly show the normality assumptions again.
  - Hopefully these are becoming familiar.


---
# Adding the difference scores

- Our assumptions concern the difference scores. 
- We showed these earlier in our calculations.
- Here we will add them to `exam_wide` for ease.

```{r}
exam_wide <- exam_wide %>%  
  mutate(
    dif = t1 - t2)
```

---
# Histograms

.pull-left[
```{r, eval=FALSE}
exam_wide %>%
  ggplot(., aes(x=dif)) +
  geom_histogram(bins = 20)
```
]

.pull-right[
```{r, echo=FALSE}
exam_wide %>%
  ggplot(., aes(x=dif)) +
  geom_histogram(bins = 20)
```
]


---
# QQ-plots

.pull-left[
```{r, eval=FALSE}
exam_wide %>%
  ggplot(., aes(sample = dif)) +
  stat_qq() +
  stat_qq_line()

```
]

.pull-right[
```{r, echo=FALSE}
exam_wide %>%
  ggplot(., aes(sample = dif)) +
  stat_qq() +
  stat_qq_line()

```
]

---
#  Shapiro-Wilks R
```{r}
shapiro.test(exam_wide$dif)
```

- Fail to reject the null, $p$ > .05

- Normality of the differences is met.

---
#  Cohen's D: Paired t
- Paired-sample t-test:

$$
D = \frac{\bar{d} - 0}{s_{d}}
$$

- $\bar{d}$ = mean of the difference scores ( $d_i$ )
- $s_{d}$ = standard deviation of the difference scores ( $d_i$ )


---
# Cohen's D in R

```{r, warning=FALSE}
library(effsize)
cohen.d(exam$stress, exam$time,
        subject = exam$ID, #<<
        paired = TRUE, #<<
        conf.level = .95)
```



---
#  Summary: Three different t-tests 

```{r tbl26, echo = FALSE}
tbl26 <- tibble::tribble(
~` `, ~`One-sample`, ~`Independent Sample`, ~`Paired (Dependent) Sample`,
"Outcome","Continuous Variable","Continuous Variable","Continuous Variable",
"Predictor","Single group vs population","Categorical: two groups","Categorical: two time points",
"Sample","One sample vs population value","Two independent groups","One group sampled at two time points",
"Measure of difference","Observed - known population value","Group 1 - Group 2","Time 1 - Time 2",
"Measure of Variability","Standard error of the mean","Pooled standard error of difference in means","Standard error of the difference in means"
)

kableExtra::kable_styling(knitr::kable(tbl26), font_size = 18)
```

