---
title: 'dapR1 (week 3): Activity2'
author: "Alex Doumas"
date: "14/09/2020"
output: html_document
---

# Means and standard deviations activity Q1

Below is a small data set. Pretend the values are scores on a test of star wars knowledge from a small sample of students. Try answering the following questions. 

$scores: [5,3,6,7,9,2,9]$

1. What is the *n* of the sample? 
2. What is the median value of the sample?
3. What are the mean and standard deviation (sd)? Try calculating them by hand.


# Means and standard deviations activity Q2

Below is some R code to generate a sample data set. Open up the activity document associated with this activity and copy and paste the code into R. Calculate the mean and sd using R functions (or try writing your own). Make a table that summarises the mean and sd you calculated.

```{r, echo=FALSE}
set.seed(10)

dat1 = rnorm(100, 10, 2)
```


# Answers.
Q1. 
  1. *n* = 7
  2. [2,3,5,6,7,9,9] -- median = 6
  3. Mean = (5+3+6+7+9+2+9)/7 = 5.857143; sd = sqrt(sum(5,3,6,7,9,2,9)/(7-1)) = 2.614065

Q2. You could use code that looks something like: 
```{r, echo=FALSE}
mydata <- tibble(
    dat1 = dat1
)
mydata %>%
  summarise(
    Variable = "Statistics Test Score",
    "Mean" = mean(dat1),
    SD = sd(dat1)
  )
```
