---
title: 'dapR1: Continuous Probability Distributions Activities - 2'
author: "Department of Psychology<br>University of Edinburgh"
date: "AY 2020-2021"
output:
  html_document:
    theme: flatly
---

## Question 1

Imagine you have a normal distribution with a mean of 100 and a sd of 18. What is the score below which 95 % of the values fall? What are the values beyond which the most extreme (both positive and negative) 5% of scores fall? What is the probability of getting a score below 115? Above 115? 

## Question 2 

Use the code below to plot a standard normal distribution (in black) and a t distribution (with df = 5; in red). 

```{r, eval=FALSE}
library(tidyverse)

ggplot(data.frame(x = c(-6, 6)), aes(x = x)) + 
  stat_function(fun = dnorm,
                geom = "line",
                color = "blue",
                args = list(mean = 0, sd=1)) +
    stat_function(fun = dt,
                geom = "line",
                color = "red",
                args = list(df=5)) + 
    labs(title="Standard normal vs. t distribution (with df = 5)")
```

Look at the code and see if you can spot where the parameter of the t distribution is specified. Modify the df of the t distribution to different values (e.g., 10, 20, 100, 1000). What happens as the df of the t distribution increases? 

<br>
<br>

---  

*Scroll down for answers*  
  
---

<div style="margin-bottom:750px;">

</div>


# Answers

1. 
    ```{r}
    #score below which 95% of values fall
    qnorm(.95, mean = 100, sd = 18)
    
    #scores below which 2.5% and 97.5% of values fall
    # (same as most extreme 5%)
    qnorm(c(.025, 0.975), mean = 100, sd = 18)
    
    # probability of getting a score below 115
    pnorm(115, mean = 100, sd = 18)
    
    # probability of getting a score above 115
    1-pnorm(115, mean = 100, sd = 18)
    ```  
    
2. As the df of the t distribution increases it becomes more and more like the standard normal distribution. 


