---
title: "<b>Week 10: Continuous Probability Distributions </b>"
subtitle: "Data Analysis for Psychology in R 1<br><br> "
author: "Alex Doumas"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
baseColor <- "#0F4C81"
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
    base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r premable, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(kableExtra)
library(patchwork)
library(moderndive)
knitr::opts_chunk$set(dev = 'svg')
```

# Course Overview

.pull-left[

```{r echo = FALSE, results='asis'}
block1_name = "Exploratory Data Analysis"
block1_lecs = c("Research design and data",
                "Describing categorical data",
                "Describing continuous data",
                "Describing relationships",
                "Functions")
block2_name = "Probability"
block2_lecs = c("Probability theory",
                "Probability rules",
                "Random variables (discrete)",
                "Random variables (continuous)",
                "Sampling")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")
course_table(block1_name,block2_name,block1_lecs,block2_lecs,week=9)
```


]

.pull-right[


```{r echo = FALSE, results='asis'}
block3_name = "Foundations of inference"
block3_lecs = c("Confidence intervals",
                "Hypothesis testing (p-values)",
                "Hypothesis testing (critical values)",
                "Hypothesis testing and confidence intervals",
                "Errors, power, effect size, assumptions")
block4_name = "Common hypothesis tests"
block4_lecs = c("One sample t-test",
                "Independent samples t-test",
                "Paired samples t-test",
                "Chi-square tests",
                "Correlation")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")
course_table(block3_name,block4_name,block3_lecs,block4_lecs,week=0)
```

]


---


# This Week's Learning Objectives
1. Understand the key difference between discrete and continuous probability distributions

2. Apply understanding of continuous probability distributions to the example of a normal distribution

3. Understand how to use a range from a continuous probability distribution

4. Introduce other continuous probability distributions

---
## Discrete vs. continuous
- Recall that a _discrete probability distribution_ describes a random variable that produces a discrete set of outcomes

--

- By contrast, a _continuous probability distribution_ describes a random variable that produces a continuous set of outcomes 

  - Temperature
  - Height
  - Reaction Time

- If you have arbitrary precision of measurement, you have a continuous random variable

--

- While a discrete probability distribution is jagged, a continuous probability distribution is smooth 
.pull-left[
```{r, echo = F, fig.height = 2.5, fig.width = 4.75}
set.seed(820)
disDat <- tibble(x = as.factor(0:5), Probability = dbinom(0:5, 5, 0.5))
conDat <- tibble(x = as.factor(1:100), Probability = rnorm(100, m=35, sd=7))

ggplot(disDat, aes(x, Probability)) + geom_bar(stat = 'identity', fill = baseColor) +
  labs(x = 'Discrete Random Variable')
```
]

.pull-right[
```{r, echo = F, fig.height = 2.5, fig.width = 4.75}
cPlot <- ggplot(conDat, aes(Probability)) + geom_density() +
  labs(x = 'Continuous Random Variable') + 
  scale_x_continuous(breaks = seq(10, 60, by=10), limits=c(10, 60)) +
  geom_hline(yintercept = 0)

cPlot
```
]

---
## Discrete vs. continuous

- Continuous probability distributions differ from discrete in two other important ways

--
    
    - $P(X=x)=0$

.center[
```{r, echo = F, fig.height=4, fig.width=6}

cPlot + geom_vline(xintercept = 42.28529, colour = 'red')

```
]

---
count: false

## Discrete vs. continuous

- Continuous probability distributions differ from discrete in two other important ways

    - $P(X=x)=0$

.center[
```{r, echo = F, message = F, fig.height=4, fig.width=6}

cPlot + geom_vline(xintercept = 42.28529, colour = 'red') + 
  coord_cartesian(xlim = c(40, 45)) + 
  scale_x_continuous(breaks=seq(40, 45, by = 1))

```
]


---
count: false

## Discrete vs. continuous

- Continuous probability distributions differ from discrete in two other important ways

    - $P(X=x)=0$

.center[
```{r, echo = F, message = F, fig.height=4, fig.width=6}

cPlot + geom_vline(xintercept = 42.28529, colour = 'red') + 
  coord_cartesian(xlim = c(41.8, 42.8)) + 
  scale_x_continuous(breaks=seq(41.8, 42.8, by = .2))

```
]

---
count: false

## Discrete vs. continuous

- Continuous probability distributions differ from discrete in two other important ways

    - $P(X=x)=0$

.center[
```{r, echo = F, message = F, fig.height=4, fig.width=6}

cPlot + geom_vline(xintercept = 42.28529, colour = 'red') + 
  coord_cartesian(xlim = c(42.27, 42.3)) + 
  scale_x_continuous(breaks=seq(42.27, 42.3, by = .005))

```
]

---
count: false

## Discrete vs. continuous

- Continuous probability distributions differ from discrete in two other important ways

    - $P(X=x)=0$

.center[
```{r, echo = F, message = F, fig.height=4, fig.width=6}

cPlot + geom_vline(xintercept = 42.28529, colour = 'red') + 
  coord_cartesian(xlim = c(42.28, 42.29)) + 
  scale_x_continuous(breaks=seq(42.28, 42.29, by = .002))

```
]

---

count: false

## Discrete vs. continuous

- Continuous probability distributions differ from discrete in two other important ways

    - $P(X=x)=0$

  - Continuous probability distributions are described using the **probability density function (PDF)**, rather than the **probability mass function**

--

- Now, let's take a look at perhaps the most widely used continuous probability distribution... 

---
## Normal distribution

.pull-left[
- A **normal distribution** (AKA the Gaussian distribution) is a continuous distribution


- It is uni-modal (one peak) and symmetrical

]

.pull-right[
```{r, echo = F, fig.height = 3, fig.width = 5}
ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
        stat_function(fun = dnorm) + geom_hline(yintercept = 0)
```
]

---
## Normal: PDF

$$
f(x;\mu, \sigma) = \frac{1}{\sigma\sqrt{2\pi}}e^{-\frac{(x - \mu)^2}{2\sigma^2}}
$$

- A bit scary!

- But the basic points are:
  
  - It is a function of data *x*
  - And *two* parameters $\mu$ and $\sigma$ (mean and SD)

--

- There is not one single normal distribution
- We have a family of different distributions that are defined by their mean, $\mu$, and standard deviation, $\sigma$


---
## The Standard Normal Distribution

- The **standard normal distribution** is a normal distribution where $\mu=0$ and $\sigma=1$

.center[
```{r, echo = F, fig.width=7, fig.height=4}
normDist <- ggplot(data.frame(x = c(-6, 6)), aes(x = x)) +
  stat_function(fun=dnorm, geom = "line", linewidth = 1.5) + 
  theme(axis.text = element_text(size = 12), 
        axis.title = element_text(size = 14, face = 'bold'))

normDist + geom_vline(xintercept = 0, linetype = 'dashed', colour = 'red')
```
]

---
## Different Normal Distributions - Adjusting $\mu$

- Adjusting $\mu$ changes where the curve is centered on the $x$-axis

.center[
```{r, echo=FALSE, fig.width=7, fig.height=4}
normDist +
  stat_function(fun = dnorm,
                geom = "line",
                col = "red",
                args = list(mean=2),
                linewidth = 1.5) +
  stat_function(fun = dnorm,
                geom = "line",
                col = baseColor,
                args = list(mean=-2),
                linewidth = 1.5) +
  annotate(geom='text', label=paste0('mu == -2'), parse = T, 
           x=-4.75, y=0.38, size = 6, colour = baseColor, hjust = 0) +
  annotate(geom='text', label=paste0('mu == 0'), parse = T, 
           x=-4.75, y=0.35, size = 6, hjust = 0) +
  annotate(geom='text', label=paste0('mu == 2'), parse = T, 
           x=-4.75, y=0.32, size = 6, colour = 'red', hjust = 0)
```
]

---
## Different Normal Distributions - Adjusting $\sigma$

- Adjusting $\sigma$ changes the shape of the curve

.center[
```{r, echo=FALSE, fig.width=7, fig.height=4}
normDist +
  stat_function(fun = dnorm,
                geom = "line",
                col = "red",
                args = list(sd=1.5),
                linewidth = 1.5) +
  stat_function(fun = dnorm,
                geom = "line",
                col = baseColor,
                args = list(sd=0.5),
                linewidth = 1.5) +
  annotate(geom='text', label=paste0('sigma == 0.5'), parse = T, 
           x=-3, y=0.71, size = 6, colour = baseColor, hjust = 0) +
  annotate(geom='text', label=paste0('sigma == 1'), parse = T, 
           x=-3, y=0.65, size = 6,  hjust = 0) +
  annotate(geom='text', label=paste0('sigma == 1.5'), parse = T, 
           x=-3, y=0.59, size = 6, colour = 'red',  hjust = 0)
```
]

---
## Properties of Normal Distributions

.pull-left[
- Properties of any normal distribution: 
  - $\approx$ 68% of area falls under 1 SD on either side of mean
]

.pull-right[
```{r, echo = F, fig.height=4, fig.width=7}

normDist + xlab("# of standard deviations from the mean") +
  geom_area(stat = "function", fun = dnorm, fill = baseColor, xlim = c(-1, 1)) +
  scale_x_continuous(breaks = seq(-4, 4, 1), limits = c(-4, 4)) + 
  theme(axis.text = element_text(size=12), axis.title = element_text(size=14, face='bold')) +
  geom_vline(xintercept = -1, colour = baseColor, linewidth = 1.5) +
  geom_vline(xintercept = 1, colour = baseColor, linewidth = 1.5) +
  annotate('text', label = '~68%', x = 0, y = 0.12, colour = 'white', size = 8)

```
]

---
count: false

## Properties of Normal Distributions

.pull-left[
- Properties of any normal distribution: 
  - $\approx$ 68% of area falls under 1 SD on either side of mean
  - $\approx$ 95% of area falls under 2 SD on either side of mean
      - _Exactly_ 95% falls under +/- **1.96 SD**
]

.pull-right[
```{r, echo = F, fig.height=4, fig.width=7}

normDist + xlab("# of standard deviations from the mean") +
  geom_area(stat = "function", fun = dnorm, fill = baseColor, xlim = c(-2, 2)) +
  scale_x_continuous(breaks = seq(-4, 4, 1), limits = c(-4, 4)) + 
  theme(axis.text = element_text(size=12), axis.title = element_text(size=14, face='bold')) +
  geom_vline(xintercept = -2, colour = baseColor, linewidth = 1.5) +
  geom_vline(xintercept = 2, colour = baseColor, linewidth = 1.5) +
  annotate('text', label = '~95%', x = 0, y = 0.12, colour = 'white', size = 8)

```
]

---
count: false

## Properties of Normal Distributions

.pull-left[
- Properties of any normal distribution:
  - $\approx$ 68% of area falls under 1 SD on either side of mean
  - $\approx$ 95% of area falls under 2 SD on either side of mean
      - _Exactly_ 95% falls under +/- **1.96 SD**
  - $\approx$ 99.75% of area falls under 3 SD on either side of mean
]

.pull-right[
```{r, echo = F, fig.height=4, fig.width=7}

normDist + xlab("# of standard deviations from the mean") +
  geom_area(stat = "function", fun = dnorm, fill = baseColor, xlim = c(-3, 3)) +
  scale_x_continuous(breaks = seq(-4, 4, 1), limits = c(-4, 4)) + 
  theme(axis.text = element_text(size=12), axis.title = element_text(size=14, face='bold')) +
  geom_vline(xintercept = -3, colour = baseColor, linewidth = 1.5) +
  geom_vline(xintercept = 3, colour = baseColor, linewidth = 1.5) +
  annotate('text', label = '~99.75%', x = 0, y = 0.12, colour = 'white', size = 8)

```
]

---
class: center, middle

## Questions?


---
## Using the PDF of the normal distribution
- Let's use the normal distribution to illustrate how continuous probability distributions work

--

- With a discrete random variable it makes sense to ask: 'What's the probability associated with a specific value of the random variable?'.
  - e.g. what the probability of getting heads on a fair coin? 

--

- With a continuous random variable it makes sense to ask about ranges of scores
  - e.g. what's the probability of sampling someone between 1.75 and 1.8 meters tall if we sample students from a university? 
  
--
  - Remember that the probability of any single value (e.g. exactly 1.764736525678943655 meters) is 0
  - The total probability (1) is divided between an infinite number of possible values that the variable could take, as the variable is continuous

---
## Using the PDF of the normal distribution
```{r, echo = F}
heightMean <- 168
heightSD <- 7.5
lowBound <- 175
hiBound <- 180
```

.pull-left[
- Let's imagine that in some course, student height is normally distributed

  + $\mu = `r heightMean`$ cm 
  + $\sigma = `r heightSD`$ cm
    
- We can ask what is the probability of sampling someone between `r lowBound` and `r hiBound` cm? 

  + This question translates to: $P(`r lowBound` \leq x \leq `r hiBound`) = ?$
  + Let's unpack this... 
]

.pull-right[
```{r, echo=FALSE, fig.height=5}
set.seed(206)
dat <- tibble(height=rnorm(200, mean = heightMean, sd = heightSD))

hPlot <- ggplot(dat, aes(height)) + 
  stat_function(fun = dnorm,
                geom = "line",
                args = list(mean = heightMean, sd=heightSD)) +
  labs(x="Height", y = "Probability Density") +
  scale_x_continuous(limits = c(135, 200)) +
  geom_hline(yintercept = 0)

hPlot
```
]

---
## Using the PDF of the normal distribution
.pull-left[

$$P(`r lowBound` \leq x \leq `r hiBound`) = ?$$

+ Let's draw these boundaries on our plot
   

]

.pull-right[
```{r, echo=FALSE, fig.height=5}
hPlot
```
]

---
## Using the PDF of the normal distribution
.pull-left[

$$P(`r lowBound` \leq x \leq `r hiBound`) = ?$$

+ Let's draw these boundaries on our plot

+ What is the value of the area under the curve between these two lines?
   

]

.pull-right[
```{r, echo=FALSE, fig.height=5}
hPlot + 
  geom_vline(xintercept = lowBound, colour = baseColor) +
  geom_vline(xintercept = hiBound, colour = baseColor)
```
]

---
## Using the PDF of the normal distribution
.pull-left[

- We get the area under a curve by calculating an integral
  
  $$\int_{a}^{b} f(a) \,dx$$
  
  + Don't worry, you don't need to know the details of integrals, but you may encounter the equation above
  
  - This equation can be read as: The integral of values falling between vertical lines $a$ and $b$ on the function $a$ of variable $x$ 
  
  - We can calculate this value using the probability density function

]

.pull-right[
```{r, echo=FALSE, fig.height=5}
hPlot + 
  geom_vline(xintercept = lowBound, colour = baseColor) +
  geom_vline(xintercept = hiBound, colour = baseColor)
```
]

---
## Using the PDF of the normal distribution
.pull-left[

- `pnorm(x, mean, sd)` 
  + *x* is the upper threshold; the function will output the probability of all values less than this
  
  + *mean* and *sd* give the parameters of the function
  
  + Returns the area under the normal distribution below x 
  
  + Remember, the normal curve changes based on the values of $\mu$ and $\sigma$, so it makes sense that this PDF requires these parameters
]

.pull-right[
```{r, echo=FALSE, fig.height=5}
hPlot + 
  geom_vline(xintercept = lowBound, colour = baseColor) +
  geom_vline(xintercept = hiBound, colour = baseColor)
```
]

---
## Using the PDF of the normal distribution
.pull-left[

- `pnorm(x, mean, sd)` 
  + *x* is the upper threshold; the function will output the probability of all values less than this
  
  + *mean* and *sd* give the parameters of the function
  
  + Returns the area under the normal distribution below x
  
  + Remember, the normal curve changes based on the values of $\mu$ and $\sigma$, so it makes sense that this PDF requires these parameters

```{r, eval=TRUE}
pnorm(180, mean=168, sd=7.5)
```

> **Test Your Understanding:** How do you interpret this output?

]


.pull-right[
```{r, echo=FALSE, fig.height=5}
hPlot + 
  geom_vline(xintercept = hiBound, colour = baseColor) +
  geom_area(stat = "function", fun = dnorm, 
            fill = baseColor, xlim = c(135, hiBound), 
            args = list(mean = heightMean, sd = heightSD),
            alpha = 0.7) +
  annotate('text', label = paste0('~', round(pnorm(hiBound, mean=heightMean, sd=heightSD),4)*100, '%'),
           x = heightMean, y = 0.02, colour = 'white', size = 8)

```
]

---
## Using the PDF of the normal distribution


.pull-left[
- We can also calculate the area under the curve below `r lowBound`:

```{r, eval=TRUE}
pnorm(175, mean=168, sd=7.5) 
```
]

.pull-right[
```{r, echo=FALSE, fig.height=5}
hPlot + 
  geom_vline(xintercept = lowBound, colour = baseColor) +
  geom_area(stat = "function", fun = dnorm, 
            fill = baseColor, xlim = c(135, lowBound), 
            args = list(mean = heightMean, sd = heightSD),
            alpha = 0.7) +
  annotate('text', label = paste0('~', round(pnorm(lowBound, mean=heightMean, sd=heightSD),4)*100, '%'),
           x = heightMean-1, y = 0.02, colour = 'white', size = 8)
```
]

--

> **Test Your Understanding:** Now you know that 94.52% of student heights fall below 180 cm, and 82.47% of student heights fall below 175 cm. How do you calculate the probability of selecting a student whose height falls between 175-180 cm?


---
## Using the PDF of the normal distribution

.pull-left[
+ $$P(175 \leq x \leq 180) = P(X<180) - P(X<175)$$

```{r, eval=TRUE}
p180 <- pnorm(180, mean=168, sd=7.5)
p175 <-  pnorm(175, mean=168, sd=7.5) 

p180-p175
```

+ So, the probability of randomly selecting a student with a height between `r lowBound` and `r hiBound` is `r round(pnorm(hiBound, mean=heightMean, sd=heightSD) - pnorm(lowBound, mean=heightMean, sd=heightSD), 2)` 

]

.pull-right[
```{r, echo=FALSE, fig.height=5}
hPlot + 
  geom_vline(xintercept = lowBound, colour = baseColor) +
  geom_vline(xintercept = hiBound, colour = baseColor) +
  geom_area(stat = "function", fun = dnorm, 
            fill = baseColor, xlim = c(135, lowBound), 
            args = list(mean = heightMean, sd = heightSD),
            alpha = 0.5) + 
  geom_area(stat = "function", fun = dnorm, 
            fill = baseColor, xlim = c(135, hiBound), 
            args = list(mean = heightMean, sd = heightSD),
            alpha = 0.5)
```
]
---
## Using the PDF of the normal distribution

```{r, echo = F}
hiSD <- 185
loSD <- 151
```

.pull-left[
- We can also ask about the probability of a sampled element having a value from one of 2+ ranges

- What is the probability that a person will have a height below `r loSD` or greater than `r hiSD`? $P(x \leq `r loSD` \:or\: x \geq `r hiSD`)$

```{r, eval=TRUE}
pnorm(151, mean=168, sd=7.5)
```

```{r, eval=TRUE}
1 - pnorm(185, mean=168, sd=7.5)
```

+ **Test your understanding:** Why are we subtracting a value from 1 here?

]

.pull-right[
```{r, echo=FALSE, fig.height = 5}
hPlot + 
  geom_vline(xintercept = loSD, colour = baseColor) +
  geom_vline(xintercept = hiSD, colour = baseColor) +
  geom_area(stat = "function", fun = dnorm, 
            fill = baseColor, xlim = c(135, loSD), 
            args = list(mean = heightMean, sd = heightSD),
            alpha = 0.7) + 
  geom_area(stat = "function", fun = dnorm, 
            fill = baseColor, xlim = c(hiSD, 200), 
            args = list(mean = heightMean, sd = heightSD),
            alpha = 0.7)

```
]

---
## Using the PDF of the normal distribution

.pull-left[
- $P(x \leq `r loSD` \cup x \geq `r hiSD`) = P(x \leq `r loSD`) + P(x \geq `r hiSD`)$ 

- $`r round(pnorm(loSD, mean=heightMean, sd=heightSD), 2)` + `r round(1-pnorm(hiSD, mean=heightMean, sd=heightSD), 2)` = `r round(pnorm(loSD, mean=heightMean, sd=heightSD)+(1-pnorm(hiSD, mean=heightMean, sd=heightSD)), 2)`$
]


.pull-right[
```{r, echo=FALSE, fig.height = 5}
hPlot + 
  geom_vline(xintercept = loSD, colour = baseColor) +
  geom_vline(xintercept = hiSD, colour = baseColor) +
  geom_area(stat = "function", fun = dnorm, 
            fill = baseColor, xlim = c(135, loSD), 
            args = list(mean = heightMean, sd = heightSD),
            alpha = 0.7) + 
  geom_area(stat = "function", fun = dnorm, 
            fill = baseColor, xlim = c(hiSD, 200), 
            args = list(mean = heightMean, sd = heightSD),
            alpha = 0.7)

```
]

---
## Using the PDF of the normal distribution
- What if I wanted to know where the 5% of the most extreme values (i.e., smallest and largest) in this distribution fall? 

--
  
   - The normal distribution is symmetric, which means that there are the same number of extreme values at the bottom and top end

--

   - This means the most extreme 5% will be the 2.5% at the bottom of the distribution and the 2.5% at the top

--

   - So our question is: What is the height below which there are only 2.5% of students, and what is the height above which there are only 2.5% of students? 

---
## Using the PDF of the normal distribution
.pull-left[

- To get these values, you can use `qnorm(p, mean, sd)` - this is the inverse function of `pnorm()`

- For a normally distributed range of heights with a mean of `r heightMean` cm and a SD of `r heightSD` cm: 

- The height below which 2.5% of students fall: 
```{r, eval=TRUE}
qnorm(.025, mean=168, sd=7.5)
```

- The height above which 2.5% of students fall: 
```{r, eval=TRUE}
qnorm(.975, mean=168, sd=7.5)
```
]

.pull-right[
```{r, echo=FALSE, message=F, fig.height=5}
hPlot + 
  geom_vline(xintercept = qnorm(.025, mean=heightMean, sd=heightSD), colour = baseColor) +
  geom_vline(xintercept = qnorm(.975, mean=heightMean, sd=heightSD), colour = baseColor) +
  scale_x_continuous(breaks = seq(130, 200, 10)) + 
  geom_area(stat = "function", fun = dnorm, 
            fill = baseColor, xlim = c(135, qnorm(.025, mean=heightMean, sd=heightSD)), 
            args = list(mean = heightMean, sd = heightSD),
            alpha = 0.7) + 
  geom_area(stat = "function", fun = dnorm, 
            fill = baseColor, xlim = c(qnorm(.975, mean=heightMean, sd=heightSD), 200), 
            args = list(mean = heightMean, sd = heightSD),
            alpha = 0.7)
  
```
]

---
## Take this knowledge forward

+ These examples might seem a bit far-fetched (when will you ever need to calculate extreme heights?), but this will be incredibly relevant when you discuss:

  + 1- and 2-tailed distributions
  + $p$ -values
  + Distributions of test statistics
    
+ You may find it helpful to come back and review these slides when you get to these topics later in the course

---
class: center, middle

## Questions?

---
## Remember $z$-scores

.pull-left[
.center[
$$
z = \frac{x - \mu}{\sigma}
$$
]
- It is quite typical to present a normal distribution in terms of ** $z$-scores**.
- $z$-scores standardise values of $x$
  - The numerator: converts $x$ to deviations from the mean
  - The denominator: scales these deviation values based on the spread of the variable (SD)
- The result is the **standard normal distribution**, also known as the $z$-distribution
]

.pull-right[
```{r, echo = F, fig.height=5}
normDist +
  scale_x_continuous(limits = c(-4, 4)) +
  labs(x = 'z-score', y = '')
```
]


---
## Standard normal vs. $t$ distribution
.pull-left[

- There are other continuous probability distributions you'll be working with next semester, such as the  $t$-distribution 
- The $t$ distribution is a bit like the $z$-distribution, but the shape differs slightly
  - When calculating $t$, we replace the population SD $(\sigma)$ with the sample SD $(s)$
  - As a result, the tails of the $t$-distribution are slightly higher to account for extra variability, or uncertainty from using an estimate $(s)$ rather than the actual population value $(\sigma)$
]

.pull-right[
```{r, echo=FALSE, fig.height=5}
normDist + 
  scale_x_continuous(limits = c(-4, 4)) +
    stat_function(fun = dt,
                geom = "line",
                color = "red",
                args = list(df=5),
                linewidth = 1.5) + 
  annotate('text', label = 'Standard Normal Distribution', hjust = 0, x=-4, y = .35, size = 4.5) +
  annotate('text', label = 't-Distribution', hjust = 0, x=-4, y = .32, colour = 'red', size = 4.5) +
  labs(x = 'Statistic', y = '')

```
]

---
## Summary of today
- Continuous probability distributions 
- The normal distribution
- Using the normal distribution to make estimates about the probability of events
- Using the normal distribution to find values at the extremes of the distribution
- The normal distribution and the $t$-distribution

--

+ Tomorrow, I'll present a live R session focused on continuous probability distributions

+ Next week, we will talk about samples and populations


---
# This week

<script src="https://cdn.jsdelivr.net/npm/iconify-icon@2.1.0/dist/iconify-icon.min.js"></script>

.pull-left[
<iconify-icon icon="clarity:tasks-solid" width="64" height="64"  style="color: #0F4C81"></iconify-icon>

## Tasks

- Attend both lectures

- Attend your lab and work together on the lab tasks

- Complete the weekly quiz
    + Opened Monday at 9am
    + Closes Sunday at 5pm
    
<!-- - Submit Formative Report A by 12 noon on Friday the 18th of October 2024 -->
<!-- - Submit Formative Report B by 12 noon on Friday the 29th of November 2024 -->
<!-- - Submit Formative Report C by 12 noon on Friday the 14th of February 2025 -->
<!-- - Submit the Assessed Report by 12 noon on Friday the 28th of March 2025 -->
]


.pull-right[
<iconify-icon icon="raphael:help" width="64" height="64"  style="color: #0F4C81"></iconify-icon>

## Support

- **Office hours**: for one-to-one support on course materials or assessments<br>(see LEARN > Course information > Course contacts)

- **Piazza**: help each other on this peer-to-peer discussion forum

- **Student Adviser**: for general support while you are at university<br>(find your student adviser on MyEd/Euclid)
]
